### 2018-11-17 PNAS-template-main.tex
Update on Overleaf.

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\title{The Geometry of Admixture}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Benjamin M Peter}
%\author[b,1,2]{Author Two} 
%\author[a]{Author Three}

\affil[a]{Affiliation One}
%\affil[b]{Affiliation Two}
%\affil[c]{Affiliation Three}

\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\nsq}[1]{\left\lVert #1 \right\rVert^2}

% Please give the surname of the lead author for the running footer
\leadauthor{Peter}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. The Significance Statement appears in the paper itself and is required for all research papers.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any conflict of interest here.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) %contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: author.two\@email.com}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Principal component analysis $|$ admixture $|$ population structure $|$ geometry $|$ $F$-statistics $|$ ancientDNA}

\begin{abstract}
Please provide an abstract of no more than 250 words in a single paragraph. Abstracts should explain to the general reader the major contributions of the article. References in the abstract must be cited in full within the abstract itself and cited in the text.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{T}In the last decade, sequencing of thousands of ancient and modern human genomes has become feasible. In particular, ancient DNA extracted from bones has revolutionized our understanding of the prehistory of Eurasia and America,  with other continents likely to follow.

With the increase in data also comes a challenge in interpreting the data; as gaps in samplings are filled and links between populations are established, paradigms relying on discrete populations become increasingly hard to apply, as hundreds of distinct units need to be joinly considered.

As such, multivariate techniques such as principal component analysis (PCA) and Structure models are widely used to visualize human population structure, but usually they are not used to do inference directly.

Instead, one usually relies on an array of other approaches including the modelling of the SFS, PSMC-like approaches or admixture graphs; all of which have the issue that they can only handle a small number (typically less than 20) populations. Thus, complex inference typically relies on a combination of exploratory analyses, followed by a number of hypothesis tests investigating small subsets of the data.

In human population genetics, a particularly fashionable choice are the F-statistics, that are used to test for admixture using 3 or 4 populations at a time. Typically, they are motivated by trees and admixture graphs, but they can be derived under arbitrary population genetic models (Peter, 2016).

However, establishing a connection between these two approaches remains difficult. 
Here, I present a connection between the exploratory PCA and the formal statistics.

\subsection*{Model}
The $F$-statistics are defined as 
$$F_2(P_A, P_B) = E(p_a - p_b)(p_a - p_b)$$
$$F_3(P_A, P_B; P_X) = E(p_a - p_x)(p_b - p_x)$$
$$F_4(P_A, P_B, P_C, P_D) = E(p_a - p_b)(p_c - p_D)$$
where $P_i, p_i$ are populations and corresponding allele frequencies, respectively. Here, we think of the expectation over all loci in the genome.

Thus, for data from $n$ individuals and $L$ sites, a natural estimator for $F_3$ is
\begin{equation}
    F_3(A, B, X) = \frac{1}{L}\sum_i^L (p_{ai} - p_{xi})(p_{bi} - p_{xi}) - \hat{h}_x
\end{equation}
where $\hat{h}_x$ is an estimator of the heterozygosity.

Now, imagine we have a matrix $X$ containing all the data, and that all individuals are diploid so that $X_{ij} \in (0,1,2)$ Then,

\begin{equation}
    F_3(A, B, X) = \frac{1}{2L}\sum_i^L (X_{ai} - X_{xi})(X_{bi} - X_{xi}) - \frac{1}{4}\sum_i X_{xi}(2-X_{xi}).
\end{equation}

Another common method to analyze population genetic data is PCA. Typically, \cite{mcvean2009} one constructs the row-normalized matrix $Y_i = X_i - \frac{1}{n}\sum_n X_i$, and performs a singular-value-decompositon, $X=UDV^T = UP$, where $P = DV^T$ is a $L \times n$ matrix of principal components, and $U$ is $L \times n$ matrix of loadings.

As $(X_{ai}-X_{xi})(X_{bi}-X_{xi}) = (X_{ai}+\mu_i -X_{xi} - \mu_i)(X_{bi} + \mu_i-X_{xi}_i-\mu_i) = (Y_{ai}-Y_{xi})(Y_{bi}-Y_{xi})$, we can rewrite $F_3$ as

\begin{equation}
    F_3(A, B, X) = \frac{1}{2L}\sum_{k=1}^K (P_{ak}  - P_{xk})(P_{bk} - P_{xk}) - h_x.
\end{equation}

That is, with the exception of the bias-correction term $h_x$, the $F$-statistic can be split into terms corresponding to individual principal components.

One property of PCA is that the PCs are orthogonal and in decreasing order, and it is commonly observed in human population genetics that only a very small number of PCs contain biologically useful information. Hence, we can approximate the data by the first $K$ PCs.

\subsection*{When is F3-negative?}
Given two samples $A$ and $B$, for which sample $X$ will we detect admixture?
As 
\begin{equation}
    2 F_3(A, B,X) = \nsq{A-X} + \nsq{B-X} - \nsq{A-B},
\end{equation}
it follows from the Pythagorean theorem that $F_3$ will be zero if  if $X$ lies on the $K$-dimensional sphere with diameter $$\norm{A,B}$, negative if it lies inside and positive outside (FIGURE).

\subsection*{The geometry of $F_4$}
Given three samples $A, B$ and $C$, for which sample $X$ will $F_4$ be zero?
First, we characterize the space for which $F_4$ is zero:
\begin{equation}
    2 F_4(A, B | C, X) = \nsq{A-X} + \nsq{B-C} - \nsq{B-X} - \nsq{A-C} = 0,
\end{equation}

This equation is equivalent to the statement that the (skew-)quadrilateral spanned by the points (A, B, C and X) is orthodiagonal. Hence, $X$ lies in the plane that goes through the point $C$, and whose norm passes through $A$ and $B$.

A special case is 
\begin{equation}
    2 F_4(A, B | A, X) =2 F_3(A, X | B) = \nsq{A-X} + \nsq{B-A} - \nsq{B-X},
\end{equation}
that arises when one considers the ``outgroup''-$F_3$-statistic or the question of which populations besides $A$ have contributed to an admixed population $X$. From the above, $F_3(A, X|B)$ is zero for samples on the plane passing through $X$ with norm perpendicular to the line passing through $A$ and $X$.

The further away $B$ is from the plane, the lower the $F_3$-statistic will be.

What can we say about $X, Y$ in
\begin{equation}
    2 F_4(A, B | X, Y) = \nsq{A-X} + \nsq{B-Y} - \nsq{A-Y} - \nsq{B-Y} = 0
    \text{?}
\end{equation}
$X, Y$ must lie on a plane perpendicular to the line segment $A-C$. 
What can we say about $X, Y$ in
\begin{equation}
    2 F_4(A, B | X, Y) = \nsq{A-X} + \nsq{B-Y} - \nsq{A-Y} - \nsq{B-Y} = k
    \text{?}
    \end{equation}
$X$ and $Y$ must lie on parallel planes with norm $A-C$ and a distance $k$ apart.

What can we say about $X, Y$ in
\begin{equation}
    2 F_4(A, X | C, Y) = \sum (A-X)(C-Y) = 0
    \text{?}
\end{equation}

As $$\sum(A-X)(C-Y) = \norm{A-X}\norm{C-Y}\cos{\theta} = 0,$$
it follows that the line segments AX and CY are orthognal. From this it follows that $X$ and $Y$ lie on orthogonal planes that pass through $A$ and $C$, respectively.

\subsubsection{F4-Ratio's in PC-space}
if we interpret $F_4(O,I,X,2)=k$ as a function of $X$, we obtain a plane that is i) orthogonal to OI and a distance of $k$ away from $P_2$. Thus, the $F_4$-statistics in the ratio
\begin{equation}
    \alpha = \frac{F_4(O,I,X,2)} {F_4(O,I,1,2)}
\end{equation}
corresponds to two parallel planes; and $\alpha$ measures the difference in distance to the plane through the point $P_2$.

\subsubsection{qpAdm}
Given a population $T$ of interest, and some outgroups (``right'' \textit{sensu} Patterson) populations $R_j$ and potential admixture sources (``left'') populations $S_i$, we aim to estimate the mixture coefficients $w_i$, s.t.

\begin{equation}
    \sum_i w_i F_4(T, S_i | R_1, R_2) = 0
\end{equation}
For fixed $T$, $R_1$, $R_2$, we know that the $F_4$ statistics lie on a plane orthogonal to $R_1, R_2$ and passing through $T$. Thus, we aim to find weights such that the weighted sum of the projection of $S_i$ onto that plane equals $T$. For more than two outgroups $R_j$, we generate a plane for each pair of populations from $R_j$, and jointly optimize in this space. This makes most sense if these planes are orthogonal to each other.

\section{Other bases}
\begin{enumerate}
    \item Treelet
    \item simple Trees
    \item Rotate for specific $F$-statistic
\end{enumerate}
\section{Spectral analysis of admixture statistics}
\begin{enumerate}
    \item Split $F$-statistics by basis vector
    \item Same $F$-stat may arise as different contributions
    \item can use clustering to infer shared history?
\end{enumerate}

\subsection*{Author Affiliations}

Max-Planck-Institute for Evolutionary Anthropology, Department of Genetics, 04103 Leipzig, Germany.

\subsection*{References}

%\begin{figure}%[tbhp]
%\centering
%\includegraphics[width=.8\linewidth]{frog}
%\caption{Placeholder image of a frog with a long example caption to show %justification setting.}
%\label{fig:frog}
%\end{figure}

%\begin{SCfigure*}[\sidecaptionrelwidth][t]
%\centering
%\includegraphics[width=11.4cm,height=11.4cm]{frog}
%\caption{This caption would be placed at the side of the figure, rather than %below it.}\label{fig:side}
%\end{SCfigure*}

\matmethods{
Let $A, B$ be populations; $L$ the number of loci and $a_l, b_l$ the number of 
derived alleles at locus $l$ when taking $n_l, m_l$ haploid individuals from populations $A$, $B$, respectively. Let $\mu_l$ be the mean allele frequency at locus $l$. The estimator for $F_2$ is 
\begin{eqnarray}
F_2(A,B) &= \sum_{l=1}^L (a_l-b_l)^2 - \frac{a_l (n_l - a_l)}{n_l^2(n_l-1)}- \frac{b_l (m_l - b_l)}{m_l^2(m_l-1)}
\end{eqnarray}

Furthermore, let $Q$ be an orthonormal basis on the space of loci, e.g. obtained through a PCA on the zero-centered data.  The square term
\begin{eqnarray}
S &=& \sum_{l=1}^L \big( (a_l - \mu_l) -(b_l -\mu_l)\big)^2\\
&=& \sum_{l=1}^L \big( \sum_k Q_{lk}P_{ka} - \sum_kQ_{lk}P_{kb}\big)^2\\
&=& \sum_{l=1}^L \left( \sum_k Q_{lk} (P_{ka} -P_{kb}) \right)^2\\
&=& \sum_{l=1}^L \left( \sum_k Q_{lk}^2 (P_{ka} -P_{kb})^2 + \sum_{k\neq k'} Q_{lk}Q_{lk'}(P_{ka} - P_{k'b})^2 \right)\\
&=& \sum_k\sum_{l=1}^L Q_{lk}^2 (P_{ka} -P_{kb})^2 + \sum_{k\neq k'}\sum_{l=1}^L Q_{lk}Q_{lk'} (P_{ka} - P_{k'b})^2\\
&=& \sum_k (P_{ka} - P_{kb})^2
\end{eqnarray}
The heterozygosity term, on the other hand, obeys
\begin{eqnarray}
H_a &=& \sum_{l=1}^L \frac{a_l (n_l - a_l)}{n_l^2(n_l-1)}\\
&=& \sum_{l=1}^L \frac{ (z_l + \mu_l) (n_l - (z_l + \mu_l))}{n_l^2(n_l-1)}\\
&=& \sum_{l=1}^L\frac{-z_l^2}{n_l^2(n_l-1)} + z_l \frac{n_l - 2\mu_l}{n_l^2(n_l-1)} + \frac{n_l\mu_l - \mu_l^2}{n_l^2(n_l-1)}\\
&\approx&  \sum_k \frac{-P_{ka}^2}{n^2(n-1)} + \sum_k P_{ka} \sum_l Q_{lk}\frac{n - 2\mu_l}{n^2(n-1)} + \sum_l\frac{n\mu_l - \mu_l^2}{n^2(n-1)}\\
&=&  -C_1 \sum_k P_{ka}^2 + \sum_k P_{ka} C_{2k}+ C_3
\end{eqnarray}
Without missing data, the constants $C_1= ({n^2(n-1)})^{-1}$; $C_{2k} =C_1\sum_l Q_{lk}(n - 2\mu_l)$ and $C_3= C_1\sum_l(n\mu_l - \mu_l^2)$ only depend on the sample size and $Q_{lk}$ and can be precomputed.

}

\showmatmethods{} % Display the Materials and Methods section

\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}

%\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliography{pnas-sample}

\end{document}

### 2021-03-19 PNAS-template-main.tex
cleaner

@@
 % Use the lineno option to display guide line numbers if required.
+
+\include{amsmath}
 
@@
 
-\title{The Geometry of Admixture}
+\title{F-stats vs PCA}
 
@@
 
+$F_3$ and $F_4$ can be written as functions of $F_2$ \cite{peter2016}: 
+\begin{align}
+F_3(P_A, P_B; P_X) &= F_2(P_A, P_X) + F_2(P_B, P_X) - F_2(P_A, P_B)\\
+F_4(P_A, P_B; P_C, P_D) &= F_2(P_A, P_C) + F_2(P_B, P_D) - F_2(P_A, P_C) - F_2(P_B, P_C)
+\end{align}
+
+In many applications, we will have a large set of $n$ samples that we wish to analyze.
+One way to proceed is to calculate all $F_2$ statistics, which will be a sufficient summary to calculate all $F_3$ and $F_4$ statistics from this data.
+
+$F_2$ can also be written as 
+$$\mathbb{E}F_2 \propto 2T_{12} - T_{11} - T_{22}$$
+
+Similarly, we may want to calculate a PCA:
+
 Thus, for data from $n$ individuals and $L$ sites, a natural estimator for $F_3$ is
@@
 
-As $(X_{ai}-X_{xi})(X_{bi}-X_{xi}) = (X_{ai}+\mu_i -X_{xi} - \mu_i)(X_{bi} + \mu_i-X_{xi}_i-\mu_i) = (Y_{ai}-Y_{xi})(Y_{bi}-Y_{xi})$, we can rewrite $F_3$ as
+As $(X_{ai}-X_{xi})(X_{bi}-X_{xi}) = (X_{ai}+\mu_i -X_{xi} - \mu_i)(X_{bi} + \mu_i-X_{xi}-\mu_i) = (Y_{ai}-Y_{xi})(Y_{bi}-Y_{xi})$, we can rewrite $F_3$ as
 
@@
 \end{equation}
-it follows from the Pythagorean theorem that $F_3$ will be zero if  if $X$ lies on the $K$-dimensional sphere with diameter $$\norm{A,B}$, negative if it lies inside and positive outside (FIGURE).
+it follows from the Pythagorean theorem that $F_3$ will be zero if  if $X$ lies on the $K$-dimensional sphere with diameter $\norm{A,B}$, negative if it lies inside and positive outside (FIGURE).
 

### 2021-04-08 data/prep_data_pca.R
fstats_pca.tex

@@ -13,6 +13,10 @@ sample_list_og = c(sample_list, "Mbuti")
 data_loc = '/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_HO_public'
 inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
 
+data_loc = '/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_1240K_public'
+inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
+
+
 f2loc1 = 'pca_test1'
 
 f2s = admixtools::extract_f2(data_loc, pops=sample_list_og, outdir=f2loc1, maxmem=64000, blgsize=.5)

### 2021-04-08 fstats_pca.tex
actual git

\documentclass[10pt,a4paper]{article}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{comment}
\usepackage{natbib}

\bibliographystyle{evolution.bst}

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}

\newcommand{\MX}{\mathbf{X}} %uncentered data
\newcommand{\MC}{\mathbf{C}} %centering
\newcommand{\MY}{\mathbf{Y}} %centered data
\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
\newcommand{\MP}{\mathbf{P}} % PCs
\newcommand{\ML}{\mathbf{L}} % loadings
\newcommand{\MK}{\mathbf{K}} % Kernel
\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix

\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
\begin{document}
\section{A geometric interpretation of $F$-statistics}
$F$-statistics have been primarily motivated by trees and admixture graphs \citep{patterson2012, peter2016}, but the calculations hold up in a much wider data space. In particular,  \cite{oteo-garcia2021} provides a thorough introduction to interpreting $F$-statistics in the \emph{data space} $\mathbb{R}^k$. Their work builds much of the foundation of this discussion, by demonstrating analogies to classical geometry. A brief summary of their key results: A population's allele frequencies can be thought of as vector in $\mathbb{R}^k$. Then, $F_2(X_1, X_2) = \normsq{X_1 - X_2}$ is the squared Euclidean distance between the populations with vectors $X_1$ and $X_2$, and $F_4(X_1, X_2 ; X_3, X_4) = \langle X_1 - X_2, X_3 - X_4 \rangle$ is the inner (scalar) product between these two vectors. Here, I will mainly use the $F$-statistic notation, but use the geometric notation where convenient.

In this paper, the main focus is to use this interpretation of $F$-statistics in the context with principal component analysis (PCA), another widely used multivariate analysis tool in population genetics \citep{cavalli-sforza1994, reich2008, novembre2008, mcvean2009}. Informally, PCA is a rotation of the data such that the axis of highest variation is the first PC. The second PC th

\section{Relationship of PCA, $F_2$ and Outgroup-$F_3$}
	Let us assume we have some genotype data $\MX$, which contains data from $n$ on the populations on the rows, and $k$ loci on columns. Each population may be represented by a single (pseduo-)haploid, diploid individual, but in general will be an aggregation of one or many samples.

	To analyze this data, we might perform a PCA. A standard way of doing this is to do first center the matrix by subtracting the mean allele frequency from each column and perform an Singular Value Decomposition (SVD):

	\begin{equation}
	\MY = \MC\MX = \mathbf{U} \MSINGULAR \mathbf{V}^T = \MP\ML
	\end{equation}

	Here, $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix and $\mathbf{I}, \mathbf{1}$ are the identity matrix and a matrix of ones, respectively. $\MP=\mathbf{U}\MSINGULAR$ is the matrix of principal components (PCs), and, $\ML=\mathbf{V}^T$ is the  matrix of loadings, respectively. Occasionally, SNPs are weighted by the square-root of their variance; for now we will assume the SNPs are unweighted, and defer discussion of weighting to a later section.

	 The principal components $\MP$ are an orthogonal matrix of size $n \times n$ that are useful for understanding population structure, the loadings are an $n \times k$ orthonormal matrix that give the contribution of each SNP along each PC, it is often useful to look for outliers that might be indicative of selection \cite[e.g]{francois2010}.

	Equivalently, we obtain the PCs by performing an eigendecomposition of the  covariance matrix of $\MY$, denoted as $\MK$:
	 \begin{equation}
 \MK = \MY \MY^T = \mathbf{U}\MEIGEN\mathbf{U}^T = \mathbf{U}\MSINGULAR^2\mathbf{U}^T =\MP\MP^T
	\end{equation} where $\MEIGEN$ is the diagonal matrix of eigenvalues.  In cases we are interested in the loadings, they can be obtained as 
	\begin{equation}
	\ML = \MEIGEN^{-1}\MP^T\MC\MX
	\end{equation}

	Let $y_{il}$ denote the genotype of the $i$-th individual at the $l$-th SNP.

	\begin{equation}
	y_{il} = x_{il} - \mu_l
	\end{equation}
	where $\mu_l$ is the mean genotype at the $l$-th locus.

	The entries of the covariance matrix $\MK$ are then

	\begin{equation}
	k_{ij} = \sum_l y_{il} y_{jl} = \sum_l (x_{il} - \mu_l)(x_{yl} - \mu_l) = F_3(\MEAN; X_i, X_j)
	\end{equation}

	It is helpful to think of $\MEAN = (\mu_1, \dots \mu_k)$ as a ``fake''-individual whose allele frequency is the sample mean of the allele frequency space, it will always lie at the origin of the PC-space.

	Thus, we can also think of a PCA as performing a decomposition of an Outgroup-$F_3$-matrix, where we set the output to the mean of all marker allele frequencies. This is good practice if we do not have any outgroups available. However, we often also have outgroup data; for example when studying early European variation, Africans are sometimes a suitable outgroup. In this case, it would be consequential to perform an eigendecomposition of a true outgroup $F_3$-matrix.

This is also a useful way to establish how we can obtain $\MP$ from $\MF$ directly: Note that the row and column means of $\MK$ are zero:
\begin{equation*}
\sum_i k_{ij}= \sum_i\sum_l (x_{il}-\mu_l)(x_{jl} - \mu_l)= \sum_l(x_{jl} - \mu_l)\left[\sum_i x_{il} -\mu_l\right] = 0 \text{.}
\end{equation*}

Since $F_2(X_1, X_2) = F_2(X_1, \mu) + F_2(X_2, \mu) - 2 F_3(\mu; X_1, X_2) \\= \normsq{X_1-\mu} + \normsq{X_2 - \mu} - 2\langle X_1 - \mu, X_2 - \mu \rangle$

(not sure if geometry notation would be easier here),

\begin{equation}
\MK = \MC\MX\MX^T\MC-\frac{1}{2} \MC \MF \MC = \MP\MP^T
\end{equation}

Thus, when we mean-center the $\MF$-matrix, we subtract the $F_2$ terms in above equations; as they are invariant with respect to the column/ row means. What remains, is directly proportional to the matrix decomposed under a standard PC.

%In summary, a PCA and an eigendecomposition of the (negative) $\MF$-matrix will give the same results, and can be thought of as an eigendecomposition of the $\MFT$-matrix of an outgroup $F_3$-matrix with the sample mean as an outgroup.

\section{Approximating uncertainty on PCA-placement}

\section{Projection using f-stats}
Suppose we have a sample $U$ we wish to project onto an existing PCA-basis made from $\MX$, and let us assume we can compute $F_2(U, X_i)$ for all $i$. The ``best'' point i
For any particular reference sample, $F_2$ places the point on the hypercircle with equation
\begin{equation}
F_2(U, X_i) = \sum_{k=1}^{n-1} (p_{ik} - u_k)^2,
\end{equation}
where $p_{ik}$ and $u_k$ are the reference and unknown coordinate on the $k$-th component, respectively. It can be shown \cite{gower1968} that the $u_k$ can be found using
\begin{equation}
\mathbf{u} = \frac{1}{2}\MEIGEN^{-1}\MP^T\mathbf{d}
\end{equation}
where $\mathbf{d}$ is a column vector with $$d_i = F_2(X_i, \MEAN) - F_2(X_i, U)$$.

Given a fixed projection, this allows us to also propagate uncertainty on a PCA plot.

\section{$F$-stats in PCA-space}
The PCA aims to reveal the axes of major variance. The data is then ``rotated'' such that these axes can be visualized. Usually, only the first few PCs are considered.

However, as $F_2$ is a squared Euclidean distance of allele frequencies, it is invariant to rotation and translation. Hence, neither mean-centering, which is a translation nor PCA-rotation will change $F_2$.

What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or the principal components $\MP$. Formally,

\begin{eqnarray}
F_2(X_i, X_j) &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
&=& \sum_{l=1}^L \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{kj}\big)^2\nonumber\\
&=& \sum_{l=1}^L \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
&=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
&=& \sum_k \left(\sum_{l=1}^L L_{kl}^2\right) (P_{ik} -P_{jk})^2 + \sum_{k\neq k'}\left(\sum_{l=1}^L L_{kl}L_{k'l}\right) (P_{ik} - P_{jk'})^2\nonumber\\
&=& \sum_k (P_{ik} - P_{jk})^2
\end{eqnarray}

In summary, the first row shows that $F_2$ on the centered data will give the same results (as distances are invariant to translations), in the second row we apply the PC-decomposition. The third row is obtained from factoring out $L_{lk}$. Row four is obtained by multiplying out the sum inside the square term for a particular $l$. We have $k$ terms when for $\binom{k}{2}$ terms for different $k$'s.  Row five is obtained by expanding the outer sum and grouping terms by $k$.The final line is obtained by recognizing that $\ML$ is an orthonormal basis; where dot products of different vectors have lengths zero.

Note that if we estimate $F_2$, unbiased estimators are obtained by subtracting the population-heterozygosities $H_i, H_j$ from the statistic. As these are scalars, they do not change above calculation.

SHORT VERSION
\begin{align}
F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2 - H_i - H_j &&\nonumber\\ 
 &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 - H_i - H_j  &=& F_2(Y_i, Y_j) \nonumber\\
 &=& \sum_k (P_{ik} - P_{jk})^2 - H_i - H_j &=& F_2(P_i, P_j)
\end{align}
As $F_3$ and $F_4$ can be written as sums of various $F_2$-terms, analogous relations apply.

\subsection{$F$-stats in 2-dimensional PC-space}
 It is useful to consider the statistics on a PCA plot. The relationships we will discuss formally only hold in the full, $n$-dimensional PCA-space where we consider all principal components. Here, we start by discussing 2-dimensional spaces. This is useful for two reasons: for one, the geometry is simpler and we can think of circles and squares as opposed to hyperspheres and other high-dimensional geometric objects and thus help us build intuition. Second, in many applications it is argued that a 2-dimensional approximation is sufficient to explain the major components of population structure. In this case, the results here will hold under the same approximation assumptions in low-dimensional PCs; if they differ substantially from each other, it is likely that not sufficiently many PCs were considered.

\subsubsection{$F_2$ in PC-space}
The $F_2$-statistic as the squared Euclidean distance is the easiest to understand, it corresponds directly to the squared distance in PCA-space. This matches our intuition that closely related populations (which have low $F_2$) will be close to each other on a PCA-plot.

\subsubsection{$F_3$ and circles}
The $F_3$-statistic becomes more interesting; as outlines above we either think of $F_3$ as ``outgroup''-$F$-stats or as admixture $F$-stats. In the admixture case, we may ask the following question: given two source populations $X_1$, $X_2$, where would admixed populations on a PCA plot lie? From theory, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \cite{brisbin2012, mcvean2009}.

Formally, we would reject admixture if $F_3$ is negative, i.e. we are looking for the space
\begin{eqnarray}
2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} \nonumber\\
      &<&0
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ iff $X_1, X_2$ and $X_x$ form a right-angled triangle. Hence, the region where $F_3$ is zero is the circle with diameter through $X_1$ and $X_2$. If $X_x$ lies inside this circle, the angle is obtuse and $F_3$ is negative, otherwise it will be positive. Similarly, if we fix $X_1$ and $X_2$ and ask where on a 2D-PCA-plot $X_2$ would lie, this space is defined by all the points for which the angle between $X_1 X_x$ and $X_2 X_x$ is obtuse.

This highlights a potential identifiability issue with $F_3$: Since all values of $F_3$ that result in the same projection will give the same value; and multiple admixture events may result in the same $F_3$-value.

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{dummy_pca.png}
	\caption{\textbf{Geometric representation of $F$-statistics on 2D-PCA-plot.} A: $F_2$ represents the squared Euclidean distance between two points in PC-space. B: Admixture-$F_3(X_x; X_1, X_2)$ is negative if $X_x$ lies in the circle specified by the diameter $X_2-X_1$}. C: $F_3(X_x; X_1, X_2)$ is negative given $X_1, X_x$ if $X_2$ is in the gray space.  D: Outgroup-$F_3$ reflects the projection of $X_2 - X_O$ on $X_1 - X_O$. E: $F_4$ is the projection of $X_3 - X_4$ on $X_1-X_2$. F: If $X_x$ is admixed between $X_1$ and $X_2$, the admixture proportions will be projected.
\end{figure}

\subsubsection{$F_4$ and right angles}
The inner-product-interpretation of $F_4$ is similar to that of $F_3$, with the change that the two vectors we consider do not involve the same population. However, a finding of $F_4(X_1, X_2; X_3, X_4) = \langle X_1 - X_2, X_3 - X4 \rangle = 0$ similarly implies that the two vectors are orthogonal, and a non-zero value reflects the projection of one vector on the other.

\subsubsection{$F_4$-ratio}
\begin{eqnarray}
\frac{F_4(X_I, X_O; X_X, X_1)}{F_4(X_I, X_O; X_2, X_1)} &=& \frac{\norm{X_I-X_O}\norm{X_X-X_1}\cos(\alpha)}{\norm{X_I-X_O}\norm{X_2-X_1}\cos(\beta)}\nonumber\\
&=&\frac{\norm{X_X-X_1}\cos(\alpha)}{\norm{X_2-X_1}\cos(\beta)}\nonumber\\
&=& \frac{\norm{X_X' - X_1'}}{\norm{X_2' - X_1'}}
\end{eqnarray}
where $\alpha$ and $\beta$ are the angles between vectors, and $X_i'$ is the projection of $X_i$ on $X_I-X_O$.

Conjecture: Thus, we are measuring the distances between the admixing populations on the projected on the axis between $X_I$ and $X_O$. This ought to be valid only if $\langle X_1 - X_1', X_2 - X_2' \rangle$ are orthogonal to each other, and to $X_OX_I$, i.e.
$F_4(X_1, X_1', X_2, X_2') = 0$

\subsection{spectral analysis of admixture statistics}
\begin{enumerate}
	\item split F-stats by PCA basis vector
	\item same F-stat value may arise with different contribution from different PCs, should hint at distinct admixture events
	\item can use clustering to infer shared history?
	\item decomposition of admixture-F3?
%	\item rank of PCA vs qpAdm
\end{enumerate}

\section{Trees and admixture graphs in PCA-space}
\subsection{Trees}
Evolutionary trees are fundamental in phylogenetic analyses, as they, on a large, scale, approximate how taxa diversify. Within a species, applying trees is also very common, but more problematic as populations frequently do not evolve as discrete lineages; instead, they admix and diversify as much more continuous processes. This is largely due to the time-scales involved, speciation events that give rise to trees might often be similarly messy, but from a distance of millions of years these issues might disappear.

Thus, when estimating trees from population genetic data, we must be very careful about whether the data is actually consistent with a tree, or belongs to some wider class of model.

Trees can be thought of as a collection of orthogonal dimensions; as drift on each branch is independent from every other branch. Thus, each sample is only 
\begin{enumerate}
	\item Trees
	\item Admixture Graphs
	\item Treelets
	\item simple trees, admixture graph
\end{enumerate}

\section{other orthogonal bases}
The most general ``model''-space for (centered) SNP-data $\MY$ is $\mathbb{R}^k$, where we allow each SNP to be in its own dimension, and treat all dimensions as independent. However, since in most analyses the number of samples $n \ll k$, we can place all SNPs in a $n$-dimensional subspace $\mathbb{R}^n$. (Could be restricted further to $[0,1]^k$, but that does not appear to add much).
If the data were normally distributed, $\MK$ has a $n$-dimensional Wishart-distribution with $k$ degrees of freedom. Since SNP are neither normal nor independent, the degrees of freedom might be considerably lower but we might still end up with something normally distributed.

\section{Technical considerations}
	\subsection{SNP weighting}
	It is clear that weighting SNP will have some effect on the resulting PCAs. Upweighting rare variants e.g. will emphasis recent events, as rare variance in the sample are more likely to be recent.

	\subsection{Missing data}

	\subsection{What is a dimension?}
	A single population at a particular point in time can be thought of as a single point in allele-frequency space, given by it's $p$-dimensional locus of allele frequencies in that population. If this population evolves for some time in isolation, allele frequencies will change due to genetic drift; i.e. the population evolves along a single tree branch in the interpretation of \cite{patterson2012}. If we now add a second population, it will behave exactly the same, and the drift in the second population will be uncorrelated to the first, i.e. it evolves in a second dimension. Thus, if we have two populations that descend from the same ancestral population in isolation, they can be thought of as evolving along orthognal dimensions from the same point. This argument is at the foundation of F-statistics.

	\section{outtakes}
	PCA from $\MX$
	\begin{equation}
	\MK = \MY \MY^T = \MC\MX\MX^T \MC = \MP\MP^T
	\end{equation}

\bibliography{main}
\end{document}

### 2021-04-08 data/pca_vs_of3.R

@@ -1,0 +1,66 @@
+require(tidyverse)
+require(admixtools)
+
+OUTGROUP = 'Mbuti'
+
+df2mat <- function(f2_df){
+    f2_df%>% select(-se) %>% 
+        arrange(pop1, pop2) %>%
+        pivot_wider(names_from=pop2, values_from=est) %>%
+        column_to_rownames('pop1') %>% as.matrix
+}
+
+double_center <- function(mat){
+    m = rowMeans(mat)
+    x = t(mat - m)-m; 
+    x = x-mean(x)
+}
+
+pca_from_mat <- function(mat){
+    df = tibble()
+    e = eigen(mat)
+    e$vectors = t(t(e$vectors) )#* e$values)
+    colnames(e$vectors) = sprintf("PC%d", 1:nrow(mat))
+    as_tibble(e$vectors) %>% mutate(pop=colnames(mat)) %>%
+        mutate_at(vars(starts_with("PC")), function(X)X *sign(X[1]))  %>%
+        select(pop, starts_with("PC"))
+
+}
+
+project_on_pcs <- function(pcs, evals, new_dists, old_dist_to_center){
+    1/evals * t(pcs[,-1)a
+}
+
+f2s = admixtools::f2_from_precomp("pca_test1/")
+f2_raw = f2(f2s, unique=F)
+pops = rownames(f2s[,,1])
+inpops = pops[pops != OUTGROUP]
+
+
+f2mat = f2_raw %>% filter(pop1 != OUTGROUP, pop2 != OUTGROUP) %>% df2mat
+f3s = f3(f2s, pop1=OUTGROUP, pop2=inpops, pop3=inpops, unique=F)
+f3mat = f3s %>% mutate(pop1=pop3) %>% select(-z, -p, -pop3) %>% df2mat
+
+
+cf2mat = -double_center(f2mat) /2
+#diag(cf2mat) = diag(cf2mat) + 1e-3
+#cf2mat = double_center(cf2mat)
+pcs = cf2mat %>% pca_from_mat
+e = eigen(cf2mat)$values
+
+OUTGROUP = 'Bulgarian'
+inpops = pops[!pops %in% c("Mbuti", OUTGROUP)]
+xf3s = f3(f2s, pop1=OUTGROUP, pop2=inpops, pop3=inpops, unique=F)
[…]

### 2021-04-08 data/prep_data_pca2.R

@@ -1,0 +1,25 @@
+require(admixtools)
+require(tidyverse)
+require(glue)
+
+anno = read_tsv("/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_HO_public.anno")
+anno = anno %>% select(iid=2, pub=4, age=6, lat=11, long=12, country=10, group=8,
+cov = 14, n_snps=15, sex=16, lib=17, qc=18)
+
+
+MISSING = c("Irish_Ulster", "Polish", "German", "Irish", "Shetlandic", "Sorb")
+
+list1 = read_table("WestEurasian_PCA.list", col_names=F)[,1] %>% unlist %>% 
+    unname
+sample_list = list1[!list1 %in% MISSING]
+sample_list_og = c(sample_list, "Mbuti")
+
+l =read_tsv("global_pca.list", col_names=F)
+
+data_loc = '/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_HO_public'
+inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
+
+f2loc1 = 'pca_test3'
+
+f2s = admixtools::extract_f2(data_loc, pops=l$X1, outdir=f2loc1, maxmiss=0.03, maxmem=64000)
+

### 2021-04-08 data/test_ibd.R

@@ -1,0 +1,14 @@
+require(tidyverse)
+
+IND = 'I0104'
+
+#ibd_all = read_tsv("./ibd.pairs_all.v1.addinfo.tsv") 
+ibd = ibd_all %>%
+    filter(iid1 == IND | iid2 == IND) %>% 
+    mutate(iid=ifelse(iid1==IND, iid2, iid1)) %>% 
+    select(-iid1, -iid2) %>% relocate(iid)
+
+a = read_csv("~/fstats_pca/data/test_I0104.bin.xz", col_types=cols(chrom='f'))
+
+b = a %>% select(chrom, map, YAM:YAMCOW) %>% pivot_longer(cols=YAM:YAMCOW) %>% 
+    filter(value>.2, name != "YAM")

### 2021-04-13 fstats_pca.tex
start intro

@@
 \documentclass[10pt,a4paper]{article}
-\usepackage[latin1]{inputenc}
-\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+%\usepackage[T1]{fontenc}
 \usepackage{amsmath}
@@
 \begin{document}
-\section{A geometric interpretation of $F$-statistics}
+\section{Introduction}
+Thanks to the widespread availability of cheap sequencing, genome-wide data sets now frequently incorporate tens of thousands of present-day individuals\cite{gnomad2020}.
+Furthermore, advances in ancient DNA techniques now allow large data sets.
+Using this data allows the inference of fine-scale population structure, but translating this wealth of data into meaningful and detailed models of population history is still a major challenge.
+
+Particularly for the analysis of ancient DNA, two approaches have been proven to be particularly useful: one are global summary analyses, such as Structure \citep{pritchard2000, alexander2009} Principal Component Analysis (PCA) \citep{cavalli-sforza1994, reich2008, novembre2008, mcvean2009} and classical multidimensional scaling (MDS) \cite{fu2016, malaspinas_bammds}.
+Typically, these methods assume that population structure is \emph{sparse}, so that a low-rank approximation with few underlying ``components'' is sufficient to model population structure See e.g.
+\cite{engelhardt2010} for a useful perspective how these approaches are related.
+
+Facing a novel data set, PCA or MDS are often the first analyses (beyond quality controls) a researcher performs, in order to obtain insights in the general population structure they are faced with.
+In order to answer more specific questions and to test specific hypotheses, the $F$-statistic framework of \cite{patterson2012} has been proven particularly powerful (see also \cite{peter2016} for a more gentle introduction).
+In the $F$-statistic framework, usually only a small number of populations are used at once, to e.g.
+test for treeness and find closely related populations.
+
+Even though these two approaches are considered in almost every ancient DNA paper, links between the inferences made from them are usually only compared qualitatively.
+In this paper, our goal is to show that PCA and $F$-statistics are in fact closely related by construction, and use a very similar summary of the data.
+
+\subsection{Introduction to $F$-statistics}
 $F$-statistics have been primarily motivated by trees and admixture graphs \citep{patterson2012, peter2016}, but the calculations hold up in a much wider data space.
@@
 
-In this paper, the main focus is to use this interpretation of $F$-statistics in the context with principal component analysis (PCA), another widely used multivariate analysis tool in population genetics \citep{cavalli-sforza1994, reich2008, novembre2008, mcvean2009}.
-Informally, PCA is a rotation of the data such that the axis of highest variation is the first PC.
-The second PC th
+\subsection{Introduction to PCA}
+
+\subsection{Introduction to MDS}
 
 \section{Relationship of PCA, $F_2$ and Outgroup-$F_3$}
-	Let us assume we have some genotype data $\MX$, which contains data from $n$ on the populations on the rows, and $k$ loci on columns.
-Each population may be represented by a single (pseduo-)haploid, diploid individual, but in general will be an aggregation of one or many samples.
+	Let us assume we have some genotype data $\MX$, which contains allele-frequency data from $n$ populations on the rows, and $k$ loci on columns.
+Each population may be represented by a single haploid, pseudo-haploid or diploid individual.
+Thus, a PCA where each row represents a single individual \cite{patterson2006} is included in this framework.
 
-	To analyze this data, we might perform a PCA.
-A standard way of doing this is to do first center the matrix by subtracting the mean allele frequency from each column and perform an Singular Value Decomposition (SVD):
+	The goal of a PCA is to find a low-dimensional representation of the data that explains most of the variance-structure in the data.
+For this purpose, we rotate the data matrix around its center point, such that the first axis aligns with the major axis of variation.
+
+	\begin{figure}
+		\includegraphics[width=\textwidth]{pca_explanation.png}
+		\caption{Basic Idea of PCA from 2D to 1D representation}
+		\label{fig:pca_explanation}
+	\end{figure}
 
@@
 \begin{equation}
-\MK = \MC\MX\MX^T\MC-\frac{1}{2} \MC \MF \MC = \MP\MP^T
+\MK = \MC\MK\MC = \frac{1}{2}\MC\big[\MF(X_1, \mu) + \MF(X_2, \mu) - \MF(X_1, X_2)\big]\MC=-\frac{1}{2} \MC \MF(X_1, X_2) \MC
 \end{equation}
+
+since $\MC\MF(X_1, X_3)\MC = 0$ for all constant $x_3$.
+In fact, this shows that if we were to mean-center any $F_3$-matrix (a standard step in multidimensional scaling) before decomposition, we retain a PCA.
 
@@
 What remains, is directly proportional to the matrix decomposed under a standard PC.
+
+\subsection{Outgroup-$F_3$-stats based MDS}
+If we have an MDS based on $F_3$,
+
+\begin{align}
+-\MC\MFT(O, X_1; X_2)\MC &= -\frac{1}{2}\MC\big[\MF(X_1, \mu) - \MF(X_2, \mu) + \MF(X_1, X_2)\big]\MC\\
+&=\frac{1}{2} \MC \MF(X_1, X_2) \MC
+\end{align}
+
+In ancient DNA, a MDS has been proposed where
+\begin{enumerate}
+	\item the off-diagonal entries are $1 - F_3(O; X_1, X_2)$ for some outgroup $O$.
+	\item the diagonal is 0.
+\end{enumerate}
+Thus, this matrix differs from that derived above in that one has been added to all off-diagonal entries; and $F_2(X_1,X_2)$ has been subtracted from the diagonal. We have therefore
+\begin{align}
+\MFT^{(Fu)} = \mathbf{1} - \MFT + \mathbf{O},
+\end{align}
+where $\mathbf{1}$ is a matrix of ones and $\mathbf{O}$ is a diagonal matrix with entries
+$$o_{ii} = F_2(O, X_i) -1$$.
+
+Centering then yields
+\begin{align}
+\MC\MFT^{(Fu)}\MC &= \MC\mathbf{1}\MC - \MC\MFT\MC + \MC\mathbf{O}\MC\nonumber\\
+                  &= \MC\MF\MC + \MC\mathbf{O}\MC\\
+\end{align}
 
@@
 What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or the principal components $\MP$. Formally,
-
-\begin{eqnarray}
-F_2(X_i, X_j) &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
-&=& \sum_{l=1}^L \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{kj}\big)^2\nonumber\\
-&=& \sum_{l=1}^L \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
-&=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
-&=& \sum_k \left(\sum_{l=1}^L L_{kl}^2\right) (P_{ik} -P_{jk})^2 + \sum_{k\neq k'}\left(\sum_{l=1}^L L_{kl}L_{k'l}\right) (P_{ik} - P_{jk'})^2\nonumber\\
-&=& \sum_k (P_{ik} - P_{jk})^2
-\end{eqnarray}
-
-In summary, the first row shows that $F_2$ on the centered data will give the same results (as distances are invariant to translations), in the second row we apply the PC-decomposition.
-The third row is obtained from factoring out $L_{lk}$.
-Row four is obtained by multiplying out the sum inside the square term for a particular $l$.
-We have $k$ terms when for $\binom{k}{2}$ terms for different $k$'s.
-Row five is obtained by expanding the outer sum and grouping terms by $k$.The final line is obtained by recognizing that $\ML$ is an orthonormal basis; where dot products of different vectors have lengths zero.
-
-Note that if we estimate $F_2$, unbiased estimators are obtained by subtracting the population-heterozygosities $H_i, H_j$ from the statistic.
-As these are scalars, they do not change above calculation.
 
@@
 
+\appendix
+\section{Derivation}
+\begin{eqnarray}
+F_2(X_i, X_j) &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
+&=& \sum_{l=1}^L \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{kj}\big)^2\nonumber\\
+&=& \sum_{l=1}^L \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
+&=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
+&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + \sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
+&=& \sum_k (P_{ik} - P_{jk})^2
+\end{eqnarray}
+
+In summary, the first row shows that $F_2$ on the centered data will give the same results (as distances are invariant to translations), in the second row we apply the PC-decomposition.
+The third row is obtained from factoring out $L_{lk}$.
+Row four is obtained by multiplying out the sum inside the square term for a particular $l$.
+We have $k$ terms when for $\binom{k}{2}$ terms for different $k$'s.
+Row five is obtained by expanding the outer sum and grouping terms by $k$.The final line is obtained by recognizing that $\ML$ is an orthonormal basis; where dot products of different vectors have lengths zero.
+
+Note that if we estimate $F_2$, unbiased estimators are obtained by subtracting the population-heterozygosities $H_i, H_j$ from the statistic.
+As these are scalars, they do not change above calculation.
 \bibliography{main}
+
 \end{document}

### 2021-04-13 fstats_pca.tex
minor updates

@@
 
+\section{Relationship of PCA, $F_2$ and Outgroup-$F_3$}
+
+The goal of this section is to give a cursory introduction to $F$-statistics, PCA and MDS, and to define notation. More detailed introductions are given in XXXXX.
+
 \subsection{Introduction to PCA}
+	Let us assume we have some genotype data summarized in a matrix $\MX$.
+Each of the $k$ columns contains the allele-frequencies at a single SNP, and we have $n$ rows; one corresponding to either a population or individual, depending on the desired resolution of the analysis.
+As a population may be represented by just one (pseudo-)haploid or diploid individual, there is no conceptual difference between these cases, but I will refer to populations as unit for analysis, for simplicity.
 
-\subsection{Introduction to MDS}
-
-\section{Relationship of PCA, $F_2$ and Outgroup-$F_3$}
-	Let us assume we have some genotype data $\MX$, which contains allele-frequency data from $n$ populations on the rows, and $k$ loci on columns.
-Each population may be represented by a single haploid, pseudo-haploid or diploid individual.
-Thus, a PCA where each row represents a single individual \cite{patterson2006} is included in this framework.
-
-	The goal of a PCA is to find a low-dimensional representation of the data that explains most of the variance-structure in the data.
-For this purpose, we rotate the data matrix around its center point, such that the first axis aligns with the major axis of variation.
+	The goal of a PCA is to find a low-dimensional representation of the data that explains most of the variance-structure in the data (see Fig. \ref{fig:pca_explanation} for an intuitive explanation).
 
@@
 		\includegraphics[width=\textwidth]{pca_explanation.png}
-		\caption{Basic Idea of PCA from 2D to 1D representation}
+		\caption{Basic Idea of PCA from 2D to 1D representation.
+A: Allele frequencies from different populations (blue dots) at two SNPs.
+A PCA is performed by centering the data (B), and rotating it (B) such that the first PC explains the majority of variation in the data, and the second PC is orthognal to the first, and explains the residual.
+A lower-dimensional approximation (in this case 1D) can be achieved by just keeping the first PC (E); which can be translated back to the original data space by inverting the rotation and centering (F).}
 		\label{fig:pca_explanation}
@@
 
+	There are several algorithms that are used to calculate a PCA in practice, the most common one relies on a singular value decomposition. Formally, 
 	\begin{equation}
@@
 
-	Here, $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix and $\mathbf{I}, \mathbf{1}$ are the identity matrix and a matrix of ones, respectively.
-$\MP=\mathbf{U}\MSINGULAR$ is the matrix of principal components (PCs), and, $\ML=\mathbf{V}^T$ is the  matrix of loadings, respectively.
-Occasionally, SNPs are weighted by the square-root of their variance; for now we will assume the SNPs are unweighted, and defer discussion of weighting to a later section.
+	Here, $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$ denoting the identity matrix and a matrix of ones, respectively.
+The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is used to reveal population structure.
+The loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times k$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \cite[e.g]{francois2010}.
 
-	 The principal components $\MP$ are an orthogonal matrix of size $n \times n$ that are useful for understanding population structure, the loadings are an $n \times k$ orthonormal matrix that give the contribution of each SNP along each PC, it is often useful to look for outliers that might be indicative of selection \cite[e.g]{francois2010}.
+	In many  implementations \citep[e.g]{patterson2006}, SNPs are weighted by the inverse of their standard deviation.
+As this weighting makes little difference in practice, I will for now assume that SNPs are unweighted, and defer discussion of weighting to a later section.
 
-	Equivalently, we obtain the PCs by performing an eigendecomposition of the  covariance matrix of $\MY$, denoted as $\MK$:
+	Equivalently, we obtain the PCs by performing an eigendecomposition of the  covariance matrix denoted as $\MK$:
 	 \begin{equation}
  \MK = \MY \MY^T = \mathbf{U}\MEIGEN\mathbf{U}^T = \mathbf{U}\MSINGULAR^2\mathbf{U}^T =\MP\MP^T
-	\end{equation} where $\MEIGEN$ is the diagonal matrix of eigenvalues.  In cases we are interested in the loadings, they can be obtained as 
+	\end{equation} where $\MEIGEN$ is the diagonal matrix with the eigenvalues of $\MK$. 
+	This algorithm does not compute the SNP-loadings. However, the $i$-th row of $L$  can be obtained from $\MP$ and the original data, whenever the eigenvalue $\lambda_i \neq 0$:
 	\begin{equation}
-	\ML = \MEIGEN^{-1}\MP^T\MC\MX
+	\ML_i = \lambda_i^{-1}\MP^T\MC\MX \text{.}
 	\end{equation}
@@
 
-	The entries of the covariance matrix $\MK$ are then
+	\subsection{PCA, $F$-statistics and midpoint rooting.}
+    The construction of the PCs through the covariance matrix $\MK$ is computationally more intensive than SVD, but it yields a simple connection to $F_3$-statistics.
 
+	Notice that the entries of the covariance matrix $\MK$ are
 	\begin{equation}
-	k_{ij} = \sum_l y_{il} y_{jl} = \sum_l (x_{il} - \mu_l)(x_{yl} - \mu_l) = F_3(\MEAN; X_i, X_j)
+	k_{ij} = \sum_l y_{il} y_{jl} = \sum_l (x_{il} - \mu_l)(x_{yl} - \mu_l) = F_3(\MEAN; X_i, X_j)\text{,}
 	\end{equation}
+	where $\MEAN = (\mu_1, \dots \mu_k)$ is the vector of the mean allele frequency at each SNP.
+One interpretation is that $\MEAN$ denotes a ``pivot''-population whose allele frequency vector is the sample mean of the allele frequency space.
+Because of this, it will always lie at the origin of the PC-space, and performing a PCA is equal to a rotation around this ``pivot''-population.
 
-	It is helpful to think of $\MEAN = (\mu_1, \dots \mu_k)$ as a ``fake''-individual whose allele frequency is the sample mean of the allele frequency space, it will always lie at the origin of the PC-space.
+	One implication of this is that $\MEAN$ strongly depends on sample composition.
+If we add many closely related populations (or, in an individual based framework, )This yields an (informal) analogy to the midpoint rooting of a tree in phylogenetics \citep[see e.g.][]{felsenstein2004}.
+In the absence of further information, a sensible choice for the root of a tree is the point that is furthest away from all the tips,
 

### 2021-04-15 fstats_pca.tex
clean up section on pca

@@
 	One implication of this is that $\MEAN$ strongly depends on sample composition.
-If we add many closely related populations (or, in an individual based framework, )This yields an (informal) analogy to the midpoint rooting of a tree in phylogenetics \citep[see e.g.][]{felsenstein2004}.
-In the absence of further information, a sensible choice for the root of a tree is the point that is furthest away from all the tips,
+If we add many closely related populations (or, in an individual based PCA, many individuals from the same population), the PCA will be skewed disproportionally towards this cluster.
 
-	Thus, we can also think of a PCA as performing a decomposition of an Outgroup-$F_3$-matrix, where we set the output to the mean of all marker allele frequencies.
-This is good practice if we do not have any outgroups available.
-However, we often also have outgroup data; for example when studying early European variation, Africans are sometimes a suitable outgroup.
-In this case, it would be consequential to perform an eigendecomposition of a true outgroup $F_3$-matrix.
+	\paragraph{note}
+	This could yield an (informal) analogy to the midpoint rooting of a tree in phylogenetics \citep[see e.g.][]{felsenstein2004}.
+In the absence of further information, a sensible choice for the root of a tree is the point that is furthest away from all the tips.
+However,  if we do have better info (i.e.
+an outgroup), that could yield more sensible results.
+A caveat is that the eigencomposition of a non-centered matrix does not appear to give the desired properties -- would need to dig substantially deeper into how pca maximises variance explains to come up with better algo.
+
+\subsection{PCA vs metric MDS on $F$-statistics}
 
@@
 
-since $\MC\MF(X_1, X_3)\MC = 0$ for all constant $x_3$.
-In fact, this shows that if we were to mean-center any $F_3$-matrix (a standard step in multidimensional scaling) before decomposition, we retain a PCA.
+since $\MC\MF(X_1, c)\MC = 0$ for all constant $c$.
+Thus, by double-centering a matrix of $\MF$-values (multiplied by $-\frac{1}{2}$), we can obtain $\MK$ and hence $\MP$.
+This is exactly what is done in classical multidimensional-scaling, and this derivation is merely a special case of a well-known method.
 
-Thus, when we mean-center the $\MF$-matrix, we subtract the $F_2$ terms in above equations; as they are invariant with respect to the column/ row means.
-What remains, is directly proportional to the matrix decomposed under a standard PC.
-
-\subsection{Outgroup-$F_3$-stats based MDS}
-If we have an MDS based on $F_3$,
-
+However, this result goes even further: consider any $F_3$-matrix, where the ``focal''-population is kept constant:
 \begin{align}
--\MC\MFT(O, X_1; X_2)\MC &= -\frac{1}{2}\MC\big[\MF(X_1, \mu) - \MF(X_2, \mu) + \MF(X_1, X_2)\big]\MC\\
-&=\frac{1}{2} \MC \MF(X_1, X_2) \MC
+\MC\MFT(O; X_1, X_2)\MC &= \frac{1}{2}\MC\big[\MF(X_1, O) + \MF(X_2, O) - \MF(X_1, X_2)\big]\MC\nonumber\\
+ &=-\frac{1}{2} \MC \MF(X_1, X_2) \MC 
 \end{align}
 
-In ancient DNA, a MDS has been proposed where
+this shows that if we were to mean-center \emph{any} $F_3$-matrix (a standard step in multidimensional scaling) before decomposition, we retain a PCA.
+
+One important detail here are the diagonals of $\MFT$; above results assumes that $F_3(O, X_i, X_i) = F_2(O, X_i)$.
+This differs from how MDS has been sometimes applied on ancient DNA-data \citep{fu2016}:
 \begin{enumerate}
@@
 \begin{align}
-\MC\MFT^{(Fu)}\MC &= \MC\mathbf{1}\MC - \MC\MFT\MC + \MC\mathbf{O}\MC\nonumber\\
-                  &= \MC\MF\MC + \MC\mathbf{O}\MC\\
+-\MC\MFT^{(Fu)}\MC &= -\MC\mathbf{1}\MC + \MC\MFT\MC - \MC\mathbf{O}\MC\nonumber\\
+                  &= -\frac{1}{2}\big[\MC\MF\MC +\MC\mathbf{O}\MC\big]\nonumber\\
+                  &= -\frac{1}{2}\big[\MC(\MF + \mathbf{O})\MC\big]
 \end{align}
+
+\paragraph{open questions}: How big is the effect of $\mathbf{O}$. Is it as expected cancelling out differences in sampling time? If so, are outgroup-$F$-stats generally preferable to PCA?
 
@@
 
-\section{Approximating uncertainty on PCA-placement}
-
-\section{Projection using f-stats}
+\subsection{Projection using f-stats}
 Suppose we have a sample $U$ we wish to project onto an existing PCA-basis made from $\MX$, and let us assume we can compute $F_2(U, X_i)$ for all $i$. The ``best'' point i
@@
 
+\paragraph{Potential Application} Use uncertainty on $F_2/F_3$ to propagate uncertainty on PCA-placement
+
 \section{$F$-stats in PCA-space}
-The PCA aims to reveal the axes of major variance. The data is then ``rotated'' such that these axes can be visualized. Usually, only the first few PCs are considered.
+In the previous section, we showed that MDS and PCA are closely related to $F$-statistics, when we consider a matrix of many populations.
+However, one of the main advantages of the $F$-stats framework is that they can be used for specific hypotheses.
+Thus, in this section I am exploring how these hypotheses relate to the placement of populations on a PCA plot.
 
-However, as $F_2$ is a squared Euclidean distance of allele frequencies, it is invariant to rotation and translation.
+Recall that informally, PCA aims to reveal the axes of major variance. To do that, we find an optimal ``rotation'' of the data, such that these axes can be visualized.
+
+As shown by e.g.
+\cite{oteo-garcia2021}, $F$-stats can be thought of as inner products in Euclidean space, and $F_2$ is an (estimated) squared Euclidean distance between two populations in allele frequency space.
+PCA includes a translation and rotation of data, but a distance is invariant to both these operations.
 Hence, neither mean-centering, which is a translation nor PCA-rotation will change $F_2$.
+What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or the principal components $\MP$.
+Formally,
 
-What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or the principal components $\MP$. Formally,
-
-SHORT VERSION
 \begin{align}
@@
  &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 - H_i - H_j  &=& F_2(Y_i, Y_j) \nonumber\\
- &=& \sum_k (P_{ik} - P_{jk})^2 - H_i - H_j &=& F_2(P_i, P_j)
+ &=& \sum_k (P_{ik} - P_{jk})^2 - H_i - H_j &=& F_2(P_i, P_j) \text{,}
 \end{align}
+where $H_i$ and $H_j$ are the bias-correction terms proposed in \cite{reich2009}. A detailed derivation of this is given in Appendix \ref{appendix:fonpc}.
 As $F_3$ and $F_4$ can be written as sums of various $F_2$-terms, analogous relations apply.
@@
 \appendix
-\section{Derivation}
+\section{Derivation}\label{appendix:fonpc}
 \begin{eqnarray}

### 2021-04-16 fstats_pca.tex
emrge

--- 2021-04-13 fstats_pca.tex
@@
 One interpretation is that $\MEAN$ denotes a ``pivot''-population whose allele frequency vector is the sample mean of the allele frequency space.
-Because of this, it will always lie at the origin of the PC-space, and performing a PCA is equal to a rotation around this ``pivot''-population.
+By definition, it lies at the origin of the PC-space, and performing a PCA is equal to a rotation around this ``pivot''-population.
 
@@
 If we add many closely related populations (or, in an individual based framework, )This yields an (informal) analogy to the midpoint rooting of a tree in phylogenetics \citep[see e.g.][]{felsenstein2004}.
-In the absence of further information, a sensible choice for the root of a tree is the point that is furthest away from all the tips,
+In the absence of any further information, a sensible choice for the root of a tree is the point that is furthest away from all the tips.
 

### 2021-04-16 fstats_pca.tex

--- 2021-04-15 fstats_pca.tex
@@
 One interpretation is that $\MEAN$ denotes a ``pivot''-population whose allele frequency vector is the sample mean of the allele frequency space.
-Because of this, it will always lie at the origin of the PC-space, and performing a PCA is equal to a rotation around this ``pivot''-population.
+By definition, it lies at the origin of the PC-space, and performing a PCA is equal to a rotation around this ``pivot''-population.
 
 	One implication of this is that $\MEAN$ strongly depends on sample composition.
-If we add many closely related populations (or, in an individual based PCA, many individuals from the same population), the PCA will be skewed disproportionally towards this cluster.
+If we add many closely related populations (or, in an individual based framework, )This yields an (informal) analogy to the midpoint rooting of a tree in phylogenetics \citep[see e.g.][]{felsenstein2004}.
+In the absence of any further information, a sensible choice for the root of a tree is the point that is furthest away from all the tips.
+
+	Thus, we can also think of a PCA as performing a decomposition of an Outgroup-$F_3$-matrix, where we set the output to the mean of all marker allele frequencies.
+This is good practice if we do not have any outgroups available.
+However, we often also have outgroup data; for example when studying early European variation, Africans are sometimes a suitable outgroup.
+In this case, it would be consequential to perform an eigendecomposition of a true outgroup $F_3$-matrix.
 

### 2021-04-20 fstats_pca.tex
intro

--- 2021-04-15 fstats_pca.tex
@@
 
+	\subsection{$F_2$ error}
+	In most $F$-statistics applications, $F_2$ is \emph{estimated} using the minimum-variance unbiased estimator \citep{reich2009}
+	\begin{subequations}
+
+	\begin{equation}
+	f_2(X_1, X_2) = \frac{1}{L}\sum_l (x_{l1} - x_{l2})^2 - \frac{1}{L}\sum_l\frac{x_{l1} (1-x_{l1})}{n_{l1}-1} - 
+	\frac{1}{L}\sum_l\frac{x_{l2} (1-x_{l2})}{n_{l2}-1} - 
+	\end{equation}
+	in contrast, as shown above, PCA can be thought as a decomposition of a matrix of uncorrected $F_2$-statistics:
+	\begin{equation}
+	F_2(X_1, X_2) = \frac{1}{L}\sum_l(x_{l1} - x_{l2})^2
+	\end{equation}
+	\end{subequations}
+
+	This leads to some issues, for example trying to perform a PCA on the matrix of $f_2$-values is not positive semidefinite, and so some principal components may be imaginary.
+One possible resolution is probabilistic PCA \citep[e.g.][]{engelhardt2010,agrawal2020}.
+
+	\begin{align}
+	    y_{ij} | \MP_{ij}, \epsilon_i &= (\MP\ML)_{ij} + \epsilon_{ij}\nonumber\\
+	    x_i & \sim N(0, \mathbf{I}) \nonumber\\
+	    \epsilon_i &\sim N(0, \sigma^2 \mathbf{I})\nonumber
+	\end{align}
+
 	\subsection{What is a dimension?}

### 2021-05-20 fstats_pca_lewontin.tex
talk

%% Author_tex.tex
%% V1.0
%% 2012/13/12
%% developed by Techset
%%
%% This file describes the coding for rsproca.cls

\documentclass[]{rsos}%%%%where rsos is the template name

%%%% *** Do not adjust lengths that control margins, column widths, etc. ***

%%%%%%%%%%% Defining Enunciations  %%%%%%%%%%%
\newtheorem{theorem}{\bf Theorem}[section]
\newtheorem{condition}{\bf Condition}[section]
\newtheorem{corollary}{\bf Corollary}[section]
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%% Article title to be placed here
\title{Insert the article title here}

\author{%%%% Author details
X. X. First author$^{1}$, X. Second author$^{2}$ and X. Third author$^{3}$}

%%%%%%%%% Insert author address here
\address{$^{1}$First author address\\
$^{2}$Second author address\\
$^{3}$Third author address}

%%%% Subject entries to be placed here %%%%
\subject{xxxxx, xxxxx, xxxx}

%%%% Keyword entries to be placed here %%%%
\keywords{xxxx, xxxx, xxxx}

%%%% Insert corresponding author and its email address}
\corres{Insert corresponding author name\\
\email{xxx@xxxx.xx.xx}}

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here.
The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%% Insert the texts which can accomdate on firstpage in the tag "fmtext" %%%%%

\begin{fmtext}
\section{Insert A head here}
%%%% Insert A head here

This demo file is intended to serve as a ``starter file''
for rsproca journal papers produced under \LaTeX\ using
rsproca.cls v1.5e.

\subsection{Insert B head here}
%%%% Insert B head here
Subsection text here.

\subsubsection{Insert C head here}
%%%% Insert C head here
Subsubsection text here.

\section{Equations}

Sample equations.

%%% Numbered equation
\begin{align}\label{1.1}
\begin{split}
\frac{\partial u(t,x)}{\partial t} &= Au(t,x) \left(1-\frac{u(t,x)}{K}\right)-B\frac{u(t-\tau,x) w(t,x)}{1+Eu(t-\tau,x)},\\
\frac{\partial w(t,x)}{\partial t} &=\delta \frac{\partial^2w(t,x)}{\partial x^2}-Cw(t,x)+D\frac{u(t-\tau,x)w(t,x)}{1+Eu(t-\tau,x)},
\end{split}
\end{align}

\begin{align}\label{1.2}
\begin{split}
\frac{dU}{dt} &=\alpha U(t)(\gamma -U(t))-\frac{U(t-\tau)W(t)}{1+U(t-\tau)},\\
\frac{dW}{dt} &=-W(t)+\beta\frac{U(t-\tau)W(t)}{1+U(t-\tau)}.
\end{split}
\end{align}

%%%% Unnumbered equation
\begin{eqnarray}
\frac{\partial(F_1,F_2)}{\partial(c,\omega)}_{(c_0,\omega_0)} = \left|
\begin{array}{ll}
\frac{\partial F_1}{\partial c} &\frac{\partial F_1}{\partial \omega} \\\noalign{\vskip3pt}
\frac{\partial F_2}{\partial c}&\frac{\partial F_2}{\partial \omega}
\end{array}\right|_{(c_0,\omega_0)}\notag\\
=-4c_0q\omega_0 -4c_0\omega_0p^2 =-4c_0\omega_0(q+p^2)>0.
\end{eqnarray}
\end{fmtext}

%%%%%%%%%%%%%%% End of first page %%%%%%%%%%%%%%%%%%%%%

\maketitle

\section{Enunciations}

\begin{theorem}\label{T0.1}
Assume that $\alpha>0, \gamma>1, \beta>\frac{\gamma+1}{\gamma-1}$.
Then there exists a small $\tau_1>0$, such that for $\tau\in
[0,\tau_1)$, if $c$ crosses $c(\tau)$ from the direction of
to  a small amplitude periodic traveling wave solution of
(2.1), and the period of $(\check{u}^p(s),\check{w}^p(s))$ is
\[
\check{T}(c)=c\cdot \left[\frac{2\pi}{\omega(\tau)}+O(c-c(\tau))\right].
\]
\end{theorem}

\begin{condition}\label{C2.2}
From (0.8) and (2.10), it holds
$\frac{d\omega}{d\tau}<0,\frac{dc}{d\tau}<0$ for $\tau\in
[0,\tau_1)$. This fact yields that the system (2.1) with delay
$\tau>0$ has the periodic traveling waves for smaller wave speed $c$
than that the system (2.1) with $\tau=0$ does. That is, the
delay perturbation stimulates an early occurrence of the traveling waves.
\end{condition}

\section{Figures \& Tables}

The output for figure is:

\begin{figure}[!h]
%\centering\includegraphics[width=2.5in]{xxxxxx.eps}
%%% where xxxxxx name represents "figurename.eps"
\caption{Insert figure caption here}
\label{fig_sim}
\end{figure}

\vspace*{-10pt}

\noindent The output for table is:

\begin{table}[!h]
\caption{An Example of a Table}%%%Table caption goes here
\label{table_example}
\begin{tabular}{llll}%%%The number of columns has to be defined here
\hline
date &Dutch policy &date &European policy \\
\hline
1988 &Memorandum Prevention &1985 &European Directive (85/339) \\
1991--1997 &{\bf Packaging Covenant I} & & \\
1994 &Law Environmental Management &1994 &European Directive (94/62) \\
1997 &Agreement Packaging and Packaging Waste & & \\
1998--2002 &{\bf Packaging Covenant II} & & \\
2003--2005 &{\bf Packaging Covenant III} & & \\
2006--2007 &{\bf Decree on Packaging and paper} & & \\\hline
\end{tabular}
\end{table}%%%End of the table

\section{Conclusion}
The conclusion text goes here.

\section*{Acknowledgment}

Insert the Acknowledgment text here.

%%%%%%%%%% Insert bibliography here %%%%%%%%%%%%%%

\begin{thebibliography}{9}

\bibitem{1} Allwood JM, Cullen JM. 2011 \textit{Sustainable materials:  with both eyes open}.
Cambridge, UK: UIT Cambridge. See \href{http://www.withbotheyesopen.com}{http://www.withbotheyesopen.com}.

\bibitem{2}  MacKay DJC. 2008  \textit{Sustainable energy:  without the hot air}.
 Cambridge, UK: UIT Cambridge. See \href{http://www.withouthotair.com}{http://www.withouthotair.com}.

\bibitem{3} Gallman PG. 2011  \textit{Green alternatives and national energy strategy: the facts
 behind the headlines}.  Baltimore,\ MD: Johns Hopkins University Press.

\bibitem{4} MacKay DJC. 2013.  Solar energy in the context of energy use, energy transportation, and
 energy storage. \textit{Proc. R. Soc. A} \textbf{371}.

\end{thebibliography}

\end{document}

### 2021-05-20 data/prep_data_pca.R

@@ -13,11 +13,17 @@ sample_list_og = c(sample_list, "Mbuti")
 data_loc = '/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_HO_public'
 inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
 
-data_loc = '/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_1240K_public'
-inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
+#data_loc = '/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_1240K_public'
+#inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
 
 
 f2loc1 = 'pca_test1'
+f2loc1 = 'pca_test1c'
 
-f2s = admixtools::extract_f2(data_loc, pops=sample_list_og, outdir=f2loc1, maxmem=64000, blgsize=.5)
+afs = admixtools::extract_afs(data_loc, pops=sample_list_og, 
+                              outdir=f2loc1,format="eigenstrat", 
+                              cols_per_chunk=1000)
+#f2s = admixtools::extract_f2(data_loc, pops=sample_list_og, 
+#                             outdir=f2loc1, maxmem=64000, blgsize=.5,
+                            # format="eigenstrat")
 

### 2021-05-20 data/talk_figs.R

@@ -1,0 +1,34 @@
+require(tidyverse)
+
+a = read_tsv("v44.3_1240K_public.anno") %>%
+    mutate(lat=as.numeric(Lat.), long=as.numeric(Long.)) %>%
+    rename(year=6, age=10) %>%
+    mutate(year=as.numeric(year), age=as.numeric(age)) %>%
+    filter(age>0, !is.na(year), !is.na(lat), !is.na(long)) %>%
+    select(lat, long, year, age)
+
+df_year = a %>% 
+    mutate(year=as.factor(year)) %>% 
+    group_by(year) %>% 
+    tally %>% 
+    mutate(n=cumsum(n))
+
+P = df_year %>% ggplot(aes(x=year, y=n)) + geom_col()
+ggsave("figures/ancient_dna_by_year.png", P, width=5, height=3)
+
+library("rnaturalearth")
+library("rnaturalearthdata")
+
+world <- ne_countries(scale = "medium", returnclass = "sf")
+
+P2 = ggplot() + 
+    geom_sf(data=world, color=NA) + 
+    theme_minimal() + 
+    geom_point(data=a, aes(color=age, x=long, y=lat)) +
+    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+    theme(legend.position = 'none') + 
+    xlab(NULL) + ylab(NULL) +
+    coord_sf(xlim=c(-170, 170), ylim=c(-55, 70)) +
+    scale_color_viridis_c(trans='log')
+
+ggsave("figures/ancient_dna_map.png", P2, width=7, height=3)

### 2021-06-23 fstats_pca.tex
*tex

@@
 \usepackage[utf8]{inputenc}
+\usepackage{fullpage}
+
 %\usepackage[T1]{fontenc}
@@
 
+\section{Results}
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
+	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
+Populations for which $F_3$ is negative are colored in red.
+B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
+C,F: Contributions of PCs 1-10 to each $F_3$-statistic.
+D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$.
+The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population.
+}
+	\label{fig:f3}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{figures/fig_f4_data.pdf}	
+	\caption{\textbf{PCA and $F_4$-statistics} A: Spectrum of select $F_4$-statistics in World data set.
+B: Projection angle representation of $F_4(X, Sardinian; Mozabite, Yoruba)$ (red) and approximations using two (blue) and ten (green) PCs.
+C: Spectrum of select $F-4$-statistics in Westeurasian data set.
+D: Scatterplot of $F_4$-projection on Finnish-Canary Islanders axis and residual PC1.
+	E: $F_4(X, \text{French}, \text{Finnish}, \text{Canary Islander})$ vs.
+prediction using two (blue) and ten(red) PCs.
+F: Percent variance explained for the projection of panel D and the first nine residual PCs.
+	}
+\label{fig:f4}
+\end{figure}
+
 \section{Technical considerations}
@@
 As these are scalars, they do not change above calculation.
+
+\section{Het term}
+\begin{align}
+H_i &= \sum_l X_{ij} (1-X_{ij})\\
+&= \sum_j (Y_{ij} + \mu_j) - (Y_{ij} + \mu_j)^2\\
+&= \sum_j Y_{ij}(1-2\mu_j) - Y_{ij}^2 + \mu_j(1-\mu_j)\\
+&= \sum_j \left[(1-2\mu_j)\sum_k P_{ik}L_{kj} - \left(\sum_k P_{ik}L_{kj}\right)^2 +  \mu_j(1-\mu_j)\right] \\
+&= \sum_k P_{ik} \sum_j (1-2\mu_j)L_{kj} - \sum_k P_{ik}^2 + M \\
+\end{align}
+with $M= \sum_j \mu_j(1-\mu_j)$
+
+because the square terms
+\begin{align}
+S &= \sum_j\left(\sum_k P_{ik}L_{kj}\right)^2 \\
+&=  \sum_j \left(\sum_k P_{ik}L_{kj}\right)\left(\sum_{k'} P_{ik'}L_{k'j}\right)\\
+&= \sum_j \left[\left(\sum_k P_{ik}^2L_{kj}^2\right) + \left(\sum_{k \neq k'} P_{ik} P_{ik'} L_{kj}L_{k'j}\right)\right]\\
+&= \left(\sum_k P_{ik}^2 \underbrace{\left[ \sum_j L_{kj}^2\right]}_1 \right) + \left(\sum_{k \neq k'} P_{ik} P_{ik'} \underbrace{\left[ \sum_j L_{kj}L_{k'j} \right]}_0\right)\\
+&= \sum_k P_{ik}^2 = \sum_k \lambda_k^2 V_{ik}^2
+\end{align}
+
 \bibliography{main}

### 2021-06-23 fstats_pca_lewontin.tex
*tex

@@
-%% Author_tex.tex
-%% V1.0
-%% 2012/13/12
-%% developed by Techset
-%%
-%% This file describes the coding for rsproca.cls
+\documentclass[12pt,a4pape, fullpage]{article}
+\usepackage[
+margin=1.5cm,
+includefoot,
+footskip=30pt,
+]{geometry}
+\usepackage[utf8]{inputenc}
+%\usepackage[T1]{fontenc}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{comment}
+\usepackage{natbib}
+\usepackage{lineno}
 
-\documentclass[]{rsos}%%%%where rsos is the template name
+\bibliographystyle{evolution.bst}
 
-%%%% *** Do not adjust lengths that control margins, column widths, etc. ***
+\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
 
-%%%%%%%%%%% Defining Enunciations  %%%%%%%%%%%
-\newtheorem{theorem}{\bf Theorem}[section]
-\newtheorem{condition}{\bf Condition}[section]
-\newtheorem{corollary}{\bf Corollary}[section]
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\MX}{\mathbf{X}} %uncentered data
+\newcommand{\MC}{\mathbf{C}} %centering
+\newcommand{\MY}{\mathbf{Y}} %centered data
+\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
+\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
+\newcommand{\MP}{\mathbf{P}} % PCs
+\newcommand{\ML}{\mathbf{L}} % loadings
+\newcommand{\MK}{\mathbf{K}} % Kernel
+\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
+\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
 
+\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
 \begin{document}
+\section{Introduction}
+About 15\% of genetic variation in humans can be explained by population structure \cite{lewontin1972, barbujani1997, rosenberg2003}, but the information contained in these 15\% is sufficient to study the genetic diversity and history in great detail \cite{edwards2003}.
+For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \cite{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
 
-%%%% Article title to be placed here
-\title{Insert the article title here}
+In Lewontin's pioneering analysis, he found that less than half (6\%), of that variation could be attributed to the continental-scale groups he called races, it seemed  which he used to claim that ''racial classification is (...) seen to be of virtually no genetic or taxonomic significance``.
 
-\author{%%%% Author details
-X. X. First author$^{1}$, X. Second author$^{2}$ and X. Third author$^{3}$}
+One related question is how discrete human populations are.
+While human genetic differentiation generally increases with geographic distance \cite{ramachandran2005, rosenberg2007}, this increase is not uniform.
+Obstacles to migration,  such as oceans, mountains or deserts do frequently cause discontinuities in population structure \cite{peter2020a}.
+Thus, while barriers to gene flow rarely are absolute, segregation policies by (perceived) ethnic or racial ancestry frequently cause local small-scale population differentiation that persist to the present day.
 
-%%%%%%%%% Insert author address here
-\address{$^{1}$First author address\\
-$^{2}$Second author address\\
-$^{3}$Third author address}
+Thus, it frequently a useful analysis tool to think of populations as discrete units.
+For example, even though the underlying population structure may be continuous, sampling is not; and when quantifying ascertainment and sampling biases, or when discussing population structure it is often helpful to pretend populations are discrete, even though the underlying structure is typically more complex.
+This leads to challenges both in data interpretation and communication, and often researchers will analyze a data set both using methods that assume population structure is discrete, and methods where this assumption does not need to be made.
 
-%%%% Subject entries to be placed here %%%%
-\subject{xxxxx, xxxxx, xxxx}
+One discrete framework for the analysis of human population structure that gained a lot of traction in the last decade are the $F$-statistics \textit{sensu} Patterson \cite{patterson2012, peter2016}.
+This framework treats populations as discrete units in the analysis, and allows for a variety of tests for treeness.
+Using this framwork, the vast majority of present-day human populations are admixed \cite{pickrell2014}.
+Yet, this framework starts with the assumption that admixture is i) rare and ii) discrete.
 
-%%%% Keyword entries to be placed here %%%%
-\keywords{xxxx, xxxx, xxxx}
+However, $F$-statistics are not restricted to discrete populations.
+Indeed, as they can be written as functions of allele frequency variances, or expected pairwise coalescence times, statistics that can be calculated under a wide range of demographic models \cite{peter2016}.
+Indeed, as they reflect inner products, they can be generalized to Euclidean space \cite{oteo2021} (or any Hilbert space, although we won't pursue that here).
+Here, I explore these links between $F$-statistics and Euclidean spaces to establish connections between $F$-statistics and PCA.
+This allows direct interpretation of admixture in scenarios where population structure might not be discrete.
 
-%%%% Insert corresponding author and its email address}
-\corres{Insert corresponding author name\\
-\email{xxx@xxxx.xx.xx}}
+Particularly for the analysis of ancient DNA, two approaches have been proven to be particularly useful: one are global summary analyses, such as Structure \citep{pritchard2000, alexander2009} Principal Component Analysis (PCA) \citep{cavalli-sforza1994, reich2008, novembre2008, mcvean2009} and classical multidimensional scaling (MDS) \cite{fu2016, malaspinas_bammds}.
+Typically, these methods assume that population structure is \emph{sparse}, so that a low-rank approximation with few underlying ``components'' is sufficient to model population structure See e.g.
+\cite{engelhardt2010} for a useful perspective how these approaches are related.
 
-%%%% Abstract text to be placed here %%%%%%%%%%%%
-\begin{abstract}
-The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here.
-The abstract text goes here. The abstract text goes here. The abstract text goes here. The abstract text goes here.
-\end{abstract}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Facing a novel data set, PCA or MDS are often the first analyses (beyond quality controls) a researcher performs, in order to obtain insights in the general population structure they are faced with.
+In order to answer more specific questions and to test specific hypotheses, the $F$-statistic framework of \cite{patterson2012} has been proven particularly powerful (see also \cite{peter2016} for a more gentle introduction).
+In the $F$-statistic framework, usually only a small number of populations are used at once, to e.g.
+test for treeness and find closely related populations.
 
-%%%%%%%%%% Insert the texts which can accomdate on firstpage in the tag "fmtext" %%%%%
+Even though these two approaches are considered in almost every ancient DNA paper, links between the inferences made from them are usually only compared qualitatively.
+In this paper, our goal is to show that PCA and $F$-statistics are in fact closely related by construction, and use a very similar summary of the data.
 
-\begin{fmtext}
-\section{Insert A head here}
-%%%% Insert A head here
+\subsection{Introduction to $F$-statistics}
+$F$-statistics have been primarily motivated by trees and admixture graphs \citep{patterson2012, peter2016}, but the calculations hold up in a much wider data space.
+In particular,  \cite{oteo-garcia2021} provides a thorough introduction to interpreting $F$-statistics in the \emph{data space} $\mathbb{R}^k$.
+Their work builds much of the foundation of this discussion, by demonstrating analogies to classical geometry.
+A brief summary of their key results: A population's allele frequencies can be thought of as vector in $\mathbb{R}^k$.
+Then, $F_2(X_1, X_2) = \normsq{X_1 - X_2}$ is the squared Euclidean distance between the populations with vectors $X_1$ and $X_2$, and $F_4(X_1, X_2 ; X_3, X_4) = \langle X_1 - X_2, X_3 - X_4 \rangle$ is the inner (scalar) product between these two vectors.
+Here, I will mainly use the $F$-statistic notation, but use the geometric notation where convenient.
 
-This demo file is intended to serve as a ``starter file''
-for rsproca journal papers produced under \LaTeX\ using
-rsproca.cls v1.5e.
+\section{Relationship of PCA, $F_2$ and Outgroup-$F_3$}
 
-\subsection{Insert B head here}
-%%%% Insert B head here
-Subsection text here.
+The goal of this section is to give a cursory introduction to $F$-statistics, PCA and MDS, and to define notation.
+A more detailed technical introduction is given in XXXXX, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
 
-\subsubsection{Insert C head here}
-%%%% Insert C head here
-Subsubsection text here.
+\subsection{Introduction to PCA}
+	Let us assume we have some genotype data summarized in a matrix $\MX$, where the entry $x_{ij}$ is the allele frequency of the $i$-th population at the $j$-th genotype.
+If we have $k$ SNPs and $n$ populations, $\MX$ will have dimension $n \times k$.
 
-\section{Equations}
+    As a population may be represented by just one (pseudo-)haploid or diploid individual, there is no conceptual difference between these cases and I will refer to populations as unit for analysis, for simplicity.
 
-Sample equations.
+    Since the allele frequencies are between zero and one, we can interpret each row $x_i$ of $\MX$ as a vector in $[0, 1]^k$, the \emph{data space} of all possible allele frequencies on our markers.
 
-%%% Numbered equation
-\begin{align}\label{1.1}
-\begin{split}
-\frac{\partial u(t,x)}{\partial t} &= Au(t,x) \left(1-\frac{u(t,x)}{K}\right)-B\frac{u(t-\tau,x) w(t,x)}{1+Eu(t-\tau,x)},\\
-\frac{\partial w(t,x)}{\partial t} &=\delta \frac{\partial^2w(t,x)}{\partial x^2}-Cw(t,x)+D\frac{u(t-\tau,x)w(t,x)}{1+Eu(t-\tau,x)},
-\end{split}
+	The goal of a PCA is to find a low-dimensional representation of this data space that retains most of the data (see Fig. \ref{fig:pca_explanation} for an intuitive explanation).
+
+	\begin{figure}
+		\includegraphics[width=\textwidth]{pca_explanation.png}
+		\caption{Basic Idea of PCA from 2D to 1D representation.
+A: Allele frequencies from different populations (blue dots) at two SNPs.
+A PCA is performed by centering the data (B), and rotating it (B) such that the first PC explains the majority of variation in the data, and the second PC is orthognal to the first, and explains the residual.
+A lower-dimensional approximation (in this case 1D) can be achieved by just keeping the first PC (E); which can be translated back to the original data space by inverting the rotation and centering (F).}
+		\label{fig:pca_explanation}
+	\end{figure}
+
+	There are several algorithms that are used to calculate a PCA in practice, the most common one relies on a singular value decomposition.
+In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
+	\begin{equation}
+	y_{il} = x_{il} - \mu_l
+	\end{equation}
+	where $\mu_l$ is the mean allele frequency at the $l$-th locus.
+
+	PCA can then be written as
+
+	\begin{equation}
+	\MY = \MC\MX = \mathbf{U} \MSINGULAR \mathbf{V}^T = \MP\ML\text{,}
+	\end{equation}
+
+	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively.
+The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is used to reveal population structure.
+The loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times k$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \cite[e.g][]{francois2010}.
+
+	In many  implementations \citep[e.g]{patterson2006}, SNPs are weighted by the inverse of their standard deviation.
+As this weighting makes little difference in practice, I will for now assume that SNPs are unweighted, and defer discussion of weighting to a later section.
+
+	Equivalently, we obtain the PCs by performing an eigendecomposition of the  covariance matrix denoted as $\MK$:
+	 \begin{equation}
+ \MK = \MY \MY^T = \mathbf{U}\MEIGEN\mathbf{U}^T = \mathbf{U}\MSINGULAR^2\mathbf{U}^T =\MP\MP^T
+	\end{equation} where $\MEIGEN$ is the diagonal matrix with the eigenvalues of $\MK$. 
+	This algorithm does not compute the SNP-loadings. However, the $i$-th row of $L$  can be obtained from $\MP$ and the original data, whenever the eigenvalue $\lambda_i \neq 0$:
+	\begin{equation}
+	\ML_i = \lambda_i^{-1}\MP^T\MC\MX \text{.}
+	\end{equation}
+
+\section{$F$-statistics in PCA-space}
+As shown by e.g.
+\cite{oteo-garcia2021}, $F$-statistics can be thought of as inner products in Euclidean space, and $F_2$ is an (estimated) squared Euclidean distance between two populations in allele frequency space.
+By performing a PCA, we just translate and rotate our data, but Euclidean distances and dot products are both invariant under both these operations.
+Hence, neither mean-centering (a translation) nor PCA (a rotation) will change $F_2$.
+What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or the principal components $\MP$.
+Formally,
+
+\begin{align}
+F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2 - H_i - H_j &&\nonumber\\ 
+ &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 - H_i - H_j  &=& F_2(Y_i, Y_j) \nonumber\\
+ &=& \sum_k (P_{ik} - P_{jk})^2 - H_i - H_j &=& F_2(P_i, P_j) \text{,}
+\end{align}
+where $H_i$ and $H_j$ are the bias-correction terms proposed in \cite{reich2009}. A detailed derivation of this is given in Appendix \ref{appendix:fonpc}.
+As $F_3$ and $F_4$ can be written as sums of $F_2$-terms, analogous relations apply.
+
+\subsection{$F$-stats in 2-dimensional PC-space}
+The transformation derived in the previous section allows us to consider the geometry of $F$-statistics in PCA-space.
+The relationships we will discuss formally only hold if we use all $n-1$ PCs.
+However, the appeal of PCA is that frequently, only a very small number $K \ll n$ of PCS contain most information that is relevant for population structure.
+
+Here, we start by discussing 2-dimensional spaces.
+This is useful for two reasons: for one, the geometry is simpler and we can think of circles and squares as opposed to $n$-balls and other high-dimensional geometric objects and thus build intuition.
+Second, in many applications it is argued that a 2-dimensional approximation is sufficient to explain the major components of population structure \cite{novembre2008}.
+In this case, the results here will hold under the same approximation assumptions in low-dimensional PCs; if they differ substantially from each other, it is likely that not sufficiently many PCs were considered.
+
+\subsubsection{$F_2$ in PC-space}
+The $F_2$-statistic is an estimate of the squared Euclidean distance is the easiest to understand, it corresponds directly to the squared distance in PCA-space.
+This matches our intuition that closely related populations (which have low $F_2$) will be close to each other on a PCA-plot.
+
+\subsubsection{$F_3$ and circles}
+The $F_3$-statistic becomes more interesting; as outlines above we either think of $F_3$ as ``outgroup''-$F$-stats or as admixture $F$-stats.
+In the admixture case, we may ask the following question: given two source populations $X_1$, $X_2$, where would admixed populations on a PCA plot lie?
+From theory, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \cite{brisbin2012, mcvean2009}.
+
+Formally, we would reject admixture if $F_3$ is negative, i.e. we are looking for the space
+\begin{eqnarray}
+2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
+      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} \nonumber\\
+      &<&0
+\end{eqnarray}
+By the Pythagorean theorem, $F_3 = 0 $ iff $X_1, X_2$ and $X_x$ form a right-angled triangle.
+Hence, the region where $F_3$ is zero is the circle with diameter through $X_1$ and $X_2$.
+If $X_x$ lies inside this circle, the angle is obtuse and $F_3$ is negative, otherwise it will be positive.
+Similarly, if we fix $X_1$ and $X_2$ and ask where on a 2D-PCA-plot $X_2$ would lie, this space is defined by all the points for which the angle between $X_1 X_x$ and $X_2 X_x$ is obtuse.
+
+This highlights a potential identifiability issue with $F_3$: Since all values of $F_3$ that result in the same projection will give the same value; and multiple admixture events may result in the same $F_3$-value.
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{dummy_pca.png}
+	\caption{\textbf{Geometric representation of $F$-statistics on 2D-PCA-plot.} A: $F_2$ represents the squared Euclidean distance between two points in PC-space.
+B: Admixture-$F_3(X_x; X_1, X_2)$ is negative if $X_x$ lies in the circle specified by the diameter $X_2-X_1$}.
+C: $F_3(X_x; X_1, X_2)$ is negative given $X_1, X_x$ if $X_2$ is in the gray space.
+D: Outgroup-$F_3$ reflects the projection of $X_2 - X_O$ on $X_1 - X_O$.
+E: $F_4$ is the projection of $X_3 - X_4$ on $X_1-X_2$.
+F: If $X_x$ is admixed between $X_1$ and $X_2$, the admixture proportions will be projected.
+\end{figure}
+
+\subsubsection{$F_4$ and right angles}
+The inner-product-interpretation of $F_4$ is similar to that of $F_3$, with the change that the two vectors we consider do not involve the same population.
+However, a finding of $F_4(X_1, X_2; X_3, X_4) = \langle X_1 - X_2, X_3 - X4 \rangle = 0$ similarly implies that the two vectors are orthogonal, and a non-zero value reflects the projection of one vector on the other.
+
+\subsubsection{$F_4$-ratio}
+\begin{eqnarray}
+\frac{F_4(X_I, X_O; X_X, X_1)}{F_4(X_I, X_O; X_2, X_1)} &=& \frac{\norm{X_I-X_O}\norm{X_X-X_1}\cos(\alpha)}{\norm{X_I-X_O}\norm{X_2-X_1}\cos(\beta)}\nonumber\\
+&=&\frac{\norm{X_X-X_1}\cos(\alpha)}{\norm{X_2-X_1}\cos(\beta)}\nonumber\\
+&=& \frac{\norm{X_X' - X_1'}}{\norm{X_2' - X_1'}}
+\end{eqnarray}
+where $\alpha$ and $\beta$ are the angles between vectors, and $X_i'$ is the projection of $X_i$ on $X_I-X_O$.
+
+Conjecture: Thus, we are measuring the distances between the admixing populations on the projected on the axis between $X_I$ and $X_O$.
+This ought to be valid only if $\langle X_1 - X_1', X_2 - X_2' \rangle$ are orthogonal to each other, and to $X_OX_I$, i.e.
+$F_4(X_1, X_1', X_2, X_2') = 0$
+
+\subsection{Higher-Dimensional Spaces}
+
+\section{Trees and admixture graphs in PCA-space}
+\subsection{Trees}
+Evolutionary trees are fundamental in phylogenetic analyses, as they, on a large, scale, approximate how taxa diversify.
+Within a species, applying trees is also very common, but more problematic as populations frequently do not evolve as discrete lineages; instead, they admix and diversify as much more continuous processes.
+This is largely due to the time-scales involved, speciation events that give rise to trees might often be similarly messy, but from a distance of millions of years these issues might disappear.
+
+Thus, when estimating trees from population genetic data, we must be very careful about whether the data is actually consistent with a tree, or belongs to some wider class of model.
+
+Trees can be thought of as a collection of orthogonal dimensions; as drift on each branch is independent from every other branch. Thus, each sample is only 
+\begin{enumerate}
+	\item Trees
+	\item Admixture Graphs
+	\item Treelets
+	\item simple trees, admixture graph
+\end{enumerate}
+
+\section{Technical considerations}
+	\subsection{SNP weighting}
+	It is clear that weighting SNP will have some effect on the resulting PCAs. Upweighting rare variants e.g. will emphasis recent events, as rare variance in the sample are more likely to be recent.
+
+	\subsection{Missing data}
+
+	\subsection{$F_2$ error}
+	In most $F$-statistics applications, $F_2$ is \emph{estimated} using the minimum-variance unbiased estimator \citep{reich2009}
+	\begin{subequations}
+
+	\begin{equation}
+	f_2(X_1, X_2) = \frac{1}{L}\sum_l (x_{l1} - x_{l2})^2 - \frac{1}{L}\sum_l\frac{x_{l1} (1-x_{l1})}{n_{l1}-1} - 
+	\frac{1}{L}\sum_l\frac{x_{l2} (1-x_{l2})}{n_{l2}-1} - 
+	\end{equation}
+	in contrast, as shown above, PCA can be thought as a decomposition of a matrix of uncorrected $F_2$-statistics:
+	\begin{equation}
+	F_2(X_1, X_2) = \frac{1}{L}\sum_l(x_{l1} - x_{l2})^2
+	\end{equation}
+	\end{subequations}
+
+	This leads to some issues, for example trying to perform a PCA on the matrix of $f_2$-values is not positive semidefinite, and so some principal components may be imaginary.
+One possible resolution is probabilistic PCA \citep[e.g.][]{engelhardt2010,agrawal2020}.
+
+	\begin{align}
+	    y_{ij} | \MP_{ij}, \epsilon_i &= (\MP\ML)_{ij} + \epsilon_{ij}\nonumber\\
+	    x_i & \sim N(0, \mathbf{I}) \nonumber\\
+	    \epsilon_i &\sim N(0, \sigma^2 \mathbf{I})\nonumber
 	\end{align}
 
-\begin{align}\label{1.2}
-\begin{split}
-\frac{dU}{dt} &=\alpha U(t)(\gamma -U(t))-\frac{U(t-\tau)W(t)}{1+U(t-\tau)},\\
-\frac{dW}{dt} &=-W(t)+\beta\frac{U(t-\tau)W(t)}{1+U(t-\tau)}.
-\end{split}
+	\subsection{qpADM}
+	In \cite{haak2015}, qpADM, a procedure to estimate admixture proportions has been proposed. qpADM aims to solve equations of the form
+
+	\begin{align}
+	\langle P_X - A, B -C \rangle &= \sum_i  \alpha_i\langle R_i - A, B - C \rangle \nonumber\\
+	&= \left\langle \sum_i  \alpha_i R_i - A, B - C \right\rangle
 	\end{align}
 
-%%%% Unnumbered equation
+	\subsection{What is a dimension?}
+	In both the PCA and $F$-statistic framework, a population at  a particular point in time can be thought of as a single point in allele-frequency space, given by the $k$-dimensional vector $v_0$ of allele frequencies at the $k$ SNPs in that population.
+If this population evolves for some time in isolation, allele frequencies will change due to genetic drift from $v_0$ to some other point $v_1$.
+Likewise, a second population with frequency $w_0$ will move to $w_1$.
+Crucially, if these populations do not interact, the changes in allele frequency, $v_1 - v_0$ and $w_1 - w_0$ will be uncorrelated \cite{patterson2012}.
+Thus, if we have two populations that descend from the same ancestral population in isolation, they can be thought of as evolving along orthognal dimensions from the same point.
+This argument is at the foundation of F-statistics.
+
+	\section{outtakes}
+	PCA from $\MX$
+	\begin{equation}
+	\MK = \MY \MY^T = \MC\MX\MX^T \MC = \MP\MP^T
+	\end{equation}
+
+\section{Discussion}
+The fa
+
+\appendix
+\section{Derivation}\label{appendix:fonpc}
 \begin{eqnarray}
-\frac{\partial(F_1,F_2)}{\partial(c,\omega)}_{(c_0,\omega_0)} = \left|
-\begin{array}{ll}
-\frac{\partial F_1}{\partial c} &\frac{\partial F_1}{\partial \omega} \\\noalign{\vskip3pt}
-\frac{\partial F_2}{\partial c}&\frac{\partial F_2}{\partial \omega}
-\end{array}\right|_{(c_0,\omega_0)}\notag\\
-=-4c_0q\omega_0 -4c_0\omega_0p^2 =-4c_0\omega_0(q+p^2)>0.
+F_2(X_i, X_j) &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
+&=& \sum_{l=1}^L \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{kj}\big)^2\nonumber\\
+&=& \sum_{l=1}^L \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
+&=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
+&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + \sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
+&=& \sum_k (P_{ik} - P_{jk})^2
 \end{eqnarray}
-\end{fmtext}
 
-%%%%%%%%%%%%%%% End of first page %%%%%%%%%%%%%%%%%%%%%
+In summary, the first row shows that $F_2$ on the centered data will give the same results (as distances are invariant to translations), in the second row we apply the PC-decomposition.
+The third row is obtained from factoring out $L_{lk}$.
+Row four is obtained by multiplying out the sum inside the square term for a particular $l$.
+We have $k$ terms when for $\binom{k}{2}$ terms for different $k$'s.
+Row five is obtained by expanding the outer sum and grouping terms by $k$.The final line is obtained by recognizing that $\ML$ is an orthonormal basis; where dot products of different vectors have lengths zero.
 
-\maketitle
-
-\section{Enunciations}
-
-\begin{theorem}\label{T0.1}
-Assume that $\alpha>0, \gamma>1, \beta>\frac{\gamma+1}{\gamma-1}$.
-Then there exists a small $\tau_1>0$, such that for $\tau\in
-[0,\tau_1)$, if $c$ crosses $c(\tau)$ from the direction of
-to  a small amplitude periodic traveling wave solution of
-(2.1), and the period of $(\check{u}^p(s),\check{w}^p(s))$ is
-\[
-\check{T}(c)=c\cdot \left[\frac{2\pi}{\omega(\tau)}+O(c-c(\tau))\right].
-\]
-\end{theorem}
-
-\begin{condition}\label{C2.2}
-From (0.8) and (2.10), it holds
-$\frac{d\omega}{d\tau}<0,\frac{dc}{d\tau}<0$ for $\tau\in
-[0,\tau_1)$. This fact yields that the system (2.1) with delay
-$\tau>0$ has the periodic traveling waves for smaller wave speed $c$
-than that the system (2.1) with $\tau=0$ does. That is, the
-delay perturbation stimulates an early occurrence of the traveling waves.
-\end{condition}
-
-\section{Figures \& Tables}
-
-The output for figure is:
-
-\begin{figure}[!h]
-%\centering\includegraphics[width=2.5in]{xxxxxx.eps}
-%%% where xxxxxx name represents "figurename.eps"
-\caption{Insert figure caption here}
-\label{fig_sim}
-\end{figure}
-
-\vspace*{-10pt}
-
-\noindent The output for table is:
-
-\begin{table}[!h]
-\caption{An Example of a Table}%%%Table caption goes here
-\label{table_example}
-\begin{tabular}{llll}%%%The number of columns has to be defined here
-\hline
-date &Dutch policy &date &European policy \\
-\hline
-1988 &Memorandum Prevention &1985 &European Directive (85/339) \\
-1991--1997 &{\bf Packaging Covenant I} & & \\
-1994 &Law Environmental Management &1994 &European Directive (94/62) \\
-1997 &Agreement Packaging and Packaging Waste & & \\
-1998--2002 &{\bf Packaging Covenant II} & & \\
-2003--2005 &{\bf Packaging Covenant III} & & \\
-2006--2007 &{\bf Decree on Packaging and paper} & & \\\hline
-\end{tabular}
-\end{table}%%%End of the table
-
-\section{Conclusion}
-The conclusion text goes here.
-
-\section*{Acknowledgment}
-
-Insert the Acknowledgment text here.
-
-%%%%%%%%%% Insert bibliography here %%%%%%%%%%%%%%
-
-\begin{thebibliography}{9}
-
-\bibitem{1} Allwood JM, Cullen JM. 2011 \textit{Sustainable materials:  with both eyes open}.
-Cambridge, UK: UIT Cambridge. See \href{http://www.withbotheyesopen.com}{http://www.withbotheyesopen.com}.
-
-\bibitem{2}  MacKay DJC. 2008  \textit{Sustainable energy:  without the hot air}.
- Cambridge, UK: UIT Cambridge. See \href{http://www.withouthotair.com}{http://www.withouthotair.com}.
-
-\bibitem{3} Gallman PG. 2011  \textit{Green alternatives and national energy strategy: the facts
- behind the headlines}.  Baltimore,\ MD: Johns Hopkins University Press.
-
-\bibitem{4} MacKay DJC. 2013.  Solar energy in the context of energy use, energy transportation, and
- energy storage. \textit{Proc. R. Soc. A} \textbf{371}.
-
-\end{thebibliography}
+Note that if we estimate $F_2$, unbiased estimators are obtained by subtracting the population-heterozygosities $H_i, H_j$ from the statistic.
+As these are scalars, they do not change above calculation.
+\bibliography{main}
 

### 2021-06-23 data/prep_data_pca.R

@@ -10,20 +10,20 @@ list1 = read_table("WestEurasian_PCA.list", col_names=F)[,1] %>% unlist %>%
 sample_list = list1[!list1 %in% MISSING]
 sample_list_og = c(sample_list, "Mbuti")
 
+sample_list %>% as.data.frame %>% write_csv("subdata/westeurasian1.poplist", col_names=F)
+
 data_loc = '/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_HO_public'
 inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
 
-#data_loc = '/mnt/ancient/ModernHuman/ReichLab/reich_public_geno_v44.3/v44.3_1240K_public'
-#inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
 
 
-f2loc1 = 'pca_test1'
-f2loc1 = 'pca_test1c'
+f2loc1 = 'westeurasian1'
+#f2loc1 = 'pca_test1c'
 
-afs = admixtools::extract_afs(data_loc, pops=sample_list_og, 
-                              outdir=f2loc1,format="eigenstrat", 
+afs = admixtools::extract_afs(data_loc, pops=sample_list, 
+                              outdir=glue("{f2loc1}_afs"),format="eigenstrat", 
                               cols_per_chunk=1000)
-#f2s = admixtools::extract_f2(data_loc, pops=sample_list_og, 
-#                             outdir=f2loc1, maxmem=64000, blgsize=.5,
-                            # format="eigenstrat")
+#f2s = admixtools::extract_f2(data_loc, pops=sample_list, 
+#                             outdir=f2loc1, maxmem=64000, blgsize=.05,
+#                             overwrite=T)
 

### 2021-06-23 data/f3_fig.R
init data analysis

@@ -1,0 +1,120 @@
+source("fscripts.R")
+library(ggpubr)
+BTHEME = theme_classic() + theme(legend.position="none")
+
+if(T){
+f2s = admixtools::read_f2("westeurasian1/")
+pcmat = pca_from_f2s(f2s)
+pcs = pca_from_pcmat(pcmat)
+idx = "Basque"
+idy = "Turkish"
+
+sample_list = pcs$pop 
+
+f3 = f3(f2s, pop3=idx, pop2=idy, pop1=sample_list) #admixture
+f3pc = f3_from_pc_matrix(pcmat, idx, idy)
+#f3 = f3(f2s, pop1=idx, pop2=idy, pop3=sample_list) #outgroup
+f4 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+
+circle = pcs %>% filter(pop %in% c(idx, idy)) %>% select(-pop)
+circle_df = colMeans(circle) %>% t %>% as.data.frame
+circle_df$radius = (dist(circle) %>% as.matrix)[1,2] / 2
+circle_df$r0 = (dist(circle[,1:2]) %>% as.matrix)[1,2] / 2
+
+pcs2 = pcs %>% left_join(f3, by=c(pop="pop1")) %>% 
+    mutate(col=as.factor(cut(est, c(-1,-1e-10,1e-10,2), labels=F) ))
+P1 = ggplot() + 
+    geom_circle(aes(x0=PC1, y0 =PC2, r=radius), data=circle_df, color='lightgray', fill='lightgray') +
+#    geom_circle(aes(x0=PC1, y0 =PC2, r=r0), data=circle_df, color='darkgray', fill='darkgray') +
+    geom_text(aes(x=PC1, y=PC2, label=pop, color=col), data=pcs2) +
+    coord_fixed()  + BTHEME
+
+f3pc = f3_from_pc_matrix(pcmat, idx, idy) %>% 
+    as.data.frame %>% rownames_to_column("pop1") %>% 
+    left_join(f3, .) %>% 
+    mutate(PC1_2=rowSums(across(PC1:PC2)), PC1_10=rowSums(across(PC1:PC10)))
+
+P2 = f3pc %>% select(pop1, est, PC1_2, PC1_10) %>% 
+    pivot_longer(starts_with("PC"), names_to='PC', values_to="f3") %>% 
[…]

### 2021-06-23 data/f4_fig.R

@@ -1,0 +1,246 @@
+source("fscripts.R")
+library(glue)
+BTHEME = theme_classic() + theme(legend.position="none")
+
+if(T){
+f2s = admixtools::read_f2("westeurasian1/")
+pcmat = pca_from_f2s(f2s)
+pcs = pca_from_pcmat(pcmat)
+idx = "Finnish"
+idy = "Canary_Islander"
+idz = "French"
+ids = c(idx, idy)
+
+sample_list = pcs$pop 
+
+f4 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+f4mat = f4 %>% 
+    select(pop1, pop2, est) %>% 
+    pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>% 
+    column_to_rownames('pop1')
+f4pc = f4_from_pc_matrix2(pcmat, idx, idy) %>%
+    left_join(f4, .) %>% 
+    filter(! pop1 %in% ids, ! pop2 %in% ids) %>%
+    mutate(PC1_2=rowSums(across(PC1:PC2)), PC1_10=rowSums(across(PC1:PC10)))
+v = pcmat[ids,] %>% diff
+pmat = t(v) %*% v / (v %*% t(v))[1,1]
+qmat = diag(length(v)) - pmat
+a = pcmat %*% pmat
+b = pcmat %*% qmat
+A = t(t(a) - a[idz,])
+B = t(t(b) - b[idz,])
+E = eigen(b %*% t(b))
+K = t(t(E$vectors) * sqrt(pmax(E$values, 0)))
+K = K %>% as.data.frame %>% mutate(pop = rownames(B))
+
+EA = eigen(a %*% t(a))
+KA = t(t(EA$vectors) * sqrt(pmax(EA$values, 0)))
+KA = KA %>% as.data.frame %>% mutate(pop = rownames(B))
+X = KA %>% select(pop, A=V1) %>% left_join(K)
+
+P3 = X %>% 
+    ggplot(aes(x=A, y=V1, label=pop, color=V2)) + 
+    geom_text(size=3) + 
+    coord_fixed() + 
+    BTHEME + 
+    scale_color_viridis_c() + 
+    xlab("proj (Finnish, Canary Islander)") + 
+    ylab("Residual PC1")
+
+f42 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=idz) %>% arrange(pop1, pop2)
[…]

### 2021-06-23 data/f_and_pca.R

@@ -1,0 +1,60 @@
+require(admixtools)
+require(tidyverse)
+require(Matrix)
+
+f3 = f3(f2s, pop3="Basque", pop2="Turkish", pop1=sample_list)
+f4 = f4(f2s, pop2="Basque", pop3="Scottish", pop4="Turkish", pop1=sample_list, pop2) %>% arrange(pop1)
+f4.2 = f4(f2s, pop4="Basque", pop3="Turkish", pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+
+f2mat = f2 %>%
+    mutate(tmp=pop2, pop2=pop1, pop1=tmp) %>% 
+    select(-tmp) %>% bind_rows(f2) %>% 
+    select(-se) %>% 
+    pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>%
+    arrange(pop1) %>%
+    column_to_rownames('pop1') %>% as.matrix
+
+f2mat %>% cmdscale(k=61) %>% as.data.frame %>% rename(PC1=V1, PC2=V2) %>% rownames_to_column("pop")
+
+cmat = t(t(f2mat - rowMeans(f2mat) ) - colMeans(f2mat)) + mean(f2mat)
+
+E = eigen(nearPD(-cmat/2)[[1]])
+
+pcs = t(t(E$vectors) * sqrt(pmax(E$values, 0)))
+rownames(pcs) = rownames(f2mat)
+pcmat = pcs
+pcs = pcs %>% as.data.frame %>% rownames_to_column("pop")
+names(pcs)[-1] = sprintf("PC%d", 1:(ncol(pcs) - 1))
+
+circle = pcs %>% filter(pop %in% c("Basque", "Turkish")) %>% select(PC1, PC2)
+radius = (dist(circle) %>% as.matrix)[1,2] / 2
+origin = colMeans(circle)
+circle_df=tibble(x0=origin[1], y0=origin[2], radius=radius, pop="0")
+
+circle34 = pcs %>% filter(pop %in% c("Basque", "Turkish")) %>% select(PC3, PC4)
+radius = (dist(circle34) %>% as.matrix)[1,2] / 2
+origin = colMeans(circle34)
+circle_df34=tibble(x0=origin[1], y0=origin[2], radius=radius, pop="0")
+
+f3_from_pc_matrix <-function(pcmat, idx="Basque", idy="Turkish"){
+    pcmat = t(pcmat)
[…]

### 2021-06-23 data/f_and_pca_worldfoci1.R

@@ -1,0 +1,87 @@
+require(admixtools)
+require(tidyverse)
+require(Matrix)
+
+f2s = admixtools::read_f2("worldfoci1/")
+
+pca_from_f2s <- function(f2s){
+    f2 = f2(f2s)
+    n = dim(f2s)[1]
+    f2mat = f2 %>%
+        mutate(tmp=pop2, pop2=pop1, pop1=tmp) %>% 
+        select(-tmp) %>% bind_rows(f2) %>% 
+        select(-se) %>% 
+        pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>%
+        arrange(pop1) %>%
+        column_to_rownames('pop1') %>% as.matrix
+
+    f2mat %>% cmdscale(k=n-1) %>% as.data.frame %>% rename(PC1=V1, PC2=V2) %>% rownames_to_column("pop")
+
+    cmat = t(t(f2mat - rowMeans(f2mat) ) - colMeans(f2mat)) + mean(f2mat)
+
+    E = eigen(nearPD(-cmat/2)[[1]])
+
+    pcs = t(t(E$vectors) * sqrt(pmax(E$values, 0)))
+    rownames(pcs) = rownames(f2mat)
+    pcmat = pcs
+    pcs = pcs %>% as.data.frame %>% rownames_to_column("pop")
+    names(pcs)[-1] = sprintf("PC%d", 1:(ncol(pcs) - 1))
+    return(pcs)
+}
+
+f3 = f3(f2s, pop3="Basque", pop2="Turkish", pop1=sample_list)
+f4 = f4(f2s, pop2="Basque", pop3="Scottish", pop4="Turkish", pop1=sample_list, pop2) %>% arrange(pop1)
+f4.2 = f4(f2s, pop4="Basque", pop3="Turkish", pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+
+f2mat = f2 %>%
+    mutate(tmp=pop2, pop2=pop1, pop1=tmp) %>% 
+    select(-tmp) %>% bind_rows(f2) %>% 
+    select(-se) %>% 
+    pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>%
+    arrange(pop1) %>%
+    column_to_rownames('pop1') %>% as.matrix
+
[…]

### 2021-06-23 data/f_and_pca_worldfoci2.R

@@ -1,0 +1,80 @@
+require(admixtools)
+require(tidyverse)
+require(Matrix)
+
+
+pca_from_f2s <- function(f2s){
+    f2 = f2(f2s)
+    n = dim(f2s)[1]
+    f2mat = f2 %>%
+        mutate(tmp=pop2, pop2=pop1, pop1=tmp) %>% 
+        select(-tmp) %>% bind_rows(f2) %>% 
+        select(-se) %>% 
+        pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>%
+        arrange(pop1) %>%
+        column_to_rownames('pop1') %>% as.matrix
+
+    cmat = t(t(f2mat - rowMeans(f2mat) ) - colMeans(f2mat)) + mean(f2mat)
+
+    E = eigen(nearPD(-cmat/2)[[1]])
+
+    pcs = t(t(E$vectors) * sqrt(pmax(E$values, 0))) 
+    rownames(pcs) = rownames(f2mat)
+    return(pcs)
+}
+
+pca_from_pcmat <- function(pcs){
+    pcs = pcs %>% as.data.frame %>% rownames_to_column("pop")
+    names(pcs)[-1] = sprintf("PC%d", 1:(ncol(pcs) - 1))
+    return(pcs)
+}
+
+
+f3_from_pc_matrix <-function(pcmat, idx="Basque", idy="Turkish"){
+    pcmat = t(pcmat)
+    f3mat = t((pcmat - pcmat[,idx]) * (pcmat - pcmat[,idy]))
+    return(f3mat)
+}
+f4_from_pc_matrix_single <-function(pcmat, A, B, C, D){
+    pcmat = t(pcmat)
+    f4mat = t((pcmat[,A] - pcmat[,B]) * (pcmat[,C] - pcmat[,D]))
+    return(f4mat)
+}
+
+f4_from_pc_matrix <-function(pcmat, idx="Basque", idz="Turkish", idy="Scottish"){
+    pcmat = t(pcmat)
+    f4mat = t((pcmat - pcmat[,idx]) * (pcmat[,idy] - pcmat[,idz]))
+    return(f4mat)
+}
+
+f4_from_pc_matrix2 <-function(pcmat, px="Basque", py="Turkish"){
+    n = nrow(pcmat)
+    x = lapply(1:n, function(id) f4_from_pc_matrix(pcmat, idx=id, idy=px, idz=py) %>% 
[…]

### 2021-06-23 data/fscripts.R

@@ -1,0 +1,109 @@
+require(admixtools)
+require(tidyverse)
+require(Matrix)
+
+
+pca_from_f2s <- function(f2s){
+    f2 = f2(f2s)
+    n = dim(f2s)[1]
+    f2mat = f2 %>%
+        mutate(tmp=pop2, pop2=pop1, pop1=tmp) %>% 
+        select(-tmp) %>% bind_rows(f2) %>% 
+        select(-se) %>% 
+        pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>%
+        arrange(pop1) %>%
+        column_to_rownames('pop1') %>% as.matrix
+
+    cmat = double_center(f2mat)
+
+    E = eigen(nearPD(-cmat/2)[[1]])
+
+    pcs = t(t(E$vectors) * sqrt(pmax(E$values, 0))) 
+    rownames(pcs) = rownames(f2mat)
+    return(pcs)
+}
+
+double_center = function(m){
+    cmat = t(t(m - rowMeans(m) ) - colMeans(m)) + mean(m)
+}
+
+pca_from_f4 <- function(f4){
+    f4mat = f4 %>% 
+        select(pop1, pop2, est) %>% 
+        pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>% 
+        column_to_rownames('pop1') %>% as.matrix %>% abs
+
+    cmat = double_center(f4mat)
+
+    E = eigen(nearPD(-cmat/2)[[1]])
+
+    pcs = t(t(E$vectors) * sqrt(pmax(E$values, 0))) 
+    rownames(pcs) = rownames(f4mat)
+    return(pcs)
+}
+
+pca_from_pcmat <- function(pcs){
+    pcs = pcs %>% as.data.frame %>% rownames_to_column("pop")
+    names(pcs)[-1] = sprintf("PC%d", 1:(ncol(pcs) - 1))
+    return(pcs)
+}
+
+
+f3_from_pc_matrix <-function(pcmat, idx="Basque", idy="Turkish"){
+    pcmat = t(pcmat)
+    f3mat = t((pcmat - pcmat[,idx]) * (pcmat - pcmat[,idy]))
+    colnames(f3mat) = sprintf("PC%d", 1:ncol(pcmat))
+    return(f3mat)
+}
+
[…]

### 2021-06-23 data/ppca.R

@@ -1,0 +1,58 @@
+library(softImpute)
+library(admixr)
+
+psvd.eig <- function(eig=test_file, ...){
+    e <- eigenstrat(eig)
+    data <- admixr::read_geno(e)  
+    labels <- names(data)
+    data <- as.matrix(data)
+    data[data==9] <- NA
+    colnames(data) <- labels
+    return(psvd(data, ...))
+}
+
+psvd <- function(data, max_miss_snp=0.5, max_miss_ind=0.0, lambda=0, type='als', rank=4, 
+                 only_pcs=TRUE, ...){
+
+    if(is.null(colnames(data)))colnames(data) <- 1:ncol(data)
+
+    #X is random allele
+    X <- data
+
+    Z <- X[rowMeans(!is.na(X)) > max_miss_snp, ]
+
+
+    ind_completeness <- colMeans(!is.na(Z)) 
+
+    if(min(ind_completeness) < max_miss_ind){
+        print("removing individuals:")
+        w <- which(ind_completeness < max_miss_ind)
+        for(i in w){
+            print(sprintf("%s: %s", colnames(data)[i], ind_completeness[i]))
+        }
+    }
+    Z <- Z[,ind_completeness >= max_miss_ind] 
+
+    print(sprintf("PCA-matrix size:  %s", dim(Z)))
+    G <- t(Z - rowMeans(Z, na.rm=T)) #- colMeans(Z, na.rm=T)  + mean(Z, na.rm=T)
+
+    fit <- softImpute(G, lambda=lambda, rank=rank, type=type, ...)
+
+    PCS = t(t(fit$u) * fit$d)
+    PCS <- data.frame(name=rownames(G), PCS)
+    names(PCS)[-1] <- sprintf("PC%s", 1:ncol(fit$u))
+
+    if(only_pcs)return(PCS)
+
+    return(list(PCS=PCS, d=fit$d, loadings=fit$v, data=data))
+}
+
+test <- function(){
+    library(tidyverse)
+    test_file="~benjamin_peter/projects/100a/plink/100a2_all/archaics"
+    pca <- psvd.eig(test_file, lambda=0, max_miss_snp=0.48, 
+                    max_miss_ind=.2,
+                        rank=6)  
+    ggplot(pca, aes(x=PC1, y=PC2, label=name)) + geom_text()     
+
+}

### 2021-06-23 data/ppca_westeurasian.R

@@ -1,0 +1,73 @@
+library(softImpute)
+library(tidyverse)
+library(admixr)
+source("ppca.R")
+
+set.seed(1)
+
+eig <- "subdata/westeurasian1"
+e <- eigenstrat(eig)
+data <- admixr::read_geno(e)  
+labels <- names(data)
+data <- as.matrix(data)
+data[data==9] <- NA
+colnames(data) <- labels
+pmiss <- colMeans(!is.na(data))
+d2 <- data[rowMeans(!is.na(data)) > .75,]
+
+if(F){
+fit <- psvd(d2, max_miss_ind=.5, thresh=1e-4, 
+            trace.it=T, rank=10)
+saveRDS(fit, "fit.rds")
+}
+
+ind = admixr::read_ind(e)
+PCS = fit$PCS
+PCS$pop = ind$label
+pcs = PCS %>% group_by(pop) %>% summarize_at(vars(PC1:PC10), .funs=mean)
+
+#get circle based on basque, turkish
+circle = pcs %>% filter(pop %in% c("Basque", "Turkish")) %>% select(PC1, PC2)
+radius = (dist(circle) %>% as.matrix)[1,2] / 2
+origin = colMeans(circle)
+circle_df=tibble(x0=origin[1], y0=origin[2], radius=radius)
+
+
+
+
+rd2 = rowMeans(d2)
+d3 = d2[!is.na(rd2),] / 2
+mu = rowMeans(d3)
+Y = d3 - mu
+SVD = svd(Y)
+lambda = SVD$d
+#lambda= lambda/sum(lambda)
+P = SVD$v
+S = t(t(P) * lambda)
+tS = t(S)
+
+
+h2 = mean(mu * (1-mu)) + mean(Y[,1] * (1-Y[,1])) - 2 * mean(Y[,1] * mu)
+h1 = mean(d3[,1] * (1-d3[,1]))
+
+T1 = sum(colSums((1 - 2 * mu) * SVD$u) * pki ) / n
+M = sum(mu * (1-mu)) / n
+SQ = sum(S[1,]^2)/ n
+h3 = T1 + M - SQ
+
+n = nrow(Y)
+npcs = ncol(Y)
+
+T1t = colSums((1-2*mu)*SVD$u) * pki
+SQt = S[1,]^2
+Mt = M *n / npcs
+
+
+f3test = function(idx, id1, id2){
[…]

### 2021-06-23 data/prep_data_pca_f4ratio.R

@@ -1,0 +1,19 @@
+require(admixtools)
+require(tidyverse)
+require(glue)
+
+
+data_loc = 'subdata/f4ratio'
+inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
+
+
+
+f2loc1 = 'f4ratio'
+
+#afs = admixtools::extract_afs(data_loc, pops=sample_list, 
+#                              outdir=glue("{f2loc1}_afs"),format="eigenstrat", 
+#                              cols_per_chunk=1000)
+f2s = admixtools::extract_f2(data_loc, 
+                             outdir=f2loc1, maxmem=64000, 
+                             overwrite=T)
+

### 2021-06-23 data/prep_data_pca_worldfocus1.R

@@ -1,0 +1,19 @@
+require(admixtools)
+require(tidyverse)
+require(glue)
+
+
+data_loc = 'subdata/worldfoci1'
+inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
+
+
+
+f2loc1 = 'worldfoci1'
+
+#afs = admixtools::extract_afs(data_loc, pops=sample_list, 
+#                              outdir=glue("{f2loc1}_afs"),format="eigenstrat", 
+#                              cols_per_chunk=1000)
+f2s = admixtools::extract_f2(data_loc, 
+                             outdir=f2loc1, maxmem=64000, 
+                             overwrite=T)
+

### 2021-06-23 data/prep_data_pca_worldfocus2.R

@@ -1,0 +1,19 @@
+require(admixtools)
+require(tidyverse)
+require(glue)
+
+
+data_loc = 'subdata/worldfoci2'
+inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
+
+
+
+f2loc1 = 'worldfoci2'
+
+#afs = admixtools::extract_afs(data_loc, pops=sample_list, 
+#                              outdir=glue("{f2loc1}_afs"),format="eigenstrat", 
+#                              cols_per_chunk=1000)
+f2s = admixtools::extract_f2(data_loc, 
+                             outdir=f2loc1, maxmem=64000, 
+                             overwrite=T)
+

### 2021-06-23 fstats_pca_lewontin.tex
better writing

@@
 \section{Introduction}
-About 15\% of genetic variation in humans can be explained by population structure \cite{lewontin1972, barbujani1997, rosenberg2003}, but the information contained in these 15\% is sufficient to study the genetic diversity and history in great detail \cite{edwards2003}.
+About 15\% of genetic variation in humans can be explained by population structure \cite{lewontin1972, barbujani1997, rosenberg2003}, but the information contained in these 15\% is sufficient to study the genetic diversity and history in great detail \cite{cavalli-sforza1994, edwards2003}.
 For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \cite{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
 
-In Lewontin's pioneering analysis, he found that less than half (6\%), of that variation could be attributed to the continental-scale groups he called races, it seemed  which he used to claim that ''racial classification is (...) seen to be of virtually no genetic or taxonomic significance``.
+However, population genetic models that adequately capture the heterogeneity in human genetic diversity remain a challenge. 
+It is also apparent that human genetic variation is largely distributed in gradients.
+In Lewontin's pioneering analysis, he found that less than half (6\%), of that variation could be attributed to the continental-scale groups. Later studies on genome-wide data show
+
+although some discontinuities, often associated with obstacles to migration such as mountain ranges or oceans, exist \cite{serre2004, ramachandran2005, rosenberg2007, peter2020}.
 
@@
 
-Thus, it frequently a useful analysis tool to think of populations as discrete units.
-For example, even though the underlying population structure may be continuous, sampling is not; and when quantifying ascertainment and sampling biases, or when discussing population structure it is often helpful to pretend populations are discrete, even though the underlying structure is typically more complex.
-This leads to challenges both in data interpretation and communication, and often researchers will analyze a data set both using methods that assume population structure is discrete, and methods where this assumption does not need to be made.
+As a consequence, tools such as Principal Component Analysis (PCA) are frequently used to visualize or model human population structure \citep{cavalli-sforza1964, cavalli-sforza1994}.
+Informally, PCA aims to extract the axes of largest variation in a data set; for population genetic data sets of e.g.
+SNPs, these axes typically reflect population structure and need to be taken into account in e.g.
+association studies \citep{price2006}.
+As population structure is frequently sparse \cite{engelhardt2010, mcvean2009}, only the first few principal components (PCs) reflect population structure, i.e.
+the 15\% of genetic variation that can be associated with structure.
+
+While PCA is able to capture continuous and complex genetic structure \citep{novembre2008, novembre2008a}, it is often hard to interpret.
+Even though human genetic diversity is to a large extent gradual, many models used to describe it have discrete components.
+This has a number of largely practical reasons; it is much easier to develop generative historical models using discrete populations \cite{adrion2020}, sampling is usually discrete \cite{mallick2016, thousand_genomes2016}, and results are often easier to communicate and interpret if populations are treated as discrete \citep{popejoy2016}.
+Yet, as with any model, this ease of interpretation comes at a loss of some of the underlying complexity.
 
 One discrete framework for the analysis of human population structure that gained a lot of traction in the last decade are the $F$-statistics \textit{sensu} Patterson \cite{patterson2012, peter2016}.
-This framework treats populations as discrete units in the analysis, and allows for a variety of tests for treeness.
-Using this framwork, the vast majority of present-day human populations are admixed \cite{pickrell2014}.
-Yet, this framework starts with the assumption that admixture is i) rare and ii) discrete.
+This framework treats populations as discrete units, assuming, as a null model, that they are related in a tree-like fashion.
+Two simple statistics, called $F_3$ and $F_4$ can be used to test whether three or four populations are related in a treelike fashion, respectively.
+If these statistics are significant, treeness is rejected in favor of \emph{admixture}.
+In recent years, more complicated methods have been developed that extend this to framework to more complex models involving many populations \cite{patterson2012, lipson2013, harney2020}, but the framework is built on the assumptions that populations are discrete, and gene flow is rare.
 
-However, $F$-statistics are not restricted to discrete populations.
-Indeed, as they can be written as functions of allele frequency variances, or expected pairwise coalescence times, statistics that can be calculated under a wide range of demographic models \cite{peter2016}.
-Indeed, as they reflect inner products, they can be generalized to Euclidean space \cite{oteo2021} (or any Hilbert space, although we won't pursue that here).
-Here, I explore these links between $F$-statistics and Euclidean spaces to establish connections between $F$-statistics and PCA.
-This allows direct interpretation of admixture in scenarios where population structure might not be discrete.
+In this paper, my goal is to develop an interpretation of $F$-statistics in the context of PCA.
+This allows direct interpretation of $F$-statistics from \emph{data space} \citep{oteo-garcia2020}, and allows us to talk about admixture in scenarios where population structure might not be discrete.
+For this purpose, I will thus briefly define PCA and $F$-statistics, and the geometry of $F$-statistics on PCA-plots.
+This allows me to develop approximations and visualizations which are illustrated on several examples.
 
-Particularly for the analysis of ancient DNA, two approaches have been proven to be particularly useful: one are global summary analyses, such as Structure \citep{pritchard2000, alexander2009} Principal Component Analysis (PCA) \citep{cavalli-sforza1994, reich2008, novembre2008, mcvean2009} and classical multidimensional scaling (MDS) \cite{fu2016, malaspinas_bammds}.
-Typically, these methods assume that population structure is \emph{sparse}, so that a low-rank approximation with few underlying ``components'' is sufficient to model population structure See e.g.
-\cite{engelhardt2010} for a useful perspective how these approaches are related.
-
-Facing a novel data set, PCA or MDS are often the first analyses (beyond quality controls) a researcher performs, in order to obtain insights in the general population structure they are faced with.
-In order to answer more specific questions and to test specific hypotheses, the $F$-statistic framework of \cite{patterson2012} has been proven particularly powerful (see also \cite{peter2016} for a more gentle introduction).
-In the $F$-statistic framework, usually only a small number of populations are used at once, to e.g.
-test for treeness and find closely related populations.
-
-Even though these two approaches are considered in almost every ancient DNA paper, links between the inferences made from them are usually only compared qualitatively.
-In this paper, our goal is to show that PCA and $F$-statistics are in fact closely related by construction, and use a very similar summary of the data.
-
-\subsection{Introduction to $F$-statistics}
-$F$-statistics have been primarily motivated by trees and admixture graphs \citep{patterson2012, peter2016}, but the calculations hold up in a much wider data space.
-In particular,  \cite{oteo-garcia2021} provides a thorough introduction to interpreting $F$-statistics in the \emph{data space} $\mathbb{R}^k$.
-Their work builds much of the foundation of this discussion, by demonstrating analogies to classical geometry.
-A brief summary of their key results: A population's allele frequencies can be thought of as vector in $\mathbb{R}^k$.
-Then, $F_2(X_1, X_2) = \normsq{X_1 - X_2}$ is the squared Euclidean distance between the populations with vectors $X_1$ and $X_2$, and $F_4(X_1, X_2 ; X_3, X_4) = \langle X_1 - X_2, X_3 - X_4 \rangle$ is the inner (scalar) product between these two vectors.
-Here, I will mainly use the $F$-statistic notation, but use the geometric notation where convenient.
-
-\section{Relationship of PCA, $F_2$ and Outgroup-$F_3$}
-
-The goal of this section is to give a cursory introduction to $F$-statistics, PCA and MDS, and to define notation.
-A more detailed technical introduction is given in XXXXX, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
-
+\section{Theory}
 \subsection{Introduction to PCA}
@@
 As a population may be represented by just one (pseudo-)haploid or diploid individual, there is no conceptual difference between these cases and I will refer to populations as unit for analysis, for simplicity.
+Note that this merely assumes that sampling, but not necessarily the underlying structure, is discrete.
 
 Since the allele frequencies are between zero and one, we can interpret each row $x_i$ of $\MX$ as a vector in $[0, 1]^k$, the \emph{data space} of all possible allele frequencies on our markers.
+The goal of a PCA is to find an optimal low-dimensional representation of $\MX$ that retains most of the variation in the data (see Fig.
+\ref{fig:pca_explanation} for an illustration).
 
-	The goal of a PCA is to find a low-dimensional representation of this data space that retains most of the data (see Fig. \ref{fig:pca_explanation} for an intuitive explanation).
-
-	\begin{figure}
+	\begin{figure}[!ht]
 		\includegraphics[width=\textwidth]{pca_explanation.png}
@@
 	There are several algorithms that are used to calculate a PCA in practice, the most common one relies on a singular value decomposition.
-In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
+In this approach, we first mean-center $\MX$ to obtain the centered matrix $\MY$
 	\begin{equation}
@@
 	\begin{equation}
-	\MY = \MC\MX = \mathbf{U} \MSINGULAR \mathbf{V}^T = \MP\ML\text{,}
+	\MY = \MC\MX = (\mathbf{U} \MSINGULAR) \mathbf{V}^T = \MP\ML\text{,}
 	\end{equation}
@@
 	In many  implementations \citep[e.g]{patterson2006}, SNPs are weighted by the inverse of their standard deviation.
-As this weighting makes little difference in practice, I will for now assume that SNPs are unweighted, and defer discussion of weighting to a later section.
+As these weights makes little difference in practice and obfuscate the connection to $F$-statistics, I will for now assume that SNPs are unweighted, and defer discussion of weighting to a later section.
 
-	Equivalently, we obtain the PCs by performing an eigendecomposition of the  covariance matrix denoted as $\MK$:
+\subsection{Introduction to $F$-statistics}
+From a modeling perspective, $F$-statistics have been primarily motivated by trees and admixture graphs \citep{patterson2012}, but can be calculated for arbitrary population genetics models for which expected pairwise coalescence times are known \citep{peter2016}.
+
+They are based on the squared Euclidean distance $F_2$ used as a tree metric, and the 
+\begin{subequations}
+	\begin{align}
+	F_2(X_1, X_2) &= \sum_i^k (x_{1i} - x_{2i})^2 = 2\mathbb{E}T_{12} - \mathbb{E}T_{11} - \mathbb{E}T_{22} &=& \normsq{X_1 - X_2}\\
+	F_3(X_1; X_2, X_3) &= \sum_i^k (x_{1i} - x_{2i})(x_{1i} - x_{3i}) &=& \langle X_1 - X_2, X_1 - X_3 \rangle\\
+	F_4(X_1 X_2; X_3, X_4) &= \sum_i^k (x_{1i} - x_{2i})(x_{3i} - x_{4i}) &=& \langle X_1 - X_2, X_3 - X_4 \rangle,
+	\end{align}
+\end{subequations}
+where the $X_i$ are rows of $\MX$ that represent all loci of population $i$, and the $T_{ij}$ denote the pairwise coalescence time of two sampled individuals from population $i$ and $j$.
+
+Here, much of the focus is not on interpreting these $F$-statistics in terms of model, but in \emph{data space}.
+For this, we interpret $F$-statistics as inner products on $\mathbb{R}^L$, and the norm $\norm{\cdot}$ and inner product $\langle\cdot, \cdot \rangle$ notation are used to make the connection to Euclidean geometry more explicit \citep[see][for a thorough introduction to this interpretation]{oteo-garcia2021}.
+
+\subsection{PCA from $F$-statistics}
+Despite their different motivation, $F$-statistics and PCA are closely related through a procedure called multidimensional scaling (MDS).
+For MDS, we consider a matrix 	$\MF, f_{ij} = F_2(X_i, X_j)$ of pairwise $F_2$-statistics.
+Then, the matrix of principal components $\MP$ can be obtained by performing an eigendecomposition of the double-centered $\MF$-matrix  \cite{gower1966}:
 \begin{equation}
- \MK = \MY \MY^T = \mathbf{U}\MEIGEN\mathbf{U}^T = \mathbf{U}\MSINGULAR^2\mathbf{U}^T =\MP\MP^T
-	\end{equation} where $\MEIGEN$ is the diagonal matrix with the eigenvalues of $\MK$. 
-	This algorithm does not compute the SNP-loadings. However, the $i$-th row of $L$  can be obtained from $\MP$ and the original data, whenever the eigenvalue $\lambda_i \neq 0$:
-	\begin{equation}
-	\ML_i = \lambda_i^{-1}\MP^T\MC\MX \text{.}
+\MP\MP^T = -\frac{1}{2} \MC \MF\MC \text{.}
 \end{equation}
 
-\section{$F$-statistics in PCA-space}
-As shown by e.g.
-\cite{oteo-garcia2021}, $F$-statistics can be thought of as inner products in Euclidean space, and $F_2$ is an (estimated) squared Euclidean distance between two populations in allele frequency space.
-By performing a PCA, we just translate and rotate our data, but Euclidean distances and dot products are both invariant under both these operations.
+\subsection{$F$-statistics in PCA-space}
+The $F$-statistics can be thought of as inner products in the Euclidean allele-frequency space.
+By performing a PCA, we just translate and rotate our data, but the dot products is invariant under both these operations.
 Hence, neither mean-centering (a translation) nor PCA (a rotation) will change $F_2$.
@@
 \begin{align}
-F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2 - H_i - H_j &&\nonumber\\ 
- &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 - H_i - H_j  &=& F_2(Y_i, Y_j) \nonumber\\
- &=& \sum_k (P_{ik} - P_{jk})^2 - H_i - H_j &=& F_2(P_i, P_j) \text{,}
+F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2 &&\nonumber\\ 
+ &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2  &=& F_2(Y_i, Y_j) \nonumber\\
+ &=& \sum_k (P_{ik} - P_{jk})^2  &=& F_2(P_i, P_j) \text{,}
 \end{align}
-where $H_i$ and $H_j$ are the bias-correction terms proposed in \cite{reich2009}. A detailed derivation of this is given in Appendix \ref{appendix:fonpc}.
-As $F_3$ and $F_4$ can be written as sums of $F_2$-terms, analogous relations apply.
+A detailed derivation of this is given in Appendix \ref{appendix:fonpc}.
+As $F_3$ and $F_4$ can be written as sums of $F_2$-terms \citep{reich2009}, analogous relations apply.
 
-\subsection{$F$-stats in 2-dimensional PC-space}
-The transformation derived in the previous section allows us to consider the geometry of $F$-statistics in PCA-space.
+\subsection{Geometric representation of $F$-statistics in PC-space}
+The transformation derived in the previous section allows us to consider the geometry of $F$-statistics in the rotated PCA-space.
 The relationships we will discuss formally only hold if we use all $n-1$ PCs.
@@
 Here, we start by discussing 2-dimensional spaces.
-This is useful for two reasons: for one, the geometry is simpler and we can think of circles and squares as opposed to $n$-balls and other high-dimensional geometric objects and thus build intuition.
-Second, in many applications it is argued that a 2-dimensional approximation is sufficient to explain the major components of population structure \cite{novembre2008}.
+This is useful for two reasons: for one, the geometry is simpler and we can think of circles as opposed to $n$-balls and other high-dimensional geometric objects.
+Second, in many applications it is argued that a 2-dimensional approximation is sufficient to explain the major components of population structure \cite[e.g.][]{novembre2008}.
 In this case, the results here will hold under the same approximation assumptions in low-dimensional PCs; if they differ substantially from each other, it is likely that not sufficiently many PCs were considered.
 
-\subsubsection{$F_2$ in PC-space}
+\subsubsection{$F_2$ in 2D PC-space}
 The $F_2$-statistic is an estimate of the squared Euclidean distance is the easiest to understand, it corresponds directly to the squared distance in PCA-space.
@@
 \subsubsection{$F_3$ and circles}
-The $F_3$-statistic becomes more interesting; as outlines above we either think of $F_3$ as ``outgroup''-$F$-stats or as admixture $F$-stats.
-In the admixture case, we may ask the following question: given two source populations $X_1$, $X_2$, where would admixed populations on a PCA plot lie?
-From theory, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \cite{brisbin2012, mcvean2009}.
+A common application of $F_3$-statistics is to test for admixture; we say that population $X_x$ is admixed from sources $X_1$ and $X_2$ if $F_3(X_x; X_1, X_2) < 0$.
+This motivates the following question: given two source populations $X_1$, $X_2$, where would we expect the admixed population $X_x$ on a PCA plot?
+Since the allele frequencies of $X_x$ are intermediate between $X_1$ and $X_2$, we would expect it to lie between these two populations \cite{brisbin2012, mcvean2009}, but we can make this notion more precise:
 
-Formally, we would reject admixture if $F_3$ is negative, i.e. we are looking for the space
 \begin{eqnarray}
-2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
+2 F_3(X_x; X_1, X_2) &=& \langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
       &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} \nonumber\\
@@
 By the Pythagorean theorem, $F_3 = 0 $ iff $X_1, X_2$ and $X_x$ form a right-angled triangle.
-Hence, the region where $F_3$ is zero is the circle with diameter through $X_1$ and $X_2$.
-If $X_x$ lies inside this circle, the angle is obtuse and $F_3$ is negative, otherwise it will be positive.
-Similarly, if we fix $X_1$ and $X_2$ and ask where on a 2D-PCA-plot $X_2$ would lie, this space is defined by all the points for which the angle between $X_1 X_x$ and $X_2 X_x$ is obtuse.
+Hence, on a 2D-PCA-plot, the region where $F_3$ is zero is the circle with diameter $\overline{X_1X_2}$ (Figure \ref{fig:fstats_theory}B).
+If $X_3$ lies inside this circle, the angle is obtuse and $F_3$ is negative, otherwise it will be positive.
+If we consider all PCs, the circle is replaced by a $n$-ball with diameter $\overline{X_1X_2}$, but otherwise the reasoning is exactly the same.
+In this case, not all populations inside a circle on a 2D-plot will have negative $F_3$, as they may be outside the $n$-ball in another dimension.
+However, all populations that map outside the circle on any PC are guaranteed to have a positive associated $F_3$-statistic.
 
-This highlights a potential identifiability issue with $F_3$: Since all values of $F_3$ that result in the same projection will give the same value; and multiple admixture events may result in the same $F_3$-value.
+Similarly, if we fix the admixed population $X_x$ and one source $X_1$, we may ask where on a 2D-PCA-plot $X_2$ would lie if $F_3(X_x; X_1, X_2) < 0$.
+This space is similarly defined by all the points for which the angle between the line segments $\overline{X_1X_x}$ and $\overline{X_2 X_x}$ is obtuse (Figure \ref{fig:fstats_theory}C).
+In this case
 
@@
 F: If $X_x$ is admixed between $X_1$ and $X_2$, the admixture proportions will be projected.
+	\label{fig:fstats_theory}
 \end{figure}
@@
 \subsection{Higher-Dimensional Spaces}
+
+\section{Results}
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
+	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
+Populations for which $F_3$ is negative are colored in red.
+B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
+C,F: Contributions of PCs 1-10 to each $F_3$-statistic.
+D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$.
+The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population.
+}
+	\label{fig:f3}
+\end{figure}
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fig_f4_data.pdf}	
+	\caption{\textbf{PCA and $F_4$-statistics} A: Spectrum of select $F_4$-statistics in World data set.
+B: Projection angle representation of $F_4(X, Sardinian; Mozabite, Yoruba)$ (red) and approximations using two (blue) and ten (green) PCs.
+C: Spectrum of select $F-4$-statistics in Westeurasian data set.
+D: Scatterplot of $F_4$-projection on Finnish-Canary Islanders axis and residual PC1.
+		E: $F_4(X, \text{French}, \text{Finnish}, \text{Canary Islander})$ vs.
+prediction using two (blue) and ten(red) PCs.
+F: Percent variance explained for the projection of panel D and the first nine residual PCs.
+	}
+	\label{fig:f4}
+\end{figure}
 

### 2021-06-24 fstats_pca.tex
figures

@@
 \usepackage{natbib}
+
+\usepackage{url}
 
@@
 \section{Results}
+The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots.
+In this section, I use these interpretation in the analysis of human genetic variation data set.
+I use two data sets based on the ``Human Origins''-SNP set (597,573 SNPs).
+Both are subsets of the Reich lab compendium data set v44.3, downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}.
+
+\paragraph{West-Eurasian data set}
+This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
+It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \cite{patterson2012}.
+Population genetic differentiation in this region is low and closely mirrors geography \cite{novembre2008}.
+
+\paragraph{World Overview data set}
+This data set of 638 individuals from 33 population contains individuals throughout the world, and is used as a sparse data set capturing much of global human genetic variation.
+This data set spans Africa, Eurasia and the Americas, and we might therefore expect the population structure to be much more sparse.
+
+I perform analyses at the level of populations to ease presentation, and because it is an assumption of $F$-statistics that the genetic variation with sampled population is independent of the variation between samples that I am focusing on here.
+I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute a matrix of $F_2$-statistics between all populations.
+To obtain a PC-decomposition I use equation XXX and the \texttt{eigen} function in \texttt{R}, and compare them with the $F_3$ and $F_4$-statistics calculated using admixtools 2.
+
+\paragraph{Admixture-$F_3$}
+As a first step, I plot the first two principal components of the Westeurasian data set (Figure \ref{fig:f3}A).
+This PCA presents two parallel clines, one from the Levant and Arabia (``BedouinB'') to the Caucasus (``Abkhasian''), and a second one from Southern (``Sardinian'') to Northeastern Europe (``Mordovian'').
+In this context, I examine $F_3(X; Basque, Turkish)$, i.e.
+a statistic that aims to ask which populations can be represented as a mixture between a Southwestern (Basque) and Southeastern (Turkish) European population.
+The -- largely Southern European -- populations for which the point estimate of these $F_3$-statistic is negative are highlighted in red.
+They both fall close to the center of the $F_3$-circle, either defined on the first two (dark grey) or all PCs (light gray).
+However, many populations inside the circle on the first 2 PCs, including English, Sardinians and Canary Islanders have positive $F_3$-values, on higher PCs, showing that the first two PCs do not capture all the genetic variation related to population structure for this data set.
+
+This is expected because for spatially continuous populations, PCA will not be sparse \cite{novembre2008a}.
+Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
+
+Thus, the main benefit of this PCA-plot is that it allows us to identify populations outside the circle (from the Levant and Caucasus), for which $F_3$ is guaranteed to be positive.
+
+\paragraph{Outgroup-$F_3$}
+
 \begin{figure}[!ht]

### 2021-06-24 fstats_pca_lewontin.tex
figures

--- 2021-06-23 fstats_pca_lewontin.tex
@@
 
-\subsubsection{$F_3$ and circles}
+\subsubsection{When is $F_3$ negative?}
 The $F_3$-statistic becomes more interesting; as outlines above we either think of $F_3$ as ``outgroup''-$F$-stats or as admixture $F$-stats.
@@
 2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
-      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} \nonumber\\
-      &<&0
+      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} 
 \end{eqnarray}
 By the Pythagorean theorem, $F_3 = 0 $ iff $X_1, X_2$ and $X_x$ form a right-angled triangle.
-Hence, the region where $F_3$ is zero is the circle with diameter through $X_1$ and $X_2$.
-If $X_x$ lies inside this circle, the angle is obtuse and $F_3$ is negative, otherwise it will be positive.
-Similarly, if we fix $X_1$ and $X_2$ and ask where on a 2D-PCA-plot $X_2$ would lie, this space is defined by all the points for which the angle between $X_1 X_x$ and $X_2 X_x$ is obtuse.
+In a 2D-PCA plot, the region where $F_3$ is zero is the circle  with diameter $\overline{X_1X_2}$, and if $X_x$ lies inside this circle, $F_3(X_x; X_1, X_2) < 0$.
+If the
 
-This highlights a potential identifiability issue with $F_3$: Since all values of $F_3$ that result in the same projection will give the same value; and multiple admixture events may result in the same $F_3$-value.
+ball, the angle is obtuse and $F_3$ is negative, otherwise it will be positive. If we approximate the PCA-space in two dimensions, the $n$-ball corresponds to a circle.
 

### 2021-07-01 fstats_pca_lewontin.tex

@@
 \usepackage{lineno}
+\usepackage{url}
 
@@
 
-\subsection{Introduction to $F$-statistics}
-$F$-statistics have been primarily motivated by trees and admixture graphs \citep{patterson2012, peter2016}, but the calculations hold up in a much wider data space.
-In particular,  \cite{oteo-garcia2021} provides a thorough introduction to interpreting $F$-statistics in the \emph{data space} $\mathbb{R}^k$.
-Their work builds much of the foundation of this discussion, by demonstrating analogies to classical geometry.
-A brief summary of their key results: A population's allele frequencies can be thought of as vector in $\mathbb{R}^k$.
-Then, $F_2(X_1, X_2) = \normsq{X_1 - X_2}$ is the squared Euclidean distance between the populations with vectors $X_1$ and $X_2$, and $F_4(X_1, X_2 ; X_3, X_4) = \langle X_1 - X_2, X_3 - X_4 \rangle$ is the inner (scalar) product between these two vectors.
-Here, I will mainly use the $F$-statistic notation, but use the geometric notation where convenient.
-
-\section{Relationship of PCA, $F_2$ and Outgroup-$F_3$}
-
-The goal of this section is to give a cursory introduction to $F$-statistics, PCA and MDS, and to define notation.
-A more detailed technical introduction is given in XXXXX, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
+\section{Theory}
+In this section, I will give a very brief introduction to $F$-statistics and PCA.
+A more detailed technical introduction of PCA is given in XXXXX, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
 
 \subsection{Introduction to PCA}
-	Let us assume we have some genotype data summarized in a matrix $\MX$, where the entry $x_{ij}$ is the allele frequency of the $i$-th population at the $j$-th genotype.
-If we have $k$ SNPs and $n$ populations, $\MX$ will have dimension $n \times k$.
+Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype.
+If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
+As a population may be represented by just one (pseudo-)haploid or diploid individual, there is no conceptual difference between these cases and I will
+refer to populations as unit for analysis, for simplicity. Since the allele
+frequencies are between zero and one, we can interpret each Population $X_i$
+of $\MX$ as a point in $[0, 1]^S$, the \emph{data space} of all possible allele frequencies on our markers.
 
-    As a population may be represented by just one (pseudo-)haploid or diploid individual, there is no conceptual difference between these cases and I will refer to populations as unit for analysis, for simplicity.
-
-    Since the allele frequencies are between zero and one, we can interpret each row $x_i$ of $\MX$ as a vector in $[0, 1]^k$, the \emph{data space} of all possible allele frequencies on our markers.
-
-	The goal of a PCA is to find a low-dimensional representation of this data space that retains most of the data (see Fig. \ref{fig:pca_explanation} for an intuitive explanation).
+The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ of the data that explains most of the variation in the data.
+$K$ is at most $n-1$, but the historical processes that generated often result in \emph{sparse} data \citep{engelhardt2010,
+patterson2012}, so that $K \ll n$; for visualization $K=2$ is frequently used (see Fig. \ref{fig:pca_explanation} for an intuitive explanation).
 
@@
 
-	There are several algorithms that are used to calculate a PCA in practice, the most common one relies on a singular value decomposition.
-In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
-	\begin{equation}
+There are several algorithms that are used to calculate a PCA in practice,
+   the most common one is based on singular value decomposition. In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
+	\begin{equation*}
 	y_{il} = x_{il} - \mu_l
-	\end{equation}
+	\end{equation*}
 	where $\mu_l$ is the mean allele frequency at the $l$-th locus.
@@
 
-	In many  implementations \citep[e.g]{patterson2006}, SNPs are weighted by the inverse of their standard deviation.
-As this weighting makes little difference in practice, I will for now assume that SNPs are unweighted, and defer discussion of weighting to a later section.
+	In many  implementations \citep[e.g]{patterson2006}, SNPs are weighted by
+    the inverse of their standard deviation. As this weighting often makes little
+    difference in practice \citep{mcvean2009}, I will assume throughout that SNPs
+    are unweighted.
 
-	Equivalently, we obtain the PCs by performing an eigendecomposition of the  covariance matrix denoted as $\MK$:
+\subsection{Introduction to $F$-statistics}
+PCA is typically used to model population structure between many populations.
+$F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time.
+
+\begin{subequations}
+	\begin{align}
+	F_2(X_1, X_2) &=& \sum_{l=1}^S(x_{1l - x{2l}})^2 &=& \normsq{X_1, X_2}\\
+	F_3(X_1; X_2, X_3) &=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \langle X_1 - X_2, X_1 - X_3 \rangle\\	
+	F_4(X_1, X_2; X_3, X_4) &=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
+	\end{align}
+\end{subequations}
+where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product.
+
+$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}, although they can be calculated under most population genetic models \citep{peter2016}.
+Most commonly, $F_3$ and $F_4$ are interpreted as admixture tests: Negative values of  $F_3(X_1; X_2, X_3) < 0$ are interpreted that $X_1$ is a mixture between populations (related to) $X_2$ and $X_3$.
+Similarly if populations are related as a tree, then $F_4(X_1, X_2; X_3, X_4) =0$.
+Alternatively, $F_2$, $F_3$ and $F_4$ are all used to measure genetic similarity.
+$F_2(X_1, X_2)$ represents the variance in allele frequency between populations $X_1$ and $X_2$, which is a measure of genetic drift.
+The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_i)$ is used if we have an unknown population $X_U$, and want to find its closest relatives from a panel of populations $X_i$.
+The highest values of $F_3$ indicate the population $X_i$ most closely related to $X_U$.
+Including an outgroup $X_0$ allows correction for sampling-time difference, which are common in applications on ancient DNA.
+Finally, $F_4(X_1, X_2; X_3, X_4)$ is used to measure the length of the internal branch on the tree connecting these four populations.
+It is used to reconstruct admixture graphs \cite{patterson2012, lipson2013} and to estimate admixture proportions \citep{petr2019, harney2020}.
+
+Here, we are interested in interpreting the $F$-statistics in the data space $\mathbb{R}^S$, and compare it with PCA.
+For a thorough introduction of interpreting $F$-statistics in data space, see \cite{oteo-garcia2021}.
+
+\paragraph{Principal components from $F$-statistics}
+The principal components can be directly calculated from $F$-statistics using multidimensional scaling.
+Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$.
+We can obtain the principal components from this matrix by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
 \begin{equation}
- \MK = \MY \MY^T = \mathbf{U}\MEIGEN\mathbf{U}^T = \mathbf{U}\MSINGULAR^2\mathbf{U}^T =\MP\MP^T
-	\end{equation} where $\MEIGEN$ is the diagonal matrix with the eigenvalues of $\MK$. 
-	This algorithm does not compute the SNP-loadings. However, the $i$-th row of $L$  can be obtained from $\MP$ and the original data, whenever the eigenvalue $\lambda_i \neq 0$:
-	\begin{equation}
-	\ML_i = \lambda_i^{-1}\MP^T\MC\MX \text{.}
+\MP\MP^T = - \frac{1}{2}\MC\MF\MC \text{.} \label{eq:mds}
 \end{equation}
@@
 \begin{align}
-F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2 - H_i - H_j &&\nonumber\\ 
- &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 - H_i - H_j  &=& F_2(Y_i, Y_j) \nonumber\\
- &=& \sum_k (P_{ik} - P_{jk})^2 - H_i - H_j &=& F_2(P_i, P_j) \text{,}
+F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2  &&\nonumber\\ 
+ &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2   &=& F_2(Y_i, Y_j) \nonumber\\
+ &=& \sum_{k=1}^n (p_{ik} - p_{jk})^2  &=& F_2(P_i, P_j) \text{,}
 \end{align}
-where $H_i$ and $H_j$ are the bias-correction terms proposed in \cite{reich2009}. A detailed derivation of this is given in Appendix \ref{appendix:fonpc}.
+A detailed derivation of this is given in Appendix \ref{appendix:fonpc}.
 As $F_3$ and $F_4$ can be written as sums of $F_2$-terms, analogous relations apply.
 
+\paragraph{Optimality of PCA}
+In most applications, we do not use all PCs, but instead use only the first $K$ PCs.
+Thus, 
+\begin{equation}
+F_2(P_i, P_j) = \sum_{k=1}^K(p_{ik} - p_{jk})^2 + \sum_{k=K+1}^n(p_{ik} - p_{jk})^2 \text{,}
+\end{equation}
+where the first sum is the PCA-approximation of $\hat{F_2}$, and the second sum is the residual or approximation error $F_2 - \hat{F_2}$.
+
+If we sum up the approximation errors over all pairs of populations, we obtain the Frobenius-norm of the error $\normsq{\MF - \hat{\MF}}_F$; it is a standard result that PCA finds the best rank-$K$ approximation so that this Frobenius-norm is minimized.
+In our context, this means that PCA using the first $K$ PCs results in approximate $F_2$-statistics such that the sum of $F_2$-distances between the approximation and full data is minimized.
+
 \subsection{$F$-stats in 2-dimensional PC-space}
-The transformation derived in the previous section allows us to consider the geometry of $F$-statistics in PCA-space.
-The relationships we will discuss formally only hold if we use all $n-1$ PCs.
-However, the appeal of PCA is that frequently, only a very small number $K \ll n$ of PCS contain most information that is relevant for population structure.
-
-Here, we start by discussing 2-dimensional spaces.
-This is useful for two reasons: for one, the geometry is simpler and we can think of circles and squares as opposed to $n$-balls and other high-dimensional geometric objects and thus build intuition.
-Second, in many applications it is argued that a 2-dimensional approximation is sufficient to explain the major components of population structure \cite{novembre2008}.
-In this case, the results here will hold under the same approximation assumptions in low-dimensional PCs; if they differ substantially from each other, it is likely that not sufficiently many PCs were considered.
+The transformation from the previous section allows us to consider the geometry
+of $F$-statistics in PCA-space. The relationships we will discuss formally only
+hold if we use all $n-1$ PCs. However, the appeal of PCA is that frequently,
+only a very small number $K \ll n$ of PCS contain most information that is
+relevant for population structure (for visualization, it is often assumed that
+$K=2$).
 
 \subsubsection{$F_2$ in PC-space}
-The $F_2$-statistic is an estimate of the squared Euclidean distance is the easiest to understand, it corresponds directly to the squared distance in PCA-space.
-This matches our intuition that closely related populations (which have low $F_2$) will be close to each other on a PCA-plot.
+The $F_2$-statistic is an estimate of the squared Euclidean distance between two
+populations. It thus corresponds to the squared distance in PCA-space, and
+reflects that closely related populations  will be close to each other on a
+PCA-plot, and have low pairwise $F_2$-statistics. In converse, if two
+populations with high $F_2$ lie on the same point on an PCA-plot, this suggests
+that substantial variation is hidden on higher PCs.
 
@@
 
-\section{Trees and admixture graphs in PCA-space}
-\subsection{Trees}
-Evolutionary trees are fundamental in phylogenetic analyses, as they, on a large, scale, approximate how taxa diversify.
-Within a species, applying trees is also very common, but more problematic as populations frequently do not evolve as discrete lineages; instead, they admix and diversify as much more continuous processes.
-This is largely due to the time-scales involved, speciation events that give rise to trees might often be similarly messy, but from a distance of millions of years these issues might disappear.
-
-Thus, when estimating trees from population genetic data, we must be very careful about whether the data is actually consistent with a tree, or belongs to some wider class of model.
-
-Trees can be thought of as a collection of orthogonal dimensions; as drift on each branch is independent from every other branch. Thus, each sample is only 
-\begin{enumerate}
-	\item Trees
-	\item Admixture Graphs
-	\item Treelets
-	\item simple trees, admixture graph
-\end{enumerate}
-
-\section{Technical considerations}
-	\subsection{SNP weighting}
-	It is clear that weighting SNP will have some effect on the resulting PCAs. Upweighting rare variants e.g. will emphasis recent events, as rare variance in the sample are more likely to be recent.
-
-	\subsection{Missing data}
-
-	\subsection{$F_2$ error}
-	In most $F$-statistics applications, $F_2$ is \emph{estimated} using the minimum-variance unbiased estimator \citep{reich2009}
-	\begin{subequations}
-
-	\begin{equation}
-	f_2(X_1, X_2) = \frac{1}{L}\sum_l (x_{l1} - x_{l2})^2 - \frac{1}{L}\sum_l\frac{x_{l1} (1-x_{l1})}{n_{l1}-1} - 
-	\frac{1}{L}\sum_l\frac{x_{l2} (1-x_{l2})}{n_{l2}-1} - 
-	\end{equation}
-	in contrast, as shown above, PCA can be thought as a decomposition of a matrix of uncorrected $F_2$-statistics:
-	\begin{equation}
-	F_2(X_1, X_2) = \frac{1}{L}\sum_l(x_{l1} - x_{l2})^2
-	\end{equation}
-	\end{subequations}
-
-	This leads to some issues, for example trying to perform a PCA on the matrix of $f_2$-values is not positive semidefinite, and so some principal components may be imaginary.
-One possible resolution is probabilistic PCA \citep[e.g.][]{engelhardt2010,agrawal2020}.
-
-	\begin{align}
-	    y_{ij} | \MP_{ij}, \epsilon_i &= (\MP\ML)_{ij} + \epsilon_{ij}\nonumber\\
-	    x_i & \sim N(0, \mathbf{I}) \nonumber\\
-	    \epsilon_i &\sim N(0, \sigma^2 \mathbf{I})\nonumber
-	\end{align}
-
-	\subsection{qpADM}
-	In \cite{haak2015}, qpADM, a procedure to estimate admixture proportions has been proposed. qpADM aims to solve equations of the form
-
-	\begin{align}
-	\langle P_X - A, B -C \rangle &= \sum_i  \alpha_i\langle R_i - A, B - C \rangle \nonumber\\
-	&= \left\langle \sum_i  \alpha_i R_i - A, B - C \right\rangle
-	\end{align}
-
 	\subsection{What is a dimension?}
@@
 
-	\section{outtakes}
-	PCA from $\MX$
-	\begin{equation}
-	\MK = \MY \MY^T = \MC\MX\MX^T \MC = \MP\MP^T
-	\end{equation}
+\section{Results}
+The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots.
+In this section, I use these interpretation in the analysis of human genetic variation data set.
+I use two data sets based on the ``Human Origins''-SNP set (597,573 SNPs).
+Both are subsets of the Reich lab compendium data set v44.3, downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}.
+
+\paragraph{West-Eurasian data set}
+This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
+It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \cite{patterson2012}.
+Population genetic differentiation in this region is low and closely mirrors geography \cite{novembre2008}.
+
+\paragraph{World Overview data set}
+This data set of 638 individuals from 33 population contains individuals throughout the world, and is used as a sparse data set capturing much of global human genetic variation.
+This data set spans Africa, Eurasia and the Americas, and we might therefore expect the population structure to be much more sparse.
+
+I perform analyses at the level of populations to ease presentation, and because it is an assumption of $F$-statistics that the genetic variation with sampled population is independent of the variation between samples that I am focusing on here.
+I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute a matrix of $F_2$-statistics between all populations.
+To obtain a PC-decomposition I use equation XXX and the \texttt{eigen} function in \texttt{R}, and compare them with the $F_3$ and $F_4$-statistics calculated using admixtools 2.
+
+\paragraph{Admixture-$F_3$}
+As a first step, I plot the first two principal components of the Westeurasian data set (Figure \ref{fig:f3}A).
+This PCA presents two parallel clines, one from the Levant and Arabia (``BedouinB'') to the Caucasus (``Abkhasian''), and a second one from Southern (``Sardinian'') to Northeastern Europe (``Mordovian'').
+In this context, I examine $F_3(X; Basque, Turkish)$, i.e.
+a statistic that aims to ask which populations can be represented as a mixture between a Southwestern (Basque) and Southeastern (Turkish) European population.
+The -- largely Southern European -- populations for which the point estimate of these $F_3$-statistic is negative are highlighted in red.
+They both fall close to the center of the $F_3$-circle, either defined on the first two (dark grey) or all PCs (light gray).
+However, many populations inside the circle on the first 2 PCs, including English, Sardinians and Canary Islanders have positive $F_3$-values, on higher PCs, showing that the first two PCs do not capture all the genetic variation related to population structure for this data set.
+
+This is expected because for spatially continuous populations, PCA will not be sparse \cite{novembre2008a}.
+Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
+
+Thus, the main benefit of this PCA-plot is that it allows us to identify populations outside the circle (from the Levant and Caucasus), for which $F_3$ is guaranteed to be positive.
+
+\paragraph{Outgroup-$F_3$}
+The Outgroup-$F_3$-statistic is commonly used to infer which population is closest in a set of reference populations.
+In Figure \ref{fig:f3}D, I present a PCA of the world data set, with populations colored according to $F_3(Mbuti; Mozabite, X_i)$, i.e.
+a statistic that is commonly interpreted as finding the population $X_i$ that is most closely related to Mozabite.
+On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment $\text{Mbuti}X_i$ onto the line through Mbuti and Mozabite (black line).
+For each population, the projection is indicated with a grey line.
+In the full data space, this line is always orthogonal to the segment Mbuti-Mozabite, but on the plot (i.e.) the subspace spanned by the first two PCs, this is only true if the relevant variation is captured by the first two PCs.
+We see that particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs.
+That the approximation of $F_3$ on
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
+	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
+Populations for which $F_3$ is negative are colored in red.
+B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
+C,F: Contributions of PCs 1-10 to each $F_3$-statistic.
+D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$.
+The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population.
+}
+	\label{fig:f3}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{figures/fig_f4_data.pdf}	
+	\caption{\textbf{PCA and $F_4$-statistics} A: Spectrum of select $F_4$-statistics in World data set.
+B: Projection angle representation of $F_4(X, Sardinian; Mozabite, Yoruba)$ (red) and approximations using two (blue) and ten (green) PCs.
+C: Spectrum of select $F-4$-statistics in Westeurasian data set.
+D: Scatterplot of $F_4$-projection on Finnish-Canary Islanders axis and residual PC1.
+	E: $F_4(X, \text{French}, \text{Finnish}, \text{Canary Islander})$ vs.
+prediction using two (blue) and ten(red) PCs.
+F: Percent variance explained for the projection of panel D and the first nine residual PCs.
+	}
+\label{fig:f4}
+\end{figure}
 
 \section{Discussion}
-The fa
+\begin{itemize}
+    \item weighting of SNPs
+    \item estimation error
+    \item propagating erros
+    \item exploratory data analysis
+    \item missing data
+    \item population vs. sample allele frequencies
+\end{itemize}
 

### 2021-07-01 fstats_pca_lewontin.tex
fig 4 text.

--- 2021-06-24 fstats_pca_lewontin.tex
@@
 
-\subsection{Higher-Dimensional Spaces}
+\section{Results}
+\subsection{$F_4$}
+Using $F_3$-statistics, I showed that we can think of the admixture test as a test of whether the admixed population lies in a particular $n$-ball, and the outgroup $F_3$-statistic can be thought of as a projection of the test populations on the line connecting the outgroup to the reference sample.
+In this section, I will develop similar interpretation of $F_4$ on PCA-plots, and to investigate sparsity.
+
+First, we investigate the sparsity in the world overview data set: We find that the vast amount of contribution to the statistics comes from the first two PCs (Figure \ref{fig:f4}A).
+For example, the correlation between $F_4(X, Y, \text{Mozabite}, \text{Yoruba})$ and its approximation using the first two PCs is 99.2\%.
+To visualize the interpretation of $F_4$ as an angle, we use statistics of the form $F_4(X, \text{Sardinian}; \text{Mozabite}, \text{Yoruba})$, which can be interpreted as the angle between the vectors Mozabite-Yoruba and $X$-Sardinian.
+In Figure \ref{fig:f4}B, I show the angle based on two (blue), ten (green) and all PCs $red$.
+I find that for most Asian and American populations the angle is very close to 90°, as would be expected if the variation between African and non-African populations is mostly orthogonal.
+On the other hand, if $X$ is an African population, the angle is lower, and much less well approximated.
+This demonstrates that this PCA-plot likely does not model within-African population structure adequately.
+
+The $F_4$-statistics for the Westeurasian data set are slightly less sparse, the correlation coefficients between $F_4(X, \text{French}; \text{Finnish}, \text{Canary Islander})$ and its approximation using the first two or three PCs is 95.5\% and 99.1\% respectively (Figure \ref{fig:f4}E).
+I also show that the interpretation of $F_4$ as a projection can be used as a useful visualization (Figure \ref{fig:f4}D).
+On the $x$-axis, I plot $\langle X; \text{Finnish}, \text{Canary Islander}\rangle$, so that the horizontal distance between all pairs of populations corresponds to their respective $F_4$-statistics $F_4(X, Y; \text{Finnish}, \text{Canary Islander})$.
+On the $Y$-axis and with the coloring I display the first two principal components of the residual, i.e.
+the genetic variation that is missed by viewing the data through this projection.
+We find that most European populations have positive values on residual PC1, and are relatively closely clustered.
+In contrast Middle Eastern and Caucasian populations have negative values on this gradient.
+This allows us to visualize that this particular $F_4$-projection does an adequate job if we are interested in describing European variation, but it fails to explain the non-European data.
+We can further quantify this by investigating the percent of variance explained on each axis (Figure \ref{fig:f4}F), where I find that the projection axis only describes around 12\% of hte variation, compared to residual PC1 with almost 30\%.
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
+	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
+Populations for which $F_3$ is negative are colored in red.
+B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
+C,F: Contributions of PCs 1-10 to each $F_3$-statistic.
+D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$.
+The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population.
+}
+	\label{fig:f3}
+\end{figure}
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fig_f4_data.pdf}	
+	\caption{\textbf{PCA and $F_4$-statistics} A: Spectrum of select $F_4$-statistics in World data set.
+B: Projection angle representation of $F_4(X, Sardinian; Mozabite, Yoruba)$ (red) and approximations using two (blue) and ten (green) PCs.
+C: Spectrum of select $F-4$-statistics in Westeurasian data set.
+D: Scatterplot of $F_4$-projection on Finnish-Canary Islanders axis and residual PC1.
+		E: $F_4(X, \text{French}, \text{Finnish}, \text{Canary Islander})$ vs.
+prediction using two (blue) and ten(red) PCs.
+F: Percent variance explained for the projection of panel D and the first nine residual PCs.
+	}
+	\label{fig:f4}
+\end{figure}
 

### 2021-07-01 data/f4_fig.R

@@ -52,7 +52,9 @@ f4pc2 = f4_from_pc_matrix(pcmat, idz, idx, idy) %>%
         as.data.frame %>% rownames_to_column("pop1") %>%
         left_join(f42, .) %>% 
         filter(! pop1 %in% ids) %>%
-        mutate(PC1_2=rowSums(across(PC1:PC2)), PC1_10=rowSums(across(PC1:PC10)))
+        mutate(PC1_2=rowSums(across(PC1:PC2)), 
+               PC1_3=rowSums(across(PC1:PC3)),
+               PC1_10=rowSums(across(PC1:PC10)))
     v = f4pc2 %>% 
         filter(pop1 %in% c("Greek", "Iranian", "Sicilian", "Estonian", "Saudi", "Georgian", "Spanish")) %>% 
         pivot_longer(PC1:PC33) %>% 
@@ -87,6 +89,13 @@ P6 = f4pc %>%
     geom_point() + 
     xlab("f4(X, FRA; FIN, CI)") + ylab("approx. f4")  + BTHEME
 
+f4all = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+f4pcall = f4_from_pc_matrix2(pcmat, px=idx, py=idy) %>%
+    #as.data.frame %>% rownames_to_column("pop1") %>%
+    left_join(f4all, .) %>% 
+    filter(! pop1 %in% ids) %>%
+    mutate(PC1_2=rowSums(across(PC1:PC2)), PC1_4=rowSums(across(PC1:PC4)))
+
 
 
 }
@@ -107,6 +116,13 @@ if(F){
         select(pop1, pop2, est) %>% 
         pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>% 
         column_to_rownames('pop1')
+    f4all = f4(f2s, pop3=idy, pop4 = idz, pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+    f4pcall = f4_from_pc_matrix2(pcmat, px=idy, py=idz) %>%
+        #as.data.frame %>% rownames_to_column("pop1") %>%
+        left_join(f4all, .) %>% 
+        filter(! pop1 %in% ids) %>%
+        mutate(PC1_2=rowSums(across(PC1:PC2)), PC1_10=rowSums(across(PC1:PC20)))
+
[…]

### 2021-07-01 fstats_pca_lewontin.tex
discussion

@@
 Hence, neither mean-centering (a translation) nor PCA (a rotation) will change $F_2$.
-What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or the principal components $\MP$.
+What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or any other orthogonal basis such as the principal components $\MP$.
 Formally,
@@
 
-\section{Results}
-\subsection{$F_4$}
-Using $F_3$-statistics, I showed that we can think of the admixture test as a test of whether the admixed population lies in a particular $n$-ball, and the outgroup $F_3$-statistic can be thought of as a projection of the test populations on the line connecting the outgroup to the reference sample.
-In this section, I will develop similar interpretation of $F_4$ on PCA-plots, and to investigate sparsity.
-
-First, we investigate the sparsity in the world overview data set: We find that the vast amount of contribution to the statistics comes from the first two PCs (Figure \ref{fig:f4}A).
-For example, the correlation between $F_4(X, Y, \text{Mozabite}, \text{Yoruba})$ and its approximation using the first two PCs is 99.2\%.
-To visualize the interpretation of $F_4$ as an angle, we use statistics of the form $F_4(X, \text{Sardinian}; \text{Mozabite}, \text{Yoruba})$, which can be interpreted as the angle between the vectors Mozabite-Yoruba and $X$-Sardinian.
-In Figure \ref{fig:f4}B, I show the angle based on two (blue), ten (green) and all PCs $red$.
-I find that for most Asian and American populations the angle is very close to 90°, as would be expected if the variation between African and non-African populations is mostly orthogonal.
-On the other hand, if $X$ is an African population, the angle is lower, and much less well approximated.
-This demonstrates that this PCA-plot likely does not model within-African population structure adequately.
-
-The $F_4$-statistics for the Westeurasian data set are slightly less sparse, the correlation coefficients between $F_4(X, \text{French}; \text{Finnish}, \text{Canary Islander})$ and its approximation using the first two or three PCs is 95.5\% and 99.1\% respectively (Figure \ref{fig:f4}E).
-I also show that the interpretation of $F_4$ as a projection can be used as a useful visualization (Figure \ref{fig:f4}D).
-On the $x$-axis, I plot $\langle X; \text{Finnish}, \text{Canary Islander}\rangle$, so that the horizontal distance between all pairs of populations corresponds to their respective $F_4$-statistics $F_4(X, Y; \text{Finnish}, \text{Canary Islander})$.
-On the $Y$-axis and with the coloring I display the first two principal components of the residual, i.e.
-the genetic variation that is missed by viewing the data through this projection.
-We find that most European populations have positive values on residual PC1, and are relatively closely clustered.
-In contrast Middle Eastern and Caucasian populations have negative values on this gradient.
-This allows us to visualize that this particular $F_4$-projection does an adequate job if we are interested in describing European variation, but it fails to explain the non-European data.
-We can further quantify this by investigating the percent of variance explained on each axis (Figure \ref{fig:f4}F), where I find that the projection axis only describes around 12\% of hte variation, compared to residual PC1 with almost 30\%.
-
-\begin{figure}[!ht]
-	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
-	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
-Populations for which $F_3$ is negative are colored in red.
-B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
-C,F: Contributions of PCs 1-10 to each $F_3$-statistic.
-D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$.
-The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population.
-}
-	\label{fig:f3}
-\end{figure}
-
-\begin{figure}[!ht]
-	\includegraphics[width=\textwidth]{figures/fig_f4_data.pdf}	
-	\caption{\textbf{PCA and $F_4$-statistics} A: Spectrum of select $F_4$-statistics in World data set.
-B: Projection angle representation of $F_4(X, Sardinian; Mozabite, Yoruba)$ (red) and approximations using two (blue) and ten (green) PCs.
-C: Spectrum of select $F-4$-statistics in Westeurasian data set.
-D: Scatterplot of $F_4$-projection on Finnish-Canary Islanders axis and residual PC1.
-		E: $F_4(X, \text{French}, \text{Finnish}, \text{Canary Islander})$ vs.
-prediction using two (blue) and ten(red) PCs.
-F: Percent variance explained for the projection of panel D and the first nine residual PCs.
-	}
-	\label{fig:f4}
-\end{figure}
-
 	\subsection{What is a dimension?}
@@
 
+\subsection{$F_4$}
+Using $F_3$-statistics, I showed that we can think of the admixture test as a test of whether the admixed population lies in a particular $n$-ball, and the outgroup $F_3$-statistic can be thought of as a projection of the test populations on the line connecting the outgroup to the reference sample.
+In this section, I will develop similar interpretation of $F_4$ on PCA-plots, and to investigate sparsity.
+
+First, we investigate the sparsity in the world overview data set: We find that the vast amount of contribution to the statistics comes from the first two PCs (Figure \ref{fig:f4}A).
+For example, the correlation between $F_4(X, Y, \text{Mozabite}, \text{Yoruba})$ and its approximation using the first two PCs is 99.2\%.
+To visualize the interpretation of $F_4$ as an angle, we use statistics of the form $F_4(X, \text{Sardinian}; \text{Mozabite}, \text{Yoruba})$, which can be interpreted as the angle between the vectors Mozabite-Yoruba and $X$-Sardinian.
+In Figure \ref{fig:f4}B, I show the angle based on two (blue), ten (green) and all PCs $red$.
+I find that for most Asian and American populations the angle is very close to 90°, as would be expected if the variation between African and non-African populations is mostly orthogonal.
+On the other hand, if $X$ is an African population, the angle is lower, and much less well approximated.
+This demonstrates that this PCA-plot likely does not model within-African population structure adequately.
+
+The $F_4$-statistics for the Westeurasian data set are slightly less sparse, the correlation coefficients between $F_4(X, \text{French}; \text{Finnish}, \text{Canary Islander})$ and its approximation using the first two or three PCs is 95.5\% and 99.1\% respectively (Figure \ref{fig:f4}E).
+I also show that the interpretation of $F_4$ as a projection can be used as a useful visualization (Figure \ref{fig:f4}D).
+On the $x$-axis, I plot $\langle X; \text{Finnish}, \text{Canary Islander}\rangle$, so that the horizontal distance between all pairs of populations corresponds to their respective $F_4$-statistics $F_4(X, Y; \text{Finnish}, \text{Canary Islander})$.
+On the $Y$-axis and with the coloring I display the first two principal components of the residual, i.e.
+the genetic variation that is missed by viewing the data through this projection.
+We find that most European populations have positive values on residual PC1, and are relatively closely clustered.
+In contrast Middle Eastern and Caucasian populations have negative values on this gradient.
+This allows us to visualize that this particular $F_4$-projection does an adequate job if we are interested in describing European variation, but it fails to explain the non-European data.
+We can further quantify this by investigating the percent of variance explained on each axis (Figure \ref{fig:f4}F), where I find that the projection axis only describes around 12\% of hte variation, compared to residual PC1 with almost 30\%.
+
 \begin{figure}[!ht]
@@
 \section{Discussion}
+Particularly for the analysis of ancient DNA, $F$-statistics have been established as a powerful tool to describe population genetic diversity, but they have a number of limitations.
+In particular, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \cite{patterson2012,harney2020}.
+As a consequence, researchers concerned about model fits may ascertain reference populations in a way that satisfies these assumptions, thus inadvertently making population structure appear sparser than it truly is.
+This is perhaps most obvious from Figure \ref{fig:f3}B, where large gaps are present.
+However, these gaps are due to sparse sampling that disappear if more populations were sampled \cite[e.g.][]{peter2020a}, not due to gaps in genetic diversity.
+If population ascertainment is not done very carefully, tools built on top of $F$-statistics, such as \texttt{qpGraph} and \texttt{qpAdm}, may thus only provide a very loose lower bound for the number of gene flow events.
+
+In contrast, the perspective on $F$-statistics in data space \citep{oteo-garcia2021} and on PCA does not require assumptions on number of admixture or gene flow events.
+Independent of any model, A population $X_x$ can be thought of as admixed between $ X_1$ and $X_2$ if it lies in the ball with diameter $\overline{X_1X_2}$.
+And independent of any models, two axes of variation that are not perpendicular to each other shows that there is some degree of shared history.
+
+The connection between the graph-based and PCA-based interpretations of $F_3$ and $F_4$ is due to concept of projection.
+As illustrated above, both $F$-statistics can be thought of as projections on any particular axis of variation.
+If that variation corresponds to a tree-branch, the interpretations align.
+A collorary to this interpretation is the importance of orthogonality, or independence for describing gene flow.
+Populations without gene-flow will evolve independently, and their changes in allele frequency will therefore be orthogonal, resulting in $F_4$-statistics of zero, and right angles on a PCA plot.
+Even though this is only true when considering all PCs, the result holds reasonably well when just looking at the first two PCs -- in our example of the world-overview data, I found that the variation in Sub-Saharan Africa is mostly -- but not completely -- uncorrelated with the variation between European and Asian populations (Figure \ref{fig:f4}A).
+
+While the data space produces a much larger model space than trees, the main drawback of the PCA-based interpretation is a lack of interpretability; it is not easy to define a generative model that could generate a complex PCA-plot.
+
+ However, despite the apparent complexity, the PC-spaces \emph{are} sparse, and using just ten or even two PCs often gives very good approximations.
+
+To make PCA and $F$-statistics more comparable in practical settings, there are a number of -- mainly statistical -- concerns that still need to be addressed in future work.
+The perhaps most obvious one is that PCA is most frequently run on individuals, whereas $F$-statistics are often calculated on populations.
+This is not a conceptual issue, as both PCA and $F$-statistics can be run on either \cite{cavalli-sforza1994}.
+Population based analyses have the advantage that they are easier to interpret and compute (current packages are ill-equipped to calculate all pairwise $F$-statistics between data sets with thousands of individuals \cite{patterson2012}).
+However, this requires the assumption that the within-population variation is independent from the between-population variation; something that is analogous to the variance partitioning based on PCs here.
+
+A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \cite{patterson2006}.
+This is only a difference of convention; $F$-statistics could also be calculated using the same weighting.
+The close connection between the two approaches developed here suggest that for most analyses, users might want to be consistent and use the same weighting for both types of analyses.
+
+The third and perhaps biggest gap are statistical issues.
+The treatment here focusses on the mean estimated $F$-statistic, but many  applications of $F$-statistics are based on hypothesis tests \cite{patterson2012}.
+This requires estimating accurate standard errors for these statistics, which is difficult since nearby SNPs will be correlated \cite{hahn2019}.
+In contrast, standard PCA does not modeljointly models the covariance matrix due to population structure and sampling.
+On the other hand, for both data sets I investigated here, the matrix $\MF$ of $F$-statistics estimated using admixtools2 is not a proper squared Euclidean distance matrix, i.e.
+it is not negative semidefinite and has imaginary PCs.
+This is not a practical when considering single $F$-statistics or PCA (for analyses here, I used a nearby matrix  \citep{higham2002} with no apparent loss of precision).
+It does however mean that tools that use matrices of $F$-statistics, such as \texttt{qpadm} or \texttt{qpgraph} may be ill-calibrated, which may partly explain why they generally have poor out-of-sample predictive power and are restricted to a few dozen samples at a time.
+A model-based framework based on probabilistic PCA \citep{meisner2021, agrawal2020, softimpute} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
+
 \begin{itemize}

### 2021-07-01 data/fscripts.R
individual-based analyses draft

@@ -1,15 +1,15 @@
 require(admixtools)
 require(tidyverse)
 require(Matrix)
+require(ggforce)
 
 
-pca_from_f2s <- function(f2s){
-    f2 = f2(f2s)
+pca_from_f2s <- function(f2s, force_nonneg=F){
+    f2 = f2(f2s, unique_only=F, sure=T) 
     n = dim(f2s)[1]
     f2mat = f2 %>%
-        mutate(tmp=pop2, pop2=pop1, pop1=tmp) %>% 
-        select(-tmp) %>% bind_rows(f2) %>% 
-        select(-se) %>% 
+        mutate(est=if(force_nonneg){pmax(est, 0)}else{est}) %>%
+        select(-se) %>%
         pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>%
         arrange(pop1) %>%
         column_to_rownames('pop1') %>% as.matrix

### 2021-07-01 data/prep_data_pca_worldfocus2.R

@@ -15,5 +15,5 @@ f2loc1 = 'worldfoci2'
 #                              cols_per_chunk=1000)
 f2s = admixtools::extract_f2(data_loc, 
                              outdir=f2loc1, maxmem=64000, 
-                             overwrite=T)
+                             overwrite=T, n_cores=100)
 

### 2021-07-01 data/f3_fig_ind.R
individual based analyses scripts

@@ -1,0 +1,121 @@
+source("fscripts.R")
+library(ggpubr)
+BTHEME = theme_classic() + theme(legend.position="none")
+
+if(T){
+f2s = admixtools::read_f2("westeurasian1ind/")
+pcmat = pca_from_f2s(f2s)
+pcs = pca_from_pcmat(pcmat)
+idx = "Basque1"
+idy = "Turkish1"
+
+sample_list = pcs$pop 
+
+f3 = f3(f2s, pop3=idx, pop2=idy, pop1=sample_list) #admixture
+f3pc = f3_from_pc_matrix(pcmat, idx, idy)
+#f3 = f3(f2s, pop1=idx, pop2=idy, pop3=sample_list) #outgroup
+f4 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+
+circle = pcs %>% filter(pop %in% c(idx, idy)) %>% select(-pop)
+circle_df = colMeans(circle) %>% t %>% as.data.frame
+circle_df$radius = (dist(circle) %>% as.matrix)[1,2] / 2
+circle_df$r0 = (dist(circle[,1:2]) %>% as.matrix)[1,2] / 2
+
+pcs2 = pcs %>% left_join(f3, by=c(pop="pop1")) %>% 
+    mutate(col=as.factor(cut(est, c(-1,-3e-4,1e-10,2), labels=F) ))
+P1 = ggplot() + 
+    geom_circle(aes(x0=PC1, y0 =PC2, r=radius), data=circle_df, color='lightgray', fill='lightgray') +
+    geom_circle(aes(x0=PC1, y0 =PC2, r=r0), data=circle_df, color='darkgray', fill='#808080') +
+    geom_point(aes(x=PC1, y=PC2, label=pop, color=col), data=pcs2) +
+    coord_fixed()  + BTHEME
+
+f3pc = f3_from_pc_matrix(pcmat, idx, idy) %>% 
+    as.data.frame %>% rownames_to_column("pop1") %>% 
+    left_join(f3, .) %>% 
+    mutate(PC1_2=rowSums(across(PC1:PC2)), PC1_10=rowSums(across(PC1:PC10)))
+
+P2 = f3pc %>% select(pop1, est, PC1_2, PC1_10) %>% 
+    pivot_longer(starts_with("PC"), names_to='PC', values_to="f3") %>% 
[…]

### 2021-07-01 data/f4_fig_ind.R

@@ -1,0 +1,183 @@
+source("fscripts.R")
+library(glue)
+BTHEME = theme_classic() + theme(legend.position="none")
+
+if(F){
+f2s = admixtools::read_f2("westeurasian1/")
+pcmat = pca_from_f2s(f2s)
+pcs = pca_from_pcmat(pcmat)
+idx = "Finnish"
+idy = "Canary_Islander"
+idz = "French"
+ids = c(idx, idy)
+
+sample_list = pcs$pop 
+
+f4 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+f4mat = f4 %>% 
+    select(pop1, pop2, est) %>% 
+    pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>% 
+    column_to_rownames('pop1')
+f4pc = f4_from_pc_matrix2(pcmat, idx, idy) %>%
+    left_join(f4, .) %>% 
+    filter(! pop1 %in% ids, ! pop2 %in% ids) %>%
+    mutate(PC1_2=rowSums(across(PC1:PC2)), PC1_10=rowSums(across(PC1:PC10)))
+v = pcmat[ids,] %>% diff
+pmat = t(v) %*% v / (v %*% t(v))[1,1]
+qmat = diag(length(v)) - pmat
+a = pcmat %*% pmat
+b = pcmat %*% qmat
+A = t(t(a) - a[idz,])
+B = t(t(b) - b[idz,])
+E = eigen(b %*% t(b))
+K = t(t(E$vectors) * sqrt(pmax(E$values, 0)))
+K = K %>% as.data.frame %>% mutate(pop = rownames(B))
+
+EA = eigen(a %*% t(a))
+KA = t(t(EA$vectors) * sqrt(pmax(EA$values, 0)))
+KA = KA %>% as.data.frame %>% mutate(pop = rownames(B))
+X = KA %>% select(pop, A=V1) %>% left_join(K)
+
+P3 = X %>% 
+    ggplot(aes(x=A, y=V1, label=pop, color=V2)) + 
+    geom_text(size=3) + 
+    coord_fixed() + 
+    BTHEME + 
+    scale_color_viridis_c() + 
+    xlab("proj (Finnish, Canary Islander)") + 
+    ylab("Residual PC1")
+
+f42 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=idz) %>% arrange(pop1, pop2)
[…]

### 2021-07-01 data/prep_data_pca_westeurasian_ind.R

@@ -1,0 +1,21 @@
+require(admixtools)
+require(tidyverse)
+require(glue)
+
+
+data_loc = 'subdata/westeurasian1ind'
+inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
+
+
+
+f2loc1 = 'westeurasian1ind'
+
+#afs = admixtools::extract_afs(data_loc, pops=sample_list, 
+#                              outdir=glue("{f2loc1}_afs"),format="eigenstrat", 
+#                              cols_per_chunk=1000)
+f2s = admixtools::extract_f2(data_loc, 
+                             inds=inds$iid,
+                             outdir=f2loc1, maxmem=500000, 
+                             n_cores=140,
+                             overwrite=T)
+

### 2021-07-01 data/prep_data_pca_worldfocus3.R

@@ -1,0 +1,21 @@
+require(admixtools)
+require(tidyverse)
+require(glue)
+
+
+data_loc = 'subdata/worldfoci2ind'
+inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
+
+
+
+f2loc1 = 'worldfoci2ind'
+
+#afs = admixtools::extract_afs(data_loc, pops=sample_list, 
+#                              outdir=glue("{f2loc1}_afs"),format="eigenstrat", 
+#                              cols_per_chunk=1000)
+f2s = admixtools::extract_f2(data_loc, 
+                             inds=inds$iid,
+                             n_cores=120,
+                             outdir=f2loc1, maxmem=240000, 
+                             overwrite=T)
+

### 2021-07-01 data/subdata/westeurasian.R
added par files

@@ -1,0 +1,7 @@
+require(tidyverse)
+system("ln -s westeurasian1.geno westeurasian1ind.geno")
+system("ln -s westeurasian1.snp westeurasian1ind.snp")
+x = read.table("westeurasian1.ind")
+x = x %>% group_by(V3) %>% mutate(V3=sprintf("%s%d", V3, order(V1)))
+write.table(x, "westeurasian1ind.ind", quote=F, col.names=F, row.names=F)
+

### 2021-07-01 data/subdata/worldfoci2.R

@@ -1,0 +1,7 @@
+require(tidyverse)
+system("ln -s worldfoci2.geno worldfoci2ind.geno")
+system("ln -s worldfoci2.snp worldfoci2ind.snp")
+x = read.table("worldfoci2.ind")
+x = x %>% group_by(V3) %>% mutate(V3=sprintf("%s%d", V3, order(V1)))
+write.table(x, "worldfoci2ind.ind", quote=F, col.names=F, row.names=F)
+

### 2021-07-01 fstats_pca_lewontin.tex
small changes

--- 2021-07-01 fstats_pca_lewontin.tex
@@
 \usepackage{url}
+\usepackage{textcomp}
 
@@
 In Figure \ref{fig:f4}B, I show the angle based on two (blue), ten (green) and all PCs $red$.
-I find that for most Asian and American populations the angle is very close to 90°, as would be expected if the variation between African and non-African populations is mostly orthogonal.
+I find that for most Asian and American populations the angle is very close to 90\textdegree, as would be expected if the variation between African and non-African populations is mostly orthogonal.
 On the other hand, if $X$ is an African population, the angle is lower, and much less well approximated.

### 2021-07-02 fstats_pca_lewontin.tex
local changes
fixed references, better intro

--- 2021-07-01 fstats_pca_lewontin.tex
@@
 ]{geometry}
-\usepackage[utf8]{inputenc}
+\usepackage[utf8x]{inputenc}
 %\usepackage[T1]{fontenc}
@@
 \section{Introduction}
-About 15\% of genetic variation in humans can be explained by population structure \cite{lewontin1972, barbujani1997, rosenberg2003}, but the information contained in these 15\% is sufficient to study the genetic diversity and history in great detail \cite{edwards2003}.
-For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \cite{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
+About 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002}, but the information contained in these 15\% is sufficient to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}.
+For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
+
+Understanding and characterizing this variation is also crucial for genomic medicine: Risk loci identified in one population perform progressively worse in more distant populations \citep{berg2019, duncan2019}; so identifying a good ``reference'' data set is crucial.
+Understanding genetic variation is also important when designing studies, in order to measure which populations studies should target \citep{fullerton2016}.
+Despite these needs, no generally accepted model for human population structure exists.
+The historical view that humans diversity can be partitioned into  distinct ``races'' \citep[e.g.][]{blumenback1798} has long been superseded in favor of more accurate and nuanced models.
+These models fail because they only explain a small amount of variation...
 
@@
 
-One related question is how discrete human populations are.
-While human genetic differentiation generally increases with geographic distance \cite{ramachandran2005, rosenberg2007}, this increase is not uniform.
-Obstacles to migration,  such as oceans, mountains or deserts do frequently cause discontinuities in population structure \cite{peter2020a}.
-Thus, while barriers to gene flow rarely are absolute, segregation policies by (perceived) ethnic or racial ancestry frequently cause local small-scale population differentiation that persist to the present day.
+Human genetic diversity has both discrete and continuous components \citep{rosenberg2002, rosenberg2005, bradburd2015,  reich2018, peter2020a}.
+On the one hand, barriers to migration such as mountain ranges, oceans or deserts lead to reduced gene flow, and thus populations on different continents had periods with little-to-no gene flow.
+On the other hand, local gene flow between neighboring populations is also pervasive, and the out-of-Africa expansion caused gradual patterns of differentiation \cite{ramachandran2005}.
+Finally, long-distance migrations, and secondary contact between diverged groups such as Neandertals and early modern humans add to the complexity of human population structure.
 
-Thus, it frequently a useful analysis tool to think of populations as discrete units.
-For example, even though the underlying population structure may be continuous, sampling is not; and when quantifying ascertainment and sampling biases, or when discussing population structure it is often helpful to pretend populations are discrete, even though the underlying structure is typically more complex.
-This leads to challenges both in data interpretation and communication, and often researchers will analyze a data set both using methods that assume population structure is discrete, and methods where this assumption does not need to be made.
+The way this complexity is frequently handled is that multiple models with different assumptions and used, each emphasizing a particular aspects of the data.
+Data-driven methods such as Principal Component Analysis (PCA) \citep{cavalli-sforza1994} structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable.
+For this more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are used, which allow for parameter estimation or hypothesis tests.
+Particularly in the analysis of human ancient DNA, a set of techniques based on $F$-statistics \textit{sensu} Patterson have risen in popularity \citep{patterson2012, peter2016}.
+This framework is based on the assumption that the relationship between three or four populations is often tree-like, and allows for a variety of tests of treeness and more complex demographic models \citep{patterson2012, harney2021}.
 
-One discrete framework for the analysis of human population structure that gained a lot of traction in the last decade are the $F$-statistics \textit{sensu} Patterson \cite{patterson2012, peter2016}.
-This framework treats populations as discrete units in the analysis, and allows for a variety of tests for treeness.
-Using this framwork, the vast majority of present-day human populations are admixed \cite{pickrell2014}.
-Yet, this framework starts with the assumption that admixture is i) rare and ii) discrete.
-
-However, $F$-statistics are not restricted to discrete populations.
-Indeed, as they can be written as functions of allele frequency variances, or expected pairwise coalescence times, statistics that can be calculated under a wide range of demographic models \cite{peter2016}.
-Indeed, as they reflect inner products, they can be generalized to Euclidean space \cite{oteo2021} (or any Hilbert space, although we won't pursue that here).
-Here, I explore these links between $F$-statistics and Euclidean spaces to establish connections between $F$-statistics and PCA.
-This allows direct interpretation of admixture in scenarios where population structure might not be discrete.
-
-Particularly for the analysis of ancient DNA, two approaches have been proven to be particularly useful: one are global summary analyses, such as Structure \citep{pritchard2000, alexander2009} Principal Component Analysis (PCA) \citep{cavalli-sforza1994, reich2008, novembre2008, mcvean2009} and classical multidimensional scaling (MDS) \cite{fu2016, malaspinas_bammds}.
-Typically, these methods assume that population structure is \emph{sparse}, so that a low-rank approximation with few underlying ``components'' is sufficient to model population structure See e.g.
-\cite{engelhardt2010} for a useful perspective how these approaches are related.
-
-Facing a novel data set, PCA or MDS are often the first analyses (beyond quality controls) a researcher performs, in order to obtain insights in the general population structure they are faced with.
-In order to answer more specific questions and to test specific hypotheses, the $F$-statistic framework of \cite{patterson2012} has been proven particularly powerful (see also \cite{peter2016} for a more gentle introduction).
-In the $F$-statistic framework, usually only a small number of populations are used at once, to e.g.
-test for treeness and find closely related populations.
-
-Even though these two approaches are considered in almost every ancient DNA paper, links between the inferences made from them are usually only compared qualitatively.
-In this paper, our goal is to show that PCA and $F$-statistics are in fact closely related by construction, and use a very similar summary of the data.
+However, the connections between PCA, $F$-statistics and demographic models are currently unclear, which makes quantitative comparisons, detecting model violations and joint interpretation of the results of these approaches difficult.
+Since both $F$-statistics and PCA are functions of expected pairwise coalescent times \citep{mcvean2009, peter2016}, this is one avenue to link these approaches.
+Here, I instead use the geometric interpretation of $F$-statistics introduced by  \cite{oteo-garcia2021} to directly visualize $F$-statsitics on a PCA plot.
 
 \section{Theory}
-In this section, I will give a very brief introduction to $F$-statistics and PCA.
-A more detailed technical introduction of PCA is given in XXXXX, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
+In this section, I will give a very brief formal introduction to $F$-statistics and PCA.
+A more detailed technical introduction of PCA is given in e.g.
+\cite{jolliffe2013}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
 
@@
 If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
-As a population may be represented by just one (pseudo-)haploid or diploid individual, there is no conceptual difference between these cases and I will
-refer to populations as unit for analysis, for simplicity. Since the allele
+As a population may be represented by just one individual, there is no conceptual difference between these cases and I will
+refer to populations as unit for analysis. Since the allele
 frequencies are between zero and one, we can interpret each Population $X_i$
-of $\MX$ as a point in $[0, 1]^S$, the \emph{data space} of all possible allele frequencies on our markers.
+of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.
 
-The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ of the data that explains most of the variation in the data.
-$K$ is at most $n-1$, but the historical processes that generated often result in \emph{sparse} data \citep{engelhardt2010,
-patterson2012}, so that $K \ll n$; for visualization $K=2$ is frequently used (see Fig. \ref{fig:pca_explanation} for an intuitive explanation).
+The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ that explains most of the variation in the data.
+$K$ is at most $n-1$, but the historical processes that generated often result in \emph{sparse} data \citep{engelhardt2010}, so that $K \ll n$ explains most ofthe variation; for visualization $K=2$ is frequently used.
 
+\begin{comment}
 	\begin{figure}
@@
 	\end{figure}
+\end{comment}
 
 There are several algorithms that are used to calculate a PCA in practice,
-   the most common one is based on singular value decomposition. In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
+   the most common one is based on singular value decomposition \citep{jolliffe2013}. In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
 	\begin{equation*}
@@
 The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is used to reveal population structure.
-The loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times k$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \cite[e.g][]{francois2010}.
+The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times k$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \citep[e.g][]{francois2010}.
 
-	In many  implementations \citep[e.g]{patterson2006}, SNPs are weighted by
+	In many  implementations \citep[e.g][]{patterson2006}, SNPs are weighted by
     the inverse of their standard deviation. As this weighting often makes little
@@
 \end{subequations}
-where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product.
+where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
+$2F_3(X_1; X_2, X_3) = 2F_2(X_2, X_3) - F_2(X_1, X_2) + F_2(X_1, X_3)$ and $2F_4(X_1, X_2; X_3, X_4) = F_2(X_1, X_3) + F_2(2_4) - F_2(X_1,X_4) - F_2(X_2, X_3)$.
 
-$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}, although they can be calculated under most population genetic models \citep{peter2016}.
-Most commonly, $F_3$ and $F_4$ are interpreted as admixture tests: Negative values of  $F_3(X_1; X_2, X_3) < 0$ are interpreted that $X_1$ is a mixture between populations (related to) $X_2$ and $X_3$.
+$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
+In a tree, $F_2(X, Y)$ measures the length of all branches between populations $X$ and $Y$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}.
+
+This interpretation is useful to understand a number of applications.
+The outgroup-$F_3$-statistic $F_3(O; U, X_i)$, for example, is useful if we have an unknown population $U$, and want to find its closest relatives from a panel of populations $X_i$.
+The highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times.
+The internal branches described by  $F_4$-statistics are frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.
+
+Most commonly however, $F_3$ and $F_4$ are used as admixture tests: Negative values of  $F_3(X_1; X_2, X_3) < 0$ correspond to a branch with negative genetic drift, which is a violation of treeness.
 Similarly if populations are related as a tree, then $F_4(X_1, X_2; X_3, X_4) =0$.
-Alternatively, $F_2$, $F_3$ and $F_4$ are all used to measure genetic similarity.
-$F_2(X_1, X_2)$ represents the variance in allele frequency between populations $X_1$ and $X_2$, which is a measure of genetic drift.
-The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_i)$ is used if we have an unknown population $X_U$, and want to find its closest relatives from a panel of populations $X_i$.
-The highest values of $F_3$ indicate the population $X_i$ most closely related to $X_U$.
-Including an outgroup $X_0$ allows correction for sampling-time difference, which are common in applications on ancient DNA.
-Finally, $F_4(X_1, X_2; X_3, X_4)$ is used to measure the length of the internal branch on the tree connecting these four populations.
-It is used to reconstruct admixture graphs \cite{patterson2012, lipson2013} and to estimate admixture proportions \citep{petr2019, harney2020}.
 
-Here, we are interested in interpreting the $F$-statistics in the data space $\mathbb{R}^S$, and compare it with PCA.
-For a thorough introduction of interpreting $F$-statistics in data space, see \cite{oteo-garcia2021}.
+To move away from trees and graph models, I build upon the geometric framework of \cite{oteo-garcia2021}.
+Here, we think of each population as a point in the data space $\mathbb{R}^S$, made up of the allele frequency at each SNP.
+Then, $F_2(X_1, X_2) = \normsq{X_1 - X_2}$ is the squared Euclidean distance between two populations $X_1$ and $X_2$, and $F_4(X_1, X_2 ; X_3, X_4) = \langle X_1 - X_2, X_3 - X_4 \rangle$ is the inner (dot) product between these two vectors.
+These dot products are useful for a variety of projections that use population structure.
 
-\paragraph{Principal components from $F$-statistics}
-The principal components can be directly calculated from $F$-statistics using multidimensional scaling.
+\subsection{Connection between PCA and $F$-statsitics}	
+\subsubsection{Principal components from $F$-statistics}
+PCA and $F$-statistics are closely related.
+In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling \citep{gower1966}.
 Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$.
@@
 
-\section{$F$-statistics in PCA-space}
-As shown by e.g.
-\cite{oteo-garcia2021}, $F$-statistics can be thought of as inner products in Euclidean space, and $F_2$ is an (estimated) squared Euclidean distance between two populations in allele frequency space.
-By performing a PCA, we just translate and rotate our data, but Euclidean distances and dot products are both invariant under both these operations.
-Hence, neither mean-centering (a translation) nor PCA (a rotation) will change $F_2$.
+\subsubsection{$F$-statistics in PCA-space}
+By performing a PCA, we rotate our data to reveal the axes of highest variation.
+However, the dot product is invariant under rotation, and $F$-statistics can be thought of as dot products.
 What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or any other orthogonal basis such as the principal components $\MP$.
@@
 In the admixture case, we may ask the following question: given two source populations $X_1$, $X_2$, where would admixed populations on a PCA plot lie?
-From theory, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \cite{brisbin2012, mcvean2009}.
+From theory, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \citep{brisbin2012, mcvean2009}.
 
@@
 
+\section{Results}
+\subsection{$F_4$}
+Using $F_3$-statistics, I showed that we can think of the admixture test as a test of whether the admixed population lies in a particular $n$-ball, and the outgroup $F_3$-statistic can be thought of as a projection of the test populations on the line connecting the outgroup to the reference sample.
+In this section, I will develop similar interpretation of $F_4$ on PCA-plots, and to investigate sparsity.
+
+First, we investigate the sparsity in the world overview data set: We find that the vast amount of contribution to the statistics comes from the first two PCs (Figure \ref{fig:f4}A).
+For example, the correlation between $F_4(X, Y, \text{Mozabite}, \text{Yoruba})$ and its approximation using the first two PCs is 99.2\%.
+To visualize the interpretation of $F_4$ as an angle, we use statistics of the form $F_4(X, \text{Sardinian}; \text{Mozabite}, \text{Yoruba})$, which can be interpreted as the angle between the vectors Mozabite-Yoruba and $X$-Sardinian.
+In Figure \ref{fig:f4}B, I show the angle based on two (blue), ten (green) and all PCs $red$.
+I find that for most Asian and American populations the angle is very close to 90\textdegree, as would be expected if the variation between African and non-African populations is mostly orthogonal.
+On the other hand, if $X$ is an African population, the angle is lower, and much less well approximated.
+This demonstrates that this PCA-plot likely does not model within-African population structure adequately.
+
+The $F_4$-statistics for the Westeurasian data set are slightly less sparse, the correlation coefficients between $F_4(X, \text{French}; \text{Finnish}, \text{Canary Islander})$ and its approximation using the first two or three PCs is 95.5\% and 99.1\% respectively (Figure \ref{fig:f4}E).
+I also show that the interpretation of $F_4$ as a projection can be used as a useful visualization (Figure \ref{fig:f4}D).
+On the $x$-axis, I plot $\langle X; \text{Finnish}, \text{Canary Islander}\rangle$, so that the horizontal distance between all pairs of populations corresponds to their respective $F_4$-statistics $F_4(X, Y; \text{Finnish}, \text{Canary Islander})$.
+On the $Y$-axis and with the coloring I display the first two principal components of the residual, i.e.
+the genetic variation that is missed by viewing the data through this projection.
+We find that most European populations have positive values on residual PC1, and are relatively closely clustered.
+In contrast Middle Eastern and Caucasian populations have negative values on this gradient.
+This allows us to visualize that this particular $F_4$-projection does an adequate job if we are interested in describing European variation, but it fails to explain the non-European data.
+We can further quantify this by investigating the percent of variance explained on each axis (Figure \ref{fig:f4}F), where I find that the projection axis only describes around 12\% of hte variation, compared to residual PC1 with almost 30\%.
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
+	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
+Populations for which $F_3$ is negative are colored in red.
+B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
+C,F: Contributions of PCs 1-10 to each $F_3$-statistic.
+D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$.
+The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population.
+}
+	\label{fig:f3}
+\end{figure}
+
 	\subsection{What is a dimension?}
@@
 Likewise, a second population with frequency $w_0$ will move to $w_1$.
-Crucially, if these populations do not interact, the changes in allele frequency, $v_1 - v_0$ and $w_1 - w_0$ will be uncorrelated \cite{patterson2012}.
+Crucially, if these populations do not interact, the changes in allele frequency, $v_1 - v_0$ and $w_1 - w_0$ will be uncorrelated \citep{patterson2012}.
 Thus, if we have two populations that descend from the same ancestral population in isolation, they can be thought of as evolving along orthognal dimensions from the same point.
@@
 This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
-It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \cite{patterson2012}.
-Population genetic differentiation in this region is low and closely mirrors geography \cite{novembre2008}.
+It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012}.
+Population genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}.
 
@@
 
-This is expected because for spatially continuous populations, PCA will not be sparse \cite{novembre2008a}.
+This is expected because for spatially continuous populations, PCA will not be sparse \citep{novembre2008a}.
 Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
@@
 Particularly for the analysis of ancient DNA, $F$-statistics have been established as a powerful tool to describe population genetic diversity, but they have a number of limitations.
-In particular, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \cite{patterson2012,harney2020}.
+In particular, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
 As a consequence, researchers concerned about model fits may ascertain reference populations in a way that satisfies these assumptions, thus inadvertently making population structure appear sparser than it truly is.
 This is perhaps most obvious from Figure \ref{fig:f3}B, where large gaps are present.
-However, these gaps are due to sparse sampling that disappear if more populations were sampled \cite[e.g.][]{peter2020a}, not due to gaps in genetic diversity.
+However, these gaps are due to sparse sampling that disappear if more populations were sampled \citep[e.g.][]{peter2020a}, not due to gaps in genetic diversity.
 If population ascertainment is not done very carefully, tools built on top of $F$-statistics, such as \texttt{qpGraph} and \texttt{qpAdm}, may thus only provide a very loose lower bound for the number of gene flow events.
@@
 While the data space produces a much larger model space than trees, the main drawback of the PCA-based interpretation is a lack of interpretability; it is not easy to define a generative model that could generate a complex PCA-plot.
+Thus, in the future the tree- and PCA-based interpretations will likely be used in conjunction.
 
@@
 The perhaps most obvious one is that PCA is most frequently run on individuals, whereas $F$-statistics are often calculated on populations.
-This is not a conceptual issue, as both PCA and $F$-statistics can be run on either \cite{cavalli-sforza1994}.
-Population based analyses have the advantage that they are easier to interpret and compute (current packages are ill-equipped to calculate all pairwise $F$-statistics between data sets with thousands of individuals \cite{patterson2012}).
+This is not a conceptual issue, as both PCA and $F$-statistics can be run on either \citep{cavalli-sforza1994}.
+Population based analyses have the advantage that they are easier to interpret and compute (current packages are ill-equipped to calculate all pairwise $F$-statistics between data sets with thousands of individuals \citep{patterson2012}).
 However, this requires the assumption that the within-population variation is independent from the between-population variation; something that is analogous to the variance partitioning based on PCs here.
 
-A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \cite{patterson2006}.
+A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \citep{patterson2006}.
 This is only a difference of convention; $F$-statistics could also be calculated using the same weighting.
@@
 The third and perhaps biggest gap are statistical issues.
-The treatment here focusses on the mean estimated $F$-statistic, but many  applications of $F$-statistics are based on hypothesis tests \cite{patterson2012}.
-This requires estimating accurate standard errors for these statistics, which is difficult since nearby SNPs will be correlated \cite{hahn2019}.
+The treatment here focusses on the mean estimated $F$-statistic, but many  applications of $F$-statistics are based on hypothesis tests \citep{patterson2012}.
+This requires estimating accurate standard errors for these statistics, which is difficult since nearby SNPs will be correlated \citep{hahn2018}.
 In contrast, standard PCA does not modeljointly models the covariance matrix due to population structure and sampling.
@@
 It does however mean that tools that use matrices of $F$-statistics, such as \texttt{qpadm} or \texttt{qpgraph} may be ill-calibrated, which may partly explain why they generally have poor out-of-sample predictive power and are restricted to a few dozen samples at a time.
-A model-based framework based on probabilistic PCA \citep{meisner2021, agrawal2020, softimpute} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
+A model-based framework based on probabilistic PCA \citep{meisner2021, agrawal2020, hastie2015} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
 

### 2021-07-02 fstats_pca_lewontin.tex
merge

--- 2021-07-01 fstats_pca_lewontin.tex
@@
 \newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
+\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
 
-\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
 \begin{document}
 \section{Introduction}
-About 15\% of genetic variation in humans can be explained by population structure \cite{lewontin1972, barbujani1997, rosenberg2003}, but the information contained in these 15\% is sufficient to study the genetic diversity and history in great detail \cite{edwards2003}.
-For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \cite{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
+About 15\% of genetic variation in humans can be explained by population structure \cite{lewontin1972, barbujani1997, rosenberg2003, mallick2016}, but the information contained in these 15\% is sufficient to study our genetic diversity and history in great detail \cite{cavalli-sforza1994, stoneking2016, reich2018}.
+Direct-to-consumer-genetics companies are using this variation to model the ancestry of millions of customers, and for some data sets, it is possible to predict an individuals' place of origin at a resolution of a few hundred kilometers \cite{novembre2008, leslie2015}.
+
+Human genetic structure is also not particularly discrete.
 

### 2021-07-05 fstats_pca_lewontin.tex
monday morning

--- 2021-07-02 fstats_pca_lewontin.tex
@@
 \newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
+\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
 
-\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
 \begin{document}
@@
 
+\begin{comment}
 Understanding and characterizing this variation is also crucial for genomic medicine: Risk loci identified in one population perform progressively worse in more distant populations \citep{berg2019, duncan2019}; so identifying a good ``reference'' data set is crucial.
@@
 In Lewontin's pioneering analysis, he found that less than half (6\%), of that variation could be attributed to the continental-scale groups he called races, it seemed  which he used to claim that ''racial classification is (...) seen to be of virtually no genetic or taxonomic significance``.
+\end{comment}
 
-Human genetic diversity has both discrete and continuous components \citep{rosenberg2002, rosenberg2005, bradburd2015,  reich2018, peter2020a}.
-On the one hand, barriers to migration such as mountain ranges, oceans or deserts lead to reduced gene flow, and thus populations on different continents had periods with little-to-no gene flow.
-On the other hand, local gene flow between neighboring populations is also pervasive, and the out-of-Africa expansion caused gradual patterns of differentiation \cite{ramachandran2005}.
-Finally, long-distance migrations, and secondary contact between diverged groups such as Neandertals and early modern humans add to the complexity of human population structure.
+Human genetic diversity has both discrete and continuous components \citep{rosenberg2002, serre2004, rosenberg2005, bradburd2018, reich2018a}.
+On the one hand, barriers to migration such as mountain ranges, oceans, or simply large distances reduce gene flow \citep{bradburd2013, peter2020a, rosenberg2005} and cause genetic differentiation.
+On the other hand, major population movements such as the out-of-Africa, Austronesian or Bantu expansions and local gene flow lead to more gradual patterns of genetic diversity \cite{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2018, racimo2020}, sometimes including partial replacement and gene flow with resident populations.
+Finally, long-distance migrations \citep{alves2015} and secondary contact and admixture between diverged populations, such as Neandertals and early modern humans \citep{green2010} add another layer.
 
-The way this complexity is frequently handled is that multiple models with different assumptions and used, each emphasizing a particular aspects of the data.
+This complex population structure is frequently handled by using multiple models with different assumptions; each emphasizing a particular aspects of the data.
 Data-driven methods such as Principal Component Analysis (PCA) \citep{cavalli-sforza1994} structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable.
-For this more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are used, which allow for parameter estimation or hypothesis tests.
+For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests.
+
 Particularly in the analysis of human ancient DNA, a set of techniques based on $F$-statistics \textit{sensu} Patterson have risen in popularity \citep{patterson2012, peter2016}.
 This framework is based on the assumption that the relationship between three or four populations is often tree-like, and allows for a variety of tests of treeness and more complex demographic models \citep{patterson2012, harney2021}.
-
 However, the connections between PCA, $F$-statistics and demographic models are currently unclear, which makes quantitative comparisons, detecting model violations and joint interpretation of the results of these approaches difficult.
@@
 In this section, I will give a very brief formal introduction to $F$-statistics and PCA.
-A more detailed technical introduction of PCA is given in e.g.
+A more detailed technical treatise of PCA is given in e.g.
 \cite{jolliffe2013}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
+Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2020a} or \cite{oteo-garcia2021} helpful.
 
@@
 The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ that explains most of the variation in the data.
-$K$ is at most $n-1$, but the historical processes that generated often result in \emph{sparse} data \citep{engelhardt2010}, so that $K \ll n$ explains most ofthe variation; for visualization $K=2$ is frequently used.
+$K$ is at most $n-1$, in which case the data is simply rotated.
+However, the  historical processes that generated genetic variation often result in \emph{sparse} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.
 
@@
 
-There are several algorithms that are used to calculate a PCA in practice,
-   the most common one is based on singular value decomposition \citep{jolliffe2013}. In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
+There are several algorithms that are used to perform PCAs, the most common one is based on singular value decomposition \citep{jolliffe2013}.
+In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
 	\begin{equation*}
@@
 
-	In many  implementations \citep[e.g][]{patterson2006}, SNPs are weighted by
+	In many  implementations \citep[e.g.][]{patterson2006}, SNPs are weighted by
     the inverse of their standard deviation. As this weighting often makes little
@@
 $F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time.
-
+The three $F$-statistics can be defined as
 \begin{subequations}
 	\begin{align}
-	F_2(X_1, X_2) &=& \sum_{l=1}^S(x_{1l - x{2l}})^2 &=& \normsq{X_1, X_2}\\
-	F_3(X_1; X_2, X_3) &=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \langle X_1 - X_2, X_1 - X_3 \rangle\\	
-	F_4(X_1, X_2; X_3, X_4) &=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
+	&F_2(X_1, X_2) &=& \sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \normsq{X_1, X_2}\\
+	&F_3(X_1; X_2, X_3) &=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \langle X_1 - X_2, X_1 - X_3 \rangle\\	
+	&F_4(X_1, X_2; X_3, X_4) &=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
 	\end{align}
@@
 where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
-$2F_3(X_1; X_2, X_3) = 2F_2(X_2, X_3) - F_2(X_1, X_2) + F_2(X_1, X_3)$ and $2F_4(X_1, X_2; X_3, X_4) = F_2(X_1, X_3) + F_2(2_4) - F_2(X_1,X_4) - F_2(X_2, X_3)$.
+\begin{subequations}
+	\begin{align}
+2F_3(X_1; X_2, X_3) &= 2F_2(X_2, X_3) - F_2(X_1, X_2) + F_2(X_1, X_3)\\
+2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)
+	\end{align}
+\end{subequations}
 
 $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
-In a tree, $F_2(X, Y)$ measures the length of all branches between populations $X$ and $Y$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}.
-
+In a tree, the squared Euclidean distance $F_2(X, Y)$ measures the length of all branches between populations $X$ and $Y$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}.
+The length is a measure of genetic drift, and is non-negative if data is generated under a tree \citep{patterson2012}.
 This interpretation is useful to understand a number of applications.
@@
 The highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times.
+The population $X_i$ with the largest value is the most closely related population out of the reference sample.
 The internal branches described by  $F_4$-statistics are frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.
 
-Most commonly however, $F_3$ and $F_4$ are used as admixture tests: Negative values of  $F_3(X_1; X_2, X_3) < 0$ correspond to a branch with negative genetic drift, which is a violation of treeness.
-Similarly if populations are related as a tree, then $F_4(X_1, X_2; X_3, X_4) =0$.
+Most commonly however, $F_3$ and $F_4$ are used as admixture tests \citep{patterson2012}: Negative values of  $F_3(X_1; X_2, X_3) < 0$ correspond to a branch with negative genetic drift, which is a violation of treeness.
+Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}.
 
@@
 
-\paragraph{Optimality of PCA}
 In most applications, we do not use all PCs, but instead use only the first $K$ PCs.
@@
 \begin{equation}
-F_2(P_i, P_j) = \sum_{k=1}^K(p_{ik} - p_{jk})^2 + \sum_{k=K+1}^n(p_{ik} - p_{jk})^2 \text{,}
+F_2(P_i, P_j) = \underbrace{\sum_{k=1}^K(p_{ik} - p_{jk})^2}_{\hat{F_2}^{(K)}(P_i, P_j)} + \sum_{k=K+1}^n(p_{ik} - p_{jk})^2 \text{.}
 \end{equation}
-where the first sum is the PCA-approximation of $\hat{F_2}$, and the second sum is the residual or approximation error $F_2 - \hat{F_2}$.
+If we sum up the approximation errors $F_2 - \hat F_2$ over all pairs of populations, we obtain the Frobenius-norm of the error $\normsq{\MF - \hat{\MF}}_F$; which is precisely the function that is minimized in MDS \citep{jolliffe2013}.
+Thus, $\hat\MF$ is the optimal approximation of $\MF$ for any $K$.
 
-If we sum up the approximation errors over all pairs of populations, we obtain the Frobenius-norm of the error $\normsq{\MF - \hat{\MF}}_F$; it is a standard result that PCA finds the best rank-$K$ approximation so that this Frobenius-norm is minimized.
-In our context, this means that PCA using the first $K$ PCs results in approximate $F_2$-statistics such that the sum of $F_2$-distances between the approximation and full data is minimized.
-
-\subsection{$F$-stats in 2-dimensional PC-space}
+\subsection{Geometry of $F$-statistics in PC-space}
 The transformation from the previous section allows us to consider the geometry
@@
 only a very small number $K \ll n$ of PCS contain most information that is
-relevant for population structure (for visualization, it is often assumed that
-$K=2$).
+relevant for population structure (for visualization
+$K=2$ is often used).
 
@@
 \begin{figure}[!ht]
-	\includegraphics[width=\textwidth]{dummy_pca.png}
+	\includegraphics[width=\textwidth]{figures/fstats_on_pca.pdf}
 	\caption{\textbf{Geometric representation of $F$-statistics on 2D-PCA-plot.} A: $F_2$ represents the squared Euclidean distance between two points in PC-space.
@@
 
-\section{Results}
-\subsection{$F_4$}
-Using $F_3$-statistics, I showed that we can think of the admixture test as a test of whether the admixed population lies in a particular $n$-ball, and the outgroup $F_3$-statistic can be thought of as a projection of the test populations on the line connecting the outgroup to the reference sample.
-In this section, I will develop similar interpretation of $F_4$ on PCA-plots, and to investigate sparsity.
-
-First, we investigate the sparsity in the world overview data set: We find that the vast amount of contribution to the statistics comes from the first two PCs (Figure \ref{fig:f4}A).
-For example, the correlation between $F_4(X, Y, \text{Mozabite}, \text{Yoruba})$ and its approximation using the first two PCs is 99.2\%.
-To visualize the interpretation of $F_4$ as an angle, we use statistics of the form $F_4(X, \text{Sardinian}; \text{Mozabite}, \text{Yoruba})$, which can be interpreted as the angle between the vectors Mozabite-Yoruba and $X$-Sardinian.
-In Figure \ref{fig:f4}B, I show the angle based on two (blue), ten (green) and all PCs $red$.
-I find that for most Asian and American populations the angle is very close to 90\textdegree, as would be expected if the variation between African and non-African populations is mostly orthogonal.
-On the other hand, if $X$ is an African population, the angle is lower, and much less well approximated.
-This demonstrates that this PCA-plot likely does not model within-African population structure adequately.
-
-The $F_4$-statistics for the Westeurasian data set are slightly less sparse, the correlation coefficients between $F_4(X, \text{French}; \text{Finnish}, \text{Canary Islander})$ and its approximation using the first two or three PCs is 95.5\% and 99.1\% respectively (Figure \ref{fig:f4}E).
-I also show that the interpretation of $F_4$ as a projection can be used as a useful visualization (Figure \ref{fig:f4}D).
-On the $x$-axis, I plot $\langle X; \text{Finnish}, \text{Canary Islander}\rangle$, so that the horizontal distance between all pairs of populations corresponds to their respective $F_4$-statistics $F_4(X, Y; \text{Finnish}, \text{Canary Islander})$.
-On the $Y$-axis and with the coloring I display the first two principal components of the residual, i.e.
-the genetic variation that is missed by viewing the data through this projection.
-We find that most European populations have positive values on residual PC1, and are relatively closely clustered.
-In contrast Middle Eastern and Caucasian populations have negative values on this gradient.
-This allows us to visualize that this particular $F_4$-projection does an adequate job if we are interested in describing European variation, but it fails to explain the non-European data.
-We can further quantify this by investigating the percent of variance explained on each axis (Figure \ref{fig:f4}F), where I find that the projection axis only describes around 12\% of hte variation, compared to residual PC1 with almost 30\%.
-
 \begin{figure}[!ht]
@@
 
-	\subsection{What is a dimension?}
-	In both the PCA and $F$-statistic framework, a population at  a particular point in time can be thought of as a single point in allele-frequency space, given by the $k$-dimensional vector $v_0$ of allele frequencies at the $k$ SNPs in that population.
-If this population evolves for some time in isolation, allele frequencies will change due to genetic drift from $v_0$ to some other point $v_1$.
-Likewise, a second population with frequency $w_0$ will move to $w_1$.
-Crucially, if these populations do not interact, the changes in allele frequency, $v_1 - v_0$ and $w_1 - w_0$ will be uncorrelated \citep{patterson2012}.
-Thus, if we have two populations that descend from the same ancestral population in isolation, they can be thought of as evolving along orthognal dimensions from the same point.
-This argument is at the foundation of F-statistics.
+\section{Results}
+The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots, and can be approximated using PCA.
+I illustrate this on two human data sets based on the ``Human Origins''-SNP set (597,573 SNPs).
+Both are based on the analyses if \citep{lazaridis2014}, and I use subsets of the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}.
 
-\section{Results}
-The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots.
-In this section, I use these interpretation in the analysis of human genetic variation data set.
-I use two data sets based on the ``Human Origins''-SNP set (597,573 SNPs).
-Both are subsets of the Reich lab compendium data set v44.3, downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}.
+\paragraph{World Overview data set}
+This data set represents global human genetic variation (638 individuals from 33 population), as used by \citep{lazaridis2014}.
+As this data set is very sparse, it may be well-approximated by an admixture graph.
 
@@
 This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
-It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012}.
-Population genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}.
+It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012, lazaridis2014}.
+Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus not particularly graph-like.
 
-\paragraph{World Overview data set}
-This data set of 638 individuals from 33 population contains individuals throughout the world, and is used as a sparse data set capturing much of global human genetic variation.
-This data set spans Africa, Eurasia and the Americas, and we might therefore expect the population structure to be much more sparse.
-
-I perform analyses at the level of populations to ease presentation, and because it is an assumption of $F$-statistics that the genetic variation with sampled population is independent of the variation between samples that I am focusing on here.
+I perform analyses at the level of populations to ease presentation.
+It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples.
 I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute a matrix of $F_2$-statistics between all populations.
-To obtain a PC-decomposition I use equation XXX and the \texttt{eigen} function in \texttt{R}, and compare them with the $F_3$ and $F_4$-statistics calculated using admixtools 2.
+To obtain a PC-decomposition I use equation \ref{eq:mds} and the \texttt{eigen} function in \texttt{R}, and compare them with the $F_3$ and $F_4$-statistics calculated using admixtools 2.
 
 \paragraph{Admixture-$F_3$}
-As a first step, I plot the first two principal components of the Westeurasian data set (Figure \ref{fig:f3}A).
-This PCA presents two parallel clines, one from the Levant and Arabia (``BedouinB'') to the Caucasus (``Abkhasian''), and a second one from Southern (``Sardinian'') to Northeastern Europe (``Mordovian'').
-In this context, I examine $F_3(X; Basque, Turkish)$, i.e.
+I visualize the admixture statistic $F_3(X; Basque, Turkish)$, on the first two PCs of the Westeurasian data set (Figure \ref{fig:f3}A). 
+%This PCA presents two parallel clines, one from the Levant and Arabia (``BedouinB'') to the Caucasus (``Abkhasian''), and a second one from Southern (``Sardinian'') to Northeastern Europe (``Mordovian'').
+This statistic aims to model In this context, I examine  i.e.
 a statistic that aims to ask which populations can be represented as a mixture between a Southwestern (Basque) and Southeastern (Turkish) European population.
@@
 We can further quantify this by investigating the percent of variance explained on each axis (Figure \ref{fig:f4}F), where I find that the projection axis only describes around 12\% of hte variation, compared to residual PC1 with almost 30\%.
-
-\begin{figure}[!ht]
-	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
-	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
-Populations for which $F_3$ is negative are colored in red.
-B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
-C,F: Contributions of PCs 1-10 to each $F_3$-statistic.
-D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$.
-The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population.
-}
-	\label{fig:f3}
-\end{figure}
 

### 2021-07-05 data/n_samples.R
monday morning

@@ -1,0 +1,35 @@
+require(tidyverse)
+a  = read_tsv("v44.3_1240K_public.anno") 
+
+z = a %>% rename(year=6,age=10,coverage=20) %>% 
+    mutate(year=as.numeric(year), age=as.numeric(age)) %>% 
+    mutate(lat=as.numeric(Lat.), long=as.numeric(Long.)) %>%
+    filter(age>0)
+
+P = z %>% filter(!is.na(year), age>0) %>% group_by(year) %>% 
+    tally %>% mutate(n=cumsum(n)) %>% 
+    ungroup %>%
+    ggplot(aes(x=as.factor(year), y=n)) + geom_col() + 
+    xlab("Publication Year") + ylab("# samples")
+ggsave("figures/ancient_dna_by_year.png", P, width=5, height=3)
+
+
+P2= z %>% mutate(coverage=as.numeric(coverage), coverage=pmin(coverage,10)) %>% 
+    ggplot(aes(x=as.numeric(coverage))) + geom_histogram(bins=100) 
+#    xlab=("Coverage") + ylab=("# samples")
+ggsave("figures/ancient_dna_coverage.png", P2, width=5, height=3)
+
+
+require(rnaturalearth)
+require(rnaturalearthdata)
+world <- ne_countries(scale = "medium", returnclass = "sf")
+P3 = ggplot() + 
+    geom_sf(data=world, color=NA) + 
+    theme_minimal() + 
+    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
+    geom_point(aes(color=age, x=long, y=lat), data=z) + 
+    scale_color_viridis_c(trans='log') +
+    xlab("") + ylab("") + 
+    theme(legend.pos='none') + 
+    coord_sf(xlim=c(-170, 160), ylim=c(-50,75))
+ggsave("figures/ancient_dna_map.png", P3, width=7, height=4, dpi=150)

### 2021-07-05 fstats_pca_lewontin.tex
Draft 1

@@
-\documentclass[12pt,a4pape, fullpage]{article}
+\documentclass[12pt,fullpage, a4paper]{article}
 \usepackage[
@@
 ]{geometry}
-\usepackage[utf8x]{inputenc}
+%\usepackage[utf8x]{inputenc}
 %\usepackage[T1]{fontenc}
@@
 \usepackage{comment}
-\usepackage{natbib}
+\usepackage[authoryear, comma]{natbib}
 \usepackage{lineno}
 \usepackage{url}
-\usepackage{textcomp}
 
-\bibliographystyle{evolution.bst}
+\linenumbers
+%\bibliographystyle{evolution.bst}
+%\bibliographystyle{plainnat}
 
@@
 \newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
-
 \newcommand{\MX}{\mathbf{X}} %uncentered data
@@
 \newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
-
+\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
+\author{Benjamin M Peter}
 \begin{document}
+	\maketitle
+\begin{abstract}
+The genetic diversity in human population is shaped by our complex history.
+$F$-statistics are a widely used tool to model genetic diversity when populations are related acording to a graph and gene flow is rare.
+Here, I use a geometric interpretation of  to interpret $F$-statistic in the context of Principal Component Analysis (PCA), showing that the two approaches are closely related.
+I show that orthogonal projections are a key concept to establish this link, and that it leads to simple geometric interpretations of $F$-statistics on a PCA-plot.
+I illustrate my results on two examples, one of local, and one of global human diversity.
+In both examples, I find that population structure is sparse, and only a few components contribute to most statistics, and I  develop novel visualizations that allow for checking the assumptions of more sophisticated models.
+My results allows interpreting  $F$-statistics in arbitrarily structured populations, to obtain  more complete and less biased models of human genetic variation.
+\end{abstract}
 \section{Introduction}
-About 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002}, but the information contained in these 15\% is sufficient to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}.
+About 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002a}, but the information contained in these 15\% is sufficient to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}.
 For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
 
-\begin{comment}
-Understanding and characterizing this variation is also crucial for genomic medicine: Risk loci identified in one population perform progressively worse in more distant populations \citep{berg2019, duncan2019}; so identifying a good ``reference'' data set is crucial.
-Understanding genetic variation is also important when designing studies, in order to measure which populations studies should target \citep{fullerton2016}.
-Despite these needs, no generally accepted model for human population structure exists.
-The historical view that humans diversity can be partitioned into  distinct ``races'' \citep[e.g.][]{blumenback1798} has long been superseded in favor of more accurate and nuanced models.
-These models fail because they only explain a small amount of variation...
-
-In Lewontin's pioneering analysis, he found that less than half (6\%), of that variation could be attributed to the continental-scale groups he called races, it seemed  which he used to claim that ''racial classification is (...) seen to be of virtually no genetic or taxonomic significance``.
-\end{comment}
-
-Human genetic diversity has both discrete and continuous components \citep{rosenberg2002, serre2004, rosenberg2005, bradburd2018, reich2018a}.
-On the one hand, barriers to migration such as mountain ranges, oceans, or simply large distances reduce gene flow \citep{bradburd2013, peter2020a, rosenberg2005} and cause genetic differentiation.
-On the other hand, major population movements such as the out-of-Africa, Austronesian or Bantu expansions and local gene flow lead to more gradual patterns of genetic diversity \cite{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2018, racimo2020}, sometimes including partial replacement and gene flow with resident populations.
-Finally, long-distance migrations \citep{alves2015} and secondary contact and admixture between diverged populations, such as Neandertals and early modern humans \citep{green2010} add another layer.
+Human genetic diversity is caused by a number of forces, and has both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}.
+Spatially isolated populations are expected to slowly diverge, particularly if they are separated by barriers to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}.
+On the other hand, major population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}.
+Local migration between neighboring populations will flatten differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and early modern humans \citep{green2010} adds further complexity.
 
@@
 	\begin{align}
-	&F_2(X_1, X_2) &=& \sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \normsq{X_1, X_2}\\
-	&F_3(X_1; X_2, X_3) &=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \langle X_1 - X_2, X_1 - X_3 \rangle\\	
-	&F_4(X_1, X_2; X_3, X_4) &=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
+	&F_2(X_1, X_2) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \normsq{X_1, X_2}\\
+	&F_3(X_1; X_2, X_3) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \langle X_1 - X_2, X_1 - X_3 \rangle\\	
+	&F_4(X_1, X_2; X_3, X_4) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
 	\end{align}
 \end{subequations}
-where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
+where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product.
+The number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently.
+Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
 \begin{subequations}
@@
 &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2   &=& F_2(Y_i, Y_j) \nonumber\\
- &=& \sum_{k=1}^n (p_{ik} - p_{jk})^2  &=& F_2(P_i, P_j) \text{,}
+&=& \sum_{k=1}^n (p_{ik} - p_{jk})^2  &=& F_2(P_i, P_j) \text{,}\label{eq:fpc}
 \end{align}
@@
 
-\subsection{Geometry of $F$-statistics in PC-space}
+\section{Material \& Methods}
+The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots, and can be approximated using PCA.
+I will explore this in detail in the next section, and illustrate it on two sample data sets that I briefly introduce here.
+Both are based on the analyses by \cite{lazaridis2014} and use the ``Human Origins''-SNP set (597,573 SNPs), based on the subset of the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}.
+The code used to create all figures and analyses will be available on [Archive to be created before publishing].
+
+\paragraph{World Overview data set}
+This data set represents global human genetic variation (638 individuals from 33 population), as used by \citep{lazaridis2014}.
+As this data set is very sparse, it may be well-approximated by an admixture graph.
+
+\paragraph{West-Eurasian data set}
+This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
+It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012, lazaridis2014}.
+Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus not particularly graph-like.
+
+\paragraph{Computation $F$-statistics and PCA}
+I perform analyses at the level of populations to ease presentation.
+It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples.
+All computations are performed in \texttt{R}.
+I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute $F$-statistics.
+To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, obtain a nearby negative semidefinite matrix using the \texttt{nearPD} function, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.
+
+\section{Results}
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fstats_on_pca.pdf}
+	\caption{\textbf{Geometric representation of $F$-statistics on 2D-PCA-plot.} A: $F_2$ represents the squared Euclidean distance between two points in PC-space.
+B: Admixture-$F_3(X_x; X_1, X_2)$ is negative if $X_x$ lies in the circle specified by the diameter $X_2-X_1$.
+C: $F_3(X_x; X_1, X_2)$ is negative given $X_1, X_x$ if $X_2$ is in the gray space.
+D: Outgroup-$F_3$ reflects the projection of $X_2 - X_O$ on $X_1 - X_O$.
+E: $F_4$ is the projection of $X_3 - X_4$ on $X_1-X_2$.
+F: Same projection, but assuming data is generated by a tree.}
+	\label{fig:geom}
+\end{figure}
+
+%\subsection{Geometry of $F$-statistics in PC-space}
 The transformation from the previous section allows us to consider the geometry
@@
 
-\subsubsection{$F_2$ in PC-space}
+\subsection{$F_2$ in PC-space}
 The $F_2$-statistic is an estimate of the squared Euclidean distance between two
-populations. It thus corresponds to the squared distance in PCA-space, and
-reflects that closely related populations  will be close to each other on a
+populations. It thus corresponds to the squared distance between populations in PCA-space, and
+reflects the intuition that closely related populations  will be close to each other on a
 PCA-plot, and have low pairwise $F_2$-statistics. In converse, if two
-populations with high $F_2$ lie on the same point on an PCA-plot, this suggests
+populations have high $F_2$ but appear on the same point on an PCA-plot, this suggests
 that substantial variation is hidden on higher PCs.
 
-\subsubsection{When is $F_3$ negative?}
-The $F_3$-statistic becomes more interesting; as outlines above we either think of $F_3$ as ``outgroup''-$F$-stats or as admixture $F$-stats.
-In the admixture case, we may ask the following question: given two source populations $X_1$, $X_2$, where would admixed populations on a PCA plot lie?
-From theory, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \citep{brisbin2012, mcvean2009}.
+\subsection{When are admixture-$F_3$ statistics negative?}
+Given two source populations $X_1$, $X_2$, can we predict which populations $X_x$ could be considered admixed between these populations based on PCA?
+Since the allele frequencies of $X_x$ are intermediate between those of $X_1$ and $X_2$, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \citep{brisbin2012, mcvean2009}.
 
-Formally, we would reject admixture if $F_3$ is negative, i.e. we are looking for the space
+The $F_3$-statistic gives a more precise interpretation: we are looking for the space where $F_3$ is negative, i.e. 
 \begin{eqnarray}
 2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
-      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} 
+      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0
 \end{eqnarray}
-By the Pythagorean theorem, $F_3 = 0 $ iff $X_1, X_2$ and $X_x$ form a right-angled triangle.
-In a 2D-PCA plot, the region where $F_3$ is zero is the circle  with diameter $\overline{X_1X_2}$, and if $X_x$ lies inside this circle, $F_3(X_x; X_1, X_2) < 0$.
-If the
+By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_1, X_2$ and $X_x$ form a right-angled triangle.
+The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_1X_2}$.
+$F_3$ is negative when the triangle is obtuse, i.e.
+$X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_1X_2}$ (Figure \ref{fig:geom}B).
 
-ball, the angle is obtuse and $F_3$ is negative, otherwise it will be positive. If we approximate the PCA-space in two dimensions, the $n$-ball corresponds to a circle.
+\paragraph{$F_3$ on a 2D-plot.} 
+ If we project this $n$-ball on a two-dimensional plot, $\overline{X_1X_2}$ will usually not align with the PCs; thus the ball may be somewhat larger.
+This geometry is perhaps easiest visualized on a globe.
+If we look at the globe from a view point parallel to the equator, both the north and south poles are visible at the very edge of the circle.
+But if we look at it from above the north pole, the north- and south-poles will be at the very same point.
 
-\begin{figure}[!ht]
-	\includegraphics[width=\textwidth]{figures/fstats_on_pca.pdf}
-	\caption{\textbf{Geometric representation of $F$-statistics on 2D-PCA-plot.} A: $F_2$ represents the squared Euclidean distance between two points in PC-space.
-B: Admixture-$F_3(X_x; X_1, X_2)$ is negative if $X_x$ lies in the circle specified by the diameter $X_2-X_1$}.
-C: $F_3(X_x; X_1, X_2)$ is negative given $X_1, X_x$ if $X_2$ is in the gray space.
-D: Outgroup-$F_3$ reflects the projection of $X_2 - X_O$ on $X_1 - X_O$.
-E: $F_4$ is the projection of $X_3 - X_4$ on $X_1-X_2$.
-F: If $X_x$ is admixed between $X_1$ and $X_2$, the admixture proportions will be projected.
-\end{figure}
+ Thus if $\hat{F}_3 \ll F_3$, the true circle will be bigger than would be predicted from a 2D-plot.
+In this case, substantial relevant genetic differentiation is ``hidden'' in the higher PCs.
+If this is the case, populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics, because they are outside the $n$-ball in higher dimensions.
+However, if a population lies outside the circle on any 2D-projection, $F_3$ is guaranteed to be bigger than 0.
 
-\subsubsection{$F_4$ and right angles}
-The inner-product-interpretation of $F_4$ is similar to that of $F_3$, with the change that the two vectors we consider do not involve the same population.
-However, a finding of $F_4(X_1, X_2; X_3, X_4) = \langle X_1 - X_2, X_3 - X4 \rangle = 0$ similarly implies that the two vectors are orthogonal, and a non-zero value reflects the projection of one vector on the other.
+\paragraph{Example}
+I visualize this on the admixture statistic $F_3(X; Basque, Turkish)$, on the first two PCs of the Westeurasian data set (Figure \ref{fig:f3}A).
+In this case, the projected $n$-ball (light gray) and circle based on 2D (dark gray) align relatively closely, but several populations inside the ball (e.g.
+Sardinian, Finnish) have, in fact, positive $F_3$-values, which reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure.
+This is expected because for spatially continuous populations, PCA will not be sparse \citep{novembre2008a}.
+Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
 
-\subsubsection{$F_4$-ratio}
-\begin{eqnarray}
-\frac{F_4(X_I, X_O; X_X, X_1)}{F_4(X_I, X_O; X_2, X_1)} &=& \frac{\norm{X_I-X_O}\norm{X_X-X_1}\cos(\alpha)}{\norm{X_I-X_O}\norm{X_2-X_1}\cos(\beta)}\nonumber\\
-&=&\frac{\norm{X_X-X_1}\cos(\alpha)}{\norm{X_2-X_1}\cos(\beta)}\nonumber\\
-&=& \frac{\norm{X_X' - X_1'}}{\norm{X_2' - X_1'}}
-\end{eqnarray}
-where $\alpha$ and $\beta$ are the angles between vectors, and $X_i'$ is the projection of $X_i$ on $X_I-X_O$.
-
-Conjecture: Thus, we are measuring the distances between the admixing populations on the projected on the axis between $X_I$ and $X_O$.
-This ought to be valid only if $\langle X_1 - X_1', X_2 - X_2' \rangle$ are orthogonal to each other, and to $X_OX_I$, i.e.
-$F_4(X_1, X_1', X_2, X_2') = 0$
+However, many populations, particularly from the Levant and Caucasus, fall outside the circle, which allows us to immediately conclude that their $F_3$-statistics must be positive.
 
@@
 
-\section{Results}
-The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots, and can be approximated using PCA.
-I illustrate this on two human data sets based on the ``Human Origins''-SNP set (597,573 SNPs).
-Both are based on the analyses if \citep{lazaridis2014}, and I use subsets of the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}.
+\subsection{$F_3$ as a projection}
+The inner product $\langle X_x - X_1, X_x - X_2 \rangle$ is closely related to the projection of one vector onto the other.
+This interpretation is useful either when calculating an outgroup-$F_3$-statistic, or when we want to find the ``best'' admixture source $X_2$ if we assume the admixed population $X_x$ and one source $X_1$ are known.
+The angle between $X_x - X_1$ and $X_x - X_2$ is obtuse if $X_2$ is in the half-plane whose boundary goes through $X_x$ and is orthogonal to $\overline{X_XX_1}$ (Figure \ref{fig:geom}C), and its value is proportional to the projection onto $X_X - X_1$.
+A common interpretation guide is to use the population which yields the most negative $F_3$-statistic as the population which is most closely related to the ``true'' source of admixture \citep{patterson2012}; on a PCA-plot this corresponds to using the population that is furthest away from $X_x$ on the direction $X_x - X_1$.
 
-\paragraph{World Overview data set}
-This data set represents global human genetic variation (638 individuals from 33 population), as used by \citep{lazaridis2014}.
-As this data set is very sparse, it may be well-approximated by an admixture graph.
+This projection argument also helps to interpret Outgroup-$F_3$-statistic on a PCA-plot (Figure \ref{fig:geom}D), but in this case we aim to find the most closely related population as that with the highest $F_3$-statistic.
 
-\paragraph{West-Eurasian data set}
-This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
-It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012, lazaridis2014}.
-Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus not particularly graph-like.
+\paragraph{Example}
+Again, these projection will be orthogonal when using the full data, and may only be approximately orthogonal when just using the first two PCs.
+In Figure \ref{fig:f3}D, I visualize this on the outgroup-$F_3$-statistic $F_3(Mbuti; Mozabite, X_i)$, i.e.
+a statistic that aims to find the population most closely related to Mozabite (a Berber ethnic group from the northern Sahara), assuming the Mbuti are an outgroup.
+On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Mozabite (black line).
+For each population, the projection is indicated with a grey line.
+In the full data space, this line is always orthogonal to the segment Mbuti-Mozabite, but on the plot (i.e.) the subspace spanned by the first two PCs, this is not necessarily the case.
+We see that particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs, and the coloring based on the full $F_3$-statistics shows that in this case, the first two PCs approximate the $F_3$-statistic very well.
+We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}E), but keep the relative ordering.
+This $F_3$-statistic is also very sparse, with e.g.
+PCs 3-5, 7 and 10 having almost zero contribution to all statistics (Figure \ref{fig:f3}F), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.
 
-I perform analyses at the level of populations to ease presentation.
-It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples.
-I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute a matrix of $F_2$-statistics between all populations.
-To obtain a PC-decomposition I use equation \ref{eq:mds} and the \texttt{eigen} function in \texttt{R}, and compare them with the $F_3$ and $F_4$-statistics calculated using admixtools 2.
+\subsection{$F_4$-statisticss as angles}
+The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct.
+As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3$ and $X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection (Figure \ref{fig:geom}E).
 
-\paragraph{Admixture-$F_3$}
-I visualize the admixture statistic $F_3(X; Basque, Turkish)$, on the first two PCs of the Westeurasian data set (Figure \ref{fig:f3}A). 
-%This PCA presents two parallel clines, one from the Levant and Arabia (``BedouinB'') to the Caucasus (``Abkhasian''), and a second one from Southern (``Sardinian'') to Northeastern Europe (``Mordovian'').
-This statistic aims to model In this context, I examine  i.e.
-a statistic that aims to ask which populations can be represented as a mixture between a Southwestern (Basque) and Southeastern (Turkish) European population.
-The -- largely Southern European -- populations for which the point estimate of these $F_3$-statistic is negative are highlighted in red.
-They both fall close to the center of the $F_3$-circle, either defined on the first two (dark grey) or all PCs (light gray).
-However, many populations inside the circle on the first 2 PCs, including English, Sardinians and Canary Islanders have positive $F_3$-values, on higher PCs, showing that the first two PCs do not capture all the genetic variation related to population structure for this data set.
+We can also see how this interpretation aligns with that of $F_4$ as the length of an internal branch on a tree : By assumption, disjoint sets of branches evolve independently \citep{cavalli-sforza1964, felsenstein1973, semple2003}.
+Since the data space is sufficiently high-dimensional, this ensures that the resulting drift trajectories will also be uncorrelated.
+Therefore, if we interpret $F_4(X_1, X_2; X_3, X_4)$ as the projection of $X_3 - X_4$ - onto $X_1 - X_2$, we can write
+$$X_3 - X_4  = (X_3 - X'_3) + (X'_3 - X'_4) + (X'_4 - X_4)\text{.}$$ Of these three branches, the first and last are orthogonal to $X_1 - X_2$ and thus the $F_4$ statistic is just the internal branch of the tree (Figure \ref{fig:geom}F).
 
-This is expected because for spatially continuous populations, PCA will not be sparse \citep{novembre2008a}.
-Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
+Since $F_4$ is a covariance, its magnitude lacks an interpretation.
+Therefore, commonly correlation coefficients are used, as there, zero means independence and one means maximum correlation.
+For $F_4$, we can write
+\begin{equation}
+\text{Cor}(X_1 - X_2, X_3 - X_4) =  \frac{\langle X_1 - X_2, X_3 - X_4 \rangle}{\norm{X_1-X_2}\norm{X_3-X_3}} = \cos(\phi),\label{eq:angle}
+\end{equation}
+where $\phi$ is the angle between $X_1 - X_2$ and $X_3 - X_4$.
+Thus, independent drift events lead to $\cos(\phi) = 0$, so that the angle is 90 degrees, whereas an angle close to zero means $\cos(\phi)\approx 1$, which means most of the genetic drift on this branch is shared.
 
-Thus, the main benefit of this PCA-plot is that it allows us to identify populations outside the circle (from the Levant and Caucasus), for which $F_3$ is guaranteed to be positive.
+\paragraph{Example}
+To illustrate the angle interpretation I again use the Westeurasian data.
+The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:f3}A), a European gradient (from Sardinian to Chuvash), and a Asian cline (from Arab to Caucasus populations).
 
-\paragraph{Outgroup-$F_3$}
-The Outgroup-$F_3$-statistic is commonly used to infer which population is closest in a set of reference populations.
-In Figure \ref{fig:f3}D, I present a PCA of the world data set, with populations colored according to $F_3(Mbuti; Mozabite, X_i)$, i.e.
-a statistic that is commonly interpreted as finding the population $X_i$ that is most closely related to Mozabite.
-On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment $\text{Mbuti}X_i$ onto the line through Mbuti and Mozabite (black line).
-For each population, the projection is indicated with a grey line.
-In the full data space, this line is always orthogonal to the segment Mbuti-Mozabite, but on the plot (i.e.) the subspace spanned by the first two PCs, this is only true if the relevant variation is captured by the first two PCs.
-We see that particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs.
-That the approximation of $F_3$ on
+This is quantified in Figure \ref{fig:f4}A, where I plot the angle corresponding to $F_4(X, Sardinian; Saudi, Georgian)$.
+For most European populations, using two PCs (green points) gives an angle close to zero, corresponding to a correlation coefficient $r>0.9$.
+Just adding PC3 (blue) gives a close approximation to the full data (red).
+This can also be seen from the spectrum of these statistics (Figure \ref{fig:f4}B), which has high loadings for the first three PCs, with minimal contributions from the higher ones.
 
-\subsection{$F_4$}
-Using $F_3$-statistics, I showed that we can think of the admixture test as a test of whether the admixed population lies in a particular $n$-ball, and the outgroup $F_3$-statistic can be thought of as a projection of the test populations on the line connecting the outgroup to the reference sample.
-In this section, I will develop similar interpretation of $F_4$ on PCA-plots, and to investigate sparsity.
+\subsection{Other projections}
+So far, I used eq.
+\ref{eq:fpc} to interpret $F$-statistics on a PC-plot, but the argument holds for \emph{any} orthonormal transformation.
+This allows for a variety of visualizations that use both $F$-statistics and PCs.
+The motivation for this is that sometimes we wish to partition the variation in the data into a subspace of interest, and an orthogonal residual space that captures the information discarded.
+Examples where analyses are restricted to such subspaces include the $F_4$-ratio test \citep{patterson2012, petr2019}, \texttt{qpWave} \citep{skoglund2015} and \texttt{qpAdm} \citep{harney2021}.
+For the $F_4$-ratio, for example, a ratio
+\begin{eqnarray}
+\alpha = \frac{F_4(R_1, R_2; X, A)}{F_4(R_1, R_2; B, A)}  \label{eq:f4ratio}
+\end{eqnarray}
+is used, which can be interpreted as projecting $X-A$ and $B-A$ onto $R_1 - R_2$.
+Thus, we can make a plot where we plot the variation on the $X$-axis along $R_1 - R_2$, and perform a PCA on the residual.
+This can be important because the residual can be used to check assumptions, e.g.
+$A - A'$ and $B - B'$ need to be orthogonal (Figure \ref{fig:geom}F).
 
-First, we investigate the sparsity in the world overview data set: We find that the vast amount of contribution to the statistics comes from the first two PCs (Figure \ref{fig:f4}A).
-For example, the correlation between $F_4(X, Y, \text{Mozabite}, \text{Yoruba})$ and its approximation using the first two PCs is 99.2\%.
-To visualize the interpretation of $F_4$ as an angle, we use statistics of the form $F_4(X, \text{Sardinian}; \text{Mozabite}, \text{Yoruba})$, which can be interpreted as the angle between the vectors Mozabite-Yoruba and $X$-Sardinian.
-In Figure \ref{fig:f4}B, I show the angle based on two (blue), ten (green) and all PCs $red$.
-I find that for most Asian and American populations the angle is very close to 90°, as would be expected if the variation between African and non-African populations is mostly orthogonal.
-On the other hand, if $X$ is an African population, the angle is lower, and much less well approximated.
-This demonstrates that this PCA-plot likely does not model within-African population structure adequately.
+\subsubsection{Example}
+In the PCA on the world overview data set, I found a seemingly linear gradient from Africans to Europeans (Figure \ref{fig:f3}D).
+I focus on this cline using an alternative projection by using  $F$-statistics of the form $F_4(X, Y; \text{Sardinian}, \text{Yoruba})$), which might e.g.
+be used if we were to quantify gene flow associated with the out-of-Africa expansion.
+These $F_4$-statistics are very well-approximated by the first two PCs, with a 99.2\% correlation between $F_4$ and its approximation using the first two PCs (Figure \ref{fig:f4}C).
 
-The $F_4$-statistics for the Westeurasian data set are slightly less sparse, the correlation coefficients between $F_4(X, \text{French}; \text{Finnish}, \text{Canary Islander})$ and its approximation using the first two or three PCs is 95.5\% and 99.1\% respectively (Figure \ref{fig:f4}E).
-I also show that the interpretation of $F_4$ as a projection can be used as a useful visualization (Figure \ref{fig:f4}D).
-On the $x$-axis, I plot $\langle X; \text{Finnish}, \text{Canary Islander}\rangle$, so that the horizontal distance between all pairs of populations corresponds to their respective $F_4$-statistics $F_4(X, Y; \text{Finnish}, \text{Canary Islander})$.
-On the $Y$-axis and with the coloring I display the first two principal components of the residual, i.e.
-the genetic variation that is missed by viewing the data through this projection.
-We find that most European populations have positive values on residual PC1, and are relatively closely clustered.
-In contrast Middle Eastern and Caucasian populations have negative values on this gradient.
-This allows us to visualize that this particular $F_4$-projection does an adequate job if we are interested in describing European variation, but it fails to explain the non-European data.
-We can further quantify this by investigating the percent of variance explained on each axis (Figure \ref{fig:f4}F), where I find that the projection axis only describes around 12\% of hte variation, compared to residual PC1 with almost 30\%.
+In Figure \ref{fig:f4}D, I show the projection $\langle X; \text{Sardinian}, \text{Yoruba} \rangle$ on the $X$-axis, which means that $F_4(X, Y; \text{Sardinian}, \text{Yoruba})$ is  is proportional to their horizontal distance between $X$ and $Y$.
+The first two residual PCs are given on the Y-axis and in the coloring; this visualization reveals some variation within Africans (with Mbuti, Biaka and Ju|'hoansi) that is largely orthogonal to this gradient, as is the variation between Europeans, Asian and the Americans.
+
+The percentage of between-population variance explained by the Sardinia-Yoruba axis (24\%) is much lower than PC1 (40\%).
+However, the cumulative variance explained of the first two axes is similar, with (52\%) explained when adding residual PC1 to the projection, compared to 55\% for the first two PCs.
+The advantage of the projection is that it displays orthogonal components more explicitly, reveals distinct structure in Africans and non-Africans and thus can be used to test assumptions of more complex models.
 
@@
 \includegraphics[width=\textwidth]{figures/fig_f4_data.pdf}	
-	\caption{\textbf{PCA and $F_4$-statistics} A: Spectrum of select $F_4$-statistics in World data set.
-B: Projection angle representation of $F_4(X, Sardinian; Mozabite, Yoruba)$ (red) and approximations using two (blue) and ten (green) PCs.
-C: Spectrum of select $F-4$-statistics in Westeurasian data set.
-D: Scatterplot of $F_4$-projection on Finnish-Canary Islanders axis and residual PC1.
-	E: $F_4(X, \text{French}, \text{Finnish}, \text{Canary Islander})$ vs.
-prediction using two (blue) and ten(red) PCs.
-F: Percent variance explained for the projection of panel D and the first nine residual PCs.
+	\caption{\textbf{PCA and $F_4$-statistics} A: Projection angle and correlation coefficient $r$ representation of $F_4(X, Sardinian; Saudi, Georgian)$ (red) in the Westeurasian data set, and approximations using two (green) and three (blue) PCs.
+	B: Spectrum of select $F_4$-statistics in the Western Eurasian data set.
+C: Spectrum of $F-4$-statistics in World data set.
+D: Scatterplot of $F_3$-projection on Sardinian-Yoruba axis  and residual PC1.
+	E: Percent variance explained from of the projection based on $F_3$ in panel D and first nine residual PCs (light gray), compared with percent variance explained by first ten PCs (dark gray).
 	}
@@
 \section{Discussion}
-Particularly for the analysis of ancient DNA, $F$-statistics have been established as a powerful tool to describe population genetic diversity, but they have a number of limitations.
-In particular, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
-As a consequence, researchers concerned about model fits may ascertain reference populations in a way that satisfies these assumptions, thus inadvertently making population structure appear sparser than it truly is.
-This is perhaps most obvious from Figure \ref{fig:f3}B, where large gaps are present.
-However, these gaps are due to sparse sampling that disappear if more populations were sampled \citep[e.g.][]{peter2020a}, not due to gaps in genetic diversity.
-If population ascertainment is not done very carefully, tools built on top of $F$-statistics, such as \texttt{qpGraph} and \texttt{qpAdm}, may thus only provide a very loose lower bound for the number of gene flow events.
+Particularly for the analysis of ancient DNA, $F$-statistics  are  a powerful tool to describe population genetic diversity.
+Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA plot.
+This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
+As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using specific models using a lower number of populations.
+It could also allows for calculation of $F$-statistics involving unsampled populations, which can be useful for checking assumptions.
 
-In contrast, the perspective on $F$-statistics in data space \citep{oteo-garcia2021} and on PCA does not require assumptions on number of admixture or gene flow events.
-Independent of any model, A population $X_x$ can be thought of as admixed between $ X_1$ and $X_2$ if it lies in the ball with diameter $\overline{X_1X_2}$.
-And independent of any models, two axes of variation that are not perpendicular to each other shows that there is some degree of shared history.
+As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. 
+However, in humans, most populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008}.
+This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow.
+In my example (Figure \ref{fig:f3}A), most Southern European populations are ``admixed'' between Basques and Turkish, but a more accurate model might be one of continuous variation where Basque and Turkish lie on one of multiple gradients; which is more directly visualized with PCA.
+There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}.
+One issue with these approaches is that they are usually restricted to at most a few dozen populations.
+As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include.
+A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data.
+However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits.
+One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs.
+However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}.
+The PCA-based  interpretation offers an alternative that trades interpretability for robustness.
+Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates naturally to models of gene flow that do not assume gene flow is rare or discrete.
+Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.
 
-The connection between the graph-based and PCA-based interpretations of $F_3$ and $F_4$ is due to concept of projection.
-As illustrated above, both $F$-statistics can be thought of as projections on any particular axis of variation.
-If that variation corresponds to a tree-branch, the interpretations align.
-A collorary to this interpretation is the importance of orthogonality, or independence for describing gene flow.
-Populations without gene-flow will evolve independently, and their changes in allele frequency will therefore be orthogonal, resulting in $F_4$-statistics of zero, and right angles on a PCA plot.
-Even though this is only true when considering all PCs, the result holds reasonably well when just looking at the first two PCs -- in our example of the world-overview data, I found that the variation in Sub-Saharan Africa is mostly -- but not completely -- uncorrelated with the variation between European and Asian populations (Figure \ref{fig:f4}A).
-
-While the data space produces a much larger model space than trees, the main drawback of the PCA-based interpretation is a lack of interpretability; it is not easy to define a generative model that could generate a complex PCA-plot.
-Thus, in the future the tree- and PCA-based interpretations will likely be used in conjunction.
-
- However, despite the apparent complexity, the PC-spaces \emph{are} sparse, and using just ten or even two PCs often gives very good approximations.
-
-To make PCA and $F$-statistics more comparable in practical settings, there are a number of -- mainly statistical -- concerns that still need to be addressed in future work.
-The perhaps most obvious one is that PCA is most frequently run on individuals, whereas $F$-statistics are often calculated on populations.
-This is not a conceptual issue, as both PCA and $F$-statistics can be run on either \citep{cavalli-sforza1994}.
-Population based analyses have the advantage that they are easier to interpret and compute (current packages are ill-equipped to calculate all pairwise $F$-statistics between data sets with thousands of individuals \citep{patterson2012}).
-However, this requires the assumption that the within-population variation is independent from the between-population variation; something that is analogous to the variance partitioning based on PCs here.
+To make this link directly applicable to data analysis, there are a number of -- primarily statistical -- concerns that will need to be addressed.
+Fist, PCA is most frequently run on individuals, whereas $F$-statistics are often calculated on populations.
+This is largely because in most workflows, PCA is run much earlier than $F$-statistics; it is a standard assumption of $F$-statistics that there is no population substructure \citep{patterson2012}, and an easy way to test that is ensure that all individuals cluster tightly on a PCA.
 
 A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \citep{patterson2006}.
-This is only a difference of convention; $F$-statistics could also be calculated using the same weighting.
+This is a difference of convention; since all SNPs have the same expectation, $F$-statistics could also be calculated using the same weighting.
 The close connection between the two approaches developed here suggest that for most analyses, users might want to be consistent and use the same weighting for both types of analyses.
@@
 The third and perhaps biggest gap are statistical issues.
-The treatment here focusses on the mean estimated $F$-statistic, but many  applications of $F$-statistics are based on hypothesis tests \citep{patterson2012}.
-This requires estimating accurate standard errors for these statistics, which is difficult since nearby SNPs will be correlated \citep{hahn2018}.
-In contrast, standard PCA does not modeljointly models the covariance matrix due to population structure and sampling.
-On the other hand, for both data sets I investigated here, the matrix $\MF$ of $F$-statistics estimated using admixtools2 is not a proper squared Euclidean distance matrix, i.e.
+The treatment here focuses on the mean estimated $F$-statistic, but many  applications of $F$-statistics are based on hypothesis tests \citep{patterson2012}.
+This requires estimating accurate standard errors for these statistics, which is difficult since nearby SNPs will be correlated due to recombination \citep{hahn2018}.
+In contrast, PCA jointly models the covariance matrix due to population structure and sampling, so if hypothesis tests are desired this will need to be incorporated.
+
+An advantage of calculating $F$-statistics based on PCs is that they yield consistent estimtates.
+For both data sets I investigated here, the matrix $\MF$ of $F$-statistics obtained  using admixtools2 is not a proper squared Euclidean distance matrix, i.e.
 it is not negative semidefinite and has imaginary PCs.
-This is not a practical when considering single $F$-statistics or PCA (for analyses here, I used a nearby matrix  \citep{higham2002} with no apparent loss of precision).
-It does however mean that tools that use matrices of $F$-statistics, such as \texttt{qpadm} or \texttt{qpgraph} may be ill-calibrated, which may partly explain why they generally have poor out-of-sample predictive power and are restricted to a few dozen samples at a time.
-A model-based framework based on probabilistic PCA \citep{meisner2021, agrawal2020, hastie2015} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
-
-\begin{itemize}
-    \item weighting of SNPs
-    \item estimation error
-    \item propagating erros
-    \item exploratory data analysis
-    \item missing data
-    \item population vs. sample allele frequencies
-\end{itemize}
+A model-based framework based on probabilistic PCA \citep{hastie2015, meisner2021, agrawal2020} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
 
@@
 As these are scalars, they do not change above calculation.
+\bibliographystyle{plainnat}
 \bibliography{main}

### 2021-07-05 data/f4_fig.R

@@ -3,13 +3,14 @@ library(glue)
 BTHEME = theme_classic() + theme(legend.position="none")
 
 if(T){
-f2s = admixtools::read_f2("westeurasian1/")
-pcmat = pca_from_f2s(f2s)
-pcs = pca_from_pcmat(pcmat)
-idx = "Finnish"
-idy = "Canary_Islander"
-idz = "French"
+    f2s = admixtools::read_f2("worldfoci2")
+    pcmat = pca_from_f2s(f2s)
+    pcs = pca_from_pcmat(pcmat)
+    idx = "Sardinian"
+    idz = "Han"
+    idy = "Yoruba"
 ids = c(idx, idy)
+n = nrow(pcs)
 
 sample_list = pcs$pop 
 
@@ -44,7 +45,7 @@ P3 = X %>%
     coord_fixed() + 
     BTHEME + 
     scale_color_viridis_c() + 
-    xlab("proj (Finnish, Canary Islander)") + 
+    xlab(glue("⟨X; {idx}, {idy}⟩")) + 
     ylab("Residual PC1")
 
 f42 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=idz) %>% arrange(pop1, pop2)
@@ -55,39 +56,46 @@ f4pc2 = f4_from_pc_matrix(pcmat, idz, idx, idy) %>%
         mutate(PC1_2=rowSums(across(PC1:PC2)), 
                PC1_3=rowSums(across(PC1:PC3)),
                PC1_10=rowSums(across(PC1:PC10)))
-    v = f4pc2 %>% 
-        filter(pop1 %in% c("Greek", "Iranian", "Sicilian", "Estonian", "Saudi", "Georgian", "Spanish")) %>% 
-        pivot_longer(PC1:PC33) %>% 
-        mutate(PC=as.integer(substr(name, 3,20))) %>% 
-        #group_by(pop1) %>% mutate(value=cumsum(value)) %>% 
-        ungroup 
-    P4= v %>%   ggplot(aes(x=PC, y=value, group=pop1, color=pop1, fill=pop1)) + 
+v = f4pc2 %>% 
+    filter(pop1 %in% c("Masai", "Basque", "Papuan", "Mbuti", "Surui", "AA", "GujaratiD")) %>% 
+    pivot_longer(PC1:PC33) %>% 
+    mutate(PC=as.integer(substr(name, 3,20))) %>% 
[…]

### 2021-07-05 data/f4ratio_fig.R

@@ -1,0 +1,190 @@
+source("fscripts.R")
+library(glue)
+BTHEME = theme_classic() + theme(legend.position="none")
+
+if(F){
+#f2s = admixtools::read_f2("f4ratio"); sample_list= rownames(f2s)[!rownames(f2s) %in% "Denisova.DG"]
+f2s = admixtools::read_f2("f4ratio", pops=sample_list )
+pcmat = pca_from_f2s(f2s)
+pcs = pca_from_pcmat(pcmat)
+idx = "Primate_Chimp"
+#idz = "Dinka.DG"
+idy = "Altai_Neanderthal.DG"
+idy = "Altai_Neanderthal.DG"
+
+#idx = "Primate_Chimp"
+#idy = "Yoruba"
+#idz = "Altai_Neanderthal.DG"
+ids = c(idx, idy)
+
+sample_list = pcs$pop 
+
+v = pcmat[ids,] %>% diff
+
+pmat = t(v) %*% v / (v %*% t(v))[1,1]
+qmat = diag(length(v)) - pmat
+a = pcmat %*% pmat
+b = pcmat %*% qmat
+A = t(t(a) - a[idz,])
+B = t(t(b) - b[idz,])
+E = eigen(b %*% t(b))
+K = t(t(E$vectors) * sqrt(pmax(E$values, 0)))
+K = K %>% as.data.frame %>% mutate(pop = rownames(B))
+
+EA = eigen(a %*% t(a))
+KA = t(t(EA$vectors) * sqrt(pmax(EA$values, 0)))
+KA = KA %>% as.data.frame %>% mutate(pop = rownames(B))
+X = KA %>% select(pop, A=V1) %>% left_join(K)
+
+
+
+f4 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+f4mat = f4 %>% 
+    select(pop1, pop2, est) %>% 
+    pivot_wider(names_from=pop2, values_from=est, values_fill=0, names_sort=T) %>% 
+    column_to_rownames('pop1')
+f4pc = f4_from_pc_matrix2(pcmat, idx, idy) %>%
+    left_join(f4, .) %>% 
+    filter(! pop1 %in% ids, ! pop2 %in% ids) %>%
+    mutate(PC1_2=rowSums(across(PC1:PC2)), PC1_3=rowSums(across(PC1:PC3)))
+
[…]

### 2021-07-07 fstats_pca_lewontin.tex
submission

@@
 \begin{abstract}
-The genetic diversity in human population is shaped by our complex history.
-$F$-statistics are a widely used tool to model genetic diversity when populations are related acording to a graph and gene flow is rare.
-Here, I use a geometric interpretation of  to interpret $F$-statistic in the context of Principal Component Analysis (PCA), showing that the two approaches are closely related.
-I show that orthogonal projections are a key concept to establish this link, and that it leads to simple geometric interpretations of $F$-statistics on a PCA-plot.
+Human  genetic diversity is shaped by our complex history.
+Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
+Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches.
+I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link.
 I illustrate my results on two examples, one of local, and one of global human diversity.
-In both examples, I find that population structure is sparse, and only a few components contribute to most statistics, and I  develop novel visualizations that allow for checking the assumptions of more sophisticated models.
-My results allows interpreting  $F$-statistics in arbitrarily structured populations, to obtain  more complete and less biased models of human genetic variation.
+In both examples, I find that population structure is sparse, and only a few components contribute to most statistics.
+Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models.
+My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
 \end{abstract}
 \section{Introduction}
-About 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002a}, but the information contained in these 15\% is sufficient to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}.
+Most genetic variation in humans is shared between all of us, but around 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002a}.
+Genome-scale data has allowed us to leverage the information contained in these 15\% to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}.
 For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
 
-Human genetic diversity is caused by a number of forces, and has both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}.
-Spatially isolated populations are expected to slowly diverge, particularly if they are separated by barriers to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}.
+However, understanding, conceptualizing and modeling this variation is far from trivial, particularly since misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies.
+Lewontin's landmark \citeyear{dobzhansky1972} paper on the apportionment of human genetic diversity was one of the first to question ``races'' not only on ethical, but also on scientific grounds.
+In Lewontin's data, less than half of the already small proportion of between-population genetic variation can be attributed to ``racial'' continental-scale groupings.
+Over the last five decades, this view has been corroborated and extended many times \citep{cann1987, cavalli-sforza1994, cann2002, rosenberg2002, reich2018a}, and we now know that human races are an arbitrary, biologically useless polyphyletic grouping that is maintained by historical and social conventions.
+
+Modern descriptions of human genetic diversity focus on the evolutionary processes that caused it, and which gives rise to  both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}.
+In isolation, populations are expected to slowly diverge, particularly if they are separated by barriers to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}.
 On the other hand, major population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}.
-Local migration between neighboring populations will flatten differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and early modern humans \citep{green2010} adds further complexity.
+Local migration between neighboring populations will flatten differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and modern humans \citep{green2010},will lead to locally increased diversity.
 
 This complex population structure is frequently handled by using multiple models with different assumptions; each emphasizing a particular aspects of the data.
-Data-driven methods such as Principal Component Analysis (PCA) \citep{cavalli-sforza1994} structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable.
+Data-driven methods such as Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable.
 For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests.
@@
 Particularly in the analysis of human ancient DNA, a set of techniques based on $F$-statistics \textit{sensu} Patterson have risen in popularity \citep{patterson2012, peter2016}.
-This framework is based on the assumption that the relationship between three or four populations is often tree-like, and allows for a variety of tests of treeness and more complex demographic models \citep{patterson2012, harney2021}.
-However, the connections between PCA, $F$-statistics and demographic models are currently unclear, which makes quantitative comparisons, detecting model violations and joint interpretation of the results of these approaches difficult.
+This framework is based on the null-assumption that the relationship between  populations is tree-like, and then uses tests involving three or four populations to probe for violations of treeness \citep{patterson2012}.
+In a further step, $F$-statistics of multiple populations can be combined to estimate parameters in more complex models \citep{patterson2012, harney2021}.
+However, the connections between PCA, $F$-statistics and explicit demographic models are often unclear, which makes quantitative comparisons, detecting model violations and joint interpretation of these approaches difficult.
 Since both $F$-statistics and PCA are functions of expected pairwise coalescent times \citep{mcvean2009, peter2016}, this is one avenue to link these approaches.
-Here, I instead use the geometric interpretation of $F$-statistics introduced by  \cite{oteo-garcia2021} to directly visualize $F$-statsitics on a PCA plot.
+Here, I instead use the geometric interpretation of $F$-statistics developed by  \cite{oteo-garcia2021} to directly visualize $F$-statistics on a PCA plot.
 
@@
 \cite{jolliffe2013}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
-Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2020a} or \cite{oteo-garcia2021} helpful.
+Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.
 
@@
 If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
-As a population may be represented by just one individual, there is no conceptual difference between these cases and I will
-refer to populations as unit for analysis. Since the allele
+As a population may be represented by just one individual, there is no conceptual difference between an individual-based and population-based analysis.
+Since the allele
 frequencies are between zero and one, we can interpret each Population $X_i$
@@
 However, the  historical processes that generated genetic variation often result in \emph{sparse} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.
-
-\begin{comment}
-	\begin{figure}
-		\includegraphics[width=\textwidth]{pca_explanation.png}
-		\caption{Basic Idea of PCA from 2D to 1D representation.
-A: Allele frequencies from different populations (blue dots) at two SNPs.
-A PCA is performed by centering the data (B), and rotating it (B) such that the first PC explains the majority of variation in the data, and the second PC is orthognal to the first, and explains the residual.
-A lower-dimensional approximation (in this case 1D) can be achieved by just keeping the first PC (E); which can be translated back to the original data space by inverting the rotation and centering (F).}
-		\label{fig:pca_explanation}
-	\end{figure}
-\end{comment}
 
@@
 The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is used to reveal population structure.
-The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times k$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \citep[e.g][]{francois2010}.
+The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times S$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \citep[e.g][]{francois2010}.
 
@@
 	\begin{align}
-	&F_2(X_1, X_2) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \normsq{X_1, X_2}\\
-	&F_3(X_1; X_2, X_3) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \langle X_1 - X_2, X_1 - X_3 \rangle\\	
-	&F_4(X_1, X_2; X_3, X_4) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
+	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \frac{1}{S}\normsq{X_1-  X_2}\\
+	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
+	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
 	\end{align}
@@
 where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product.
-The number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently.
+The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently.
+Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
+\begin{subequations}
+\begin{align}
+\langle a, b \rangle &= \sum_i a_ib_i\\
+\langle a, b \rangle &= \norm{a}\norm{b}\cos{\phi}\\
+\langle a, a \rangle &= \normsq{a}\\
+\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
+\end{align}
+\end{subequations}
+where $\phi$ is the angle between $a$ and $b$.
+
  Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
@@
 	\begin{align}
-2F_3(X_1; X_2, X_3) &= 2F_2(X_2, X_3) - F_2(X_1, X_2) + F_2(X_1, X_3)\\
-2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)
+2F_3(X_1; X_2, X_3) &= F_2(X_2, X_3) - F_2(X_1, X_2) - F_2(X_1, X_3)\label{eq:f3fromf2}\\
+2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
 	\end{align}
@@
 $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
-In a tree, the squared Euclidean distance $F_2(X, Y)$ measures the length of all branches between populations $X$ and $Y$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}.
+In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of all branches between populations $X_1$ and $X_2$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}.
 The length is a measure of genetic drift, and is non-negative if data is generated under a tree \citep{patterson2012}.
@@
 PCA and $F$-statistics are closely related.
-In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling \citep{gower1966}.
+In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling, which, for squared Euclidean ($F_2$)-distances, leads to an identical decomposition to PCA \citep{gower1966}.
 Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$.
@@
 \end{align}
-A detailed derivation of this is given in Appendix \ref{appendix:fonpc}.
-As $F_3$ and $F_4$ can be written as sums of $F_2$-terms, analogous relations apply.
+A derivation of this change-of-basis is given in Appendix \ref{appendix:fonpc}.
+As $F_3$ and $F_4$ can be written as sums of $F_2$-terms (Eqs. \ref{eq:f3fromf2}, \ref{eq:f4fromf2}), analogous relations apply.
 
-In most applications, we do not use all PCs, but instead use only the first $K$ PCs.
+In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
 Thus, 
@@
 If we sum up the approximation errors $F_2 - \hat F_2$ over all pairs of populations, we obtain the Frobenius-norm of the error $\normsq{\MF - \hat{\MF}}_F$; which is precisely the function that is minimized in MDS \citep{jolliffe2013}.
-Thus, $\hat\MF$ is the optimal approximation of $\MF$ for any $K$.
+Thus, $\hat\MF^{(K)}$ is an optimal sparse approximation of $\MF$ for any $K$.
 
 \section{Material \& Methods}
-The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots, and can be approximated using PCA.
-I will explore this in detail in the next section, and illustrate it on two sample data sets that I briefly introduce here.
-Both are based on the analyses by \cite{lazaridis2014} and use the ``Human Origins''-SNP set (597,573 SNPs), based on the subset of the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}.
+The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation in PCA-space, which can be approximated on PCA plots.
+In the next section I explore this connection in detail, and illustrate it on two sample data sets that I briefly introduce here.
+Both are based on the analyses by \cite{lazaridis2014}.
+The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs).
+SNPs with missing data in any population are excluded.
 The code used to create all figures and analyses will be available on [Archive to be created before publishing].
 
-\paragraph{World Overview data set}
+\paragraph{``World'' data set}
 This data set represents global human genetic variation (638 individuals from 33 population), as used by \citep{lazaridis2014}.
@@
 It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012, lazaridis2014}.
-Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus not particularly graph-like.
+Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus may not be particularly graph-like.
 
@@
 I perform analyses at the level of populations to ease presentation.
-It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples.
+It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples \citep{patterson2012}.
 All computations are performed in \texttt{R}.
@@
  Thus if $\hat{F}_3 \ll F_3$, the true circle will be bigger than would be predicted from a 2D-plot.
-In this case, substantial relevant genetic differentiation is ``hidden'' in the higher PCs.
-If this is the case, populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics, because they are outside the $n$-ball in higher dimensions.
-However, if a population lies outside the circle on any 2D-projection, $F_3$ is guaranteed to be bigger than 0.
+In this case, substantial relevant genetic differentiation is ``hidden'' in the higher PCs, and populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics.
+This is because they are outside the $n$-ball in higher dimensions.
+The converse interpretation is more strict: if a population lies outside the circle on \emph{any} 2D-projection, $F_3$ is guaranteed to be bigger than 0.
 
 \paragraph{Example}
-I visualize this on the admixture statistic $F_3(X; Basque, Turkish)$, on the first two PCs of the Westeurasian data set (Figure \ref{fig:f3}A).
+As an example, I visualize the admixture statistic $F_3(X; \text{Basque}, \text{Turkish})$, on the first two PCs of the Westeurasian data set (Figure \ref{fig:f3}A).
 In this case, the projected $n$-ball (light gray) and circle based on 2D (dark gray) align relatively closely, but several populations inside the ball (e.g.
-Sardinian, Finnish) have, in fact, positive $F_3$-values, which reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure.
+Sardinian, Finnish) have, in fact, positive $F_3$-values.
+This reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure.
 This is expected because for spatially continuous populations, PCA will not be sparse \citep{novembre2008a}.
@@
 
-\subsection{$F_3$ as a projection}
-The inner product $\langle X_x - X_1, X_x - X_2 \rangle$ is closely related to the projection of one vector onto the other.
+\subsection{$F$-statistics as projections}
+The inner product $\langle X_x - X_1, X_x - X_2 \rangle$ is closely related to the projection of a vector onto another one.
 This interpretation is useful either when calculating an outgroup-$F_3$-statistic, or when we want to find the ``best'' admixture source $X_2$ if we assume the admixed population $X_x$ and one source $X_1$ are known.
@@
 \paragraph{Example}
-Again, these projection will be orthogonal when using the full data, and may only be approximately orthogonal when just using the first two PCs.
-In Figure \ref{fig:f3}D, I visualize this on the outgroup-$F_3$-statistic $F_3(Mbuti; Mozabite, X_i)$, i.e.
+Again, these projection will be orthogonal when using the full data, and may only be approximately orthogonal when approximated using the first two PCs.
+In Figure \ref{fig:f3}D, I visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Mozabite}, X_i)$, i.e.
 a statistic that aims to find the population most closely related to Mozabite (a Berber ethnic group from the northern Sahara), assuming the Mbuti are an outgroup.
@@
 For each population, the projection is indicated with a grey line.
-In the full data space, this line is always orthogonal to the segment Mbuti-Mozabite, but on the plot (i.e.) the subspace spanned by the first two PCs, this is not necessarily the case.
+In the full data space, this line is always orthogonal to the segment Mbuti-Mozabite, but on the plot (i.e.
+the subspace spanned by the first two PCs), this is not necessarily the case.
 We see that particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs, and the coloring based on the full $F_3$-statistics shows that in this case, the first two PCs approximate the $F_3$-statistic very well.
@@
 
-\subsection{$F_4$-statisticss as angles}
+\subsection{$F_4$-statistics as angles}
 The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct.
-As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3$ and $X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection (Figure \ref{fig:geom}E).
+As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection (Figure \ref{fig:geom}E).
 
@@
 $$X_3 - X_4  = (X_3 - X'_3) + (X'_3 - X'_4) + (X'_4 - X_4)\text{.}$$ Of these three branches, the first and last are orthogonal to $X_1 - X_2$ and thus the $F_4$ statistic is just the internal branch of the tree (Figure \ref{fig:geom}F).
+It also suggests a number of diagnostic $F$-statistics that check assumptions; for example if the tree holds, then
+$F_4(X_3, X_3'; X_4, X_4') = 0 $.
 
@@
 The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:f3}A), a European gradient (from Sardinian to Chuvash), and a Asian cline (from Arab to Caucasus populations).
-
-This is quantified in Figure \ref{fig:f4}A, where I plot the angle corresponding to $F_4(X, Sardinian; Saudi, Georgian)$.
-For most European populations, using two PCs (green points) gives an angle close to zero, corresponding to a correlation coefficient $r>0.9$.
-Just adding PC3 (blue) gives a close approximation to the full data (red).
-This can also be seen from the spectrum of these statistics (Figure \ref{fig:f4}B), which has high loadings for the first three PCs, with minimal contributions from the higher ones.
+This is quantified in Figure \ref{fig:f4}A, where I plot the angle corresponding to $F_4(X, \text{Sardinian}; \text{Saudi}, \text{Georgian})$.
+For most European populations, using two PCs (green points) gives an angle close to zero, corresponding to a correlation coefficient between the two clines of $r>0.9$.
+Just adding PC3 (blue), however, shows that the parallelism of the clines is spurious: Using three PCs or the full data (red) shows that most correlations are low.
+I arrive at a similar interpretation from the spectrum of these statistics (Figure \ref{fig:f4}B), which has high loadings for the first three PCs, with minimal contributions from the higher ones.
 
@@
 In Figure \ref{fig:f4}D, I show the projection $\langle X; \text{Sardinian}, \text{Yoruba} \rangle$ on the $X$-axis, which means that $F_4(X, Y; \text{Sardinian}, \text{Yoruba})$ is  is proportional to their horizontal distance between $X$ and $Y$.
-The first two residual PCs are given on the Y-axis and in the coloring; this visualization reveals some variation within Africans (with Mbuti, Biaka and Ju|'hoansi) that is largely orthogonal to this gradient, as is the variation between Europeans, Asian and the Americans.
+The first two residual PCs are given on the Y-axis and in the coloring; this visualization reveals some variation within Africans (with Mbuti, Biaka and Ju$\vert$'hoansi) that is largely orthogonal to this gradient, as is the variation between Europeans, Asian and the Americans.
 
-The percentage of between-population variance explained by the Sardinia-Yoruba axis (24\%) is much lower than PC1 (40\%).
-However, the cumulative variance explained of the first two axes is similar, with (52\%) explained when adding residual PC1 to the projection, compared to 55\% for the first two PCs.
-The advantage of the projection is that it displays orthogonal components more explicitly, reveals distinct structure in Africans and non-Africans and thus can be used to test assumptions of more complex models.
+The percentage of between-population variance explained by the Sardinia-Yoruba axis (24\%) is much lower than that of the first PC (40\%, Figure \ref{fig:f4}E).
+However, the cumulative variance explained by the first two axes is similar, with (52\%) explained when adding residual PC1 to the projection, compared to 55\% for the first two PCs.
+The advantage of specifying one axis is that it displays the orthogonal components more explicitly, reveals distinct structure in Africans and non-Africans and thus can be used to test assumptions of more complex models.
 
@@
 As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. 
-However, in humans, most populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008}.
+However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008}.
 This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow.
@@
 The PCA-based  interpretation offers an alternative that trades interpretability for robustness.
-Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates naturally to models of gene flow that do not assume gene flow is rare or discrete.
+Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow.
 Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.
@@
 
-A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \citep{patterson2006}.
-This is a difference of convention; since all SNPs have the same expectation, $F$-statistics could also be calculated using the same weighting.
+A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \citep{patterson2006, patterson2012}.
+This is a difference of convention; since $F$-statistics are summed over SNPs with the same expectation, $F$-statistics could also be calculated using the same weighting.
 The close connection between the two approaches developed here suggest that for most analyses, users might want to be consistent and use the same weighting for both types of analyses.
@@
 &=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
-&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + \sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
+&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
 &=& \sum_k (P_{ik} - P_{jk})^2

### 2021-07-07 data/f3_fig_ind.R

@@ -2,7 +2,7 @@ source("fscripts.R")
 library(ggpubr)
 BTHEME = theme_classic() + theme(legend.position="none")
 
-if(T){
+if(F){
 f2s = admixtools::read_f2("westeurasian1ind/")
 pcmat = pca_from_f2s(f2s)
 pcs = pca_from_pcmat(pcmat)
@@ -48,7 +48,7 @@ P5 = f3pc %>% pivot_longer(PC1:PC10) %>%
     ylab("f3(X; Tur., Bas.)") + 
     scale_x_continuous(breaks=1:10) + BTHEME
 }
-if(F){
+if(T){
 f2s = admixtools::read_f2("worldfoci2ind")
 pcmat = pca_from_f2s(f2s)
 pcs = pca_from_pcmat(pcmat)
@@ -83,9 +83,8 @@ df = tibble(x=pcs$PC1,
 pcs2 = pcs %>% left_join(f3, by=c(pop="pop3")) 
 P3 = ggplot() + 
     geom_abline(intercept=intercept, slope=slope) +
-    #geom_circle(aes(x0=PC1, y0 =PC2, r=radius), data=circle_df, color='lightgray', fill='lightgray') +
-    #geom_circle(aes(x0=PC1, y0 =PC2, r=r0), data=circle_df, color='darkgray', fill='darkgray') +
-    geom_text(aes(x=PC1, y=PC2, label=pop, color=est), data=pcs2) +
+    #geom_text(aes(x=PC1, y=PC2, label=pop, color=est), data=pcs2) +
+    geom_point(aes(x=PC1, y=PC2, label=pop, color=est), data=pcs2) +
     scale_color_viridis_c(direction=1) + 
     geom_segment(data=df, aes(x=x, y=y, xend=xend, yend=yend), alpha=.1, color='lightgray') +
     coord_fixed() + BTHEME
@@ -93,7 +92,7 @@ P3 = ggplot() +
 
 P4 = f3pc %>% select(pop3, est, PC1_2, PC1_10, PC1_20) %>% 
     filter(!pop3 %in% c(idx, idy)) %>%
-    pivot_longer(starts_with("PC"), names_to='PC', values_to="f3") %>% 
+#    pivot_longer(starts_with("PC"), names_to='PC', values_to="f3") %>% 
     ggplot(aes(x=est, y=f3, color=PC)) + geom_point() + geom_abline() + 
[…]

### 2021-07-07 data/f4_fig_ind.R

@@ -2,8 +2,8 @@ source("fscripts.R")
 library(glue)
 BTHEME = theme_classic() + theme(legend.position="none")
 
-if(F){
-f2s = admixtools::read_f2("westeurasian1/")
+if(T){
+f2s = admixtools::read_f2("westeurasian1ind/")
 pcmat = pca_from_f2s(f2s)
 pcs = pca_from_pcmat(pcmat)
 idx = "Finnish"
@@ -99,7 +99,7 @@ f4pcall = f4_from_pc_matrix2(pcmat, px=idx, py=idy) %>%
 
 
 }
-if(T){
+if(F){
     f2s = admixtools::read_f2("worldfoci2ind")
     pcmat = pca_from_f2s(f2s)
     pcs = pca_from_pcmat(pcmat)
@@ -175,8 +175,8 @@ if(T){
 #ggsave( "f4_iwesteurasia2.png", P4, width=1.25, height=4, scale=1.5)
 #ggsave( "f4_iwesteurasia3.png", P5, width=1.3, height=1, scale=1.5)
 #ggsave( "f4_iwesteurasia4.png", P6, width=1.3, height=1, scale=1.5)
-ggsave( "f4_iworld1.png", P1, width=1.75, height=4, scale=1.5)
-ggsave( "f4_iworld2.png", P2, width=1.25, height=4, scale=1.5)
+#ggsave( "f4_iworld1.png", P1, width=1.75, height=4, scale=1.5)
+#ggsave( "f4_iworld2.png", P2, width=1.25, height=4, scale=1.5)
 
 
 

### 2021-07-11 fstats_pca_lewontin.tex

@@
 In this section, I will give a very brief formal introduction to $F$-statistics and PCA.
-A more detailed technical treatise of PCA is given in e.g.
-\cite{jolliffe2013}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
+A more detailed technical treatise of PCA is given by e.g.
+\cite{jolliffe2013} and \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
 Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.
@@
 A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \citep{patterson2006, patterson2012}.
-This is a difference of convention; since $F$-statistics are summed over SNPs with the same expectation, $F$-statistics could also be calculated using the same weighting.
+This is a difference of convention \citep{cavalli-sforza1975}; since $F$-statistics are summed over SNPs with the same expectation, $F$-statistics could also be calculated using the same weighting.
 The close connection between the two approaches developed here suggest that for most analyses, users might want to be consistent and use the same weighting for both types of analyses.

### 2021-07-12 fstats_pca_lewontin.tex
better discussion

@@
 	\begin{align}
-	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \frac{1}{S}\normsq{X_1-  X_2}\\
+	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
+	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
 	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
@@
 \langle a, b \rangle &= \sum_i a_ib_i\\
-\langle a, b \rangle &= \norm{a}\norm{b}\cos{\phi}\\
+\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
 \langle a, a \rangle &= \normsq{a}\\
@@
 \end{align}
-A derivation of this change-of-basis is given in Appendix \ref{appendix:fonpc}.
+A derivation of this change-of-basis is given in Appendix \ref{appendix:fonpc}, Equation \ref{eq:changeofbasis}.
 As $F_3$ and $F_4$ can be written as sums of $F_2$-terms (Eqs. \ref{eq:f3fromf2}, \ref{eq:f4fromf2}), analogous relations apply.
@@
 SNPs with missing data in any population are excluded.
-The code used to create all figures and analyses will be available on [Archive to be created before publishing].
+The code used to create all figures and analyses will be available on \url{https://github.com/BenjaminPeter/fstats_pca}.
 
@@
 
-\paragraph{Computation $F$-statistics and PCA}
+\paragraph{Computing $F$-statistics and PCA}
 I perform analyses at the level of populations to ease presentation.
@@
 All computations are performed in \texttt{R}.
-I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute $F$-statistics.
+I use \texttt{admixtools 2.0.0} (\url{https://github.com/uqrmaie1/admixtools}) to compute $F$-statistics.
 To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, obtain a nearby negative semidefinite matrix using the \texttt{nearPD} function, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.
@@
 2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
-      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0
+      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0 \label{eq:f3neg}
 \end{eqnarray}
@@
 $F_3$ is negative when the triangle is obtuse, i.e.
-$X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_1X_2}$ (Figure \ref{fig:geom}B).
+$X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_1X_2}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).
 
@@
 This is because they are outside the $n$-ball in higher dimensions.
-The converse interpretation is more strict: if a population lies outside the circle on \emph{any} 2D-projection, $F_3$ is guaranteed to be bigger than 0.
+The converse interpretation is more strict: if a population lies outside the circle on \emph{any} 2D-projection, $F_3$ is guaranteed to be bigger than 0 (see Equation \ref{eq:f3circle} in the Appendix).
 
@@
 This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
-As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using specific models using a lower number of populations.
-It could also allows for calculation of $F$-statistics involving unsampled populations, which can be useful for checking assumptions.
+As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, (e.g using a lower number of populations).
+It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are explained solely by higher PCs, and so whenever such contradictions arise, and so higher PCs will be informative for population structure.
+Previous interpretation of PCA in the context of population genetic models have primarily focused on the PCs, which can be derived analytically for trees \citep{cavalli-sforza1975} and homogeneous spatial models \citep{novembre2008a}.
+My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs.
+One consequence is that the results here are not impacted by sample ascertainment and sample sizes \citep{mcvean2009, novembre2008a}, which are common concerns in the interpretation of PCA.
+However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis.
+From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data
 
 As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
-However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008}.
+However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}.
 This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow.
@@
 \appendix
-\section{Derivation}\label{appendix:fonpc}
+\section{Derivations}\label{appendix:fonpc}
+\setcounter{equation}{0}
+\renewcommand{\theequation}{\thesection\arabic{equation}}
+Depending on a readers' background in linear algebra, these results may appear elementary; I include them here for reference and because they were not obvious to me at the onset of this project.
+\paragraph{$F$-statistics are invariant under a change-of-basis}
 \begin{eqnarray}
@@
 &=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
-&=& \sum_k (P_{ik} - P_{jk})^2
+&=& \sum_k (P_{ik} - P_{jk})^2 \label{eq:changeofbasis}
 \end{eqnarray}
@@
 As these are scalars, they do not change above calculation.
+
+\paragraph{The region of negative $F_3$-statistics is a $n$-ball}
+Without loss of generality, assume that $X_1 = (r, 0, 0, \dots)$ and $X_2 = (-r, 0, 0, \dots)$, and let us assume that $X_x$ has coordinates $(x_1, x_2, \dots, x_S)$ Assuming $F_3(X_x; X_1, X_2) = 0$, equation \ref{eq:f3neg} becomes
+\begin{align}
+2 F_3(X_x; X_1, X_2) &=  \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} = 0 \nonumber\\
+&= \left[(x_1-r)^2 + \sum_{i=2}^S x_i^2\right] + \left[(x_1 + r)^2 + \sum_{i=2}^S x_i^2\right] - 4r^2 \nonumber\\
+&= 2 \left[\sum_{i=1}^S x_i^2 + r^2 + x_1r - x_1r\right] - 4r^2\nonumber\\
+F_3(X_x; X_1, X_2) & = -r^2 + \sum_{i=1}^S x_i^2 = -r^2 + \normsq{X_x} = 0  \text{,} \label{eq:f3sphere}
+\end{align}
+which is the equation of a $n$-sphere with radius $r$ and center at the origin, as assumed from the placing of $X_1$ and $X_2$.
+Now, assume that $F_3$ is negative, i.e.
+$F_3(X_x; X_1, X_2) = -k < 0$.
+Moving $r^2$ to the left we obtain
+\begin{equation}
+r^2 - k = \normsq{X_x} \text{,}
+\end{equation}
+which is another $n$-sphere with a smaller radius, showing that all points inside the $n$-sphere will have negative $F_3$-values.
+
+\paragraph{If a population lies outside the circle of this $n$-Sphere in any 2D-projection, $F_3$ is positive} 
+Assume the center of the $n$-sphere $C = \frac{X_1 + X_2}{2} = (c_1, c_2, \dots c_S)$, and $X_x = (x_1, x_2, \dots x_S)$.  Then, 
+\begin{align}
+F_3(X_x; X_1, X_2) &= \normsq{X_x - C} - r^2 \nonumber\\
+&= \underbrace{(x_1 - c_1)^2 + (x_2 - c_2)^2}_{>r^2} + \underbrace{\sum_{i=3}^S (x_i - c_i)^2}_{\geq 0} - r^2 \nonumber\\
+& > 0 \text{.}\label{eq:f3circle}
+\end{align}
+The condition $(x_1 - c_1)^2 + (x_2 - c_2)^2 >r^2$ is satisfied whenever $X_x$ is outside the circle obtained from projecting the $n$-sphere on the first two dimensions.
+An analogous argument applies for any low-dimensional representation.
+
 \bibliographystyle{plainnat}

### 2021-07-16 fstats_pca_lewontin.tex
typos
better explanation on some technical things

@@
 	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively.
-The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is used to reveal population structure.
-The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times S$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \citep[e.g][]{francois2010}.
+For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is contains information about population structure.
+The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times S$, its rows give the contribution of each SNP to each PC, it is often used to look for outliers, which might be indicative of selection \citep[e.g][]{francois2010}.
 
 	In many  implementations \citep[e.g.][]{patterson2006}, SNPs are weighted by
-    the inverse of their standard deviation. As this weighting often makes little
+    the inverse of their standard deviations. As this weighting often makes little
     difference in practice \citep{mcvean2009}, I will assume throughout that SNPs
@@
 	\begin{align}
-2F_3(X_1; X_2, X_3) &= F_2(X_2, X_3) - F_2(X_1, X_2) - F_2(X_1, X_3)\label{eq:f3fromf2}\\
+2F_3(X_1; X_2, X_3) &=  F_2(X_1, X_2) + F_2(X_1, X_3) - F_2(X_2, X_3)\label{eq:f3fromf2}\\
 2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
@@
 Thus, 
+\begin{align}
+F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
+&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
+\end{align}
+In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error.
+If we sum up the squared approximation errors over all pairs of populations, we obtain 
 \begin{equation}
-F_2(P_i, P_j) = \underbrace{\sum_{k=1}^K(p_{ik} - p_{jk})^2}_{\hat{F_2}^{(K)}(P_i, P_j)} + \sum_{k=K+1}^n(p_{ik} - p_{jk})^2 \text{.}
+\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
 \end{equation}
-If we sum up the approximation errors $F_2 - \hat F_2$ over all pairs of populations, we obtain the Frobenius-norm of the error $\normsq{\MF - \hat{\MF}}_F$; which is precisely the function that is minimized in MDS \citep{jolliffe2013}.
-Thus, $\hat\MF^{(K)}$ is an optimal sparse approximation of $\MF$ for any $K$.
+where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements.
+This is precisely the function that is minimized in MDS \citep{jolliffe2013}.
+In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the approximation error of all $F_2$-statistics.
 
@@
 
-\paragraph{West-Eurasian data set}
+\paragraph{Western Eurasian data set}
 This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
@@
 of $F$-statistics in PCA-space. The relationships we will discuss formally only
-hold if we use all $n-1$ PCs. However, the appeal of PCA is that frequently,
+hold if we use all PCs. However, the appeal of PCA is that frequently,
 only a very small number $K \ll n$ of PCS contain most information that is
@@
 
-The $F_3$-statistic gives a more precise interpretation: we are looking for the space where $F_3$ is negative, i.e. 
+The $F_3$-statistic gives a more precise interpretation: we are looking for the space where $F_3$ is negative, i.e. the set of points $X_x$ for which 
 \begin{eqnarray}
 2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
-      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0 \label{eq:f3neg}
+      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0 \label{eq:f3neg}\text{.}
 \end{eqnarray}
 By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_1, X_2$ and $X_x$ form a right-angled triangle.
-The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_1X_2}$.
+The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_1X_2}$ (The overline notation denotes a line segment).
 $F_3$ is negative when the triangle is obtuse, i.e.
@@
 
-\paragraph{$F_3$ on a 2D-plot.} 
- If we project this $n$-ball on a two-dimensional plot, $\overline{X_1X_2}$ will usually not align with the PCs; thus the ball may be somewhat larger.
+\paragraph{$F_3$ on a 2D PCA-plot.} 
+ If we project this $n$-ball on a two-dimensional plot, $\overline{X_1X_2}$ will usually not align with the PCs; thus the ball may be somewhat larger than it appears on the plot.
 This geometry is perhaps easiest visualized on a globe.
@@
 \paragraph{Example}
-As an example, I visualize the admixture statistic $F_3(X; \text{Basque}, \text{Turkish})$, on the first two PCs of the Westeurasian data set (Figure \ref{fig:f3}A).
+As an example, I visualize the admixture statistic $F_3(X; \text{Basque}, \text{Turkish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A).
 In this case, the projected $n$-ball (light gray) and circle based on 2D (dark gray) align relatively closely, but several populations inside the ball (e.g.
@@
 
-However, many populations, particularly from the Levant and Caucasus, fall outside the circle, which allows us to immediately conclude that their $F_3$-statistics must be positive.
+However, many populations, particularly from Western Asia and the Caucasus, fall outside the circle. This allows us to immediately conclude that their $F_3$-statistics must be positive.
 
@@
 The angle between $X_x - X_1$ and $X_x - X_2$ is obtuse if $X_2$ is in the half-plane whose boundary goes through $X_x$ and is orthogonal to $\overline{X_XX_1}$ (Figure \ref{fig:geom}C), and its value is proportional to the projection onto $X_X - X_1$.
-A common interpretation guide is to use the population which yields the most negative $F_3$-statistic as the population which is most closely related to the ``true'' source of admixture \citep{patterson2012}; on a PCA-plot this corresponds to using the population that is furthest away from $X_x$ on the direction $X_x - X_1$.
+This can be useful for interpretation: if potential source pulations are related according to a tree, the admixture axis will be orthogonal to the tree branches connecting potential source populations, and the population with the most negative $F$-statistic will be the most likely source \citep{patterson2012}.
 
-This projection argument also helps to interpret Outgroup-$F_3$-statistic on a PCA-plot (Figure \ref{fig:geom}D), but in this case we aim to find the most closely related population as that with the highest $F_3$-statistic.
+On the other hand, if population relationships are more gradual, the interpretation is different: The lowest $F_3$-statistics will correspond to the populations where the $n$-sphere is biggest, i.e.
+the ends of the cline.
+This is a practical issuel for example \cite{patterson2012} observed that many Western Eurasian populations have the most negative $F_3$-statistics if one of the source population is from Europe, and one from South America, which can be thought of a gradient in relationships throughout Eurasia and the Americas.
+Depending on whether we think of population relationships as more clinal or tree-like, negative $F_3$-statistics will thus have very different interpretations.
+
+This projection argument also helps to interpret Outgroup-$F_3$-statistic on a PCA-plot (Figure \ref{fig:geom}D).
+In this context, we interpret the statistic $F_3(O; U, X_i)$ as the projection of $X_i$ onto the line segment $\overline{OU}$ from the outgroup $O$ to the test population $U$.
+The population $X_i$ with the largest $F_3$-statistic will have the largest projected segment, and is thus interpreted as the population most closely related to $U$.
 
 \paragraph{Example}
-Again, these projection will be orthogonal when using the full data, and may only be approximately orthogonal when approximated using the first two PCs.
+Again, these projections will be orthogonal when using the full data, and may only be approximately orthogonal when approximated using the first two PCs.
 In Figure \ref{fig:f3}D, I visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Mozabite}, X_i)$, i.e.
@@
 the subspace spanned by the first two PCs), this is not necessarily the case.
-We see that particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs, and the coloring based on the full $F_3$-statistics shows that in this case, the first two PCs approximate the $F_3$-statistic very well.
+The coloring is based on the $F_3$-statistic calculated from all the data.
+In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs.
 We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}E), but keep the relative ordering.
@@
 \paragraph{Example}
-To illustrate the angle interpretation I again use the Westeurasian data.
+To illustrate the angle interpretation I again use the Western Eurasian data.
 The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:f3}A), a European gradient (from Sardinian to Chuvash), and a Asian cline (from Arab to Caucasus populations).
@@
 \includegraphics[width=\textwidth]{figures/fig_f4_data.pdf}	
-	\caption{\textbf{PCA and $F_4$-statistics} A: Projection angle and correlation coefficient $r$ representation of $F_4(X, Sardinian; Saudi, Georgian)$ (red) in the Westeurasian data set, and approximations using two (green) and three (blue) PCs.
+	\caption{\textbf{PCA and $F_4$-statistics} A: Projection angle and correlation coefficient $r$ representation of $F_4(X, Sardinian; Saudi, Georgian)$ (red) in the Western Eurasian data set, and approximations using two (green) and three (blue) PCs.
 	B: Spectrum of select $F_4$-statistics in the Western Eurasian data set.
@@
 A model-based framework based on probabilistic PCA \citep{hastie2015, meisner2021, agrawal2020} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
+\newpage
+\bibliographystyle{plainnat}
+\bibliography{main}
 
+\newpage
 \appendix
@@
 
-\bibliographystyle{plainnat}
-\bibliography{main}
-
 \end{document}

### 2021-08-27 fstats_pca_lewontin.tex
revision part 1

@@
 However, understanding, conceptualizing and modeling this variation is far from trivial, particularly since misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies.
-Lewontin's landmark \citeyear{dobzhansky1972} paper on the apportionment of human genetic diversity was one of the first to question ``races'' not only on ethical, but also on scientific grounds.
-In Lewontin's data, less than half of the already small proportion of between-population genetic variation can be attributed to ``racial'' continental-scale groupings.
-Over the last five decades, this view has been corroborated and extended many times \citep{cann1987, cavalli-sforza1994, cann2002, rosenberg2002, reich2018a}, and we now know that human races are an arbitrary, biologically useless polyphyletic grouping that is maintained by historical and social conventions.
+Lewontin's landmark \citeyear{dobzhansky1972} paper on the apportionment of human genetic diversity was one of the first to question ``races'' not only on ethical grounds, but also based on genetic data.
+Based on 15 blood group loci, less than half of the already small proportion of between-population genetic variation can be attributed to ``racial'' continental-scale groupings.
+Over the last five decades, this view has been corroborated and extended many times \citep{cann1987, cavalli-sforza1994, cann2002, rosenberg2002a, reich2018a}, and we now know that human races cannot be motivated genetically, but rather are a polyphyletic grouping that is maintained by historical and social conventions.
 
@@
 
-This complex population structure is frequently handled by using multiple models with different assumptions; each emphasizing a particular aspects of the data.
+The interplay of these demographic processes leads to complex genetic structure that is difficult to model using standard population genetic models: Human populations are typically non-discrete, not at equilibrium and not randomly mating.
+One analysis paradigm is to use different tools based on different sets of assumptions; each emphasizing a particular aspects of the data.
 Data-driven methods such as Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable.
-For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests.
+For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests, but with the drawback that strong assumptions about models need to be assumed.
 
@@
 If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
-As a population may be represented by just one individual, there is no conceptual difference between an individual-based and population-based analysis.
+As a population may be represented by just one individual, all results are valid if we treat each individual as its own population.
 Since the allele
@@
 
-The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ that explains most of the variation in the data.
+The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ that explains the most important variation in the data.
 $K$ is at most $n-1$, in which case the data is simply rotated.
@@
 
-	In many  implementations \citep[e.g.][]{patterson2006}, SNPs are weighted by
-    the inverse of their standard deviations. As this weighting often makes little
-    difference in practice \citep{mcvean2009}, I will assume throughout that SNPs
-    are unweighted.
+	There are a number of choices when constructing a PCA from genetic data, for example SNPs are sometimes standardized by the inverse of their standard deviations \citep{patterson2006}, or low-quality samples are projected onto PCs created from high-quality samples \citep{patterson2006}.
+In order to simplify the relationship between PCA and $F$-statistics, I assume no standardization and no missing data.
 
@@
 \paragraph{``World'' data set}
-This data set represents global human genetic variation (638 individuals from 33 population), as used by \citep{lazaridis2014}.
-As this data set is very sparse, it may be well-approximated by an admixture graph.
+This data set is a subset of the ``World Foci'' data set of \cite{lazaridis2014}, where I removed samples which are not permitted for free reuse.
+These populations span the globe and roughly represents global human genetic variation (638 individuals from 33 population) As this data set is very sparse, with often thousands of kilometers between adjacent sampling locations, I speculate that gene flow between these populations may not be particularly common; and their structure may therefore be well-approximated by an admixture graph.
+A file with all individuals used and their assigned population is given in \textbf{Supplementary File 1.}
 
@@
 This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
-It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012, lazaridis2014}.
-Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus may not be particularly graph-like.
+It is motivated by the analysis of \cite{lazaridis2014}, who used it as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}.
+Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}.
+We thus might think that gene flow between these populations is common \citep{ralph2013}, and a discrete model such as a tree or an admixture graph might be a rather poor reflection of this data.
+A file with all individuals used and their assigned population is given in \textbf{Supplementary File 2.}
 
@@
 I use \texttt{admixtools 2.0.0} (\url{https://github.com/uqrmaie1/admixtools}) to compute $F$-statistics.
-To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, obtain a nearby negative semidefinite matrix using the \texttt{nearPD} function, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.
+To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.
+A squared Euclidean distance matrix such as $\MF$ has all negative or zero eigenvalues (i.e.
+$\MF$ is negative-semidefinite).
+However as $F_2$-statistics are estimates, some eigenvalues might be slightly positive, which would lead to imaginary PCs.
+I avoid this by using the \texttt{nearPD}-function in R that ensures all eigenvalues have the correct sign.
 
@@
 The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct.
-As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection (Figure \ref{fig:geom}E).
+As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection, multiplied by the distances $\norm{X_1 - X_2}$ and $ \norm{X_3 - X_4}$ (Figure \ref{fig:geom}E).
 
@@
 As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, (e.g using a lower number of populations).
-It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are explained solely by higher PCs, and so whenever such contradictions arise, and so higher PCs will be informative for population structure.
+It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are explained solely by higher PCs, and so whenever such contradictions arise, higher PCs will be informative for population structure.
 Previous interpretation of PCA in the context of population genetic models have primarily focused on the PCs, which can be derived analytically for trees \citep{cavalli-sforza1975} and homogeneous spatial models \citep{novembre2008a}.
@@
 A model-based framework based on probabilistic PCA \citep{hastie2015, meisner2021, agrawal2020} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
+
 \newpage
@@
 \begin{eqnarray}
-F_2(X_i, X_j) &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
-&=& \sum_{l=1}^L \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{kj}\big)^2\nonumber\\
-&=& \sum_{l=1}^L \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
-&=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
-&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
+F_2(X_i, X_j) &=& \sum_{l=1}^S \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
+&=& \sum_{l=1}^S \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{jk}\big)^2\nonumber\\
+&=& \sum_{l=1}^S \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
+&=& \sum_{l=1}^S \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
+&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^S L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
 &=& \sum_k (P_{ik} - P_{jk})^2 \label{eq:changeofbasis}

### 2021-08-27 fstats_pca_lewontin.tex
revision part 1

@@
 In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error.
+I will omit the superscript when the exact number of PCs is not relevant.
 If we sum up the squared approximation errors over all pairs of populations, we obtain 
@@
 populations have high $F_2$ but appear on the same point on an PCA-plot, this suggests
-that substantial variation is hidden on higher PCs.
+that substantial variation is hidden on higher PCs, and we may want to investigate the PCs which contribute a large distance to this particular $F_2$-statistic in order to better understand and visualize their relationship.
 

### 2021-09-06 fstats_pca_lewontin.tex
revisions

@@
 \section{Introduction}
-Most genetic variation in humans is shared between all of us, but around 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002a}.
-Genome-scale data has allowed us to leverage the information contained in these 15\% to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}.
-For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
+As in other species, the genetic diversity of human populations  is shaped by their environment and history over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994, schraiber2015}.
+In turn, studying the diversity patterns in humans can be used to reconstruct the demographic and evolutionary history of our species.
+As such, population genetic theory is essential to understand how this works.
 
-However, understanding, conceptualizing and modeling this variation is far from trivial, particularly since misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies.
-Lewontin's landmark \citeyear{dobzhansky1972} paper on the apportionment of human genetic diversity was one of the first to question ``races'' not only on ethical grounds, but also based on genetic data.
-Based on 15 blood group loci, less than half of the already small proportion of between-population genetic variation can be attributed to ``racial'' continental-scale groupings.
-Over the last five decades, this view has been corroborated and extended many times \citep{cann1987, cavalli-sforza1994, cann2002, rosenberg2002a, reich2018a}, and we now know that human races cannot be motivated genetically, but rather are a polyphyletic grouping that is maintained by historical and social conventions.
+\paragraph{processes}
+Over time, genetic drift will slowly change allele frequencies and as a result, isolated populations are expected to slowly diverge.
+In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance.
+However, isolation-by-distance patterns are not uniform, but shaped by geography, particularly barriers   to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}.
+In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}.
+Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity.
 
 Modern descriptions of human genetic diversity focus on the evolutionary processes that caused it, and which gives rise to  both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}.
-In isolation, populations are expected to slowly diverge, particularly if they are separated by barriers to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}.
-On the other hand, major population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}.
-Local migration between neighboring populations will flatten differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and modern humans \citep{green2010},will lead to locally increased diversity.
+The interplay of these demographic processes leads to the complex genetic structure that we observe in present-day people, and which we aim to reconstruct from genetic data, which is challenging since we cannot expect to devise a single model that captures all complexity.
 
-The interplay of these demographic processes leads to complex genetic structure that is difficult to model using standard population genetic models: Human populations are typically non-discrete, not at equilibrium and not randomly mating.
-One analysis paradigm is to use different tools based on different sets of assumptions; each emphasizing a particular aspects of the data.
-Data-driven methods such as Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable.
-For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests, but with the drawback that strong assumptions about models need to be assumed.
+\paragraph{Data-driven approaches}
+One commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions; each emphasizing a particular aspects of the data.
+Particularly for large data sets with several thousand samples, exploratory, data-driven methods such as population trees, Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are a first step aimed at providing a simple, low-dimensional summary of the data while making minimal assumptions about the underlying population structure.
 
-Particularly in the analysis of human ancient DNA, a set of techniques based on $F$-statistics \textit{sensu} Patterson have risen in popularity \citep{patterson2012, peter2016}.
-This framework is based on the null-assumption that the relationship between  populations is tree-like, and then uses tests involving three or four populations to probe for violations of treeness \citep{patterson2012}.
-In a further step, $F$-statistics of multiple populations can be combined to estimate parameters in more complex models \citep{patterson2012, harney2021}.
-However, the connections between PCA, $F$-statistics and explicit demographic models are often unclear, which makes quantitative comparisons, detecting model violations and joint interpretation of these approaches difficult.
-Since both $F$-statistics and PCA are functions of expected pairwise coalescent times \citep{mcvean2009, peter2016}, this is one avenue to link these approaches.
-Here, I instead use the geometric interpretation of $F$-statistics developed by  \cite{oteo-garcia2021} to directly visualize $F$-statistics on a PCA plot.
+\paragraph{Model-based appraoches}
+While PCA and related tools give a good overview of the data, they are not designed to answer specific research questions and cannot be used for model testing or parameter estimates.
+For this purpose, methods based on explicit demographic models  are often used \citep{gutenkunst2009, kamm2015, excoffier2013}.
+The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modelling assumptions such as that populations are discrete, randomly mating, and at equilibrium, with the hope that despite these assumptions being violated, the resulting model fits provide a sufficiently accurate description to answer the research questions.
+
+\paragraph{F-stats}
+However, when the number of populations exceeds a few dozen,  even codifying reasonable population models including all data can be prohibitively difficult.
+One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset.
+However, this has the drawback that a large proportion of the data may be unused.
+An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.
+
+The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}.
+I safe the formal definition for later; but the easiest way to motivate them assumes that populations are related as a tree, where the edges measure how much genetic drift has occurred.
+In this case, $F_2$ measures the genetic distance between populations, $F_3$ corresponds to an external, and $F_4$ to an internal branch of a tree ) \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.
+
+In most applications, these $F$-statistics are estimated from some data, and then used as tests of treeness.
+The canonical alternative model is an admixture graph or phylogenetic network \citep{patterson2012}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A).
+Typically, analyses proceed by testing specific hypotheses abot gene flow using $F$-statistics \citep{reich2009, green2010, lazaridis2014}.
+Increasingly commonly, $F$-statistics are also incorporated into tools that jointly fit more complicated models \citep[e.g.
+][]{patterson2012,harney2021}.
+However, admixture graphs are not the only alternative model that explain violations of treeness.
+Expected $F$-statistics can be calculated for a wide range of demographic models \citep{peter2016}, or even in abstract Euclidean spaces \citep{oteo-garcia2021}, so that model fit remains a major concern.
+
+\paragraph{PCA}
+With minimal assumptions, PCA is one of the most widely used data summary and data exploration techniques.
+One way to think about PCA is that it aims to create a low-dimensional summary of the data that retains a maximum of information about the underlying structure in the data (Again, a formal description will be given in the next section).
+It has been widely used and independently discovered in many disciplines.
+In population genetics, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize global genetic diversity in the pre-genomic era \citep{cavalli-sforza1994}.
+Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale population genetic data set.
+This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g.
+remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses.
+Theoretically, the PCA-decomposition has been studied in terms of trees \citep{cavalli-sforza1975}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}.
+
+\paragraph{F-stats and PCA}
+However, the connections between PCA and $F$-statistics remains poorly understood, and the goal of this paper is to investigate this connection.
+To do so, I use the geometric interpretation of $F$-statistics developed by \cite{oteo-garcia2021} who defined them in arbitrary Euclidean spaces.
+We make use of the empirical finding that while the allele-frequency space is frequently high-dimensional, for many empirical data sets, a low number of dimensions are sufficient to characterize population structure, and hence $F$-statistics.
+This allows for approximate interpretations of $F$-statistics on PCA plots, and I use two sample data sets to evaluate these approximations in practice.
 
 \section{Theory}
-In this section, I will give a very brief formal introduction to $F$-statistics and PCA.
-A more detailed technical treatise of PCA is given by e.g.
+In this section, I will introduce the mathematics and notations for $F$-statistics and PCA.
+A comprehensive  treatise on PCA is given by e.g.
 \cite{jolliffe2013} and \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
@@
 
-\subsection{Introduction to PCA}
+\subsection{Formal Definition of PCA}
 Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype.
@@
 
-\subsection{Introduction to $F$-statistics}
+\subsection{Formal Definition to $F$-statistics}
 PCA is typically used to model population structure between many populations.
@@
 \begin{figure}[!ht]
-	\includegraphics[width=\textwidth]{figures/fstats_on_pca.pdf}
-	\caption{\textbf{Geometric representation of $F$-statistics on 2D-PCA-plot.} A: $F_2$ represents the squared Euclidean distance between two points in PC-space.
-B: Admixture-$F_3(X_x; X_1, X_2)$ is negative if $X_x$ lies in the circle specified by the diameter $X_2-X_1$.
-C: $F_3(X_x; X_1, X_2)$ is negative given $X_1, X_x$ if $X_2$ is in the gray space.
-D: Outgroup-$F_3$ reflects the projection of $X_2 - X_O$ on $X_1 - X_O$.
-E: $F_4$ is the projection of $X_3 - X_4$ on $X_1-X_2$.
-F: Same projection, but assuming data is generated by a tree.}
+	\includegraphics[width=\textwidth]{figures/fstats_pca_vs_tree.pdf}
+	\caption{\textbf{Representation of $F$-statistics on tree and 2D-PCA-plot.} The schematics show four populations and their representation using a tree (top row) or a 2D-PCA plot.
+A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space.
+B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthoginal projection of $X_1 - X_3$ onto $X_1$-$X_4$.
+C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or the orthogonal projection of $X_2 - X_3$ on $X_1 - X4$.
+D: $F_4(X_1, X_2; X_3, X_4)$ The two paths from $X_1$ to $X_2$ and $X_3$ and $X_4$ are non-overlapping in the tree, which corresponds to orthogonal vectors in PCA-space.}
 	\label{fig:geom}
+\end{figure}
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fstats_admixture_pca.pdf}
+	\caption{\textbf{Admixture representation on 2D-PCA-plot.} The schematics show four populations and their representation using an admixture graph (A) or a 2D-PCA plot.
+A: Admixture graph, with population $X_y$ originating as an admixture of $X_2$ and $X_3$, with $X_2$ contributing proportion $\alpha$.
+Subsequent drift (red branch) will change allele frequency to sampled admixture population $X_x$.
+B: PCA representation of the scenario in A.
+$X_y$ originates on the segment connecting $X_2$ and $X_3$, and subsequent drift may move it in a random direction.
+C: $F_3(X_x; X_2, X_3)$ and negative region (light gray circle).
+$F_4(X_1, X_x; X_3, X_4)$ will no longer be zero (compare to Figure \ref{fig:geom}D).
+}
+	\label{fig:admix}
 \end{figure}

### 2021-09-08 fstats_pca_lewontin.tex
revision figures

@@
 \newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
+\newcommand{\vectorproj}[2][]{\textit{proj}_{#1}#2}
+
 \title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
@@
 
+\subsection{Formal Definition to $F$-statistics}
+Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ loci.
+Let $x_{il}$ denote the frequency at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$.
+As $X_i$ will be the only data summary for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.
+
+The three $F$-statistics can then be defined as
+\begin{subequations}
+	\begin{align}
+	F_2(X_1, X_2) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
+	\\
+	F_3(X_1; X_2, X_3) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) \\	
+	F_4(X_1, X_2; X_3, X_4) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) 	\text{,}
+	\end{align}
+\end{subequations}
+The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
+\begin{subequations}
+	\begin{align}
+	2F_3(X_1; X_2, X_3) &=  F_2(X_1, X_2) + F_2(X_1, X_3) - F_2(X_2, X_3)\label{eq:f3fromf2}\\
+	2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
+	\end{align}
+\end{subequations}
+
+As highlighted in the introduction, $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
+In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of all branches between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch between four nodes, respectively (Figure \ref{fig:geom}C).
+Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero.
+The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}.
+
+Thinking of $F$-statistics as branch lengths is useful to understand a number of applications: In particular, one common task is to find the population most closely related to an unknown sample $X_U$ \citep{raghavan2014}.
+One way to do that is using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ denotes an outgroup, and the $X_i$ are a panel of populations that are candidates for the closest match.
+The  highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times.
+The intuition is given in Figure \ref{fig:outgroupf3}A; where the outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ is highlighted.
+It represents the length of the branch from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the statistic.
+In contrast to a simple genetic distance, the sample time has no effect: The branch between $X_U$ and $X_2$, would be shorter than to one between $X_U$ and $X_1$, but the path to the shared junction and hence the $F_3$-statistic would be the same.
+Larger sets of $F_3$ and $F_4$-statistics are also frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.
+
+Most commonly however, $F_3$ and $F_4$ are used as tests of treeness \citep{patterson2012}: Negative $F_3$-values correspond to a branch with negative genetic drift, which is a violation of treeness.
+Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}.
+The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A, where (the typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$.
+Over time, genetic drift will change $X_y$ to $X_x$, which is the population we observe.
+This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact equations can be found in ][]{peter2016}.
+While admixture graphs are commonly considered, other alternative models exist.
+In \citep{peter2016}, I showed that $F_4$ is sensitive to many non-symmetric population structure models.
+$F_3$ is more robust, but may also be negative if substantial population substructure exists.
+
+\subsubsection{Geometric interpretation of $F$-statistics}
+For very large data sets, it is hard to justify a specific demographic models that includes all data, but we will typically still be able to explore such data e.g.
+using PCA.
+One way to justify $F$-statistics is to interpret the populations $X_i$ geometrically as points or vectors in the $S$-dimensionsl \emph{data space} $\mathbb{R}^S$.
+\cite{oteo-garcia2021} showed that in this case, the $F$-statistics can be thought of as inner (or dot) products, and that all properties and tests related to treeness can be derived from this larger space.
+In particular the $F$-statistics can be defined as
+\begin{subequations}
+	\begin{align}
+	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
+	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
+	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
+	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
+	\end{align}
+\end{subequations}
+where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product.
+Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
+\begin{subequations}
+	\begin{align}
+	\langle a, b \rangle &= \sum_i a_ib_i\\
+	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
+	\langle a, a \rangle &= \normsq{a}\\
+	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
+	\end{align}
+\end{subequations}
+where $\phi$ is the angle between $a$ and $b$. \cite{oteo-garcia2021} showed that in this framework, $F$-statistics are closely related to vector projections
+\begin{equation}
+\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
+\end{equation}
+which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$.
+
+The drawback of the geometric approach of \cite{oteo-garcia2021} is that we have to deal with an very high-dimensional space, as the number of SNPs is frequently in the millions.
+However, it has been commonly observed that population structure is quite low-dimensional, with the first few PCs providing a good approximation of the underlying population structure.
+Therefore, we may hope that PCA could yield a reasonable approximation of the data space.
+
+\begin{figure}[!ht]
+	\includegraphics[width=.5\textwidth]{figures/outgroup_f3.pdf}
+	\caption{\textbf{Outgroup-$F_3$-statistics}}
+	\label{fig:outgroupf3}
+\end{figure}
+
 \subsection{Formal Definition of PCA}
-Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype.
+PCA is a common way of summarizing genetic data, and so a large number of variation of PCA exist, e.g.
+in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis.
+The version of PCA I describe here is set up in a way that the similarities to the $F$-statistics-framework are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets.
+In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}.
+The differences this causes will be addressed in the discussion.
+
+Let us again assume we have genotype data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype.
 If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
-As a population may be represented by just one individual, all results are valid if we treat each individual as its own population.
-Since the allele
-frequencies are between zero and one, we can interpret each Population $X_i$
-of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.
+Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
+of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.
 
-The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ that explains the most important variation in the data.
+One way one can define PCA is that it aims to find a $K$-dimensional subspace of the data space that retains the most important variation.
 $K$ is at most $n-1$, in which case the data is simply rotated.
-However, the  historical processes that generated genetic variation often result in \emph{sparse} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.
+However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.
 
@@
 	\begin{equation}
-	\MY = \MC\MX = (\mathbf{U} \MSINGULAR) \mathbf{V}^T = \MP\ML\text{,}
+	\MY = \MC\MX = (\mathbf{U} \MSINGULAR) \mathbf{V}^T = \MP\ML\text{,} \label{eq:svd}
 	\end{equation}
@@
 	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively.
-For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is contains information about population structure.
-The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times S$, its rows give the contribution of each SNP to each PC, it is often used to look for outliers, which might be indicative of selection \citep[e.g][]{francois2010}.
+For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is contains information about population structure.
+The SNP loadings $\ML=\mathbf{V}^T$ form an orthonormal basis of size $n \times S$, its rows give the contribution of each SNP to each PC.
+It is often used to look for outliers, which might be indicative of selection \citep[e.g][]{duforet-frebourg2016}.
+Alternatively, the PCs can also be motivated as an eigendecomposition of the covariance matrix $\MY\MY^T$.
+This can be motivated from (\ref{eq:svd}):
 
-	There are a number of choices when constructing a PCA from genetic data, for example SNPs are sometimes standardized by the inverse of their standard deviations \citep{patterson2006}, or low-quality samples are projected onto PCs created from high-quality samples \citep{patterson2006}.
-In order to simplify the relationship between PCA and $F$-statistics, I assume no standardization and no missing data.
-
-\subsection{Formal Definition to $F$-statistics}
-PCA is typically used to model population structure between many populations.
-$F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time.
-The three $F$-statistics can be defined as
-\begin{subequations}
-	\begin{align}
-	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
-	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
-	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
-	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
-	\end{align}
-\end{subequations}
-where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product.
-The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently.
-Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
-\begin{subequations}
-\begin{align}
-\langle a, b \rangle &= \sum_i a_ib_i\\
-\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
-\langle a, a \rangle &= \normsq{a}\\
-\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
-\end{align}
-\end{subequations}
-where $\phi$ is the angle between $a$ and $b$.
-
- Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
-\begin{subequations}
-	\begin{align}
-2F_3(X_1; X_2, X_3) &=  F_2(X_1, X_2) + F_2(X_1, X_3) - F_2(X_2, X_3)\label{eq:f3fromf2}\\
-2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
-	\end{align}
-\end{subequations}
-
-$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
-In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of all branches between populations $X_1$ and $X_2$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}.
-The length is a measure of genetic drift, and is non-negative if data is generated under a tree \citep{patterson2012}.
-This interpretation is useful to understand a number of applications.
-The outgroup-$F_3$-statistic $F_3(O; U, X_i)$, for example, is useful if we have an unknown population $U$, and want to find its closest relatives from a panel of populations $X_i$.
-The highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times.
-The population $X_i$ with the largest value is the most closely related population out of the reference sample.
-The internal branches described by  $F_4$-statistics are frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.
-
-Most commonly however, $F_3$ and $F_4$ are used as admixture tests \citep{patterson2012}: Negative values of  $F_3(X_1; X_2, X_3) < 0$ correspond to a branch with negative genetic drift, which is a violation of treeness.
-Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}.
-
-To move away from trees and graph models, I build upon the geometric framework of \cite{oteo-garcia2021}.
-Here, we think of each population as a point in the data space $\mathbb{R}^S$, made up of the allele frequency at each SNP.
-Then, $F_2(X_1, X_2) = \normsq{X_1 - X_2}$ is the squared Euclidean distance between two populations $X_1$ and $X_2$, and $F_4(X_1, X_2 ; X_3, X_4) = \langle X_1 - X_2, X_3 - X_4 \rangle$ is the inner (dot) product between these two vectors.
-These dot products are useful for a variety of projections that use population structure.
+	\begin{equation}
+	\MY\MY^T = \MP\ML\ML^T\MP^T = \MP\MP^T,
+	\end{equation}
+	since $\ML\ML^T=\mathbf{I}$.
 
@@
 This is precisely the function that is minimized in MDS \citep{jolliffe2013}.
-In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the approximation error of all $F_2$-statistics.
+In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the sum of approximation errors of all $F_2$-statistics.
+
+\subsubsection{$F$-statistics and projection on PCA}
+One of the easiest ways of dealing with missing data in PCA is to calculate the principal components (equation \ref{eq:svd}) only on a subset of the data with no missingness, and then to \emph{project} the lower quality samples with high missingness onto this PCA.
+The simplest way to do this is to note that
+
+\begin{equation*}
+\MY\ML^T = \MP\ML\ML^T = \MP,
+\end{equation*}
+and so a new (centered) population $Y_{\text{new}}$ can be projected onto an existing PC simply by post-multiplying it with $\ML^T$: $$P_{\text{proj}} = Y_{\text{new}}\ML^T;$$ the $k$-th entry of $P_{\text{proj}}$ gives the coordinates of the new sample on the $k$-th PC.
+However, it is likely that $Y_{\text{new}}$ lies outside the variation of the original sample.
+In this case, there is a projection error
+$$\normsq{Y_{\text{new}} - P_{\text{proj}}\ML }.$$ If we project with missing data, a similar projection can be used where we remove the rows from $Y_{\text{new}}$ and $\ML$ where data in $Y_{\text{new}}$ is missing \citep{patterson2006}.
+
+%Equivalently, if we have $F_2$-distances from the 
+%$$p_{\text{new}} = \frac{1}{2} \MSINGULAR^{-2}P^T d,$$
+%where $d_i = \normsq{X_i} - F_2(X_i, X_{\text{new}})$ \citep{gower1968}. In this case,
+
+Thus, if we compare the $F$-statistic of a projected sample, we have
+
+\begin{equation}
+F_2(X_i, X_{\text{new}}) = F_2(Y_i, Y_{\text{new}}) = F_2(P_i, P_{\text{proj}}) + F_2(P_{\text{proj}}\ML, Y_{\text{new}}),
+\end{equation}
+because the projection error and projection are orthogonal to each other.
 
@@
 \subsection{When are admixture-$F_3$ statistics negative?}
-Given two source populations $X_1$, $X_2$, can we predict which populations $X_x$ could be considered admixed between these populations based on PCA?
-Since the allele frequencies of $X_x$ are intermediate between those of $X_1$ and $X_2$, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \citep{brisbin2012, mcvean2009}.
+Consider the admixture scenario in Figure \ref{fig:admix}A, where population $X_y$ is the result of a mixture of $X_2$ and $X_3$, and subsequent drift changes the allele frequencies of the admixed population from $X_y$ to the sampled population $X_x$.
+How is such a scenario displayed on a PCA?
+ince the allele frequencies of $X_y$ are a linear combination of $X_2$ and $X_3$, it will lie on the line segment connecting these two populations (Figure \ref{fig:admix}B).
+In fact, if $X_x$ (and other populations related to $X_x$) was not part of the construction of the PCA, and is instead projected onto it, the drift from $X_y$ to $X_x$ will be independent of the population on the PCA, and $X_x$ will be at the same location on a PCA plot as $X_y$, which can be used to predict the admixture proportions \citep{brisbin2012, mcvean2009, oteo-garcia2021}.
 
-The $F_3$-statistic gives a more precise interpretation: we are looking for the space where $F_3$ is negative, i.e. the set of points $X_x$ for which 
+If $X_x$, or populations more closely related to $X_x$ than $X_2$ and $X_3$, are included in the construction of the PCA, this is no longer true, and $X_x$ will project on a different spot than $X_y$ (Figure \ref{fig:admix}B).
+Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might be considered admixed between them?
+
+One way to address this question is to consider the space for which $F_3$ is negative, i.e. 
 \begin{eqnarray}
-2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
-      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0 \label{eq:f3neg}\text{.}
+2 F_3(X_x; X_2, X_3) &=& 2\langle  X_x - X_2, X_x - X_3 \rangle \nonumber\\
+      &=& \normsq{X_x - X_2} + \normsq{X_x - X_3}  - \normsq{X_2 - X_3} < 0 \label{eq:f3neg}\text{.}
 \end{eqnarray}
-By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_1, X_2$ and $X_x$ form a right-angled triangle.
-The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_1X_2}$ (The overline notation denotes a line segment).
+By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_2, X_3$ and $X_x$ form a right-angled triangle.
+The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_2X_3}$ (The overline denotes a line segment).
 $F_3$ is negative when the triangle is obtuse, i.e.
-$X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_1X_2}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).
+$X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_2X_3}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).
 
 \paragraph{$F_3$ on a 2D PCA-plot.} 
- If we project this $n$-ball on a two-dimensional plot, $\overline{X_1X_2}$ will usually not align with the PCs; thus the ball may be somewhat larger than it appears on the plot.
+ If we project this $n$-ball on a two-dimensional plot, $\overline{X_2X_3}$ will usually not align with the PCs; thus the ball may be somewhat larger than it appears on the plot.
 This geometry is perhaps easiest visualized on a globe.
@@
 This reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure.
-This is expected because for spatially continuous populations, PCA will not be sparse \citep{novembre2008a}.
 Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
 
-However, many populations, particularly from Western Asia and the Caucasus, fall outside the circle. This allows us to immediately conclude that their $F_3$-statistics must be positive.
+However, many populations, particularly from Western Asia and the Caucasus, fall outside the circle.
+This allows us to immediately conclude that their $F_3$-statistics must be positive; and we should not consider them as a mixture between Basques and Turks.
 
@@
 
-\subsection{$F$-statistics as projections}
-The inner product $\langle X_x - X_1, X_x - X_2 \rangle$ is closely related to the projection of a vector onto another one.
-This interpretation is useful either when calculating an outgroup-$F_3$-statistic, or when we want to find the ``best'' admixture source $X_2$ if we assume the admixed population $X_x$ and one source $X_1$ are known.
-The angle between $X_x - X_1$ and $X_x - X_2$ is obtuse if $X_2$ is in the half-plane whose boundary goes through $X_x$ and is orthogonal to $\overline{X_XX_1}$ (Figure \ref{fig:geom}C), and its value is proportional to the projection onto $X_X - X_1$.
+\subsection{$F_3$-statistics as projections}
+Outgorup-$F_3$-statistic motivate a comparison of $F_3$-statistics to projections (Equation \ref{eq:proj}), consider again the case displayed in Figure \ref{fig:outgroupf3}A, where the goal is to find the population $X_i$ that is closest to an unknown population $X_U$, with respect to an outgroup $X_O$, which we can do using the statistic $F_3(X_O; X_U, X_i)$.
+On a PCA-plot, we can visualize this $F_3$-statistic as the projection of the vector $X_i - X_O$ onto $X_U - X_O$:
+$$\vectorproj[X_U - X_O]{X_i - X_O} = \frac{F_3(X_O; X_U, X_i)}{F_2(X_O; X_U)}(X_U - X_O).$$
+
+Of the right-hand-side terms, $X_U-X_O$ gives the direction of the resulting vector, and the $F_2$-term in the denominator is a normalizing constant.
+Neiter of these terms depend on $X_i$, which justifies the argument that the $F_3$-statistic and length of the projected vector are proportional to each other, and can thus be interpreted similarly.
+Thus, the outgroup-$F_3$-statistic is larger for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in the example in Figure \ref{fig:outgroupf3}B this is $X_3$.
+
+%geometric interpretations of which are, for $F_3$ and $F_4$ given in Figure \ref{fig:geom}F-H.
+The term in parentheses gives the direction of the resulting vectors, and the the numerator of the right-hand side is the $F$-statistic.
+The denominator is the distance between the last two arguments, in many applications we compare different $F$-statistics with repeated populations, and so it is useful to think of $F$-statistics as representing (i.e.
+being proportional to) projections.
+This yields a simple representation of $F$-statistics on PCA plots that again, is exact when the first two PCs explain all relevant population structure, and approximate otherwise.
+In particular $F_3(X_1; X_3, X_4)$ can be thought of as the projection of $X_3 - X_1$ onto $X_4 - X_1$ (Figure \ref{fig:geom}F), or how far, along the axis from $X_1$ to $X_4$ would $X_3$ project.
+Similarly, $F_4(X_1, X_4; X_2, X_3)$ corresponds to the projection of $X_2 - X_3$ onto $X_1 - X_4$, and corresponds to the shorter line segment in Figure \ref{fig:geom}G.
+For a different configuration of the arguments, $F_4(X_1, X_2; X_3, X_4) $, we see that $X_1$ and $X_2$ project to the same spot on $X_3 - X_4$, which means that the two arguments are orthogonal Figure \ref{fig:geom}H, precisely what we want to test using an $F_4$-test \citep{patterson2012}.
+
+ %The angle between $X_x - X_1$ and $X_x - X_2$ is obtuse if $X_2$ is in the half-plane whose boundary goes through $X_x$ and is orthogonal to $\overline{X_XX_1}$ (Figure \ref{fig:geom}C), and its value is proportional to the projection onto $X_X - X_1$.
 This can be useful for interpretation: if potential source pulations are related according to a tree, the admixture axis will be orthogonal to the tree branches connecting potential source populations, and the population with the most negative $F$-statistic will be the most likely source \citep{patterson2012}.
 
-On the other hand, if population relationships are more gradual, the interpretation is different: The lowest $F_3$-statistics will correspond to the populations where the $n$-sphere is biggest, i.e.
+%On the other hand, if population relationships are more gradual, the interpretation is different: The lowest $F_3$-statistics will correspond to the populations where the $n$-sphere is biggest, i.e.
 the ends of the cline.
@@
 
-This projection argument also helps to interpret Outgroup-$F_3$-statistic on a PCA-plot (Figure \ref{fig:geom}D).
-In this context, we interpret the statistic $F_3(O; U, X_i)$ as the projection of $X_i$ onto the line segment $\overline{OU}$ from the outgroup $O$ to the test population $U$.
-The population $X_i$ with the largest $F_3$-statistic will have the largest projected segment, and is thus interpreted as the population most closely related to $U$.
-
 \paragraph{Example}
-Again, these projections will be orthogonal when using the full data, and may only be approximately orthogonal when approximated using the first two PCs.
+Again, these projections will be orthogonal when using the full data, and may only be approximately orthogonal when approximated using the first few PCs.
 In Figure \ref{fig:f3}D, I visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Mozabite}, X_i)$, i.e.
@@
 the subspace spanned by the first two PCs), this is not necessarily the case.
-The coloring is based on the $F_3$-statistic calculated from all the data.
-In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs.
+The coloring is based on the $F_3$-statistic calculated from all the data, with brighter values indicating higher $F_3$-statistics.
+In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.
 We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}E), but keep the relative ordering.
-This $F_3$-statistic is also very sparse, with e.g.
-PCs 3-5, 7 and 10 having almost zero contribution to all statistics (Figure \ref{fig:f3}F), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.
+I also find that many PCs, e.g.
+PCs 3-5, 7 and 10 have almost zero contribution to all $F_3$-statistics (Figure \ref{fig:f3}F), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.
 
@@
 The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct.
-As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection, multiplied by the distances $\norm{X_1 - X_2}$ and $ \norm{X_3 - X_4}$ (Figure \ref{fig:geom}E).
+$F_4$-statistics that correspond to a branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot (Figure \ref{fig:geom}G), again with the caveat that we need to scale it by a constant.
+If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, or that the two populations map to the same point (Figure \ref{fig:geom}H).
+In the case of an admixture graph, this is no longer the case: Population $X_x$ in Figure \ref{fig:admix}D does \emph{not} map to the same point as $X_1$ or $X_2$ do, implying that statistics of the form $F_4(X_1, X_x; X_3, X_4) \neq 0$.
 
-We can also see how this interpretation aligns with that of $F_4$ as the length of an internal branch on a tree : By assumption, disjoint sets of branches evolve independently \citep{cavalli-sforza1964, felsenstein1973, semple2003}.
+%We can also see how this interpretation aligns with that of $F_4$ as the length of an internal branch on a tree : By assumption, disjoint sets of branches evolve independently \citep{cavalli-sforza1964, felsenstein1973, semple2003}.
 Since the data space is sufficiently high-dimensional, this ensures that the resulting drift trajectories will also be uncorrelated.
 Therefore, if we interpret $F_4(X_1, X_2; X_3, X_4)$ as the projection of $X_3 - X_4$ - onto $X_1 - X_2$, we can write
-$$X_3 - X_4  = (X_3 - X'_3) + (X'_3 - X'_4) + (X'_4 - X_4)\text{.}$$ Of these three branches, the first and last are orthogonal to $X_1 - X_2$ and thus the $F_4$ statistic is just the internal branch of the tree (Figure \ref{fig:geom}F).
+%$$X_3 - X_4  = (X_3 - X'_3) + (X'_3 - X'_4) + (X'_4 - X_4)\text{.}$$ Of these three branches, the first and last are orthogonal to $X_1 - X_2$ and thus the $F_4$ statistic is just the internal branch of the tree (Figure \ref{fig:geom}F).
 It also suggests a number of diagnostic $F$-statistics that check assumptions; for example if the tree holds, then

### 2021-09-08 data/f4_fig.R

@@ -209,7 +209,7 @@ ids = c(idx, idy, idz)
 
 }
 
-if(T){
+if(F){
 f2s = admixtools::read_f2("f4ratio"); sample_list= rownames(f2s)[!rownames(f2s) %in% "Denisova.DG"]
 f2s = admixtools::read_f2("f4ratio", pops=sample_list )
 pcmat = pca_from_f2s(f2s)

### 2021-09-08 data/prep_data_pca_f4ratio.R

@@ -3,12 +3,12 @@ require(tidyverse)
 require(glue)
 
 
-data_loc = 'subdata/f4ratio'
+data_loc = 'subdata/ancient'
 inds = read_table2(glue("{data_loc}.ind"), col_names=c("iid","sex", "pop"))
 
 
 
-f2loc1 = 'f4ratio'
+f2loc1 = 'ancient'
 
 #afs = admixtools::extract_afs(data_loc, pops=sample_list, 
 #                              outdir=glue("{f2loc1}_afs"),format="eigenstrat", 

### 2021-10-15 fstats_pca_lewontin.tex
revision text updates

@@
 \section{Introduction}
-As in other species, the genetic diversity of human populations  is shaped by their environment and history over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994, schraiber2015}.
-In turn, studying the diversity patterns in humans can be used to reconstruct the demographic and evolutionary history of our species.
-As such, population genetic theory is essential to understand how this works.
+As most species, the genetic diversity of human populations is influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, nielsen2020}.
+Population genetic models use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.
 
-\paragraph{processes}
-Over time, genetic drift will slowly change allele frequencies and as a result, isolated populations are expected to slowly diverge.
-In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance.
-However, isolation-by-distance patterns are not uniform, but shaped by geography, particularly barriers   to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}.
-In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}.
-Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity.
+In particular, in isolated populations genetic drift will slowly change allele frequencies.
+As a result, isolated populations are expected to slowly differentiate \citep{wahlund1928,cavalli-sforza1975}.
+In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985a}.
+However, isolation-by-distance patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}.
+In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}.
+Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016} and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2019}.
 
-Modern descriptions of human genetic diversity focus on the evolutionary processes that caused it, and which gives rise to  both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}.
-The interplay of these demographic processes leads to the complex genetic structure that we observe in present-day people, and which we aim to reconstruct from genetic data, which is challenging since we cannot expect to devise a single model that captures all complexity.
+The interplay of these demographic processes leads to the complex genetic structure observed in present-day human populations \citep{the2015, mallick2016} with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a} that we model in order to reconstruct human demographic history.
+This is challenging because particularly for large and heterogeneous data sets we cannot expect to devise a single model that captures all processes.
+A commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions.
+each emphasizing particular aspects of the data.
 
-\paragraph{Data-driven approaches}
-One commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions; each emphasizing a particular aspects of the data.
-Particularly for large data sets with several thousand samples, exploratory, data-driven methods such as population trees, Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are a first step aimed at providing a simple, low-dimensional summary of the data while making minimal assumptions about the underlying population structure.
-
-\paragraph{Model-based appraoches}
-While PCA and related tools give a good overview of the data, they are not designed to answer specific research questions and cannot be used for model testing or parameter estimates.
-For this purpose, methods based on explicit demographic models  are often used \citep{gutenkunst2009, kamm2015, excoffier2013}.
-The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modelling assumptions such as that populations are discrete, randomly mating, and at equilibrium, with the hope that despite these assumptions being violated, the resulting model fits provide a sufficiently accurate description to answer the research questions.
+A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}.
+Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}).
+However, these methods are not designed to answer specific research questions, and are limited in their ability to estimate biologically meaningful parameters.
+For this purpose, methods based on explicit demographic models are often used that aim to fit a specified or estimated model of divergence, migration and genetic drift to the data \citep{gutenkunst2009, excoffier2013, kamm2015}.
+The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modeling assumptions such as that populations are discrete, randomly mating, or at equilibrium.
+While in most cases these assumptions are violated to some extent and cannot be verified, but we hope that the resulting model fits provide a sufficiently accurate answer for specific research questions.
 
 \paragraph{F-stats}
-However, when the number of populations exceeds a few dozen,  even codifying reasonable population models including all data can be prohibitively difficult.
-One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset.
+However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult.
+One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}.
 However, this has the drawback that a large proportion of the data may be unused.
@@
 The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}.
-I safe the formal definition for later; but the easiest way to motivate them assumes that populations are related as a tree, where the edges measure how much genetic drift has occurred.
-In this case, $F_2$ measures the genetic distance between populations, $F_3$ corresponds to an external, and $F_4$ to an internal branch of a tree ) \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.
+I save the formal definition for later; but the easiest way to motivate them assumes that populations are related as a tree, where the edge lengths measure how much genetic drift has occurred.
+\citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.
 
-In most applications, these $F$-statistics are estimated from some data, and then used as tests of treeness.
-The canonical alternative model is an admixture graph or phylogenetic network \citep{patterson2012}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A).
-Typically, analyses proceed by testing specific hypotheses abot gene flow using $F$-statistics \citep{reich2009, green2010, lazaridis2014}.
-Increasingly commonly, $F$-statistics are also incorporated into tools that jointly fit more complicated models \citep[e.g.
-][]{patterson2012,harney2021}.
-However, admixture graphs are not the only alternative model that explain violations of treeness.
-Expected $F$-statistics can be calculated for a wide range of demographic models \citep{peter2016}, or even in abstract Euclidean spaces \citep{oteo-garcia2021}, so that model fit remains a major concern.
-
-\paragraph{PCA}
-With minimal assumptions, PCA is one of the most widely used data summary and data exploration techniques.
-One way to think about PCA is that it aims to create a low-dimensional summary of the data that retains a maximum of information about the underlying structure in the data (Again, a formal description will be given in the next section).
-It has been widely used and independently discovered in many disciplines.
-In population genetics, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize global genetic diversity in the pre-genomic era \citep{cavalli-sforza1994}.
-Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale population genetic data set.
-This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g.
-remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses.
-Theoretically, the PCA-decomposition has been studied in terms of trees \citep{cavalli-sforza1975}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}.
+In most applications, these $F$-statistics are estimated from  data, and then used as tests of treeness.
+In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations.
+The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A).
+However, admixture graphs are not the only plausible alternative model, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.
 
 \paragraph{F-stats and PCA}
-However, the connections between PCA and $F$-statistics remains poorly understood, and the goal of this paper is to investigate this connection.
-To do so, I use the geometric interpretation of $F$-statistics developed by \cite{oteo-garcia2021} who defined them in arbitrary Euclidean spaces.
-We make use of the empirical finding that while the allele-frequency space is frequently high-dimensional, for many empirical data sets, a low number of dimensions are sufficient to characterize population structure, and hence $F$-statistics.
-This allows for approximate interpretations of $F$-statistics on PCA plots, and I use two sample data sets to evaluate these approximations in practice.
+The practical issue addressed in this study is how $F$-statistics can be reconciled with one of the most widely used data-driven  techniques, PCA.
+One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal components, PCs) retaining a maximum of the variance present in the data.
+In population genetics, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize global genetic diversity \citep{cavalli-sforza1994}.
+Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale data sets.
+%This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g.
+remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses.
+The PCA-decomposition has been studied for a number of population genetic models including trees \citep{cavalli-sforza1975}, spatially continuous structure \citep{novembre2008a}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}.
+Here, in order to link PCA to $F$-statistics, I interpret both of them geometrically in \emph{allele frequency space}, i.e.
+as functions of a high-dimensional Euclidean space.
+For $F$-statistics, this interpretation was recently developed by \cite{oteo-garcia2021}, and for PCA it follows naturally from the interpretation of approximating a high-dimensional space with a low-dimensional one.
+
+In the next section, I will formally derive the connection between $F$-statistics and PCA, and show how $F$-statistics can be interpreted geometrically, with a particular emphasis on two-dimensional PCA plots.
+In the results section, I will then discuss how some of the most common applications of $F$-statistics manifest themselves on a PCA, and illustrate them on two example data sets.
 
@@
 A comprehensive  treatise on PCA is given by e.g.
-\cite{jolliffe2013} and \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
+\cite{jolliffe2013} a useful primer on the mathematics is \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
 Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.
 
-\subsection{Formal Definition to $F$-statistics}
+\subsection{Formal Definition of $F$-statistics}
 Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ loci.
 Let $x_{il}$ denote the frequency at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$.
-As $X_i$ will be the only data summary for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.
+As $X_i$ will be the only data summary considered here for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.
 
@@
 As highlighted in the introduction, $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
-In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of all branches between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch between four nodes, respectively (Figure \ref{fig:geom}C).
+In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch between four nodes, respectively (Figure \ref{fig:geom}C).
 Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero.
@@
 Over time, genetic drift will change $X_y$ to $X_x$, which is the population we observe.
-This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact equations can be found in ][]{peter2016}.
-While admixture graphs are commonly considered, other alternative models exist.
-In \citep{peter2016}, I showed that $F_4$ is sensitive to many non-symmetric population structure models.
-$F_3$ is more robust, but may also be negative if substantial population substructure exists.
+This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.
 
 \subsubsection{Geometric interpretation of $F$-statistics}
-For very large data sets, it is hard to justify a specific demographic models that includes all data, but we will typically still be able to explore such data e.g.
-using PCA.
-One way to justify $F$-statistics is to interpret the populations $X_i$ geometrically as points or vectors in the $S$-dimensionsl \emph{data space} $\mathbb{R}^S$.
-\cite{oteo-garcia2021} showed that in this case, the $F$-statistics can be thought of as inner (or dot) products, and that all properties and tests related to treeness can be derived from this larger space.
-In particular the $F$-statistics can be defined as
+An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare.
+Recently, \cite{oteo-garcia2021} showed that these assumptions are not necessary by re-deriving $F$-statistics in a geometric framework.
+Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$.
+In this case, the $F$-statistics can be thought of as inner (or dot) products, and that all properties and tests related to treeness can be derived from this larger space.
+In particular the $F$-statistics can be written as
 \begin{subequations}
@@
 \end{subequations}
-where $\phi$ is the angle between $a$ and $b$. \cite{oteo-garcia2021} showed that in this framework, $F$-statistics are closely related to vector projections
+where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
 \begin{equation}
@@
 The drawback of the geometric approach of \cite{oteo-garcia2021} is that we have to deal with an very high-dimensional space, as the number of SNPs is frequently in the millions.
-However, it has been commonly observed that population structure is quite low-dimensional, with the first few PCs providing a good approximation of the underlying population structure.
-Therefore, we may hope that PCA could yield a reasonable approximation of the data space.
+However, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data \citep{patterson2006}.
+Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.
 
@@
 \subsection{Formal Definition of PCA}
-PCA is a common way of summarizing genetic data, and so a large number of variation of PCA exist, e.g.
+PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g.
 in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis.
-The version of PCA I describe here is set up in a way that the similarities to the $F$-statistics-framework are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets.
+The version of PCA I describe here is set up in a way that the similarities to the $F$-statistics framework are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets.
 In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}.
@@
 
-Let us again assume we have genotype data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype.
+Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype.
 If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
@@
 
-One way one can define PCA is that it aims to find a $K$-dimensional subspace of the data space that retains the most important variation.
+One way PCA can be motivated is that it aims to find a $K$-dimensional subspace of the data space that retains most  variation in the data.
 $K$ is at most $n-1$, in which case the data is simply rotated.
@@
 	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively.
-For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is contains information about population structure.
+For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and contains information about population structure.
 The SNP loadings $\ML=\mathbf{V}^T$ form an orthonormal basis of size $n \times S$, its rows give the contribution of each SNP to each PC.
 It is often used to look for outliers, which might be indicative of selection \citep[e.g][]{duforet-frebourg2016}.
-Alternatively, the PCs can also be motivated as an eigendecomposition of the covariance matrix $\MY\MY^T$.
+Alternatively, the PCs can also be obtained from an eigendecomposition of the covariance matrix $\MY\MY^T$.
 This can be motivated from (\ref{eq:svd}):
@@
 \subsubsection{Principal components from $F$-statistics}
-PCA and $F$-statistics are closely related.
+PCA, as defined above, and $F$-statistics are closely related.
 In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling, which, for squared Euclidean ($F_2$)-distances, leads to an identical decomposition to PCA \citep{gower1966}.
@@
 By performing a PCA, we rotate our data to reveal the axes of highest variation.
-However, the dot product is invariant under rotation, and $F$-statistics can be thought of as dot products.
+However, the dot product is invariant under rotation, and $F$-statistics can be thought of as dot products \citep{oteo-garcia2021}.
 What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or any other orthogonal basis such as the principal components $\MP$.
@@
 In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error.
-I will omit the superscript when the exact number of PCs is not relevant.
+I will omit the superscript of $\hat{F_2}$ when the exact number of PCs is not relevant.
 If we sum up the squared approximation errors over all pairs of populations, we obtain 
@@
 
-\subsubsection{$F$-statistics and projection on PCA}
+\subsubsection{$F$-statistics and samples projected onto PCA}
 One of the easiest ways of dealing with missing data in PCA is to calculate the principal components (equation \ref{eq:svd}) only on a subset of the data with no missingness, and then to \emph{project} the lower quality samples with high missingness onto this PCA.
@@
 \end{equation*}
-and so a new (centered) population $Y_{\text{new}}$ can be projected onto an existing PC simply by post-multiplying it with $\ML^T$: $$P_{\text{proj}} = Y_{\text{new}}\ML^T;$$ the $k$-th entry of $P_{\text{proj}}$ gives the coordinates of the new sample on the $k$-th PC.
+and so a new (centered) population $Y_{\text{new}}$ can be projected onto an existing PCA simply by post-multiplying it with $\ML^T$: $$P_{\text{proj}} = Y_{\text{new}}\ML^T;$$ the $k$-th entry of $P_{\text{proj}}$ gives the coordinates of the new sample on the $k$-th PC.
 However, it is likely that $Y_{\text{new}}$ lies outside the variation of the original sample.
 In this case, there is a projection error
-$$\normsq{Y_{\text{new}} - P_{\text{proj}}\ML }.$$ If we project with missing data, a similar projection can be used where we remove the rows from $Y_{\text{new}}$ and $\ML$ where data in $Y_{\text{new}}$ is missing \citep{patterson2006}.
+$$\normsq{Y_{\text{new}} - P_{\text{proj}}\ML } =  F_2(P_{\text{proj}}\ML, Y_{\text{new}}).$$ If we project with missing data, a similar projection can be used where we remove the rows from $Y_{\text{new}}$ and $\ML$ where data in $Y_{\text{new}}$ is missing \citep{patterson2006}.
 
@@
 
-\begin{equation}
-F_2(X_i, X_{\text{new}}) = F_2(Y_i, Y_{\text{new}}) = F_2(P_i, P_{\text{proj}}) + F_2(P_{\text{proj}}\ML, Y_{\text{new}}),
-\end{equation}
-because the projection error and projection are orthogonal to each other.
+\begin{align}
+F_2(X_i, X_{\text{new}}) &= F_2(Y_i, Y_{\text{new}})\nonumber\\ 
+&= F_2(P_i, P_{\text{proj}}) + F_2(P_{\text{proj}}\ML, Y_{\text{new}})\nonumber\\
+&= \hat{F_2}(P_i, P_j) + {\epsilon(P_i, P_j)} + F_2(P_{\text{proj}}\ML, Y_{\text{new}}) \label{eq:approx}.
+\end{align}
+The second row follows because the projection error and projection are orthogonal to each other.
+The main implication of equation \ref{eq:approx} is that we do not need to worry about distinguishing the projection and approximation errors.
 
@@
 This data set is a subset of the ``World Foci'' data set of \cite{lazaridis2014}, where I removed samples which are not permitted for free reuse.
-These populations span the globe and roughly represents global human genetic variation (638 individuals from 33 population) As this data set is very sparse, with often thousands of kilometers between adjacent sampling locations, I speculate that gene flow between these populations may not be particularly common; and their structure may therefore be well-approximated by an admixture graph.
+These populations span the globe and roughly represents global human genetic variation (638 individuals from 33 population) As adjacent sampling locations are often thousands of kilometers apart, I speculate that gene flow between these populations may not be particularly common; and their structure may therefore be well-approximated by an admixture graph.
 A file with all individuals used and their assigned population is given in \textbf{Supplementary File 1.}
@@
 Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}.
-We thus might think that gene flow between these populations is common \citep{ralph2013}, and a discrete model such as a tree or an admixture graph might be a rather poor reflection of this data.
+I thus speculate that gene flow between these populations is common \citep{ralph2013}, and a discrete model such as a tree or an admixture graph might be a rather poor reflection of this data.
 A file with all individuals used and their assigned population is given in \textbf{Supplementary File 2.}
@@
 \paragraph{Computing $F$-statistics and PCA}
-I perform analyses at the level of populations to ease presentation.
-It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples \citep{patterson2012}.
 All computations are performed in \texttt{R}.
@@
 \subsection{$F_2$ in PC-space}
-The $F_2$-statistic is an estimate of the squared Euclidean distance between two
-populations. It thus corresponds to the squared distance between populations in PCA-space, and
-reflects the intuition that closely related populations  will be close to each other on a
-PCA-plot, and have low pairwise $F_2$-statistics. In converse, if two
-populations have high $F_2$ but appear on the same point on an PCA-plot, this suggests
-that substantial variation is hidden on higher PCs, and we may want to investigate the PCs which contribute a large distance to this particular $F_2$-statistic in order to better understand and visualize their relationship.
+The $F_2$-statistic is an estimate of the squared allele-frequency distance between two
+populations.
+On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations.
+In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will be close to each other on a
+PCA-plot, and have low pairwise $F_2$-statistics.
+However, since  equation \ref{eq:fpc} is a sum of squared (non-negative) terms over PCs, the $F_2$-distance on a PCA-plot will always be an underestimate of the full distance.
+Thus, PCA might project two populations with high $F_2$ very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations.
+In converse, populations that are distant on a PCA-plot are guaranteed to also have a large $F_2$-distance.
 
 \subsection{When are admixture-$F_3$ statistics negative?}
-Consider the admixture scenario in Figure \ref{fig:admix}A, where population $X_y$ is the result of a mixture of $X_2$ and $X_3$, and subsequent drift changes the allele frequencies of the admixed population from $X_y$ to the sampled population $X_x$.
+Consider the admixture scenario in Figure \ref{fig:admix}A, where population $X_y$ is the result of a mixture of $X_2$ and $X_3$, and subsequent drift changes the allele frequencies of the admixed population from $X_y$ to $X_x$, for which we have data.
 How is such a scenario displayed on a PCA?
-ince the allele frequencies of $X_y$ are a linear combination of $X_2$ and $X_3$, it will lie on the line segment connecting these two populations (Figure \ref{fig:admix}B).
-In fact, if $X_x$ (and other populations related to $X_x$) was not part of the construction of the PCA, and is instead projected onto it, the drift from $X_y$ to $X_x$ will be independent of the population on the PCA, and $X_x$ will be at the same location on a PCA plot as $X_y$, which can be used to predict the admixture proportions \citep{brisbin2012, mcvean2009, oteo-garcia2021}.
+Since the allele frequencies of $X_y$ are a linear combination of $X_2$ and $X_3$, it will lie on the line segment connecting these two populations (Figure \ref{fig:admix}B), at a location predicted by the admixture proportions.
+Subsequent drift will change the allele frequency of $X_x$, and so in general it might fall on a different point on a PCA-plot.
+An exception occurs when $X_x$ (and no other populations related to $X_x$) are not part of the construction of the PCA, so that $F_2(X_x, X_y)$ is orthogonal to all PCs.
+In this case, $X_x$ and $X_y$ project to the same point, and the location on the PCA  can be used to predict the admixture proportions \citep{brisbin2012, mcvean2009, oteo-garcia2021}.
+However, if $X_x$, is included in the construction of the PCA, or if some gene flow occurred between $X_x$ and any of the populations used to construct the PCA, $X_x$ and $X_y$ will project on different spots (Figure \ref{fig:admix}B).
 
-If $X_x$, or populations more closely related to $X_x$ than $X_2$ and $X_3$, are included in the construction of the PCA, this is no longer true, and $X_x$ will project on a different spot than $X_y$ (Figure \ref{fig:admix}B).
 Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might be considered admixed between them?
-
 One way to address this question is to consider the space for which $F_3$ is negative, i.e.
@@
 $F_3$ is negative when the triangle is obtuse, i.e.
-$X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_2X_3}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).
+$X_x$ could be considered admixed if it lies inside the $n$-ball with diameter $\overline{X_2X_3}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).
 
@@
 
- Thus if $\hat{F}_3 \ll F_3$, the true circle will be bigger than would be predicted from a 2D-plot.
-In this case, substantial relevant genetic differentiation is ``hidden'' in the higher PCs, and populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics.
+ Thus if $\hat{F}_3 \ll F_3$, the ``true'' circle will be bigger than what would be predicted from a 2D-plot, and populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics.
 This is because they are outside the $n$-ball in higher dimensions.
@@
 So far, I used eq.
-\ref{eq:fpc} to interpret $F$-statistics on a PC-plot, but the argument holds for \emph{any} orthonormal transformation.
-This allows for a variety of visualizations that use both $F$-statistics and PCs.
-The motivation for this is that sometimes we wish to partition the variation in the data into a subspace of interest, and an orthogonal residual space that captures the information discarded.
-Examples where analyses are restricted to such subspaces include the $F_4$-ratio test \citep{patterson2012, petr2019}, \texttt{qpWave} \citep{skoglund2015} and \texttt{qpAdm} \citep{harney2021}.
-For the $F_4$-ratio, for example, a ratio
+\ref{eq:fpc} to interpret $F$-statistics on a PC-plot, but the argument holds for \emph{any} orthonormal projection of the data space.
+This is useful in particular for estimates of admixture proportions, which are often done in a small reference space \citep{patterson2012, petr2019, harney2021, oteo-garcia2021}.
+
+The simplest approach is the $F_4$-ratio to infer the admixture sources of population $X$ as 
 \begin{eqnarray}
-\alpha = \frac{F_4(R_1, R_2; X, A)}{F_4(R_1, R_2; B, A)}  \label{eq:f4ratio}
+\alpha = \frac{F_4(R_1, R_2; X, A)}{F_4(R_1, R_2; B, A)} = \frac{\vectorproj[R_1 - R_2]{X - A}}{\vectorproj[R_1 - R_2]{B - A}} \label{eq:f4ratio},
 \end{eqnarray}
-is used, which can be interpreted as projecting $X-A$ and $B-A$ onto $R_1 - R_2$.
-Thus, we can make a plot where we plot the variation on the $X$-axis along $R_1 - R_2$, and perform a PCA on the residual.
-This can be important because the residual can be used to check assumptions, e.g.
-$A - A'$ and $B - B'$ need to be orthogonal (Figure \ref{fig:geom}F).
+which can be interpreted as projecting $X-A$ and $B-A$ onto $R_1 - R_2$ and measuring their relative proportions \citep{oteo-garcia2021}.
+\texttt{qpAdm} extends this approach to a higher-dimensional reference space and multiple potential source populations.
+One open practical question in many applications is which reference and putative source populations to use \citep{harney2021}.
+The theory developed here suggests some possible visualizations that may address this issue.
 
 \subsubsection{Example}
-In the PCA on the world overview data set, I found a seemingly linear gradient from Africans to Europeans (Figure \ref{fig:f3}D).
+In the PCA on the world overview data set, I found a gradient from Africans to Europeans (Figure \ref{fig:f3}D).
 I focus on this cline using an alternative projection by using  $F$-statistics of the form $F_4(X, Y; \text{Sardinian}, \text{Yoruba})$), which might e.g.
-be used if we were to quantify gene flow associated with the out-of-Africa expansion.
+be used in an $F_4$-ratio.
 These $F_4$-statistics are very well-approximated by the first two PCs, with a 99.2\% correlation between $F_4$ and its approximation using the first two PCs (Figure \ref{fig:f4}C).
 
-In Figure \ref{fig:f4}D, I show the projection $\langle X; \text{Sardinian}, \text{Yoruba} \rangle$ on the $X$-axis, which means that $F_4(X, Y; \text{Sardinian}, \text{Yoruba})$ is  is proportional to their horizontal distance between $X$ and $Y$.
-The first two residual PCs are given on the Y-axis and in the coloring; this visualization reveals some variation within Africans (with Mbuti, Biaka and Ju$\vert$'hoansi) that is largely orthogonal to this gradient, as is the variation between Europeans, Asian and the Americans.
+In Figure \ref{fig:f4}D, I show the projection $\langle X; \text{Sardinian}, \text{Yoruba} \rangle$ on the $X$-axis, which means that the horizontal difference between any pair of population is proportional to their $F_4$-statistic relative to Sardinians and Yorubans.
+We can also ask what variation is not represented by performinc a PCA on the residual of this projection, the first two residual PCs are given on the Y-axis and in the coloring.
+This visualization reveals that variation within Africans (with Mbuti, Biaka and Ju$\vert$'hoansi, top right) and the variation in East Asians and Americans are largely orthogonal to this projection axis, and so Sardinians and Yoruba would be poor references if we were interested in studying East Asian genetic variation.
 
@@
 \section{Discussion}
-Particularly for the analysis of ancient DNA, $F$-statistics  are  a powerful tool to describe population genetic diversity.
-Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA plot.
+Particularly for the analysis of human genetic variation, $F$-statistics  are  a powerful tool to describe population genetic diversity.
+Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA-plot.
 This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
-As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, (e.g using a lower number of populations).
-It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are explained solely by higher PCs, and so whenever such contradictions arise, higher PCs will be informative for population structure.
-Previous interpretation of PCA in the context of population genetic models have primarily focused on the PCs, which can be derived analytically for trees \citep{cavalli-sforza1975} and homogeneous spatial models \citep{novembre2008a}.
+As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, e.g using a lower number of populations.
+It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are either due to variation on higher PCs, or due to differences in assumptions about normalizations or population groupings.
+Previous interpretation of PCA in the context of population genetic models have focused on simple models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{franccois2021}.
 My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs.
-One consequence is that the results here are not impacted by sample ascertainment and sample sizes \citep{mcvean2009, novembre2008a}, which are common concerns in the interpretation of PCA.
+One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA.
 However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis.
-From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data
+From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.
 
@@
 
-To make this link directly applicable to data analysis, there are a number of -- primarily statistical -- concerns that will need to be addressed.
-Fist, PCA is most frequently run on individuals, whereas $F$-statistics are often calculated on populations.
-This is largely because in most workflows, PCA is run much earlier than $F$-statistics; it is a standard assumption of $F$-statistics that there is no population substructure \citep{patterson2012}, and an easy way to test that is ensure that all individuals cluster tightly on a PCA.
+The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized.
+In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies.
+In contrast, most data analyses have to grapple with missing data, SNP are often weighted according to their allele frequency and observed, individual-level genotypes are used as the basis of PCA.
 
-A second difference is that frequently, rare SNPs are weighted higher in PCA, whereas all SNPs are weighted the same for $F$-statistics \citep{patterson2006, patterson2012}.
-This is a difference of convention \citep{cavalli-sforza1975}; since $F$-statistics are summed over SNPs with the same expectation, $F$-statistics could also be calculated using the same weighting.
-The close connection between the two approaches developed here suggest that for most analyses, users might want to be consistent and use the same weighting for both types of analyses.
+The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}}.
+Alternatively, samples with large amounts of missing data are projected onto PCAs computed without missing data \citep{patterson2006}.
+In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation.
+These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}}.
 
-The third and perhaps biggest gap are statistical issues.
-The treatment here focuses on the mean estimated $F$-statistic, but many  applications of $F$-statistics are based on hypothesis tests \citep{patterson2012}.
-This requires estimating accurate standard errors for these statistics, which is difficult since nearby SNPs will be correlated due to recombination \citep{hahn2018}.
-In contrast, PCA jointly models the covariance matrix due to population structure and sampling, so if hypothesis tests are desired this will need to be incorporated.
+The normalization of SNPs is similarly a matter of convention.
+The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted.
+The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}.
 
-An advantage of calculating $F$-statistics based on PCs is that they yield consistent estimtates.
-For both data sets I investigated here, the matrix $\MF$ of $F$-statistics obtained  using admixtools2 is not a proper squared Euclidean distance matrix, i.e.
-it is not negative semidefinite and has imaginary PCs.
-A model-based framework based on probabilistic PCA \citep{hastie2015, meisner2021, agrawal2020} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
+The third issue of individual-based vs population-based analysis is similarly a matter of interpretation.
+For $n$ samples, the number of possible $F$-statistics is on the order of $n^4$, and so the number of statistics is kept low by grouping individuals into populations.
+This is, however, not necessary, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
+Here, individual-based PCA can provide some guidance for grouping samples: Since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on a PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.
+
+The final difference between the $F$-statistic-based PCA used here and individual-based PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
+The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
+Probabilistic PCA is one class of approaches that aim to separate the population structure from individual-level noise \citep{agrawal2020}, and it seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.
+
+Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions that are partially arbitrary, and partially  explained by the focus of PCA on exploratory data analysis.
+Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if PCA and $F$-statistics are used in a way such that the results are comparable to each other.
 

### 2021-10-29 data/change_of_basis.R
revised figures

@@ -1,0 +1,39 @@
+source("fscripts.R")
+library(glue)
+BTHEME = theme_classic() + theme(legend.position="none")
+
+f2s = admixtools::read_f2("westeurasian1")
+pcmat = pca_from_f2s(f2s)
+pcs = pca_from_pcmat(pcmat)
+idw = "AA"
+idx = "Sardinian"
+idy = "Basque"
+idz = "Han"
+
+ids = c(idx, idy)
+n = nrow(pcs)
+
+f2mat = f2(f2s, unique_only=F) %>% select(-se) %>% 
+    pivot_wider(names_from=pop2, values_from=est) %>%
+    column_to_rownames('pop1') %>% as.matrix
+cmat = double_center(-f2mat)
+
+a = treelet::Run_JTree(cmat, n-1, 1:(n-1))
+B = lapply(a$basis, function(b){
+               rownames(b) = rownames(cmat)
+               colnames(b) = rownames(cmat)
+               b %>% as.data.frame %>%
+               rownames_to_column('pop1')
+    })
+
+
+M = B %>% 
+    bind_rows(.id='lvl') %>% 
+    mutate(lvl=as.integer(lvl)) %>% 
+    pivot_longer(-1:-2, names_to='pop2', values_to='bval')
+
+B = a$basis[[n-1]]
+x = function(b)ifelse( abs(b) < 1e-5, NA, (b - min(b)) / diff(range(b)))
+S = apply(B, 2,x)
+
+

### 2021-10-29 fstats_pca_lewontin.tex

@@
 ]{geometry}
-%\usepackage[utf8x]{inputenc}
+\usepackage[utf8]{inputenc}
 %\usepackage[T1]{fontenc}
@@
 \section{Introduction}
-As most species, the genetic diversity of human populations is influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, nielsen2020}.
-Population genetic models use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.
+As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2020}.
+In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.
 
-In particular, in isolated populations genetic drift will slowly change allele frequencies.
-As a result, isolated populations are expected to slowly differentiate \citep{wahlund1928,cavalli-sforza1975}.
+The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a}.
+In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}.
 In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985a}.
-However, isolation-by-distance patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}.
+However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}.
 In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}.
-Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016} and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2019}.
+Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2019}.
 
-The interplay of these demographic processes leads to the complex genetic structure observed in present-day human populations \citep{the2015, mallick2016} with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a} that we model in order to reconstruct human demographic history.
-This is challenging because particularly for large and heterogeneous data sets we cannot expect to devise a single model that captures all processes.
+Particularly for large and heterogeneous data sets, disentangling all these processes is challenging, and we cannot expect to devise a single model catching both broad strokes and minute details of human history.
 A commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions.
@@
 The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modeling assumptions such as that populations are discrete, randomly mating, or at equilibrium.
-While in most cases these assumptions are violated to some extent and cannot be verified, but we hope that the resulting model fits provide a sufficiently accurate answer for specific research questions.
+While in most cases these assumptions are violated to some extent and cannot be verified, we hope that the resulting model fits provide sufficiently accurate answers to specific research questions.
 
-\paragraph{F-stats}
+\paragraph{F-statistics}
 However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult.
 One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}.
-However, this has the drawback that a large proportion of the data may be unused.
+However, this has the drawback that a large proportion of the available data remains unused.
 An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.
@@
 The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}.
-I save the formal definition for later; but the easiest way to motivate them assumes that populations are related as a tree, where the edge lengths measure how much genetic drift has occurred.
+Formal definition will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches.
 \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.
 
-In most applications, these $F$-statistics are estimated from  data, and then used as tests of treeness.
+In most applications, $F$-statistics are estimated from  data, and then used as tests of treeness.
 In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations.
@@
 
-\paragraph{F-stats and PCA}
-The practical issue addressed in this study is how $F$-statistics can be reconciled with one of the most widely used data-driven  techniques, PCA.
-One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal components, PCs) retaining a maximum of the variance present in the data.
-In population genetics, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize global genetic diversity \citep{cavalli-sforza1994}.
+\paragraph{F-statistics and PCA}
+The practical issue addressed in this study is how $F$-statistics can be reconciled with PCA, one of the most widely used data-driven modeling  techniques.
+One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal component, PC) retaining a maximum of the variance present in the data.
+To understand population structure, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize genetic diversity \citep{cavalli-sforza1994}.
 Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale data sets.
@@
 remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses.
-The PCA-decomposition has been studied for a number of population genetic models including trees \citep{cavalli-sforza1975}, spatially continuous structure \citep{novembre2008a}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}.
+The PCA-decomposition has been studied for a number of explicit population genetic models including trees \citep{cavalli-sforza1975}, spatially continuous structure \citep{novembre2008a}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}.
 Here, in order to link PCA to $F$-statistics, I interpret both of them geometrically in \emph{allele frequency space}, i.e.
@@
 In the next section, I will formally derive the connection between $F$-statistics and PCA, and show how $F$-statistics can be interpreted geometrically, with a particular emphasis on two-dimensional PCA plots.
-In the results section, I will then discuss how some of the most common applications of $F$-statistics manifest themselves on a PCA, and illustrate them on two example data sets.
+In the Results section, I will then discuss how some of the most common applications of $F$-statistics manifest themselves on a PCA, and illustrate them on two example data sets, before ending with a discussion.
 
@@
 A comprehensive  treatise on PCA is given by e.g.
-\cite{jolliffe2013} a useful primer on the mathematics is \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}.
+\cite{jolliffe2013}, a useful primer on the mathematics is \cite{pachter2014}, and a helpful guide to interpretation is \cite{cavalli-sforza1994}.
 Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.
@@
 
-The three $F$-statistics can then be defined as
+The three $F$-statistics are defined as
 \begin{subequations}
@@
 
-As highlighted in the introduction, $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
-In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch between four nodes, respectively (Figure \ref{fig:geom}C).
+$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
+In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch, respectively (Figure \ref{fig:geom}C).
 Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero.
@@
 
-Thinking of $F$-statistics as branch lengths is useful to understand a number of applications: In particular, one common task is to find the population most closely related to an unknown sample $X_U$ \citep{raghavan2014}.
+Thinking of $F$-statistics as branch lengths is useful for a number of applications, including building multi-population models \citep{patterson2012, lipson2013}, estimating admixture proportions \citep{petr2019, harney2021} and finding the population most closely related to an unknown sample (``Outgroup''-$F_3$-statistic).
+
+%In particular, one common task is to find the population most closely related to an unknown sample $X_U$ \citep{raghavan2014}.
 One way to do that is using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ denotes an outgroup, and the $X_i$ are a panel of populations that are candidates for the closest match.
-The  highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times.
+The largest value of $F_3$ indicates the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times.
 The intuition is given in Figure \ref{fig:outgroupf3}A; where the outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ is highlighted.
@@
 
-Most commonly however, $F_3$ and $F_4$ are used as tests of treeness \citep{patterson2012}: Negative $F_3$-values correspond to a branch with negative genetic drift, which is a violation of treeness.
-Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}.
-The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A, where (the typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$.
-Over time, genetic drift will change $X_y$ to $X_x$, which is the population we observe.
+Most commonly however, $F_3$ and $F_4$ are used as tests of treeness \citep{patterson2012}: Negative $F_3$-values correspond to a branch with negative genetic drift, which is not allowed under the null assumption of a tree-like population relationship.
+Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{buneman1974, patterson2012}.
+
+The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A.
+Here, the *typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$.
+Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe.
 This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.
@@
 An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare.
-Recently, \cite{oteo-garcia2021} showed that these assumptions are not necessary by re-deriving $F$-statistics in a geometric framework.
+Recently, \cite{oteo-garcia2021} showed re-deriving $F$-statistics in a geometric framework, showing that these assumptions are not necessary.
 Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$.
-In this case, the $F$-statistics can be thought of as inner (or dot) products, and that all properties and tests related to treeness can be derived from this larger space.
+In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space.
 In particular the $F$-statistics can be written as
@@
 which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$.
+Thinking of $F$-statistics as projections also holds on trees: In e.g.
+a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:admix}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$.
+The external branches are independent populations, and so they are expected to drift orthogonally to each other.
 
@@
 	\includegraphics[width=.5\textwidth]{figures/outgroup_f3.pdf}
-	\caption{\textbf{Outgroup-$F_3$-statistics}}
+	\caption{\textbf{Outgroup-$F_3$-statistics}. The circle with a dot denotes a right angle.}
 	\label{fig:outgroupf3}
@@
 in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis.
-The version of PCA I describe here is set up in a way that the similarities to the $F$-statistics framework are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets.
+The version of PCA I use here is set up such that the similarities to  $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets.
 In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}.
@@
 
-Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype.
+Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ of many populations in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype.
 If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
@@
 
-There are several algorithms that are used to perform PCAs, the most common one is based on singular value decomposition \citep{jolliffe2013}.
+There are several algorithms that are used to perform PCAs, the most common one is based on singular value decomposition \citep[e.g.][]{jolliffe2013}.
 In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
@@
 I will omit the superscript of $\hat{F_2}$ when the exact number of PCs is not relevant.
-If we sum up the squared approximation errors over all pairs of populations, we obtain 
+If we sum up the squared approximation errors over all pairs of populations in our sample, we obtain 
 \begin{equation}
@@
 and so a new (centered) population $Y_{\text{new}}$ can be projected onto an existing PCA simply by post-multiplying it with $\ML^T$: $$P_{\text{proj}} = Y_{\text{new}}\ML^T;$$ the $k$-th entry of $P_{\text{proj}}$ gives the coordinates of the new sample on the $k$-th PC.
-However, it is likely that $Y_{\text{new}}$ lies outside the variation of the original sample.
+However, it is likely that $Y_{\text{new}}$ lies outside the variation of the original samples.
 In this case, there is a projection error
-$$\normsq{Y_{\text{new}} - P_{\text{proj}}\ML } =  F_2(P_{\text{proj}}\ML, Y_{\text{new}}).$$ If we project with missing data, a similar projection can be used where we remove the rows from $Y_{\text{new}}$ and $\ML$ where data in $Y_{\text{new}}$ is missing \citep{patterson2006}.
+$$\normsq{Y_{\text{new}} - P_{\text{proj}}\ML } =  F_2(P_{\text{proj}}\ML, Y_{\text{new}}).$$ If we project with missing data, a similar projection can be used where we remove the rows from $Y_{\text{new}}$ and $\ML$ where data in $Y_{\text{new}}$ is missing, and add a scaling factor \citep{patterson2006}.
 
@@
 The second row follows because the projection error and projection are orthogonal to each other.
-The main implication of equation \ref{eq:approx} is that we do not need to worry about distinguishing the projection and approximation errors.
+The main implication of equation \ref{eq:approx} is that both truncation and projection introduce some error, and that $ \hat{F_2}(P_i, P_j)$ will be a good approximation to  $F_2(P_i, P_j)$ only if both errors are small.
 
@@
 To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.
-A squared Euclidean distance matrix such as $\MF$ has all negative or zero eigenvalues (i.e.
-$\MF$ is negative-semidefinite).
-However as $F_2$-statistics are estimates, some eigenvalues might be slightly positive, which would lead to imaginary PCs.
+The right-hand side matrix of equation \ref{eq:mds} is supposed to have non-negative eigenvalues (i.e.
+$-\MC\MF\MC$ is positive-semidefinite).
+However as $F_2$-statistics are estimates, some eigenvalues might be slightly negative, which would lead to imaginary PCs.
 I avoid this by using the \texttt{nearPD}-function in R that ensures all eigenvalues have the correct sign.
@@
 	\includegraphics[width=\textwidth]{figures/fstats_pca_vs_tree.pdf}
-	\caption{\textbf{Representation of $F$-statistics on tree and 2D-PCA-plot.} The schematics show four populations and their representation using a tree (top row) or a 2D-PCA plot.
+	\caption{\textbf{Representation of $F$-statistics on trees and 2D-PCA-plots.} The schematics show four populations and their representation using a tree (top row) or a 2D-PCA plot (bottom row).
 A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space.
-B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthoginal projection of $X_1 - X_3$ onto $X_1$-$X_4$.
+B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthogonal projection of $X_1 - X_3$ onto $X_1$-$X_4$.
 C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or the orthogonal projection of $X_2 - X_3$ on $X_1 - X4$.
@@
 %\subsection{Geometry of $F$-statistics in PC-space}
-The transformation from the previous section allows us to consider the geometry
-of $F$-statistics in PCA-space. The relationships we will discuss formally only
-hold if we use all PCs. However, the appeal of PCA is that frequently,
-only a very small number $K \ll n$ of PCS contain most information that is
-relevant for population structure (for visualization
-$K=2$ is often used).
+The transformation from the previous section allows us to consider the geometry of $F$-statistics in PCA-space.
+The relationships we will discuss formally only hold if we use all PCs.
+However, the appeal of PCA is that frequently, only a very small number $K \ll n$ of PCS contain most information that is
+relevant for population structure, in which case the geometric interpretations become very simple.
+Thus, throughout the schematic figures, I assume that two PCs are sufficient to characterize population structure.
+In the data applications I evaluate how deviations of this assumption my manifest themselves in PCA plots.
 
@@
 On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations.
-In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will be close to each other on a
-PCA-plot, and have low pairwise $F_2$-statistics.
-However, since  equation \ref{eq:fpc} is a sum of squared (non-negative) terms over PCs, the $F_2$-distance on a PCA-plot will always be an underestimate of the full distance.
-Thus, PCA might project two populations with high $F_2$ very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations.
+In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will fall close to each other on a PCA-plot, and have low pairwise $F_2$-statistics.
+However, since $F_2$ can be written as  a sum of squared (non-negative) terms for each PC (eq.
+\ref{eq:fpc}), the distance on a PCA-plot will always be an underestimate of the full $F_2$-distance.
+Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations.
 In converse, populations that are distant on a PCA-plot are guaranteed to also have a large $F_2$-distance.
@@
 \subsection{When are admixture-$F_3$ statistics negative?}
-Consider the admixture scenario in Figure \ref{fig:admix}A, where population $X_y$ is the result of a mixture of $X_2$ and $X_3$, and subsequent drift changes the allele frequencies of the admixed population from $X_y$ to $X_x$, for which we have data.
+Consider again the admixture scenario in Figure \ref{fig:admix}A, where population $X_y$ is the result of a mixture of $X_2$ and $X_3$, and subsequent drift changes the allele frequencies of the admixed population from $X_y$ to $X_x$.
 How is such a scenario displayed on a PCA?
@@
 Subsequent drift will change the allele frequency of $X_x$, and so in general it might fall on a different point on a PCA-plot.
-An exception occurs when $X_x$ (and no other populations related to $X_x$) are not part of the construction of the PCA, so that $F_2(X_x, X_y)$ is orthogonal to all PCs.
-In this case, $X_x$ and $X_y$ project to the same point, and the location on the PCA  can be used to predict the admixture proportions \citep{brisbin2012, mcvean2009, oteo-garcia2021}.
-However, if $X_x$, is included in the construction of the PCA, or if some gene flow occurred between $X_x$ and any of the populations used to construct the PCA, $X_x$ and $X_y$ will project on different spots (Figure \ref{fig:admix}B).
+An exception occurs when $X_x$ (and no other populations related to $X_x$) are not part of the construction of the PCA, so that $X_x - X_y$ is orthogonal to all PCs, i.e.
+$$\langle X_x - X_y, X_i - X_j \rangle = \langle X_x - X_y, P_i \rangle = 0$$ for all populations $i,j \leq n$.
+In this case, $X_x$ and $X_y$ project to the same point, and the location on the PCA  can directly be used to predict the admixture proportions \citep{mcvean2009, brisbin2012, oteo-garcia2021}.
+However, if either $X_x$, is included in the construction of the PCA, or if some gene flow occurred between $X_x$ and any of the populations used to construct the PCA, $X_x$ and $X_y$ may project on different spots (Figure \ref{fig:admix}B).
 
@@
 As an example, I visualize the admixture statistic $F_3(X; \text{Basque}, \text{Turkish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A).
-In this case, the projected $n$-ball (light gray) and circle based on 2D (dark gray) align relatively closely, but several populations inside the ball (e.g.
-Sardinian, Finnish) have, in fact, positive $F_3$-values.
+In this case, the projected $n$-ball (light gray) and circle based on two dimensions (dark gray) have similar sizes.
+However, several populations that appear inside the circles (e.g.
+Sardinian, Finnish) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball.
 This reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure.
@@
 \begin{figure}[!ht]
-	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
+	\includegraphics[width=\textwidth]{figures/fig_data_europe.pdf}
 	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
@@
 
-\subsection{$F_3$-statistics as projections}
-Outgorup-$F_3$-statistic motivate a comparison of $F_3$-statistics to projections (Equation \ref{eq:proj}), consider again the case displayed in Figure \ref{fig:outgroupf3}A, where the goal is to find the population $X_i$ that is closest to an unknown population $X_U$, with respect to an outgroup $X_O$, which we can do using the statistic $F_3(X_O; X_U, X_i)$.
+\subsection{Outgroup-$F_3$-statistics as projections}
+A common application of $F_3$-statistics is, given an unknown sample $X_U$, to find the most closely related population among a reference panel $(X_i)$ \citep{raghavan2014}.
+This is done  using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ is an outgroup.
+The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (\ref{fig:outgroupf3}A) The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.
+
+To make sense of outgorup-$F_3$-statistics in the PCA context, I use the association of $F_3$-statistics to projections (Equation \ref{eq:proj}):
 On a PCA-plot, we can visualize this $F_3$-statistic as the projection of the vector $X_i - X_O$ onto $X_U - X_O$: 
-$$\vectorproj[X_U - X_O]{X_i - X_O} = \frac{F_3(X_O; X_U, X_i)}{F_2(X_O; X_U)}(X_U - X_O).$$
+$$\vectorproj[X_U - X_O]{X_i - X_O} =F_3(X_O; X_U, X_i) \frac{X_U - X_O}{F_2(X_O; X_U)}.$$
 
-Of the right-hand-side terms, $X_U-X_O$ gives the direction of the resulting vector, and the $F_2$-term in the denominator is a normalizing constant.
-Neiter of these terms depend on $X_i$, which justifies the argument that the $F_3$-statistic and length of the projected vector are proportional to each other, and can thus be interpreted similarly.
-Thus, the outgroup-$F_3$-statistic is larger for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in the example in Figure \ref{fig:outgroupf3}B this is $X_3$.
+On the right-hand-side terms only the $F_3$ term depends on the $X_i$.
+The fraction can be thought of as a normalizing constant,  which justifies the argument that the $F_3$-statistic and length of the projected vector are proportional to each other, and can thus be interpreted similarly.
+Thus, the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in the example in Figure \ref{fig:outgroupf3}B this is $X_3$.
 
@@
 \paragraph{Example}
-Again, these projections will be orthogonal when using the full data, and may only be approximately orthogonal when approximated using the first few PCs.
-In Figure \ref{fig:f3}D, I visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Mozabite}, X_i)$, i.e.
-a statistic that aims to find the population most closely related to Mozabite (a Berber ethnic group from the northern Sahara), assuming the Mbuti are an outgroup.
-On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Mozabite (black line).
+In Figure \ref{fig:f3}D, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Gujarati}, X_i)$, in i.e.
+a statistic that aims to find the population most closely related to Gujarati (from Western India), assuming the Mbuti are an outgroup to all populations.
+On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Gujarati (black line).
 For each population, the projection is indicated with a grey line.
-In the full data space, this line is always orthogonal to the segment Mbuti-Mozabite, but on the plot (i.e.
+In the full data space, this line is always orthogonal to the segment Mbuti-Gujarati, but on the plot (i.e.
 the subspace spanned by the first two PCs), this is not necessarily the case.
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figures/fig_f4_data.pdf}	
+\includegraphics[width=\textwidth]{figures/fig_data_world.pdf}	
 	\caption{\textbf{PCA and $F_4$-statistics} A: Projection angle and correlation coefficient $r$ representation of $F_4(X, Sardinian; Saudi, Georgian)$ (red) in the Western Eurasian data set, and approximations using two (green) and three (blue) PCs.
@@
 
-The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}}.
+The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}.
 Alternatively, samples with large amounts of missing data are projected onto PCAs computed without missing data \citep{patterson2006}.
 In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation.
-These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}}.
+These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}.
 

### 2021-11-01 fstats_pca_lewontin.tex
discussion reordering

@@
 \section{Discussion}
-Particularly for the analysis of human genetic variation, $F$-statistics  are  a powerful tool to describe population genetic diversity.
+Particularly for the analysis of human genetic variation with a large number of individuals with heterogeneous relationships, $F$-statistics  are  a powerful tool to describe population genetic diversity.
 Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA-plot.
 This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
-As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, e.g using a lower number of populations.
-It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are either due to variation on higher PCs, or due to differences in assumptions about normalizations or population groupings.
+As PCA is often ran in an early step in data analysis, this may aid in generation of hypotheses that can be more directly evaluated using generative models, typically using a lower number of populations.
+It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots.
+In many cases, differences between the two data summaries will be due to variation on higher PCs.
+In this case, plotting additional PCs, or further subsetting the data to a more local set of populations seems prudent.
+
+The other cause for disagreements between $F$-statistics and PCA are differences in assumptions.
+The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized.
+In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies.
+In contrast, most data analyses have to grapple with missing data, SNP are often weighted according to their allele frequency and observed, individual-level genotypes are used as the basis of PCA.
+
+The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}.
+Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}.
+In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation.
+These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}, and a PCA-based approach to estimating $F$-statistics may provide higher accuracy than naive approaches because it allows using data from related populations.
+
+The normalization of SNPs is similarly a matter of convention.
+The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted.
+The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}.
+As genetic differentiation between human populations is low, the  normalization used may matter little in practice \citep{felsenstein1973}, but using consistent normalization between $F$-statistics and PCA-based analysis seems advisable since it removes normalization as a source of errors.
+.
+
+The third issue of individual-based vs population-based analysis is similarly a matter of interpretation.
+For $n$ samples, the number of possible $F$-statistics is on the order of $n^4$, and so the number of statistics is kept low by grouping individuals into populations.
+This is, however, not necessary, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
+Here, individual-based PCA can provide some guidance for grouping samples: Since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on a PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.
+
+The final difference between the $F$-statistic-based PCA used here and individual-based PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
+The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
+Probabilistic PCA is one class of approaches that aim to separate the population structure from individual-level noise \citep{agrawal2020}, and it seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.
+
+Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions that are partially arbitrary, and partially  explained by the focus of PCA on exploratory data analysis.
+Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if PCA and $F$-statistics are used in a way such that the results are comparable to each other.
+
 Previous interpretation of PCA in the context of population genetic models have focused on simple models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{franccois2021}.
@@
 Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.
-
-The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized.
-In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies.
-In contrast, most data analyses have to grapple with missing data, SNP are often weighted according to their allele frequency and observed, individual-level genotypes are used as the basis of PCA.
-
-The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}.
-Alternatively, samples with large amounts of missing data are projected onto PCAs computed without missing data \citep{patterson2006}.
-In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation.
-These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}.
-
-The normalization of SNPs is similarly a matter of convention.
-The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted.
-The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}.
-
-The third issue of individual-based vs population-based analysis is similarly a matter of interpretation.
-For $n$ samples, the number of possible $F$-statistics is on the order of $n^4$, and so the number of statistics is kept low by grouping individuals into populations.
-This is, however, not necessary, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
-Here, individual-based PCA can provide some guidance for grouping samples: Since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on a PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.
-
-The final difference between the $F$-statistic-based PCA used here and individual-based PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
-The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
-Probabilistic PCA is one class of approaches that aim to separate the population structure from individual-level noise \citep{agrawal2020}, and it seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.
-
-Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions that are partially arbitrary, and partially  explained by the focus of PCA on exploratory data analysis.
-Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if PCA and $F$-statistics are used in a way such that the results are comparable to each other.
 

### 2021-11-01 data/f3_fig.R

@@ -6,7 +6,7 @@ library(glue)
 R = 1.5
 BTHEME = theme_classic() + theme(legend.position="none")
 
-if(T){
+if(F){
 f2s = admixtools::read_f2("westeurasian1/")
 pcmat = pca_from_f2s(f2s)
 pcs = pca_from_pcmat(pcmat)
@@ -66,12 +66,12 @@ ggsvg(P1, "f3_westeurasia1.svg", width=4 * R, height=2.5 * R)
 ggsvg(P2, "f3_westeurasia2.svg", width=1.5 * R, height=1.25 * R)
 ggsvg(P5, "f3_westeurasia3.svg", width=1.5 * R, height=1.25 * R)
 }
-if(F){
+if(T){
 f2s = admixtools::read_f2("worldfoci2")
 pcmat = pca_from_f2s(f2s)
 pcs = pca_from_pcmat(pcmat)
 idx = "Mbuti"
-idy = "Sardinian"
+idy = "Surui"
 
 sample_list = pcs$pop 
 

### 2021-11-01 fstats_pca_lewontin.tex
other proj

--- 2021-10-29 fstats_pca_lewontin.tex
@@
 \linenumbers
-%\bibliographystyle{evolution.bst}
+\bibliographystyle{genetics}
 %\bibliographystyle{plainnat}
@@
 Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models.
-My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
+My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.d
 \end{abstract}
@@
 In the Results section, I will then discuss how some of the most common applications of $F$-statistics manifest themselves on a PCA, and illustrate them on two example data sets, before ending with a discussion.
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fstats_pca_vs_tree.pdf}
+	\caption{\textbf{Representation of $F$-statistics on trees and 2D-PCA-plots.} The schematics show four populations and their representation using a tree (top row) or a 2D-PCA plot (bottom row).
+A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space.
+B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthogonal projection of $X_1 - X_3$ onto $X_1$-$X_4$.
+C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or the orthogonal projection of $X_2 - X_3$ on $X_1 - X4$.
+D: $F_4(X_1, X_2; X_3, X_4)$ The two paths from $X_1$ to $X_2$ and $X_3$ and $X_4$ are non-overlapping in the tree, which corresponds to orthogonal vectors in PCA-space.}
+	\label{fig:geom}
+\end{figure}
 
@@
 Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.
-
-\begin{figure}[!ht]
-	\includegraphics[width=.5\textwidth]{figures/outgroup_f3.pdf}
-	\caption{\textbf{Outgroup-$F_3$-statistics}. The circle with a dot denotes a right angle.}
-	\label{fig:outgroupf3}
-\end{figure}
 
@@
 \begin{figure}[!ht]
-	\includegraphics[width=\textwidth]{figures/fstats_pca_vs_tree.pdf}
-	\caption{\textbf{Representation of $F$-statistics on trees and 2D-PCA-plots.} The schematics show four populations and their representation using a tree (top row) or a 2D-PCA plot (bottom row).
-A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space.
-B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthogonal projection of $X_1 - X_3$ onto $X_1$-$X_4$.
-C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or the orthogonal projection of $X_2 - X_3$ on $X_1 - X4$.
-D: $F_4(X_1, X_2; X_3, X_4)$ The two paths from $X_1$ to $X_2$ and $X_3$ and $X_4$ are non-overlapping in the tree, which corresponds to orthogonal vectors in PCA-space.}
-	\label{fig:geom}
-\end{figure}
-
-\begin{figure}[!ht]
 	\includegraphics[width=\textwidth]{figures/fstats_admixture_pca.pdf}
@@
 \paragraph{Example}
-As an example, I visualize the admixture statistic $F_3(X; \text{Basque}, \text{Turkish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A).
+As an example, I visualize the admixture statistic $F_3(X; \text{Sardinian}, \text{Finnish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A).
 In this case, the projected $n$-ball (light gray) and circle based on two dimensions (dark gray) have similar sizes.
 However, several populations that appear inside the circles (e.g.
-Sardinian, Finnish) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball.
-This reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure.
-Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
+Basque, Canary Islandersh) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball.
+This reveals that the first two PCs do not capture all the genetic variation relevant for  European population structure.
+Consequently, approximating $F_3$ by the first two or even ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
 
-However, many populations, particularly from Western Asia and the Caucasus, fall outside the circle.
-This allows us to immediately conclude that their $F_3$-statistics must be positive; and we should not consider them as a mixture between Basques and Turks.
+However, many populations, particularly from Western Asia and the Caucasus, on the right-hand side of the plot, fall outside the circle.
+This allows us to immediately conclude that their $F_3$-statistics must be positive; and we should not consider them as a mixture between Sardinians and Fins.
 
@@
 	\includegraphics[width=\textwidth]{figures/fig_data_europe.pdf}
-	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative.
+	\caption{\textbf{PCA and $F$-statistics for the Western Eurasian data set} A: PCA-biplot; the light grey circle denotes the region for which $F_3(X; \text{Sardinian}, \text{Finnish})$ may be negative, the dark circle is based on just the first two PCs.
 Populations for which $F_3$ is negative are colored in red.
-B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
-C,F: Contributions of PCs 1-10 to each $F_3$-statistic.
-D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$.
-The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population.
-}
+B: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum.
+C: Boxplot of contributions of PCs 1-10 to each $F_3$-statistic.
+D: Projection angle and correlation interpretation of $F_4(X, \text{Saudi}; \text{Sardinian}, \text{Finnish})$ based on two PCs (green), three PCs (blue) or full data (red).
+E: Contribution of the first ten PCs to select $F_4$-statistics, with the first three PCs containing the majority of contributions.}
 	\label{fig:f3}
@@
 
+\begin{figure}[!ht]
+	\includegraphics[width=.8\textwidth]{figures/outgroup_f3.pdf}
+	\caption{\textbf{Outgroup-$F_3$-statistics}.
+Interpretation of outgroup $F_3$-statistic on a tree (A) and PCA-plot (B).
+The red segment represents $F_3(X_O; X_U, X_3)$ and the dashed blue segment reflects $F_3(X_O; X_U, X_1)$ and $F_3(X_O; X_U, X_2)$, which have the same value.
+	}
+	\label{fig:outgroupf3}
+\end{figure}
 \subsection{Outgroup-$F_3$-statistics as projections}
@@
 
-To make sense of outgorup-$F_3$-statistics in the PCA context, I use the association of $F_3$-statistics to projections (Equation \ref{eq:proj}):
+To make sense of outgroup-$F_3$-statistics in the PCA context, I use the association of $F_3$-statistics to projections (Equation \ref{eq:proj}):
 On a PCA-plot, we can visualize this $F_3$-statistic as the projection of the vector $X_i - X_O$ onto $X_U - X_O$: 
@@
 Thus, the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in the example in Figure \ref{fig:outgroupf3}B this is $X_3$.
+
+\begin{figure}[!ht]
+	\includegraphics[width=\textwidth]{figures/fig_data_world.pdf}	
+	\caption{\textbf{PCA and $F$-statistics for the World data set} A: Visualization of Outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X)$ on a PCA-biplot.
+The color of points correspond to the value of the $F_3$ statistic, with brighter yellows indicating higher values, i.e.
+higher similarity to Sardinians.
+The $F_3$-projection axis is given by a black line, the projection of populations onto this axis by thin gray lines.
+In the full data space, these projection are orthogonal to the axis.
+B: Projection along the axis Sardinian-Mbuti (X-axis), and PCA on residual of this projection (PC1 on Y-axis, PC2 as coloring).
+C: Approximation of $F_3(\text{Mbuti}; \text{Sardinian}, X)$ using the first two (blue) and first ten (red) PCs, respectively.
+D: Contributions of first ten PCs to all statistics of the form $F_3(\text{Mbuti}; \text{Sardinian}, X)$.
+E: Contributions of the first ten PCs to select $F_4$-statistics.
+	}	
+	\label{fig:f4}
+\end{figure}
 
@@
 \paragraph{Example}
-In Figure \ref{fig:f3}D, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Gujarati}, X_i)$, in i.e.
-a statistic that aims to find the population most closely related to Gujarati (from Western India), assuming the Mbuti are an outgroup to all populations.
-On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Gujarati (black line).
+In Figure \ref{fig:f4}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, in i.e.
+a statistic that aims to find the population most closely related to Sardinian (a Mediterranean Island), assuming the Mbuti are an outgroup to all populations in the data set.
+On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Sardinians (black line).
 For each population, the projection is indicated with a grey line.
-In the full data space, this line is always orthogonal to the segment Mbuti-Gujarati, but on the plot (i.e.
+In the full data space, this line is always orthogonal to the segment Mbuti-Sardinian, but on the plot (i.e.
 the subspace spanned by the first two PCs), this is not necessarily the case.
 The coloring is based on the $F_3$-statistic calculated from all the data, with brighter values indicating higher $F_3$-statistics.
-In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.
-We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}E), but keep the relative ordering.
+In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas (cluster in the top left of the plot) project very close to orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.
+We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:fA}C), but keep the relative ordering.
 I also find that many PCs, e.g.
-PCs 3-5, 7 and 10 have almost zero contribution to all $F_3$-statistics (Figure \ref{fig:f3}F), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.
+PCs 3-5, 7 and 10 have almost zero contribution to all $F_3$-statistics (Figure \ref{fig:f3}D), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.
 
 \subsection{$F_4$-statistics as angles}
-The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct.
+One interpretation of $F_4$ on PCA plots is similar to that of $F_3$; as a projection of one vector onto another, with the difference that now all four points may be distinct.
 $F_4$-statistics that correspond to a branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot (Figure \ref{fig:geom}G), again with the caveat that we need to scale it by a constant.
-If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, or that the two populations map to the same point (Figure \ref{fig:geom}H).
-In the case of an admixture graph, this is no longer the case: Population $X_x$ in Figure \ref{fig:admix}D does \emph{not} map to the same point as $X_1$ or $X_2$ do, implying that statistics of the form $F_4(X_1, X_x; X_3, X_4) \neq 0$.
+If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e.
+that $X_1$ and $X_2$  map to the same point on the projection axis $\overline{X_3X_4}$ (Figure \ref{fig:geom}H).
+In the case of an admixture graph, this is no longer the case: Both population $X_y$ and $X_x$ in Figure \ref{fig:admix}D do \emph{not} map to the same point as $X_1$ or $X_2$ do, implying that statistics of the form $F_4(X_1, X_x; X_3, X_4) \neq 0$.
 
@@
 It also suggests a number of diagnostic $F$-statistics that check assumptions; for example if the tree holds, then
-$F_4(X_3, X_3'; X_4, X_4') = 0 $.
 
@@
 \begin{equation}
-\text{Cor}(X_1 - X_2, X_3 - X_4) =  \frac{\langle X_1 - X_2, X_3 - X_4 \rangle}{\norm{X_1-X_2}\norm{X_3-X_3}} = \cos(\phi),\label{eq:angle}
+\text{Cor}(X_1 - X_2, X_3 - X_4) =  \frac{F_4( X_1, X_2; X_3, X_4) }{\norm{X_1-X_2}\norm{X_3-X_3}} = \cos(\phi),\label{eq:angle}
 \end{equation}
@@
 \paragraph{Example}
-To illustrate the angle interpretation I again use the Western Eurasian data.
-The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:f3}A), a European gradient (from Sardinian to Chuvash), and a Asian cline (from Arab to Caucasus populations).
-This is quantified in Figure \ref{fig:f4}A, where I plot the angle corresponding to $F_4(X, \text{Sardinian}; \text{Saudi}, \text{Georgian})$.
-For most European populations, using two PCs (green points) gives an angle close to zero, corresponding to a correlation coefficient between the two clines of $r>0.9$.
-Just adding PC3 (blue), however, shows that the parallelism of the clines is spurious: Using three PCs or the full data (red) shows that most correlations are low.
-I arrive at a similar interpretation from the spectrum of these statistics (Figure \ref{fig:f4}B), which has high loadings for the first three PCs, with minimal contributions from the higher ones.
+To illustrate the angle interpretation I return to the Western Eurasian data.
+The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:f3}A), a European gradient (from Sardinian to Finnish and Chuvash), and a Asian cline from Arab populations (top right) to the Caucasus  (bottom right).
+This is quantified in Figure \ref{fig:f3}D, where I plot the angle corresponding to $F_4(X, \text{Saudi}; \text{Sardinian}, \text{Finnish})$.
+For most Asian populations, using two PCs (green points) gives an angle close to zero, corresponding to a correlation coefficient between the two clines of $r>0.9$.
+Just adding a third PC (blue), however, shows that the clines are not, in fact, parallel, and the correlation for most populations is low.
+The finding that three PCs are necessary to explain this data can also be seen from the spectrum of these $F_4$-statistics (Figure \ref{fig:f3}E), which have high contributions only from the first three PCs.
 
@@
 So far, I used eq.
-\ref{eq:fpc} to interpret $F$-statistics on a PC-plot, but the argument holds for \emph{any} orthonormal projection of the data space.
+\ref{eq:fpc} to interpret $F$-statistics on a PCA-plot, but the argument holds for \emph{any} orthonormal projection of the data space.
 This is useful in particular for estimates of admixture proportions, which are often done in a small reference space \citep{patterson2012, petr2019, harney2021, oteo-garcia2021}.
@@
 which can be interpreted as projecting $X-A$ and $B-A$ onto $R_1 - R_2$ and measuring their relative proportions \citep{oteo-garcia2021}.
+One important assumption is that there is no gene flow between $X$ and the reference populations after gene flow \citep{petr2019}.
+One way to verify this is by inspecting the \emph{null space} of the projection, i.e.
+all vectors orthogonal to $R_1 - R_2$.
+One way this should manifest is that $A, B$ and $X$ should be colinear to each other on a PCA plot.
+
 \texttt{qpAdm} extends this approach to a higher-dimensional reference space and multiple potential source populations.
 One open practical question in many applications is which reference and putative source populations to use \citep{harney2021}.
+
 The theory developed here suggests some possible visualizations that may address this issue.
@@
 \subsubsection{Example}
-In the PCA on the world overview data set, I found a gradient from Africans to Europeans (Figure \ref{fig:f3}D).
+Let as use the World data set (Figure \ref{fig:f4}A), and use Sardinian and Mbuti as references for admixture estimation.
+One way we can do that is to project all populations $X$ onto this axis in the data (Figure \ref{fig:f4}B), so that the distance along the $X$-axis between any pairs of populations $X_1, X_2$ is proportional to $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$.
+
+On the Y-axis, we may now plot the first PC of the Null Space of this projection, i.e.
+the variation that is not along the axis from Sardinians to Mbuti.
+This variation largely follows an East-West gradient across Eurasia, with Siberian, East Asian and American populations having the lowest values (Figure \ref{fig:f4}B).
+
+Populations suitable for $F_4$-ratios should now be colinear on such a plot.
+One example are the Tubalar, which appear roughly intermediate between the Kalash (from Pakistan) and the Surui (from Brazil)
+ found a gradient from Africans to Europeans (Figure \ref{fig:f3}D).
 I focus on this cline using an alternative projection by using  $F$-statistics of the form $F_4(X, Y; \text{Sardinian}, \text{Yoruba})$), which might e.g.
@@
 These $F_4$-statistics are very well-approximated by the first two PCs, with a 99.2\% correlation between $F_4$ and its approximation using the first two PCs (Figure \ref{fig:f4}C).
-
-In Figure \ref{fig:f4}D, I show the projection $\langle X; \text{Sardinian}, \text{Yoruba} \rangle$ on the $X$-axis, which means that the horizontal difference between any pair of population is proportional to their $F_4$-statistic relative to Sardinians and Yorubans.
-We can also ask what variation is not represented by performinc a PCA on the residual of this projection, the first two residual PCs are given on the Y-axis and in the coloring.
-This visualization reveals that variation within Africans (with Mbuti, Biaka and Ju$\vert$'hoansi, top right) and the variation in East Asians and Americans are largely orthogonal to this projection axis, and so Sardinians and Yoruba would be poor references if we were interested in studying East Asian genetic variation.
-
-The percentage of between-population variance explained by the Sardinia-Yoruba axis (24\%) is much lower than that of the first PC (40\%, Figure \ref{fig:f4}E).
-However, the cumulative variance explained by the first two axes is similar, with (52\%) explained when adding residual PC1 to the projection, compared to 55\% for the first two PCs.
-The advantage of specifying one axis is that it displays the orthogonal components more explicitly, reveals distinct structure in Africans and non-Africans and thus can be used to test assumptions of more complex models.
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figures/fig_data_world.pdf}	
-	\caption{\textbf{PCA and $F_4$-statistics} A: Projection angle and correlation coefficient $r$ representation of $F_4(X, Sardinian; Saudi, Georgian)$ (red) in the Western Eurasian data set, and approximations using two (green) and three (blue) PCs.
-	B: Spectrum of select $F_4$-statistics in the Western Eurasian data set.
-C: Spectrum of $F-4$-statistics in World data set.
-D: Scatterplot of $F_3$-projection on Sardinian-Yoruba axis  and residual PC1.
-	E: Percent variance explained from of the projection based on $F_3$ in panel D and first nine residual PCs (light gray), compared with percent variance explained by first ten PCs (dark gray).
-	}
-\label{fig:f4}
-\end{figure}
 
@@
 \newpage
-\bibliographystyle{plainnat}
 \bibliography{main}

### 2021-11-04 fstats_pca_lewontin.tex
discussion

--- 2021-11-01 fstats_pca_lewontin.tex
@@
 Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA-plot.
+
+Previous interpretation of PCA in the context of population genetic models have focused on explicit models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{francois2021}.
+My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs.
+One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA.
+However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis.
+From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.
+
+\subsection{more PCs}
 This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
@@
 
+\subsection{Assumptions}
 The other cause for disagreements between $F$-statistics and PCA are differences in assumptions.
@@
 
+\subsubsection{missing data}
 The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}.
@@
 
+\subsubsection{normalization}
 The normalization of SNPs is similarly a matter of convention.
@@
 The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}.
-As genetic differentiation between human populations is low, the  normalization used may matter little in practice \citep{felsenstein1973}, but using consistent normalization between $F$-statistics and PCA-based analysis seems advisable since it removes normalization as a source of errors.
-.
+As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but should be explored in future work \citep{felsenstein1973}.
 
-The third issue of individual-based vs population-based analysis is similarly a matter of interpretation.
-For $n$ samples, the number of possible $F$-statistics is on the order of $n^4$, and so the number of statistics is kept low by grouping individuals into populations.
-This is, however, not necessary, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
-Here, individual-based PCA can provide some guidance for grouping samples: Since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on a PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.
+\subsubsection{estimated vs. observed allele frequencies}
+The third difference between $F$-statistics and standard  PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
+The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
+Probabilistic PCA is one class of approaches that aim to separate the population structure from sampling noise \citep[e.g.][]{agrawal2020}.
+It seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.
 
-The final difference between the $F$-statistic-based PCA used here and individual-based PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
-The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
-Probabilistic PCA is one class of approaches that aim to separate the population structure from individual-level noise \citep{agrawal2020}, and it seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.
+\subsubsection{individual vs. population-based analyses}
+The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.
+This is not necessary; population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
+Thus, more commonly the choice between individual-based and population-based approaches is in the goal of an analysis.
+Grouping individuals into populations introduces a major assumption, that can e.g.
+be justified using an individual-based PCA.
+In particular, since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on an individual-based PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.
 
-Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions that are partially arbitrary, and partially  explained by the focus of PCA on exploratory data analysis.
-Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if PCA and $F$-statistics are used in a way such that the results are comparable to each other.
+\subsubsection{summary}
+Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions, or due to different goals for a particular analysis.
+Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if  conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.
 
-Previous interpretation of PCA in the context of population genetic models have focused on simple models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{franccois2021}.
-My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs.
-One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA.
-However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis.
-From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.
+\subsection{broader scope}
 

### 2021-11-04 data/world_map_talks.R

@@ -1,0 +1,71 @@
+library("rnaturalearth")
+library("rnaturalearthdata")
+library(ggrepel)
+world <- ne_countries(scale = "medium", returnclass = "sf")
+source("fscripts.R")
+library(ggpubr)
+BTHEME = theme_classic() + theme(legend.position="none")
+
+
+anno = read_tsv("v44.3_HO_public.anno") %>%
+    select(iid=2,
+           age=6,
+           pop=8,
+           lat=11,
+           long=12,
+           src=13,
+           cov=14) %>%
+    mutate(lat = as.numeric(lat),
+           long=as.numeric(long),
+           age = as.numeric(age),
+           cov = as.numeric(cov))
+
+ind_eu = read_table("subdata/westeurasian1.ind", col_names=c("iid", "sex", "pop"))
+ind_world = read_table("subdata/worldfoci2.ind", col_names=c("iid", "sex", "pop"))
+
+pops_eu = ind_eu %>% left_join(anno) %>% select(pop, lat, long) %>% 
+    distinct() %>%
+    group_by(pop) %>% summarize_all(mean)
+pops_world = ind_world %>% left_join(anno) %>% select(pop, lat, long) %>% distinct() %>%
+    group_by(pop) %>% summarize_all(mean)
+
+
+#mapp
+
+plot_map_eu <- function(){
+    P1 = ggplot() +
+        geom_sf(data=world, color=NA) +
+        theme_minimal() +
+        geom_point(data=pops_eu, aes(x=long, y=lat)) +
+        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+        theme(legend.position = 'none') +
+        xlab(NULL) + ylab(NULL) +
+        coord_sf(xlim=range(pops_eu$long,na.rm=T), ylim=range(pops_eu$lat, na.rm=T)) +
+        scale_color_viridis_c(trans='log')
+}
+
+P1 = plot_map_eu() + 
+    geom_text_repel(data=pops_eu, aes(x=long, y=lat, label=pop), 
+                    size=2, segment.color='darkgrey') 
+ggsave("figures/sample_map_eu.png", P1, width=4, height=3)
+ggsave("figures/sample_map_eu2.png", plot_map_eu(), width=4, height=3)
+
+plot_map_world <- function(){
+    P2 = ggplot() +
+        geom_sf(data=world, color=NA) +
+        theme_minimal() +
[…]

### 2021-11-04 fstats_pca_lewontin.tex
better discussion

@@
 
-Previous interpretation of PCA in the context of population genetic models have focused on explicit models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{francois2021}.
-My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs.
-One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA.
-However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis.
-From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.
-
 \subsection{more PCs}
@@
 
+\subsection{Previous Interpretations}
+Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2021}.
+My interpretation here is different in that the utility of PCA is to simplify the geometry of the data.
+One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA.
+However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis.
+From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.
+
+The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
+On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
+Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.
+
 \subsection{broader scope}
@@
 As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
-However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}.
-This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow.
-In my example (Figure \ref{fig:f3}A), most Southern European populations are ``admixed'' between Basques and Turkish, but a more accurate model might be one of continuous variation where Basque and Turkish lie on one of multiple gradients; which is more directly visualized with PCA.
+However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}.
+This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness.
+A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians.
+In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.
+
+This trade-off between $F$-statistics, which allow for significance teste tension between
+
 There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}.

### 2021-11-04 fstats_pca_lewontin.tex

--- 2021-11-01 fstats_pca_lewontin.tex
@@
 Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA-plot.
+
+\subsection{more PCs}
 This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
@@
 The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}.
-As genetic differentiation between human populations is low, the  normalization used may matter little in practice \citep{felsenstein1973}, but using consistent normalization between $F$-statistics and PCA-based analysis seems advisable since it removes normalization as a source of errors.
-.
+As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but should be explored in future work \citep{felsenstein1973}.
 
-The third issue of individual-based vs population-based analysis is similarly a matter of interpretation.
-For $n$ samples, the number of possible $F$-statistics is on the order of $n^4$, and so the number of statistics is kept low by grouping individuals into populations.
-This is, however, not necessary, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
-Here, individual-based PCA can provide some guidance for grouping samples: Since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on a PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.
+\subsubsection{estimated vs. observed allele frequencies}
+The third difference between $F$-statistics and standard  PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
+The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
+Probabilistic PCA is one class of approaches that aim to separate the population structure from sampling noise \citep[e.g.][]{agrawal2020}.
+It seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.
 
-The final difference between the $F$-statistic-based PCA used here and individual-based PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
-The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
-Probabilistic PCA is one class of approaches that aim to separate the population structure from individual-level noise \citep{agrawal2020}, and it seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.
+\subsubsection{individual vs. population-based analyses}
+The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.
+This is not necessary; population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
+Thus, more commonly the choice between individual-based and population-based approaches is in the goal of an analysis.
+Grouping individuals into populations introduces a major assumption, that can e.g.
+be justified using an individual-based PCA.
+In particular, since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on an individual-based PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.
 
-Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions that are partially arbitrary, and partially  explained by the focus of PCA on exploratory data analysis.
-Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if PCA and $F$-statistics are used in a way such that the results are comparable to each other.
+\subsubsection{summary}
+Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions, or due to different goals for a particular analysis.
+Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if  conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.
 
-Previous interpretation of PCA in the context of population genetic models have focused on simple models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{franccois2021}.
-My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs.
-One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA.
+\subsection{Previous Interpretations}
+Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2021}.
+My interpretation here is different in that the utility of PCA is to simplify the geometry of the data.
+One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA.
 However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis.
@@
 
+The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
+On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
+Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.
+
+\subsection{broader scope}
+
 As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
-However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}.
-This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow.
-In my example (Figure \ref{fig:f3}A), most Southern European populations are ``admixed'' between Basques and Turkish, but a more accurate model might be one of continuous variation where Basque and Turkish lie on one of multiple gradients; which is more directly visualized with PCA.
+However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}.
+This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness.
+A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians.
+In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.
+
+This trade-off between $F$-statistics, which allow for significance teste tension between
+
 There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}.

### 2021-11-08 fstats_pca_lewontin.tex
update formating, move to biblatex

@@
 \usepackage{comment}
-\usepackage[authoryear, comma]{natbib}
 \usepackage{lineno}
@@
 
+%\usepackage[authoryear, comma]{natbib}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{mbe}
+\usepackage[style=authoryear,
+			natbib=true,
+			backend=biber,
+			maxbibnames=10,
+			maxcitenames=2,
+			uniquename=false,
+			uniquelist=false,
+			isbn=false]{biblatex}
+\addbibresource{main.bib}
+%biblatex stuff
+\DeclareFieldFormat[article]{citetitle}{#1}
+\DeclareFieldFormat[article]{title}{#1} 
+\renewbibmacro{in:}{}
+\DeclareFieldFormat{pages}{#1}
+\renewcommand{\labelnamepunct}{\addcolon\space}
+\AtEveryBibitem{\clearlist{language}}
+\AtEveryBibitem{\clearfield{month}}
+\AtEveryCitekey{\clearfield{month}}
+\AtEveryBibitem{\ifentrytype{article}{
+		\clearfield{url}
+		\clearfield{urlyear}
+	    \clearfield{language}}{}}
+\AtEveryBibitem{\ifentrytype{book}{
+		\clearfield{url}
+		\clearfield{urlyear}
+		\clearfield{language}}{}}
+
+	\renewbibmacro*{volume+number+eid}{%
+		\printfield{volume}%
+		%  \setunit*{\adddot}% DELETED
+		\setunit*{\addnbspace}% NEW (optional); there's also \addnbthinspace
+		\printfield{number}%
+		\setunit{\addcomma\space}%
+		\printfield{eid}}
+	\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
+
 \linenumbers
-\bibliographystyle{genetics}
-%\bibliographystyle{plainnat}
 
@@
 Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models.
-My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.d
+My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
 \end{abstract}
 \section{Introduction}
-As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2020}.
+As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017}.
 In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.
@@
 In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}.
-In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985a}.
+In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985}.
 However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}.
 In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}.
-Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2019}.
+Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2018}.
 
@@
 In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas (cluster in the top left of the plot) project very close to orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.
-We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:fA}C), but keep the relative ordering.
+We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}C), but keep the relative ordering.
 I also find that many PCs, e.g.
@@
 \subsubsection{missing data}
-The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}.
+The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
 Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}.
@@
 \newpage
-\bibliography{main}
+%\bibliography{main}
+\printbibliography
 

### 2021-11-09 fstats_pca_lewontin.tex
reorganize discussion

@@
 \section{Introduction}
-As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017}.
+As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}.
 In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.
 
-The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a}.
+The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a, gopalan2022}.
 In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}.
@@
 In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}.
-Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2018}.
+Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{gopalan2022}.
 
@@
 
+\subsection{Previous Interpretations}
+Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2010, francois2021}.
+My interpretation here is different in that the utility of PCA is to simplify the geometry of the data.
+One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA.
+However, a very skewed sampling distribution will increases the likelihood that more, or different PCs will have to be included in the analysis.
+
+The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
+On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
+Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.
+
 \subsection{more PCs}
-This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
+My focus on the geometry of the data allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
 As PCA is often ran in an early step in data analysis, this may aid in generation of hypotheses that can be more directly evaluated using generative models, typically using a lower number of populations.
@@
 \subsubsection{summary}
-Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions, or due to different goals for a particular analysis.
-Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if  conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.
+The version of PCA used here differs from that proposed by \cite{patterson2006}, and thus some care will be required to directly extend the interpretations developed here to individual-based PCAs.
+However, the differences are largely due to conventions, and particularly for studies where the description of population structure is a major focus, results might be easier to interpret if conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.
 
-\subsection{Previous Interpretations}
-Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2021}.
-My interpretation here is different in that the utility of PCA is to simplify the geometry of the data.
-One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA.
-However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis.
-From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.
+\subsection{The Apportionment of Human Diversity}
+Most genetic variation in humans is shared between all of us, but the around 15\% that can be explained by population structure can be leveraged to study our history and diversity in great detail \citep{dobzhansky1972, cavalli-sforza1964, reich2018a}.
+For some data sets it is possible to predict an individuals' origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
 
-The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
-On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
-Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.
+However, understanding, conceptualizing and modeling this variation is far from trivial, and misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies.
+Lewontin’s landmark 1972 paper on the apportionment of human genetic diversity was one of the first to quantify how little of between-population genetic variation could be attributed to ``racial'' continental-scale groupings.
+Over the last five decades, this view has been corroborated, refined and extended many times \citep{cann1987, cavalli-sforza1994, barbujani1997, rosenberg2002a}.
 
-\subsection{broader scope}
+From a practical perspective, formulating hypotheses and designing studies in terms of discrete populations with ``uniform'' genetic backgrounds is often sensible, as it enables  e.g.
+prediction of phenotypes \citep{berg2019, yair2021}, inference of demographic parameters and schematic models of human genetic history \citep{patterson2012}.
+This is also the case for $F$-statistics, which are motivated by trees.
+They assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
+However, these simplification do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of deemphasizing people that do not rigidly fall into predefined genetic groups.
 
-As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
  However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}.
@@
 In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.
-
-This trade-off between $F$-statistics, which allow for significance teste tension between
 
@@
 Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.
+
+\subsubsection{cost of simplificatio}
+
+However, these simplification do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of excluding people that do not rigidly fall into predefined genetic groups.
+By framing admixture in the continuous context of PCA, the approach developed here shows one way how this trade-off between interpretability and accuracy can be mitigated.
 

### 2021-11-10 fstats_pca_lewontin.tex
discussion

@@
 
- However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}.
+In many parts of the world, and particularly at more local scales, distinctions between populations begin to blur, and everyone could be considered  admixed to some degree \citep{pickrell2014}.
 This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness.
@@
 
-There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}.
+This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}.
 One issue with these approaches is that they are usually restricted to at most a few dozen populations.
@@
 However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits.
-One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs.
-However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}.
-The PCA-based  interpretation offers an alternative that trades interpretability for robustness.
+The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as we do not need
+
+.
 Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow.

### 2021-11-10 fstats_pca_lewontin.tex
ratio fig labels

@@
 \ref{eq:fpc} to interpret $F$-statistics on a PCA-plot, but the argument holds for \emph{any} orthonormal projection of the data space.
-This is useful in particular for estimates of admixture proportions, which are often done in a small reference space \citep{patterson2012, petr2019, harney2021, oteo-garcia2021}.
+This is useful in particular for estimates of admixture proportions, which are often done as projections into a low-dimensional reference space defined by $F$-statistics \citep{patterson2012, petr2019, harney2021, oteo-garcia2021}.
 
-The simplest approach is the $F_4$-ratio to infer the admixture sources of population $X$ as 
+For example, a common way to estimate admixture proportion is the $F_4$-ratio:
 \begin{eqnarray}
-\alpha = \frac{F_4(R_1, R_2; X, A)}{F_4(R_1, R_2; B, A)} = \frac{\vectorproj[R_1 - R_2]{X - A}}{\vectorproj[R_1 - R_2]{B - A}} \label{eq:f4ratio},
+\alpha = \frac{F_4(R_1, R_2; X_X, X_1)}{F_4(R_1, R_2; X_2, X_1)} = \frac{\vectorproj[R_1 - R_2]{X_X - X_1}}{\vectorproj[R_1 - R_2]{X_2 - X_1}} \label{eq:f4ratio},
 \end{eqnarray}
-which can be interpreted as projecting $X-A$ and $B-A$ onto $R_1 - R_2$ and measuring their relative proportions \citep{oteo-garcia2021}.
-One important assumption is that there is no gene flow between $X$ and the reference populations after gene flow \citep{petr2019}.
-One way to verify this is by inspecting the \emph{null space} of the projection, i.e.
-all vectors orthogonal to $R_1 - R_2$.
-One way this should manifest is that $A, B$ and $X$ should be colinear to each other on a PCA plot.
+which can be interpreted as projecting $X_X-X_1$ and $X_2-X_1$ onto $R_1 - R_2$ and the ratio of the lengths gives the proportion of $X_x$ contributed by $X_1$  \citep{oteo-garcia2021}.
 
-\texttt{qpAdm} extends this approach to a higher-dimensional reference space and multiple potential source populations.
-One open practical question in many applications is which reference and putative source populations to use \citep{harney2021}.
+The admixture graph motivating this statistic is visualized in Figure \ref{fig:ratio}A, and the PCA-like interpretation in Figure \ref{fig:ratio}B.
+In both panels, the solid gray lines is the projection axis, and the dotted lines give the residual, i.e.
+the branches or genetic variation that is ignored by the projection.
 
-The theory developed here suggests some possible visualizations that may address this issue.
+The PCA-like projection can be used to visualize admixture proportions, as the horizontal position of $X_X$ relative to $X_1$ and $X_2$ (red dashed line vs black line) directly represents the estimated admixture proportion $\alpha$.
+In addition, the residuals can be used to verify assumptions of the admixture graph model.
+In particular, since $X_X$ arises as a linear combination of $X_1$ and $X_2$, if admixture is recent we might expect the three populations to be collinear; if they are not this means that either of the populations experienced gene flow from some other population which might bias results \citep{petr2019}.
+
+In addition, the external tree branches $X_1 - X'_1$ and $X_2 - X'_2$ are disjoint which means they should be orthogonal.
+One a one-dimensional residual plot (Figure \ref{fig:ratio)B) this can not be verified, but the statistic
+\begin{equation}
+F_4(X_1, X'_1; X_2, X'_2) = 0
+\end{equation}
+can be calculated for all samples.
 
 \subsubsection{Example}
-Let as use the World data set (Figure \ref{fig:f4}A), and use Sardinian and Mbuti as references for admixture estimation.
-One way we can do that is to project all populations $X$ onto this axis in the data (Figure \ref{fig:f4}B), so that the distance along the $X$-axis between any pairs of populations $X_1, X_2$ is proportional to $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$.
-
-On the Y-axis, we may now plot the first PC of the Null Space of this projection, i.e.
-the variation that is not along the axis from Sardinians to Mbuti.
-This variation largely follows an East-West gradient across Eurasia, with Siberian, East Asian and American populations having the lowest values (Figure \ref{fig:f4}B).
-
-Populations suitable for $F_4$-ratios should now be colinear on such a plot.
-One example are the Tubalar, which appear roughly intermediate between the Kalash (from Pakistan) and the Surui (from Brazil)
- found a gradient from Africans to Europeans (Figure \ref{fig:f3}D).
-I focus on this cline using an alternative projection by using  $F$-statistics of the form $F_4(X, Y; \text{Sardinian}, \text{Yoruba})$), which might e.g.
-be used in an $F_4$-ratio.
-These $F_4$-statistics are very well-approximated by the first two PCs, with a 99.2\% correlation between $F_4$ and its approximation using the first two PCs (Figure \ref{fig:f4}C).
+I use the World data set as an example, using Sardinian and Mbuti as references populations (Figure \ref{fig:f4}B).
+The data is the same as in  the PCA (Figure \ref{fig:f4}A), but it is now rotated such that the axis between the reference population (black line in Figure \ref{fig:f4}A) is now reflected by the x-axis.
+For any pair of populations $X_1$ $X_2$, their horizontal distance reflects $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$ and the relative horizontal distance corresponds exactly to $F_4$-ratio admixture estimates.
+For many sets of populations, this is of course not sensible, and just looking at the first PC of the residual shows many examples where the populations are not collinear.
+For example, on the x-axis, the South American Surui are between Papuans and Georgians, but since the Surui clearly are not on the line between Papuans and Georgians, this cannot be the result of admixture.
 

### 2021-11-10 fstats_pca_lewontin.tex
all changes

@@
 The finding that three PCs are necessary to explain this data can also be seen from the spectrum of these $F_4$-statistics (Figure \ref{fig:f3}E), which have high contributions only from the first three PCs.
+\begin{figure}[!ht]
+	\includegraphics[width=.7\textwidth]{figures/fig_f4_ratio.pdf}	
+	\caption{\textbf{Admixture proportion estimates} A: Visualization of the admixture graph scenario used to estimate the ancestry proportion $\alpha$ contributed from $X_1$ to $X_X$, using references $R_1$ and $R_2$.
+The full grey line corresponds to the projection axis, and the dotted grey lines corresponds to the branches ignored in the projections.
+The admixture proportion $\alpha$ corresponds to the length of the dashed red line relative to the black line between $X_1$ and $X_2$.
+B: The same scenario, but in Euclidean space, $X_1$, $X_2$ and $X_X$ align on a line both in the (low-dimensional approximation of the) residual space and on the projection axis.
+	}	
+	\label{fig:ratio}
+\end{figure}
 
@@
 In addition, the external tree branches $X_1 - X'_1$ and $X_2 - X'_2$ are disjoint which means they should be orthogonal.
-One a one-dimensional residual plot (Figure \ref{fig:ratio)B) this can not be verified, but the statistic
+One a one-dimensional residual plot (Figure \ref{fig:ratio}B) this can not be verified, but the statistic
 \begin{equation}
@@
 
-\subsection{Previous Interpretations}
+\subsection{The Geometry of Admixture}
 Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2010, francois2021}.
@@
 
-\subsection{more PCs}
 My focus on the geometry of the data allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots.
@@
 
-\subsubsection{missing data}
+\subsubsection{Missing data}
 The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
@@
 
-\subsubsection{normalization}
+\subsubsection{Normalization}
 The normalization of SNPs is similarly a matter of convention.
@@
 
-\subsubsection{estimated vs. observed allele frequencies}
+\subsubsection{Estimated vs. observed allele frequencies}
 The third difference between $F$-statistics and standard  PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
@@
 
-\subsubsection{individual vs. population-based analyses}
+\subsubsection{Individual vs. population-based analyses}
 The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.
@@
 
-\subsubsection{summary}
+\subsubsection{Summary}
 The version of PCA used here differs from that proposed by \cite{patterson2006}, and thus some care will be required to directly extend the interpretations developed here to individual-based PCAs.
@@
 However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits.
-The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as we do not need
-
-.
-Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow.
-Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.
-
-\subsubsection{cost of simplificatio}
-
-However, these simplification do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of excluding people that do not rigidly fall into predefined genetic groups.
-By framing admixture in the continuous context of PCA, the approach developed here shows one way how this trade-off between interpretability and accuracy can be mitigated.
+The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not to assume that gene flow is rare, and visualizing both the projection used for estimating admixture proportion, and the residual space, can be used to highlight scenarios where the assumptions are violated.
 

### 2021-11-11 fstats_pca_lewontin.tex
supp files, updated fig

@@
 I illustrate my results on two examples, one of local, and one of global human diversity.
-In both examples, I find that population structure is sparse, and only a few components contribute to most statistics.
+In both examples, I find that just using the first few PCs provides good population structure is sparse, and only a few components contribute to most statistics.
 Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models.
@@
 
+\section{Supplementary File Description}
+Both supplementary files give the individual-identifier, sex and population label for each sample.
+
 \newpage

### 2021-11-11 data/f4_fig.R

@@ -42,10 +42,10 @@ KA = KA %>% as.data.frame %>% mutate(pop = rownames(B))
 X = KA %>% select(pop, A=V1) %>% left_join(K)
 
 P3 = X %>% 
-    #mutate(V2 = ifelse(pop %in% c(idx, idy), NA, V2)) %>%
-    #ggplot(aes(x=A, y=V1, label=pop, color=V2)) + 
-    mutate(A = ifelse(pop %in% c(idx, idy), NA, A)) %>%
-    ggplot(aes(x=V1, y=V2, label=pop, color=A)) + 
+    mutate(V2 = ifelse(pop %in% c(idx, idy), NA, V2)) %>%
+    ggplot(aes(x=A, y=V1, label=pop, color=V2)) + 
+    #mutate(A = ifelse(pop %in% c(idx, idy), NA, A)) %>%
+    #ggplot(aes(x=V1, y=V2, label=pop, color=A)) + 
     geom_hline(yintercept=X$V1[X$pop == idx], color='lightgray') +
     geom_text_repel() + 
     geom_point() +

### 2021-11-11 data/f3_fig.R
updated figures

@@ -6,7 +6,7 @@ library(glue)
 R = 1.5
 BTHEME = theme_classic() + theme(legend.position="none")
 
-if(F){
+if(T){
 f2s = admixtools::read_f2("westeurasian1/")
 pcmat = pca_from_f2s(f2s)
 pcs = pca_from_pcmat(pcmat)
@@ -71,7 +71,7 @@ f2s = admixtools::read_f2("worldfoci2")
 pcmat = pca_from_f2s(f2s)
 pcs = pca_from_pcmat(pcmat)
 idx = "Mbuti"
-idy = "Surui"
+idy = "Sardinian"
 
 sample_list = pcs$pop 
 

### 2021-11-11 data/f4_fig.R

@@ -8,7 +8,7 @@ if(T){
     f2s = admixtools::read_f2("worldfoci2")
     pcmat = pca_from_f2s(f2s)
     pcs = pca_from_pcmat(pcmat)
-    idx = "Surui"
+    idx = "Sardinian"
     idz = "Han"
     idy = "Mbuti"
 ids = c(idx, idy)

### 2021-11-15 fstats_pca_lewontin.tex
updated figures
submission before journal specific changes

@@
 \begin{abstract}
-Human  genetic diversity is shaped by our complex history.
-Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
-Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches.
+Human  genetic diversity is shaped by our complex history. Data-driven methods such as Principal Component Analysis (PCA) are an important population genetic tool to understand this method. 
+Here, I contrast PCA with a set of statistics motivated by trees ($F$-statistics).
+Here, I show that these two methods are closely related, and I derive explicit connections between the two approaches.
 I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link.
@@
 
-\paragraph{F-statistics}
+%\paragraph{F-statistics}
 However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult.
@@
 The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}.
-Formal definition will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches.
+Formal definitions will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches.
 \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.
@@
 The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A).
-However, admixture graphs are not the only plausible alternative model, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.
+However, admixture graphs are not the only plausible alternative models, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.
 
@@
 The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A.
-Here, the *typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$.
+Here, the (typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$.
 Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe.
@@
 An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare.
-Recently, \cite{oteo-garcia2021} showed re-deriving $F$-statistics in a geometric framework, showing that these assumptions are not necessary.
+Recently, \cite{oteo-garcia2021}  re-derived $F$-statistics in a geometric framework, showing that these assumptions are not necessary.
 Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$.
@@
 \end{subequations}
-where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
+where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to the vector projection
 \begin{equation}
@@
 Thinking of $F$-statistics as projections also holds on trees: In e.g.
-a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:admix}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$.
-The external branches are independent populations, and so they are expected to drift orthogonally to each other.
+a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:geom}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$.
+On trees, all disjoint paths are independent (i.e.
+orthogonal) from each other, and thus the external branches vanish under the projection.
 
@@
 PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g.
-in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis.
+in how SNPs are standardized, how missing data is treated, what algorithm is used, or whether we use individuals or populations as units of analysis.
 The version of PCA I use here is set up such that the similarities to $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets.
-In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}.
+In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies.
+In contrast, many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}.
 The differences this causes will be addressed in the discussion.
@@
 If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
-Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
-of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.
+Since the allele frequencies are between zero and one, we can interpret each population $X_i$
+of $\MX$ as a point in $\mathbb{R}^S$.
 
-One way PCA can be motivated is that it aims to find a $K$-dimensional subspace of the data space that retains most  variation in the data.
-$K$ is at most $n-1$, in which case the data is simply rotated.
+PCA allows us to approximate the points in the high-dimensional allele frequency space by a $K$-dimensional subspace of the  data.
+If all PCs are considered, $K=n-1$, in which case the data is simply rotated.
 However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.
@@
 SNPs with missing data in any population are excluded.
-The code used to create all figures and analyses will be available on \url{https://github.com/BenjaminPeter/fstats_pca}.
+The code used to write this paper, create all figures and analyses is  available on \url{https://github.com/BenjaminPeter/fstats_pca}.
 
@@
 This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
-It is motivated by the analysis of \cite{lazaridis2014}, who used it as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}.
+\cite{lazaridis2014} used this data set as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}.
 Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}.
@@
 However as $F_2$-statistics are estimates, some eigenvalues might be slightly negative, which would lead to imaginary PCs.
-I avoid this by using the \texttt{nearPD}-function in R that ensures all eigenvalues have the correct sign.
+I avoid this by setting all negative eigenvalues to zero.
 
@@
 Thus, throughout the schematic figures, I assume that two PCs are sufficient to characterize population structure.
-In the data applications I evaluate how deviations of this assumption my manifest themselves in PCA plots.
+In the data applications I evaluate how deviations of this assumption may manifest themselves in PCA plots.
 
@@
 populations.
-On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations.
+On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations (Figure \ref{fig:geom}E).
 In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will fall close to each other on a PCA-plot, and have low pairwise $F_2$-statistics.
@@
 \ref{eq:fpc}), the distance on a PCA-plot will always be an underestimate of the full $F_2$-distance.
-Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations.
+Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations, and likely more PCs need to be investigated to understand how these populations are related to each other.
 In converse, populations that are distant on a PCA-plot are guaranteed to also have a large $F_2$-distance.
@@
 
-Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might be considered admixed between them?
-One way to address this question is to consider the space for which $F_3$ is negative, i.e.
+Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might have negative $F_3$-statistics? This condition can be written as  
 \begin{eqnarray}
@@
 However, several populations that appear inside the circles (e.g.
-Basque, Canary Islandersh) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball.
+Basque, Canary Islanders) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball.
 This reveals that the first two PCs do not capture all the genetic variation relevant for  European population structure.
@@
 \begin{figure}[!ht]
-	\includegraphics[width=.8\textwidth]{figures/outgroup_f3.pdf}
+	\includegraphics[width=.6\textwidth]{figures/outgroup_f3.pdf}
 	\caption{\textbf{Outgroup-$F_3$-statistics}.
@@
 This is done  using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ is an outgroup.
-The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (\ref{fig:outgroupf3}A) The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.
+The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (Figure \ref{fig:outgroupf3}A).
+The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.
 
@@
 
-On the right-hand-side terms only the $F_3$ term depends on the $X_i$.
-The fraction can be thought of as a normalizing constant,  which justifies the argument that the $F_3$-statistic and length of the projected vector are proportional to each other, and can thus be interpreted similarly.
-Thus, the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in the example in Figure \ref{fig:outgroupf3}B this is $X_3$.
+Of the right-hand-side terms, only the $F_3$ term depends on the $X_i$.
+The fraction can be thought of as a normalizing constant,  so the $F_3$-statistic is proportional to the length of  the projected vector, and they can thus be interpreted similarly.
+This means, that the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in Figure \ref{fig:outgroupf3} this is $X_3$.
 
@@
 The $F_3$-projection axis is given by a black line, the projection of populations onto this axis by thin gray lines.
-In the full data space, these projection are orthogonal to the axis.
+In the full allele frequency  space, these projection are orthogonal to the axis.
 B: Projection along the axis Sardinian-Mbuti (X-axis), and PCA on residual of this projection (PC1 on Y-axis, PC2 as coloring).
@@
 \paragraph{Example}
-In Figure \ref{fig:f4}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, in i.e.
+In Figure \ref{fig:f4}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, i.e.
 a statistic that aims to find the population most closely related to Sardinian (a Mediterranean Island), assuming the Mbuti are an outgroup to all populations in the data set.
@@
 For each population, the projection is indicated with a grey line.
-In the full data space, this line is always orthogonal to the segment Mbuti-Sardinian, but on the plot (i.e.
+In the full allele frequency space, this line is always orthogonal to the segment Mbuti-Sardinian, but on the plot (i.e.
 the subspace spanned by the first two PCs), this is not necessarily the case.
@@
 $F_4$-statistics that correspond to a branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot (Figure \ref{fig:geom}G), again with the caveat that we need to scale it by a constant.
-If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e.
+If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ which implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e.
 that $X_1$ and $X_2$  map to the same point on the projection axis $\overline{X_3X_4}$ (Figure \ref{fig:geom}H).
@@
 Just adding a third PC (blue), however, shows that the clines are not, in fact, parallel, and the correlation for most populations is low.
-The finding that three PCs are necessary to explain this data can also be seen from the spectrum of these $F_4$-statistics (Figure \ref{fig:f3}E), which have high contributions only from the first three PCs.
+The finding that three PCs are necessary to explain this data can also be seen from the spectrum of these $F_4$-statistics (Figure \ref{fig:f3}E), which have high contributions from the first three PCs.
 \begin{figure}[!ht]
-	\includegraphics[width=.7\textwidth]{figures/fig_f4_ratio.pdf}	
+	\includegraphics[width=.6\textwidth]{figures/fig_f4_ratio.pdf}	
 	\caption{\textbf{Admixture proportion estimates} A: Visualization of the admixture graph scenario used to estimate the ancestry proportion $\alpha$ contributed from $X_1$ to $X_X$, using references $R_1$ and $R_2$.
@@
 So far, I used eq.
-\ref{eq:fpc} to interpret $F$-statistics on a PCA-plot, but the argument holds for \emph{any} orthonormal projection of the data space.
+\ref{eq:fpc} to interpret $F$-statistics on a PCA-plot, but the argument holds for \emph{any} orthonormal projection of the allele frequency space.
 This is useful in particular for estimates of admixture proportions, which are often done as projections into a low-dimensional reference space defined by $F$-statistics \citep{patterson2012, petr2019, harney2021, oteo-garcia2021}.
@@
 I use the World data set as an example, using Sardinian and Mbuti as references populations (Figure \ref{fig:f4}B).
-The data is the same as in  the PCA (Figure \ref{fig:f4}A), but it is now rotated such that the axis between the reference population (black line in Figure \ref{fig:f4}A) is now reflected by the x-axis.
-For any pair of populations $X_1$ $X_2$, their horizontal distance reflects $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$ and the relative horizontal distance corresponds exactly to $F_4$-ratio admixture estimates.
+The data is the same as in  the PCA (Figure \ref{fig:f4}A), but it is now rotated such that the line between the reference population (black line in Figure \ref{fig:f4}A) now corresponds to the x-axis.
+For any pair of populations $X_1$ and $X_2$, their horizontal distance reflects $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$ and the relative horizontal distance corresponds exactly to $F_4$-ratio admixture estimates.
 For many sets of populations, this is of course not sensible, and just looking at the first PC of the residual shows many examples where the populations are not collinear.
-For example, on the x-axis, the South American Surui are between Papuans and Georgians, but since the Surui clearly are not on the line between Papuans and Georgians, this cannot be the result of admixture.
+For example, if we just focus on the x-axis, the South American Surui appear between Papuans and Georgians.
+But since the Surui clearly are not on the line between Papuans and Georgians, this plot reveals that this should not be interpreted as the result of admixture.
 
@@
 My interpretation here is different in that the utility of PCA is to simplify the geometry of the data.
-One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA.
+One consequence is that the results here are not directly impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA.
 However, a very skewed sampling distribution will increases the likelihood that more, or different PCs will have to be included in the analysis.
 
-The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
+The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
 On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
@@
 \subsubsection{Missing data}
-The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
+The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing missing data have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
 Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}.
 In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation.
-These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}, and a PCA-based approach to estimating $F$-statistics may provide higher accuracy than naive approaches because it allows using data from related populations.
+These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}.
+The theory developed here suggests that developing a PCA-based $F$-statistics framework might be more powerful, as related populations could be used to fill in missing data.
 
 \subsubsection{Normalization}
-The normalization of SNPs is similarly a matter of convention.
-The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted.
-The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}.
+In PCA, SNPs are typically normalized to have expected variance of one; a step that is omitted in calculating $F$-statistics \citep{patterson2006}.
+The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable.
+which holds regardless of weighting.
+Thus, normalization of SNPs is largely a matter of convention, for  $F$-statistics the dependency on  additional samples (through  mean allele frequencies) is often unwanted, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics such as \texttt{qpAdm} \citep{patterson2012, harney2021}.
 As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but should be explored in future work \citep{felsenstein1973}.
@@
 However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits.
-The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not to assume that gene flow is rare, and visualizing both the projection used for estimating admixture proportion, and the residual space, can be used to highlight scenarios where the assumptions are violated.
+The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not need to assume that gene flow is rare.
 
@@
 
-\newpage
 %\bibliography{main}

### 2021-12-14 fstats_pca_lewontin.tex
local

@@
 \begin{abstract}
-Human  genetic diversity is shaped by our complex history. Data-driven methods such as Principal Component Analysis (PCA) are an important population genetic tool to understand this method. 
-Here, I contrast PCA with a set of statistics motivated by trees ($F$-statistics).
+Human  genetic diversity is shaped by our complex history.
+In a common analysis paradigm, data-driven methods such as Principal Component Analysis (PCA) are used in exploratory data analyses, which are then followed up by more formal statistical treatments using $F$-statistics.
 Here, I show that these two methods are closely related, and I derive explicit connections between the two approaches.

### 2022-01-13 fstats_pca_lewontin.tex
bla

@@
 Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe.
-This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.
+This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in][]{peter2016}.
+
+In many applications, it is useful to distinguish between $F$-statistics as parameters (e.
+g.
+length of a branch that is estimated) and an estimated $f$-statistic from data, denoted by convention by a lower case $f$ \citep{patterson2012}.
+Because PCA in population genetics is most commonly done on complete data without an estimation step, I do not make this distinction and use (upper-case) $F$-statistics throughout the manuscript.
 

### 2022-01-13 talk_leipzig.tex
bla

\documentclass[11pt]{beamer}
\beamertemplatenavigationsymbolsempty
\setbeamertemplate{footline}[frame number]
\usefonttheme[onlymath]{serif}

\makeatletter
\newlength\beamerleftmargin
\setlength\beamerleftmargin{\Gm@lmargin}
\makeatother

\usepackage{tikz}   
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[absolute,overlay]{textpos}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{unicode-math}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{array}
\usepackage{comment}

\usetheme{Warsaw}

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
\setbeamercolor{framesource}{fg=gray}
\setbeamerfont{framesource}{size=\tiny}

\newcommand{\MX}{\mathbf{X}} %uncentered data
\newcommand{\MC}{\mathbf{C}} %centering
\newcommand{\MY}{\mathbf{Y}} %centered data
\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
\newcommand{\MP}{\mathbf{P}} % PCs
\newcommand{\ML}{\mathbf{L}} % loadings
\newcommand{\MK}{\mathbf{K}} % Kernel
\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix

\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel

\newcommand{\POP}[1]{X_{#1}}
\newcommand{\POPX}{\POP{X}}
\newcommand{\POPA}{\POP{A}}
\newcommand{\POPO}{\POP{O}}
\newcommand{\q}[1]{\hat{p}_{#1}}
\newcommand{\HAP}[1]{I_{#1}}
\newcommand{\HET}[1]{H_{#1}}
%\newcommand{\PCOAL}[2]{c_{#1\downarrow #2}}
\newcommand{\PCOAL}[2]{f}
\newcommand{\FX}[1]{F_{#1}}

\newcommand{\source}[1]{\begin{textblock*}{4cm}(8.7cm,8.6cm)
		\begin{beamercolorbox}[ht=0.5cm,right]{framesource}
			\usebeamerfont{framesource}\usebeamercolor[fg]{framesource} {#1}
		\end{beamercolorbox}
\end{textblock*}}

\begin{document}
	\author{Benjamin Peter}
	\title[PCA and F-stats]{Connections between $F$-statistics and Principal Component Analysis}

	%\subtitle{}
	%\logo{}
	%\institute{}
	%\date{}
	%\subject{}
	%\setbeamercovered{transparent}
	%\setbeamertemplate{navigation symbols}{}
	\begin{frame}[plain]
	\maketitle
\end{frame}

\begin{comment}
{
\usebackgroundtemplate{\includegraphics[width=1.4\paperwidth]{figures/bjorklund_art.jpeg}}%
\begin{frame}
	%\frametitle{bla}
	\source{@tombjorklund}
\end{frame}
}
\end{comment}

\begin{comment}

\begin{frame}
\frametitle{Population structure and ancient DNA}
\begin{center}
		\hspace*{-1.8\beamerleftmargin}
		\includegraphics<1>[width=\textwidth]{data/figures/ancient_dna_by_year.png}
		\includegraphics<2>[width=1.05\paperwidth]{data/figures/ancient_dna_map.png}		
\end{center}
		 \source{\url{https://reich.hms.harvard.edu/}}
\end{frame}

\begin{frame}[plain,c]
\begin{center}
	\Huge Conceptualizing Complex Population Structure remains a major challenge
\end{center}
\end{frame}
\end{comment}

\begin{frame}{Example: Map}
\includegraphics[width=\textwidth]{figures/sikora2019_fig1ab.png}
\source{Sikora et al. 2019}
\end{frame}

\begin{frame}{Example: PCA}
\includegraphics[width=.8\textwidth]{figures/sikora2019fig1c.png}
	\source{Sikora et al. 2019}
\end{frame}

\begin{frame}{Example: F-statistics}
\includegraphics[width=1.1\textwidth]{figures/sikora20019fige3.png}
	\source{Sikora et al. 2019}
\end{frame}

\begin{frame}{Example: Map}
	\includegraphics[width=\textwidth]{figures/hajdinjak2021fig1.png}
	\source{Hajdinjak et al. 2021}
\end{frame}

\begin{frame}{Example: PCA}
	\includegraphics[width=.8\textwidth]{figures/hajdinjak2021fige2.png}
	\source{Hajdinjak et al. 2021}
\end{frame}

\begin{frame}{Example: F-statistics}
	\begin{center}
	\includegraphics[width=.6\textwidth]{figures/hajdinjak2021fig2.png}	
	\end{center}

	\source{Hajdinjak et al. 2021}
\end{frame}

\begin{frame}
\frametitle{Goals of this talk}
\begin{itemize}
	\item Establish conceptual links between frameworks
	\begin{enumerate}
		\item How can we interpret PCA in context of $F$-stats?
		\item How can we interpret $F$-stats in the context of PCA?		
	\end{enumerate}
	\item (Use established links to improve data interpretation)
\end{itemize}
\only<2>{
\begin{alertblock}{Focus on intuition}
	Details in terms of estimation, normalization, missing data will be glossed over
\end{alertblock}
}
\end{frame}

\begin{frame}
	\frametitle{F-statistics: Trees vs. Admixture Graphs}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			\includegraphics<1-2>[width=\textwidth]{figures/tree_intro.pdf}
			\includegraphics<3>[width=\textwidth]{figures/f2_intro.pdf}			
		\end{column}
	\begin{column}{0.5\textwidth}
		\includegraphics<2->[width=\textwidth]{figures/admixture_graph_intro.pdf}
	\end{column}

	\end{columns}
\end{frame}

\begin{frame}
\frametitle{$F$-statistics}
  \begin{tabular} {m{8.5cm}  |m{2cm}  }
	Definition & Branch length \\%& Path  \\
	\hline
	\vspace{6pt}$$\FX2(\POP1,\POP2) = \sum_l (X_{il} - X_{jl})^2 $$
	&\vspace{6pt}\includegraphics[scale=0.4]{figures/f2_internal.pdf}        
	%&\vspace{6pt}\includegraphics[scale=0.5]{./f2_admixture.eps}        
	\\ 
	\hline
	 \only<2->{\small{$$\FX3(\POP{x}; \POP1, \POP2) = \sum_l(X_{xl} - X_{1l})(X_{xl} - X_{2l}) $$}
	&\includegraphics[scale=0.4]{figures//f3_internal.pdf}       
	%&\vspace{6pt}\includegraphics[scale=0.5]{./f3_admixture.eps}        
	\\ 
	\hline}
\only<3>{``Admixture''-$F_3$-statistic: If data is generated by a tree-like relationship, $F_3(P_X; P_1, P_2) \geq 0$
	&\vspace{6pt}\includegraphics[scale=0.4]{figures//f3_admixture.pdf}      
	\\ 
	\hline
}
\only<4>{``Outgroup''-$F_3$-statistic: Most similar pops have highest $F_3(P_2; P_X, P_1)$
	&\vspace{6pt}\includegraphics[scale=0.4]{figures//f3_outgroup.pdf}      
	\\ 
	\hline
}
	\only<5>{$$\FX4^{(B)}(\POP1; \POP2; \POP3, \POP4) = \sum_l(X_{1l} - X_{3l})(X_{2l} - X_{4l})$$
	&\vspace{6pt}\includegraphics[scale=0.4]{figures//f4_internal_branch.pdf}        
	%&\vspace{6pt}\includegraphics[scale=0.5]{./f4b_admixture.pdf}        
	\\ 
	\hline
	}
	\only<6>{
	$$\FX4^{(T)}(\POP1; \POP2; \POP3, \POP4) = \sum_l(X_{1l} - X_{2l})(X_{3l} - X_{4l})$$
	&\vspace{6pt}\includegraphics[scale=0.4]{figures//f4_internal.pdf}        
	%&\vspace{6pt}\includegraphics[scale=0.5]{./f4_admixture.eps} \\ 
}
\end{tabular}

		\source{Patterson et al. 2012; Peter 2016}

\end{frame}

\begin{comment}
\begin{frame}
\frametitle{Principal Component Analysis}
\begin{columns}
	\begin{column}{0.5\textwidth}
		\includegraphics[width=1.1\textwidth]{figures/hajdinjak2021_fige2_no_legend.png}
	\end{column}
	\begin{column}{0.5\textwidth}
		\includegraphics<2>[width=1\textwidth]{figures/mcvean2009_fig3a.png}			
		\includegraphics<3->[width=1\textwidth]{figures/mcvean2009_fig3ab.png}		
		\includegraphics<4->[width=1\textwidth]{figures/mcvean2009_fig4b.png}
		\source{McVean, 2009}
	\end{column}
\end{columns}
\end{frame}
\end{comment}
	\begin{frame}
\frametitle{Principal Component Analysis}
	\begin{center}
		\includegraphics<1>{figures/pca1.pdf}
		\includegraphics<2>{figures/pca2.pdf}
		\includegraphics<3>{figures/pca3.pdf}
		\includegraphics<4>{figures/pca4.pdf}
		\includegraphics<5->{figures/pca5.pdf}						
		%\includegraphics<6>{figures/pca6.pdf}						
	\end{center}
\end{frame}

\begin{comment}
\begin{frame}
\frametitle{How to find PCs}
\begin{columns}
\begin{column}{0.4\textwidth}
	\includegraphics[width=\textwidth]{figures/pca3.pdf}
\end{column}
\begin{column}{0.7\textwidth}
	\begin{itemize}
		\item<1-> Singular Value Decomposition: $\MY = (\mathbf{U}\mathbf{D})\mathbf{L} = \mathbf{P}\mathbf{L}$
		\item<2-> Eigendecomposition of $\MY\MY^T$: $\MY\MY^T = \mathbf{U}\mathbf{D}^2\mathbf{U}^T = \MP\MP^T$
		\only<3>{\item $(\MY\MY^T)_{ij}$}
		\item<4-> $(\MY\MY^T)_{ij} = \sum_l (x_{il} - \mu_l)(x_{jl} - \mu_l)$
		\item<5-> \small{$\FX3(\POP{x}; \POP1, \POP2) = \sum_l(X_{1l}-X_{xl})(X_{2l}-X_{xl})$}
		\item<6-> $(\MY\MY^T)_{ij} = F_3(\MEAN; \MX_i, \MX_j)$
	\end{itemize}
\end{column}			
\end{columns}
\vspace{30px}
\only<6>{
\begin{alertblock}{Observation}
	PCA is equivalent to outgroup-$F_3$-analysis with sample mean as outgroup
\end{alertblock}
}
\end{frame}
\end{comment}

\begin{frame}
\frametitle{(metric) Multi-Dimensional Scaling (MDS)}

\begin{columns}
\begin{column}{0.5\textwidth}
\includegraphics<1>{figures/pca1.pdf}
\includegraphics<2>{figures/mds1.pdf}
\includegraphics<3->{figures/mds2.pdf}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics<4>{figures/mds3.pdf}		
\includegraphics<5>{figures/mds4.pdf}		
\includegraphics<6>{figures/mds5.pdf}		
\includegraphics<7>{figures/mds6.pdf}					
\end{column}
\end{columns}
		\only<7->{
	\begin{alertblock}{Observation}
		MDS on $F_2$ is equivalent to PCA
\end{alertblock}}
\end{frame}

\begin{frame}
	\frametitle{Constructing PCA}
	Let $\MX$ be a matrix of genotypes, $\MC$ a centering matrix that subtracts row means and $\MY = \MC\MX$ be the centered genotype matrix, and $\MF$ be a matrix of $F_2$-statistics.

	The following constructions of PCA are equivalent:
	\begin{itemize}
		\item Singular Value Decomposition of $\MY$
		\item Eigendecomposition of $\MY\MY^T$
		\item Eigendecomposition of $-\frac{1}{2}\MC\MF\MC$ 
	\end{itemize}
	\only<2->{
	\begin{alertblock}{Observation}
		$\MF$ is a sufficient statistic to compute a PCA
\end{alertblock}}
\end{frame}

\begin{comment}
\begin{frame}
\frametitle{PCA is MDS on $\MF$}

\begin{itemize}
\item<1-> PCA is decomposition of Covariance matrix: $\MY\MY^T$
\item<2-> Consider $\MF; f_{ij} = F_2(X_i, X_j) = \sum_l (X_{li}^2 -X_{lj})^2$
\item<3-> Consider $\MF; f_{ij} = \sum_l X_{li}^2 + \sum_l X_{lj}^2 - 2 \sum_l X_{li}X_{lj}$
\item<4-> MDS is Eigendecomposition of $-\frac{1}{2}\MC\MF\MC$
\only<5>{\item<5> $\MC\MF\MC = \MC\MX_i^2\MC + \MC\MX_j^2\MC - 2\MC\MX\MX^T\MC$}
\item<6-> $\MC\MF\MC = \underbrace{\MC\MX_i^2\MC}_{0}  + \underbrace{\MC\MX_j^2\MC}_{0} - 2\underbrace{\MC\MX\MX^T\MC}_{\MY\MY^T}$
\end{itemize}

\only<7>{
\begin{alertblock}{Observation}
PCA is equivalent to MDS on $\MF$
\end{alertblock}
}
\end{frame}

\begin{frame}
\frametitle{PCA is MDS on Outgroup$\MFT$}

\begin{itemize}
\item<1-> PCA is decomposition of Covariance matrix: $\MY\MY^T$
\item<2-> $\MFT(O); f_{ij} = F_3(O; X_i, X_j)$
\item<2-> $f_ij = \sum_l \big[ o_l^2 - o_l X_{li} - o_l X_{li} +  X_{li} X_{lj} \big]$
\item<3-> $\MC\MFT\MC = \underbrace{\MC O^2\MC}_{0}  - 
\underbrace{\MC O\MX_i\MC}_{0} -
\underbrace{\MC O\MX_j\MC}_{0} + 
\underbrace{\MC\MX\MX^T\MC}_{\MY\MY^T}$
\end{itemize}

\only<5>{
\begin{alertblock}{Observation}
Decomposition of \emph{any} centered $F_3$-matrix is equivalent to PCA.
\end{alertblock}
}
\end{frame}

\begin{frame}
\frametitle{0-diagonal MDS}
In aDNA applications (e.g. Fu et al 2016), MDS is performed on
\small{
$$\mathbf{M} = 
\begin{pmatrix}
0                     & 1 - f_3(O; X_1, X_2) & \dots & 1 - f_3(O; X_1, X_n)\\
 1 - f_3(O; X_1, X_2) & 0 &                    \dots & 1 - f_3(O; X_2, X_n)\\
 \vdots &             \vdots                 & \ddots& 1 - f_3(O; X_i, X_n)\\
 1 - f_3(O; X_1, X_n) & 1 - f_3(O; X_2, X_n)   & \dots & 0
\end{pmatrix}
$$}\source{Fu et al. 2016}
It can be shown that 
$$\MC\mathbf{M}\MC = \MC\MF\MC + \MC\mathbf{O}\MC$$
with $o_{ii} = f_2(O, X_i) - 1; o_{ij} = 0$
\end{frame}

\begin{frame}
\frametitle{How can we interpret PCA in context of $F$-stats?}
\begin{itemize}
	\item PCA is decomposition of $\MFT(\MEAN, X_1, X_2)$-matrix
	\item PCA is a function of (unnormalized)-$\MF$-matrix
	\begin{itemize}
		\item same information content
		\item differences in analysis choices (individual/population, normalization, estimate) need to be justified
	\end{itemize}
	\item<2-> setting diagonal allows some leeway
	\item<3-> \emph{Conjecture}: PCA is analogous to midpoint-rooted phylogenetic analyses; Outgroup-$F_3$ allow other rootings.

\end{itemize}
\end{frame}
\end{comment}

\begin{frame}
	\frametitle{Geometric Interpretation}
	\includegraphics[width=\textwidth]{figures/oteo_header.png}

	\begin{columns}
		\begin{column}{.4\textwidth}
	   \includegraphics<2->[width=\textwidth]{figures/oteo_fig1a.png}
		\end{column}
	\begin{column}{.6\textwidth}

	\only<2>{\begin{align*}
	F_2(X_1, X_2) &= \sum_l (X_{1l} - X_{2l})^2\\
	F_3(X_1; X_2, X_3) &= \sum_l (X_{1l} - X_{2l})(X_{1l}-X_{13})\\
	F_4(X_1, X_2; X_3, X_4) &= \sum_l (X_{1l} - X_{2l})(X_{3l}-X_{4l})
	\end{align*}}
	\only<3->{\begin{align*}
	F_2(X_1, X_2) &= \normsq{X_{1l} - X_{2l}}\\
	F_3(X_1; X_2, X_3) &= \langle X_1 - X_2, X_1 - X_3\rangle\\
	F_4(X_1, X_2; X_3, X_4) &= \langle X_1 - X_2, X_3 - X_4\rangle\end{align*}}
\end{column}	
\end{columns}
\only<4->{
	\begin{alertblock}{Interpretation}
		Think of $F$-statistics as inner products in L-dimensional \emph{data space}.
	\end{alertblock}
}
\end{frame}

\begin{frame}
\frametitle{$F$-statistics on PCA-plot}
\begin{itemize}
	\item Recall that PCA is just rotation around centroid
	\item<3-> Distances (such as $F_2$) are invariant to  rotation
\end{itemize}
\begin{center}
	\includegraphics<2>{figures/pca2.pdf}
	\includegraphics<2>{figures/pca4.pdf}						
\end{center}
\begin{itemize}
	\item<4-> $$F_2(X_1, X_2) = \sum_{\text{loci}}(x_{1l} - x_{2l})^2$$
	\item<5-> $$F_2(X_1, X_2) = \sum_{\text{PCs}}(x_{1p} - x_{2p})^2$$	
\end{itemize}
\only<6->{
	\begin{alertblock}{Observation}
		$F_2$ can be decomposed into contributions from different principal components
	\end{alertblock}
}
\end{frame}

\begin{frame}
\frametitle{$F$-statistics on PCA-plot}
\begin{columns}
	\begin{column}{0.5\textwidth}
		\includegraphics<1>{figures/pca1.pdf}
		\includegraphics<2->{figures/pca1b.pdf}		
	\end{column}
	\begin{column}{0.5\textwidth}
		\begin{itemize}
			\item<3-> $F$-statistics have a geometrical representation on PCA-plot
			\item<3-> Exact only if we use \emph{all} PCs
			\item<4-> If first two PCs capture relevant population structure, $F$-statistics will be well approximated
		\end{itemize}		
	\end{column}
\end{columns}
\end{frame}

\begin{frame}
	\frametitle{Technical interlude: Does PCA provide a good approximation?}
	\pause
	\begin{align*}
	F_2(P_i, P_j) &= \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+ \sum_{k=K+1}^n(p_{ik} - p_{jk})^2\nonumber\pause\\
	&= {\hat{F_2}^{(K)}(P_i, P_j)} &+ {\epsilon^{(K)}(P_i, P_j)}
	\end{align*}
	\pause
	The total error is then
	\begin{equation}
	\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
	\end{equation}\pause
	\begin{alertblock}{Observation}
		$\hat{F}_2$ provides optimal approximation in the sense that total squared error between \emph{all} $F_2$-statistics is minimized.
	\end{alertblock}
\end{frame}

\begin{frame}
	\frametitle{Technical interlude 2: What about projections?}
	Recall that the basic way we project a sample onto a PCA is by multiplying with $\ML$
	\begin{align*}
	F_2(X_i, X_{\text{new}}) &= F_2(Y_i, Y_{\text{new}})\nonumber\\ 
	&= F_2(P_i, P_{\text{proj}}) + F_2(P_{\text{proj}}\ML, Y_{\text{new}})\nonumber\pause\\
	&= \hat{F_2}(P_i, P_j) + {\epsilon(P_i, P_j)} + F_2(P_{\text{proj}}\ML, Y_{\text{new}}) \label{eq:approx}.
	\end{align*}
	\pause
	\begin{alertblock}{Observation}
		Projections adds another error-term that will be small if the projected sample is well-represented by variation of reference sample.
	\end{alertblock}
\end{frame}

\begin{frame}
	\frametitle{Global Data}
	\includegraphics[width=\paperwidth]{data/figures/sample_map_world.png}
\end{frame}
\begin{frame}
	\frametitle{Global PCA}
	\begin{center}
		\begin{tikzpicture}
		\node (img1) {\includegraphics[height=5cm]{data/figures/talk/world_pca.png}};
		\node (img2) at (img1.south)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_world2.png}};
		\end{tikzpicture}		
	\end{center}
\end{frame}
\begin{frame}
	\frametitle{Westeurasian Data}
	\includegraphics{data/figures/sample_map_eu.png}
\end{frame}

\begin{frame}
	\frametitle{Westeurasian PCA}
\begin{tikzpicture}
\node (img1)[xshift=-1cm] {\includegraphics[height=6cm]{data/figures/talk/eu_pca.png}};
\node (img2) at (img1.south east)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_eu2.png}};
\end{tikzpicture}	
\end{frame}

\begin{frame}
\frametitle{$F_2$-statistic on PCA-plot}
\begin{columns}
	\begin{column}{0.5\textwidth}
		\includegraphics<1>{figures/f3_on_pca_1b.pdf}
		\includegraphics<2>{figures/f2_on_pca.pdf}
	\end{column}
	\begin{column}{0.6\textwidth}
		\begin{itemize}
			\item $F_2(X_1, X_2) = \sum_l(X_{1l}-X_{2l})^2$
			\item $F_2(X_1, X_2) = \normsq{X_1 -  X_2}$			
		\end{itemize}		
	\end{column}
\end{columns}
\end{frame}

\begin{frame}
	\frametitle{Distance from Sardinians}
	\begin{center}
		\begin{tikzpicture}
		\node (img1) {\includegraphics[height=5cm]{data/figures/talk/world_pca_highlightf2.png}};
		\node (img2) at (img1.south)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_world2.png}};\pause
		\node (img3) at (img1.west)[xshift=-1cm] {\includegraphics[height=6cm]{data/figures/talk/world_f2.png}};
		\end{tikzpicture}		
	\end{center}
\end{frame}

\begin{frame}
\frametitle{Admixed populations ($F_3$) on PCA-plot}
\begin{columns}
	\begin{column}{0.4\textwidth}
		\includegraphics<1>{figures/f3_on_pca_1b.pdf}
		\includegraphics<2-4>{figures/f3_on_pca_1.pdf}		
		\includegraphics<5>{figures/f3_on_pca_1d.pdf}				
	\end{column}
	\begin{column}{0.6\textwidth}
		\begin{itemize}
			\item<1-> Given $X_1, X_2$, which pops have $F_3 < 0$?	
			\item<2-> $F_3(Y; X_1, X_2) = 0$ is a circle!
%			\item<3-> Samples outside circle will always have positive $F_3$
			\item<5> Circle is larger when considering more PCs
		\end{itemize}

	\only<1>{\begin{center}
			\includegraphics[scale=1]{figures//f3_admixture.pdf}
		\end{center}}  		
		\includegraphics<4>[width=\textwidth]{figures/mcvean2009_fig4b.png}
	\end{column}
\end{columns}
\end{frame}

\begin{frame}
	\frametitle{Westeurasian PCA}
	\begin{tikzpicture}
	\node (img1)[xshift=-1cm] {\includegraphics[height=6cm]{data/figures/talk/eu_pca.png}};
	\node (img2) at (img1.south east)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_eu2.png}};
	\end{tikzpicture}	
\end{frame}
\begin{frame}
	\frametitle{Westeurasian PCA}
	\begin{tikzpicture}
	\node (img1)[xshift=-1cm] {\includegraphics[height=6cm]{data/figures/talk/eu_pca_f3.png}};
	\node (img2) at (img1.south east)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_eu_f3.png}};
	\end{tikzpicture}	
\end{frame}

\begin{frame}
	\frametitle{Westeurasian PCA}
	\begin{tikzpicture}
	\node (img1)[xshift=-1cm] {\includegraphics[height=6cm]{data/figures/talk/eu_pca_f32.png}};
	\node (img2) at (img1.south east)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_eu_f3.png}};
	\end{tikzpicture}	
\end{frame}

\begin{frame}
\frametitle{Summary Admixture $F_3$}
\begin{columns}
	\begin{column}{.5\textwidth}
\begin{enumerate}[<+->]
	\item $F_3 < 0 \Leftrightarrow $ $n$-ball on PCA
	\item on 2D-plot samples might be in circle, but positive $F_3$
	\item Samples outside circle always have $F_3 > 0$

\end{enumerate}
	\end{column}
	\begin{column}{.7\textwidth}
\includegraphics[width=\textwidth]{data/figures/talk/eu_pca_f32.png}
\end{column}
\end{columns}

\end{frame}

\begin{comment}
\begin{frame}
\frametitle{Admixture $F_3$-stats on PCA-plot}
\begin{columns}
	\begin{column}{0.4\textwidth}
		\includegraphics<1>{figures/f3_on_pca_2b.pdf}
		\includegraphics<2->{figures/f3_on_pca_2.pdf}		
	\end{column}
	\begin{column}{0.6\textwidth}
		\begin{itemize}
			\item<1-> Given $X_1, X_x$, which pops $X_2$ have $F_3 < 0$?	
			\item<2-> $F_3$ is 0 if $(X_x; X_1), (X_x; X_2)$ form a right angle!
			\item<3-> Inner (dot) product: $F_3(X_x; X_1, X_2) = \langle X_x - X_1, X_x - X_2\rangle$
		\end{itemize}		
	\end{column}
\end{columns}
\end{frame}
\end{comment}

\begin{frame}
	\frametitle{Outgroup $F_3$-stats on PCA-plot}
	\begin{columns}
		\begin{column}{0.4\textwidth}
			\includegraphics<1-2>{figures/f3_on_pca_3b.pdf}
			\includegraphics<3>{figures/f3_on_pca_3.pdf}		
			\includegraphics<4->{figures/f3_on_pca_3c.pdf}			
		\end{column}
		\begin{column}{0.6\textwidth}
			\begin{itemize}
				\item<1-> Given $X_O, X_1$, which pop $X_2$ has highest $F_3$?
			\end{itemize}
		\only<1>{\begin{center}
				\includegraphics[scale=1]{figures/f3_outgroup.pdf}
		\end{center}}
		\begin{itemize}	
	\item<2-> As $F_3 = \langle X_O - X_1, X_O - X_2 \rangle$, project $\overline{X_OX_2}$ on $\overline{X_OX_1}$
	\item<3-> $F_3$ is proportional to segment in dir $X_0 \rightarrow X_1$
	\item<4-> Pop furthest on axis has highest $F_3$.
\end{itemize}

		\end{column}
	\end{columns}
\end{frame}

\begin{frame}
\frametitle{$F_3$(Mbuti; Mozabite, X))}
\begin{center}
\begin{tikzpicture}
\node (img1) {\includegraphics[height=5cm]{data/figures/talk/world_pca.png}};
\node (img2) at (img1.south)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_worldf32.png}};
\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}
	\frametitle{$F_3$(Mbuti; Mozabite, X))}
	\begin{center}
		\begin{tikzpicture}
		\node (img1) {\includegraphics[height=5cm]{data/figures/talk/world_pca2.png}};
		\node (img2) at (img1.south)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_worldf3.png}};
		\end{tikzpicture}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{$F_3$(Mbuti; Mozabite, X))}
	\begin{center}
		\begin{tikzpicture}
		\node (img1) {\includegraphics[height=5cm]{data/figures/talk/world_pca3.png}};
		\node (img2) at (img1.south)[yshift=-1.1cm] {\includegraphics[height=5cm]{data/figures/sample_map_worldf3.png}};
		\end{tikzpicture}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Outgroup-$F_3$}
	\begin{columns}
		\begin{column}{.5\textwidth}
			\includegraphics<1->[width=\textwidth]{data/figures/talk/outgroup_f31.png}
		\end{column}
		\begin{column}{.5\textwidth}
			\includegraphics<2->[width=\textwidth]{data/figures/talk/outgroup_f32.png}
		\end{column}
	\end{columns}
	\begin{itemize}
		\item <3-> Outgroup-$F_3$ corresponds to projection on axis from Outgroup to test population
		\item <4-> Two PCs give excellent approximation
		\item <5-> Offset by constant due to local population structure
	\end{itemize}
\end{frame}

\begin{frame}
\frametitle{ $F_4$-stats on PCA-plot}
\begin{columns}
	\begin{column}{0.4\textwidth}
		\includegraphics<1>{figures/f4_on_tree3.pdf}
		\includegraphics<2->{figures/f4_on_tree2.pdf}
	\end{column}
	\begin{column}{0.7\textwidth}
		\begin{itemize}
			\item<1-> $F_4(X_1, X_2 ; X_3, X_4)$ is internal branch of a tree
			\item<2> $F_4(X_1, X_2 ; X_3, X_4) \Leftrightarrow \langle X_1 - X_2; X_3 - X_4\rangle$			
			\item<2> $F_4$ is zero if $X_1 -X_2$ and $X_3 -X_4$  are orthogonal
		\end{itemize}	
	\end{column}
\end{columns}
\end{frame}

\begin{frame}
	\frametitle{ $F_4$-stats on PCA-plot}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			\includegraphics{figures/f4_on_tree2.pdf}
		\end{column}
		\begin{column}{0.5\textwidth}	
			\includegraphics{figures/f4_on_pca2.pdf}	
		\end{column}
	\end{columns}
\end{frame}

\begin{frame}
	\frametitle{ $F_4$-stats on PCA-plot}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			\includegraphics{figures/f4_on_tree2.pdf}
		\end{column}
		\begin{column}{0.5\textwidth}	
			\includegraphics{figures/f4_on_pca.pdf}	
		\end{column}
	\end{columns}
\begin{equation*}
\text{Cor}(X_1 - X_2, X_3 - X_4) =  \frac{\langle X_1 - X_2, X_3 - X_4 \rangle}{\norm{X_1-X_2}\norm{X_3-X_4}} = \cos(\phi),\label{eq:angle}
\end{equation*}
\end{frame}

\begin{frame}
	\frametitle{Two Gradients in Westeurasian PCA}
\begin{columns}
	\begin{column}{.7\textwidth}
		$$F_4(\text{Georgian}, \text{Saudi}; \text{Sardinian}, X)$$
\includegraphics<1-2>[width=\textwidth]{data/figures/talk/anglepca.png}
\includegraphics<3>[width=\textwidth]{data/figures/talk/anglepca13.png}
	\end{column}
	\begin{column}{.4\textwidth}
	\includegraphics<2->[width=\textwidth]{data/figures/talk/angle.pdf}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Where does Orthogonality come from?}
\begin{columns}
	\begin{column}{0.5\textwidth}
		\includegraphics<1->{figures/f3_nonorthogonal.pdf}
	\end{column}
	\begin{column}{0.5\textwidth}
		\includegraphics<2>{figures/f3_orthogonal.pdf}
	\end{column}
\end{columns}
\end{frame}

\begin{frame}
	\frametitle{How can we interpret $F$-stats in the context of PCA?}
	\begin{itemize}[<+->]
		\item $F$-stats have geometric interpretation that aide in interpretation of PCA
		\item Inconsistenty between $F$-stats and PCA hint at important structure in higher PCs
		\item exact in $n$-dimensional, approximate in 2D
		\item $F_3$ is a (hyper)-circle
		\item $F_4$ is angle and tests orthogonality on PCA
		\item Orthogonality stems from independence of drift in distinct parts of history
	\end{itemize}
\end{frame}

\begin{comment}
\begin{frame}
	\frametitle{Thank You!}
	\includegraphics[width=\textwidth]{figures/group_photo.png}
\end{frame}
\end{comment}

\end{document}

### 2022-01-26 fstats_pca_lewontin.tex
part1 revisions

--- 2021-11-15 fstats_pca_lewontin.tex
@@
 \begin{abstract}
-Human  genetic diversity is shaped by our complex history. Data-driven methods such as Principal Component Analysis (PCA) are an important population genetic tool to understand this method. 
-Here, I contrast PCA with a set of statistics motivated by trees ($F$-statistics).
-Here, I show that these two methods are closely related, and I derive explicit connections between the two approaches.
+Human  genetic diversity is shaped by our complex history.
+Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics \textit{sensu} Patterson.
+I show that these two perspectives are closely related, and I derive explicit connections between the two approaches.
 I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link.
 I illustrate my results on two examples, one of local, and one of global human diversity.
-In both examples, I find that just using the first few PCs provides good population structure is sparse, and only a few components contribute to most statistics.
+In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and thus PCA-plots are effective to predict $F$-statistics.
 Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models.
-My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
+My results extend $F$-statistics to non-discrete populations, moving towards more complete descriptions of human genetic variation.
 \end{abstract}
@@
 In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}.
-Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{gopalan2022}.
+Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{boca2020}.
 
 Particularly for large and heterogeneous data sets, disentangling all these processes is challenging, and we cannot expect to devise a single model catching both broad strokes and minute details of human history.
-A commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions.
-each emphasizing particular aspects of the data.
+A commonly used analysis paradigm is thus to combine tools based on different sets of assumptions, each emphasizing particular aspects of the data.
 
 A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}.
-Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}).
-However, these methods are not designed to answer specific research questions, and are limited in their ability to estimate biologically meaningful parameters.
-For this purpose, methods based on explicit demographic models are often used that aim to fit a specified or estimated model of divergence, migration and genetic drift to the data \citep{gutenkunst2009, excoffier2013, kamm2015}.
-The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modeling assumptions such as that populations are discrete, randomly mating, or at equilibrium.
-While in most cases these assumptions are violated to some extent and cannot be verified, we hope that the resulting model fits provide sufficiently accurate answers to specific research questions.
+Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}.
+These methods are limited in their ability to estimate biologically meaningful parameters, but provide useful summaries and visualizations.
+These analyses are then complemented with methods based on explicit demographic models, which are used to estimate parameters or test hypotheses \citep{gutenkunst2009, excoffier2013, kamm2015}.
 
-%\paragraph{F-statistics}
-However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult.
+\paragraph{F-statistics}
+When the number of populations exceeds a few dozen, even codifying reasonable population models can be prohibitively difficult.
 One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}.
 However, this has the drawback that a large proportion of the available data remains unused.
-An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.
-
-The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}.
-Formal definitions will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches.
+An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to focus on the relationship between two, three or four populations, commonly using $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}.
+Formal definition will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches.
 \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.
@@
 The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A).
-However, admixture graphs are not the only plausible alternative models, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.
+However, admixture graphs are not the only plausible alternative model, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.
 
 \paragraph{F-statistics and PCA}
-The practical issue addressed in this study is how $F$-statistics can be reconciled with PCA, one of the most widely used data-driven modeling  techniques.
+The practical issue addressed in this study is how $F$-statistics can be compared with PCA, one of the most widely used data-driven modeling  techniques.
 One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal component, PC) retaining a maximum of the variance present in the data.
@@
 	\includegraphics[width=\textwidth]{figures/fstats_pca_vs_tree.pdf}
-	\caption{\textbf{Representation of $F$-statistics on trees and 2D-PCA-plots.} The schematics show four populations and their representation using a tree (top row) or a 2D-PCA plot (bottom row).
+	\caption{\textbf{Representation of $F$-statistics on trees and 2D-PCA-plots.} The schematics show four populations and their representation using an (arbitrarily rooted) tree (top row) or a 2D-PCA plot (bottom row).
 A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space.
@@
 \subsection{Formal Definition of $F$-statistics}
-Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ loci.
-Let $x_{il}$ denote the frequency at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$.
+Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ biallelic loci.
+Let $x_{il}$ denote the frequency of an arbitrary allele at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$.
 As $X_i$ will be the only data summary considered here for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.
@@
 In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch, respectively (Figure \ref{fig:geom}C).
-Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero.
 The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}.
+Crucially, this means that $F_4$ will be zero for pairs of populations from non-overlapping clades, which means that the tree lacks the branch corresponding to this statistic (as in Figure \ref{fig:geom}D).
 
@@
 The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A.
-Here, the (typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$.
+Here, the *typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$.
 Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe.
-This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.
+In this case, all $F_4$-statistics involving $X_y$ and both admixture sources will be non-zero, and, in some cases, $F_3(X_y; X_2, X_3)$ will be negative \citep[exact conditions can be found in ][]{peter2016}.
 
@@
 An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare.
-Recently, \cite{oteo-garcia2021}  re-derived $F$-statistics in a geometric framework, showing that these assumptions are not necessary.
 Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$.
@@
 \end{subequations}
-where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to the vector projection
+where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
 \begin{equation}
@@
 Thinking of $F$-statistics as projections also holds on trees: In e.g.
-a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:geom}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$.
-On trees, all disjoint paths are independent (i.e.
-orthogonal) from each other, and thus the external branches vanish under the projection.
+a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:admix}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$.
+The external branches are independent populations, and so they are expected to drift orthogonally to each other.
 
-The drawback of the geometric approach of \cite{oteo-garcia2021} is that we have to deal with an very high-dimensional space, as the number of SNPs is frequently in the millions.
-However, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data \citep{patterson2006}.
+The geometric approach of \cite{oteo-garcia2021} assumes a very high-dimensional space, as each SNP in a data set adds a dimension, and data sets have commonly a million or more SNPs.
+Such high-dimensional spaces are hard to visualize and analyze, however, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data \citep{patterson2006}.
 Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.
@@
 PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g.
-in how SNPs are standardized, how missing data is treated, what algorithm is used, or whether we use individuals or populations as units of analysis.
+in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis \citep{cavalli-sforza1994, patterson2006}.
 The version of PCA I use here is set up such that the similarities to  $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets.
-In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies.
-In contrast, many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}.
+In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}.
 The differences this causes will be addressed in the discussion.
@@
 If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$.
-Since the allele frequencies are between zero and one, we can interpret each population $X_i$
-of $\MX$ as a point in $\mathbb{R}^S$.
+Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
+of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.
 
-PCA allows us to approximate the points in the high-dimensional allele frequency space by a $K$-dimensional subspace of the  data.
-If all PCs are considered, $K=n-1$, in which case the data is simply rotated.
+One way PCA can be motivated is that it aims to find a $K$-dimensional subspace of the data space that retains most  variation in the data.
+$K$ is at most $n-1$, in which case the data is simply rotated.
 However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.
@@
 PCA, as defined above, and $F$-statistics are closely related.
-In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling, which, for squared Euclidean ($F_2$)-distances, leads to an identical decomposition to PCA \citep{gower1966}.
-Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$.
-We can obtain the principal components from this matrix by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
+It is a classical result that PCA is equivalent to multidimensional scaling using squared Euclidean distances \citep{gower1966}.
+Since $F_2$-distances are squared Euclidean, we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$.
+Multidimensional scaling then proceeds by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
 \begin{equation}
@@
 \end{equation}
+Although we arrive at $\MP$ from a very different angle, as long as we make the same choices about normalization and units of analysis, we will get the exact same results.
 
@@
 \begin{align}
-F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
-&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
+F_2(P_i, P_j) = \underbrace{\sum_{k=1}^K(p_{ik} - p_{jk})^2}_{\hat{F_2}^{(K)}(P_i, P_j)} + \underbrace{\sum_{k=K+1}^n(p_{ik} - p_{jk})^2}_{\epsilon^{(K)}(P_i, P_j)} \text{.}
 \end{align}
@@
 SNPs with missing data in any population are excluded.
-The code used to write this paper, create all figures and analyses is  available on \url{https://github.com/BenjaminPeter/fstats_pca}.
+The code used to create all figures and analyses will be available on \url{https://github.com/BenjaminPeter/fstats_pca}.
 
@@
 This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.
-\cite{lazaridis2014} used this data set as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}.
+It is motivated by the analysis of \cite{lazaridis2014}, who used it as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}.
 Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}.
@@
 However as $F_2$-statistics are estimates, some eigenvalues might be slightly negative, which would lead to imaginary PCs.
-I avoid this by setting all negative eigenvalues to zero.
+I avoid this by using the \texttt{nearPD}-function in R that ensures all eigenvalues have the correct sign.
 
@@
 	\caption{\textbf{Admixture representation on 2D-PCA-plot.} The schematics show four populations and their representation using an admixture graph (A) or a 2D-PCA plot.
-A: Admixture graph, with population $X_y$ originating as an admixture of $X_2$ and $X_3$, with $X_2$ contributing proportion $\alpha$.
+A: Admixture graph, with population $X_y$ originating as an admixture of $X_2$ (with proportion $1-\alpha$)and $X_3$ (proportion ($\alpha$), with $X_2$ contributing proportion $\alpha$.
 Subsequent drift (red branch) will change allele frequency to sampled admixture population $X_x$.
@@
 Thus, throughout the schematic figures, I assume that two PCs are sufficient to characterize population structure.
-In the data applications I evaluate how deviations of this assumption may manifest themselves in PCA plots.
+In the data applications I evaluate how deviations of this assumption my manifest themselves in PCA plots.
 
@@
 populations.
-On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations (Figure \ref{fig:geom}E).
+On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations.
 In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will fall close to each other on a PCA-plot, and have low pairwise $F_2$-statistics.
@@
 \ref{eq:fpc}), the distance on a PCA-plot will always be an underestimate of the full $F_2$-distance.
-Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations, and likely more PCs need to be investigated to understand how these populations are related to each other.
+Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations.
 In converse, populations that are distant on a PCA-plot are guaranteed to also have a large $F_2$-distance.
@@
 Since the allele frequencies of $X_y$ are a linear combination of $X_2$ and $X_3$, it will lie on the line segment connecting these two populations (Figure \ref{fig:admix}B), at a location predicted by the admixture proportions.
-Subsequent drift will change the allele frequency of $X_x$, and so in general it might fall on a different point on a PCA-plot.
+Subsequent drift will change the allele frequencies of $X_x$ (to say, $X_y$), and so in general it might fall on a different point on a PCA-plot.
 An exception occurs when $X_x$ (and no other populations related to $X_x$) are not part of the construction of the PCA, so that $X_x - X_y$ is orthogonal to all PCs, i.e.
@@
 
-Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might have negative $F_3$-statistics? This condition can be written as  
+Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might be considered admixed between them?
+One way to address this question is to consider the space for which $F_3$ is negative, i.e.
 \begin{eqnarray}
@@
 However, several populations that appear inside the circles (e.g.
-Basque, Canary Islanders) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball.
+Basque, Canary Islandersh) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball.
 This reveals that the first two PCs do not capture all the genetic variation relevant for  European population structure.
@@
 This is done  using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ is an outgroup.
-The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (Figure \ref{fig:outgroupf3}A).
-The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.
+The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (\ref{fig:outgroupf3}A) The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.
 
@@
 
-Of the right-hand-side terms, only the $F_3$ term depends on the $X_i$.
-The fraction can be thought of as a normalizing constant,  so the $F_3$-statistic is proportional to the length of  the projected vector, and they can thus be interpreted similarly.
-This means, that the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in Figure \ref{fig:outgroupf3} this is $X_3$.
+On the right-hand-side terms only the $F_3$ term depends on the $X_i$.
+The fraction can be thought of as a normalizing constant,  which justifies the argument that the $F_3$-statistic and length of the projected vector are proportional to each other, and can thus be interpreted similarly.
+Thus, the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in the example in Figure \ref{fig:outgroupf3}B this is $X_3$.
 
@@
 The $F_3$-projection axis is given by a black line, the projection of populations onto this axis by thin gray lines.
-In the full allele frequency  space, these projection are orthogonal to the axis.
+In the full data space, these projection are orthogonal to the axis.
 B: Projection along the axis Sardinian-Mbuti (X-axis), and PCA on residual of this projection (PC1 on Y-axis, PC2 as coloring).
@@
 \paragraph{Example}
-In Figure \ref{fig:f4}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, i.e.
+In Figure \ref{fig:f4}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, in i.e.
 a statistic that aims to find the population most closely related to Sardinian (a Mediterranean Island), assuming the Mbuti are an outgroup to all populations in the data set.
@@
 For each population, the projection is indicated with a grey line.
-In the full allele frequency space, this line is always orthogonal to the segment Mbuti-Sardinian, but on the plot (i.e.
+In the full data space, this line is always orthogonal to the segment Mbuti-Sardinian, but on the plot (i.e.
 the subspace spanned by the first two PCs), this is not necessarily the case.
 The coloring is based on the $F_3$-statistic calculated from all the data, with brighter values indicating higher $F_3$-statistics.
-In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas (cluster in the top left of the plot) project very close to orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.
+In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia and the Americas  project almost orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.
 We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}C), but keep the relative ordering.
@@
 One interpretation of $F_4$ on PCA plots is similar to that of $F_3$; as a projection of one vector onto another, with the difference that now all four points may be distinct.
-$F_4$-statistics that correspond to a branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot (Figure \ref{fig:geom}G), again with the caveat that we need to scale it by a constant.
-If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ which implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e.
+$F_4$-statistics that correspond to an internal branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot (Figure \ref{fig:geom}G), again with the caveat that we need to scale it by a constant.
+If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e.
 that $X_1$ and $X_2$  map to the same point on the projection axis $\overline{X_3X_4}$ (Figure \ref{fig:geom}H).
@@
 To illustrate the angle interpretation I return to the Western Eurasian data.
-The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:f3}A), a European gradient (from Sardinian to Finnish and Chuvash), and a Asian cline from Arab populations (top right) to the Caucasus  (bottom right).
+The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:f3}A), a European gradient (from Sardinian to Finnish and Chuvash), and an Asian cline from Arab populations (top right) to the Caucasus  (bottom right).
 This is quantified in Figure \ref{fig:f3}D, where I plot the angle corresponding to $F_4(X, \text{Saudi}; \text{Sardinian}, \text{Finnish})$.
@@
 Just adding a third PC (blue), however, shows that the clines are not, in fact, parallel, and the correlation for most populations is low.
-The finding that three PCs are necessary to explain this data can also be seen from the spectrum of these $F_4$-statistics (Figure \ref{fig:f3}E), which have high contributions from the first three PCs.
+The finding that three PCs are necessary to explain this data can also be seen from the spectrum of these $F_4$-statistics (Figure \ref{fig:f3}E), which have high contributions only from the first three PCs.
 \begin{figure}[!ht]
 	\includegraphics[width=.6\textwidth]{figures/fig_f4_ratio.pdf}	
-	\caption{\textbf{Admixture proportion estimates} A: Visualization of the admixture graph scenario used to estimate the ancestry proportion $\alpha$ contributed from $X_1$ to $X_X$, using references $R_1$ and $R_2$.
+	\caption{\textbf{Admixture proportion estimates} A: Visualization of the admixture graph scenario used to estimate the proportion $\alpha$ contributed from $X_1$ to $X_X$, using references $R_1$ and $R_2$.
 The full grey line corresponds to the projection axis, and the dotted grey lines corresponds to the branches ignored in the projections.
@@
 So far, I used eq.
-\ref{eq:fpc} to interpret $F$-statistics on a PCA-plot, but the argument holds for \emph{any} orthonormal projection of the allele frequency space.
+\ref{eq:fpc} to interpret $F$-statistics on a PCA-plot, but the argument holds for \emph{any} orthonormal projection in the data space.
 This is useful in particular for estimates of admixture proportions, which are often done as projections into a low-dimensional reference space defined by $F$-statistics \citep{patterson2012, petr2019, harney2021, oteo-garcia2021}.
 
-For example, a common way to estimate admixture proportion is the $F_4$-ratio:
+For example, a common way to estimate admixture proportion $\alpha$ of $X_1$ is the $F_4$-ratio:
 \begin{eqnarray}
@@
 In addition, the external tree branches $X_1 - X'_1$ and $X_2 - X'_2$ are disjoint which means they should be orthogonal.
-One a one-dimensional residual plot (Figure \ref{fig:ratio}B) this can not be verified, but the statistic
+On a one-dimensional residual plot (Figure \ref{fig:ratio}B) this can not be verified, but the statistic
 \begin{equation}
@@
 I use the World data set as an example, using Sardinian and Mbuti as references populations (Figure \ref{fig:f4}B).
-The data is the same as in  the PCA (Figure \ref{fig:f4}A), but it is now rotated such that the line between the reference population (black line in Figure \ref{fig:f4}A) now corresponds to the x-axis.
-For any pair of populations $X_1$ and $X_2$, their horizontal distance reflects $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$ and the relative horizontal distance corresponds exactly to $F_4$-ratio admixture estimates.
+The data is the same as in  the PCA (Figure \ref{fig:f4}A), but it is now rotated such that the axis between the reference population (black line in Figure \ref{fig:f4}A) is now reflected by the x-axis.
+For any pair of populations $X_1$ $X_2$, their horizontal projection distance reflects $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$ and the relative horizontal distance corresponds exactly to $F_4$-ratio admixture estimates.
 For many sets of populations, this is of course not sensible, and just looking at the first PC of the residual shows many examples where the populations are not collinear.
-For example, if we just focus on the x-axis, the South American Surui appear between Papuans and Georgians.
-But since the Surui clearly are not on the line between Papuans and Georgians, this plot reveals that this should not be interpreted as the result of admixture.
+For example, on the x-axis, the South American Surui are between Papuans and Georgians, but since the Surui clearly are not on the line between Papuans and Georgians, this cannot be the result of admixture.
 
@@
 My interpretation here is different in that the utility of PCA is to simplify the geometry of the data.
-One consequence is that the results here are not directly impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA.
-However, a very skewed sampling distribution will increases the likelihood that more, or different PCs will have to be included in the analysis.
+One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA, but, a very skewed sampling distribution will increases the likelihood that more, or different PCs will have to be included in the analysis.
 
-The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
+The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
 On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
@@
 The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized.
-In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies.
-In contrast, most data analyses have to grapple with missing data, SNP are often weighted according to their allele frequency and observed, individual-level genotypes are used as the basis of PCA.
+In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted, and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies.
+In contrast, most data analyses have to grapple with missing data, SNPs are often weighted according to their allele frequencies, and observed, individual-level genotypes are used as the basis of PCA.
 
 \subsubsection{Missing data}
-The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing missing data have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
+The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
 Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}.
 In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation.
-These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}.
-The theory developed here suggests that developing a PCA-based $F$-statistics framework might be more powerful, as related populations could be used to fill in missing data.
+These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}, and a PCA-based approach to estimating $F$-statistics may provide higher accuracy than naive approaches because it allows using data from related populations.
 
 \subsubsection{Normalization}
-In PCA, SNPs are typically normalized to have expected variance of one; a step that is omitted in calculating $F$-statistics \citep{patterson2006}.
-The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable.
-which holds regardless of weighting.
-Thus, normalization of SNPs is largely a matter of convention, for  $F$-statistics the dependency on  additional samples (through  mean allele frequencies) is often unwanted, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics such as \texttt{qpAdm} \citep{patterson2012, harney2021}.
+The normalization of SNPs is similarly a matter of convention.
+The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted.
+The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}.
 As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but should be explored in future work \citep{felsenstein1973}.
@@
 \subsubsection{Estimated vs. observed allele frequencies}
-The third difference between $F$-statistics and standard  PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
-The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
+The third difference between $F$-statistics and PCA is on the usage of estimated allele frequencies versus individual-based genotypes.
+The fact that PCA does not distinguish between sampling error and the underlying structure is a well-known drawback of PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise.
 Probabilistic PCA is one class of approaches that aim to separate the population structure from sampling noise \citep[e.g.][]{agrawal2020}.
@@
 From a practical perspective, formulating hypotheses and designing studies in terms of discrete populations with ``uniform'' genetic backgrounds is often sensible, as it enables  e.g.
-prediction of phenotypes \citep{berg2019, yair2021}, inference of demographic parameters and schematic models of human genetic history \citep{patterson2012}.
+prediction of phenotypes \citep{berg2019, yair2021}, inference of demographic parameters, and schematic models of human genetic history \citep{patterson2012}.
 This is also the case for $F$-statistics, which are motivated by trees.
 They assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
-However, these simplification do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of deemphasizing people that do not rigidly fall into predefined genetic groups.
+However, these simplifications do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of deemphasizing people that do not rigidly fall into predefined genetic groups.
 
@@
 However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits.
-The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not need to assume that gene flow is rare.
+The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not to assume that gene flow is rare, and visualizing both the projection used for estimating admixture proportion, and the residual space, can be used to highlight scenarios where the assumptions are violated.
 
+\section{Conclusion}
+\begin{itemize}
+	\item $F$-stats and PCA are partially redundant
+	\item Population structure is sparse - trees	
+	\item Formulating hypotheses
+	\item Orthogonality as a central concept
+	\item Orthogonality of scales (within-population variation) is independent of between-population variation
+	\item Orthogonality of isolation (two locations have little exchange)
+\end{itemize}
 \section{Supplementary File Description}
@@
 
+\newpage
 %\bibliography{main}

### 2022-01-26 fstats_pca_lewontin.tex

@@
 My results extend $F$-statistics to non-discrete populations, moving towards more complete descriptions of human genetic variation.
-\end{abstract}
 \section{Introduction}
@@
 
-\paragraph{F-statistics}
 When the number of populations exceeds a few dozen, even codifying reasonable population models can be prohibitively difficult.
@@
 
-The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
+The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
 On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.

### 2022-01-26 response.txt
response no fig

pdf:

1: fixed
2: fixed
3: changed reconciled to compared on first occurence. Kept second occurance
(p15) since there I explicitly talk about apparent contradictions
4: fixed
5: I agree with this point and delete the sentence
6: I changed this to: "Crucially, this means that $F_4$ will be zero for pairs
of populations from non-overlapping clades."
7: I was not thinking about computing here, but about interpretation. 
High-dimensional spaces are hard to interpret and analyze, which is why PCA is
useful. I rephrased this to make it hopefully more clear:
The geometric approach of \cite{oteo-garcia2021} assumes a very high-dimensional space, as each SNP in a data set adds a dimension, and data sets have commonly a million or more SNPs. Such high-dimensional spaces are hard to visualize and analyze, however, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data
8: Removed due to change in response to (7)
9: I added the sentence:
"It turns out that decomposing this matrix is equivalent to  calculating the PCA
from the original data, as long as we make the same choices about normalization
and units of analysis."
10: Changed to "Subsequent drift will change the allele frequencies of $X_x$ (to
say, $X_y$),"
11: I increased the size of alpha and added it to the caption
12: I added labels, as suggested
13: changed to "project almost orthogonally"
14: changed to "internal branch"
15: fixed
16: fixed
17: fixed as suggested
18: fixed
19: fixed
20: changed to but, (...)
21: fixed
22: fixed
23: revised
24: fixed
25: fixed

Noah:
1. The title sounds a bit more like that of a review article. Consider something more precise, e.g. “A geometric interpretation of the relationship between F2, F3, and F4 statistics and principal components analysis,” or “On the relationship between F2, F3, and F4 statistics and principal components analysis.”
TODO

2. At 1.46-1.70, the introduction seems to set up the F statistics as a preferred alternative to methods such as PCA and structure. However, this reasoning is unconvincing, as the various methods are all useful methods that can contribute in different ways. You could simply say that PCA is used to accomplish task X, and F statistics are used to accomplish task Y. It is motivation enough that one wants to understand how these different methods relate to each other.
This section seems to also comment on the use of F statistics to build models with large numbers of populations from calculations with 3 and 4 populations, as a contrast to (and possibly preferable to?) coalescent-style models. However, the paper does not really have much to say about these topics, and if anything, it is negative about them in the final paragraph. That makes the introduction strange in setting up F statistics as preferred to other methods but then deciding that they are problematic. I would suggest deferring any mention of the models of large numbers of populations to the Discussion. As in the previous paragraph, it’s enough to just want to understand the methods better.
RESPONSE:
I can see your point. My intention here was to simply give a brief overview of
the different methods used, and explain why one typically combines multiple
approaches by giving their main benefits and drawbacks. I can see how this is
not entirely relevant, and shortened this paragraph to remove the
benefits/drawback part.

3. In Fig 1 caption, please comment on the case in which the true relationship for four populations is ((X1,X2),(X3,X4)). Is it always assumed that species must have a relationship (((X1,X2),X3),X4) so that the symmetric case is never considered? 
RESPONSE:
Since F-statistics are distances, they correspond to unrooted trees, i.e. the
rooting is arbitrary. I changed the caption to "using an (arbitrarily rooted)
tree".

4. The paper is somewhat missing a grand summary that clarifies its main contributions. It seems that F3 and F4 computations are sometimes interpreted, problematically, as definitive tests of admixture. The paper shows that they may be redundant with PCA, or they may represent tests of particular hypotheses about placement of points in a PCA plot. I would be interested to see a table that describes two columns e.g. “ways F tests are interpreted in the literature” and “ways the geometric relationship to PCA suggests F tests should be interpreted.” We need some clarity on the meaning and appropriate use of these methods, and this paper has potential to provide a nice summary.
Some reviewers of empirical papers in human genomics reflexively expect authors to augment studies that interpret population relationships using PCA, MDS, Structure, Admixture, etc. with F tests. I gather from this paper that F tests are accomplishing the same kind of post-processing of a PCA plot as, say, the so-called average linkage distance (e.g. Kopelman et al. 2009 BMC Genetics for an example). It would be great if the paper could clarify.
TODO

5. In Section 5.3, the first three paragraphs seem to be setting up a significant and strong conclusion, but the last two paragraphs do not really deliver one. Move the discussion of technical consequences of complexities in human population structure for the qp methods earlier. The two paragraphs about Lewontin’s paper are not really used to make the subsequent points; authors for the special issue should not feel obligated to extensively discuss Lewontin 1972 if the relationships to their papers are tenuous.

Specific comments (author’s line numbers rather than journal’s)
8 “to understand this method” – unclear
changed to "tools to understand this variation"
9 Second consecutive sentence beginning with “Here”
Removed both Here
10 Clarify that these are F2, F3, and F4 statistics and not Wright’s F
statistics 
added "sensu patterson"
13 “good population structure is sparse” – phrase is unclear, as is the meaning of “good”
14 “most statistics” – what is meant?
changed to:
"In both examples, I find that the first few PCs are sufficient to approximate
most $F$-statistics, and thus PCA-plots are effective to predict
$F$-statistics."

17 “less biased” in what sense? Note that there is a “bias” in the paper in that a systematic underestimation occurs from summing only some of the PCs.
REVISE

37 There is a mismatch of the citations in this line and the examples cited. Consider citing specific papers for each topic rather than in a block of citations.

41 Boca et al. 2020 J Math Biol is possibly appropriate where you cite Gopalan et al.
46 Cite T Jombart et al. 2009 Heredity.
51 “not designed to answer specific research questions” – clarify this overly vague statement, which is not supportable.
changed to "These methods are limited in their ability to estimate biologically
meaningful parameters, but provide useful summaries and visualizations"

106 Clarify assumptions. SNPs are assumed to be biallelic? One allele is chosen arbitrarily to be the allele at which the frequency is measured? All populations are assumed to be typed at all loci? Are SNPs monomorphic in a set of 3 or 4 populations considered? Are all SNPs used irrespective of whether they are monomorphic?
I added that SNPs are biallelic and we use an arbitrary allele. Missing data is
explicitly treated later.

115 Give an example of what is meant by “branches that do not exist in the tree.”
Clarified to: 
Crucially, this means that $F_4$ will be zero for pairs of populations from non-overlapping clades, which means that the tree lacks the branch corresponding to this statistic

129 “F4-statistics that are non-zero” – which F4 statistics?
Clarified to
In this case, all $F_4$-statistics involving $X_y$ and both admixture sources will be non-zero, and, in some cases, $F_3(X_y; X_2, X_3)$ will be negative

131 The journal will likely insist that any section that has subsections needs to have two subsections. At present, 2.1 has only one.

138 Clarify that each SNP contributes a dimension. I do not see the high dimensionality as a drawback. The geometric framework is very intuitive, even if it involves many dimensions. Rather than framing the high dimension as a weakness, frame the dimensionality reduction as a way to make the interpretations more intuitive.
See response to reviewer 1, comment 7 who had an analogous comment

148 Sentence needs some references. These subtle details are important to anyone who wants to run a PCA, and it would be helpful to list the main places where the differences are discussed.
Added references to Cavalli-Sforza/Patterson et al. 2006. I am not sure I can
exhaustively list references, since many packages exist, and many packages have
multiple options for normalizations or dealing with missing data.
178 This connection between PCA and MDS is not understandable enough unless one already knows it – please clarify.
I decided not to re-derive this result (see also response to reviewer 1 in
previous review), and merely state the result. I rewrote this section to better
reflect that I state a known result
Page 7 Second line of eq. 10 is an approximation.
Changed the typesetting to better reflect that this is the definition of the
approximation
237 By “supposed” do you mean “assumed”? Clarify.
changed to "expected"
238 Explain why being estimates may make the values negative.
changed to "sampling noise might make some of them slightly negative"

273-282 How about a schematic figure here.
TODO

Fig 3 caption – refer to specific equations that are referenced.
I am not quite sure what you mean with "specific equations here

Typo in denominator of equation 14: X3-X4 not X3-X3.
Fixed

355 Explain the significance of this holding for any orthonormal projection.
This is done in the follow-up section: This is useful in particular (...)

362-366 This text seems to be arguing for a non-identifiability of the admixture proportion from the position of the points if not collinear.
The point here is more that collinearity is an implicit assumption in admixture
proportion inference

394-397 Explain. Why are these concerns rendered unimportant?
TODO

404 I don’t quite follow – why is this the expectation?
Changed to: Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs as there local differentiation, which is typically reflected on higher PCs, becomes important.

430 What might this PCA-based F statistics framework look like?

472 These is an odd group of citations; I would use Cavalli et al. 1994 and the Jobling et al. book detailed discussions of population history. Since classic apportionment does not give us a way to think about ancient DNA, you’d want to refer to Witt et al. in the special issue if you are wanting to connect Lewontin’s paper to ancient DNA such as in Reich 2018.
482 Cite Novembre 2022 here.
TODO

Doc:
-Throughout the document, the distinction between the expectations of the statistics (F2, F3, F4 in Patterson notation) and the statistics themselves (f2, f3, f4 in Patterson notation) is mostly avoided. Overall, I think it is the right decision given the manuscript's focus, but in some places it could lead to confusion. As one example, in line 71-73, the claim about F3 being restricted to be non-negative under a tree structure applies to F3 but not necessarily f3. Please address this distinction somewhere (presumably in the intro or theory sections, up to your discretion) and explain it in relation to the manuscript's setting in allele-frequency space.
Added a paragraph after the definition of F-statistics:
Commonly, a distinction is made between statistics estimated from data (denoted with lowercase-$f$), and theoretical quantities (defined in eq. 1). I do not make this distinction, but will explicitly mention when I analyze statistics calculated from data.

-lines 259-260 - this point is phrased somewhat vaguely. Please consider a more precise phrasing.
TODO

-lines 273-277 - The globe analogy is useful, but it needs to be explicitly connected to the problem. (For example, state that in the example, the north and south poles represent population allele frequencies, the globe represents the space in which F3<=0, etc.) Another schematic of this scenario, though perhaps challenging to draw, would be useful, for example with two points in 3D space defining a ball, two orthogonal slices representing projections, and the circles (in 2d) defined by the positions of the points in the projections.
TODO

-In the same section about F3 on a 2D PCA plot, please explain how to compute the true (light gray in figure 3) and apparent (dark gray in figure 3) circles. The intuition for the dark gray circle comes across via figure 2, but it will not be obvious to many readers how to think of the light circle. A rigorous calculation may not be necessary. One intuitive explanation (if the following is a correct reading!) might be that one computes the distance between the populations in the full space, then draws a circle with that diameter centered on the midpoint between the positions of the points in the projection. It would be helpful to contrast this with an explanation of the dark grey circle as centered in the same spot, but with a diameter defined by the distance between the points in the current 2D projection, which is guaranteed to be no greater than the full distance.
TODO

-In your eq A4, plus lines 278-282, you note that there is a strict interpretation that if the point lies outside the circle in any one projection, then it is guaranteed to be outside the n-ball. You also state that it is possible for it to lie inside the circle on any one biplot yet still have positive F3. I think it is also true that it can lie inside the (light gray) circle on all biplots and still have positive F3. (e.g. in 3D, the volume contained in the intersection of the cylinders defined by the circles from the 2D projections will be a bicylinder that completely contains the ball and has a bit of extra volume too.) Please address this.
TODO

-I am confused by the qualifier "i.e. it is a test statistic" in lines 328-329. Isn't it just an F4 computed for a set of populations for which no corresponding branch exists in the tree? How does testing relate?
This uses the notation I introduced in my previous F-stats paper. I agree it's
not particularly useful and I removed it.

Figure 3D: I am intrigued by the apparently big change from 2 to 3 PCs---can we learn anything worth a comment from that?
It simply means that the 3rd PC is relevant for this data set. I added this
sentence: "Both results indicate that adding a third PC would give a much better
description of the data, and the relationship between within-European variation
to Saudis in particular."

Lines 476-478: Please change the phrasing to make it broader, for example changing to something like "However, understanding, conceptualizing, and modeling this variation is far from trivial, particularly in a historical context in which mistaken ideas about human variation have been used to justify racist, eugenic, and even genocidal policies." I am suggesting the change because in the previous phrasing "models of human genetic differentation", one could read it as though e.g. Madison Grant was misinterpreting the meaning of an admixture graph, rather than working in a totally bankrupt framework.
I appreciate this comment, and changed the MS accordingly.

-lines 478-480: This phrasing seems to emphasize the ratio of (among continent) / (total - among population), as opposed to the typical headline of (among population) / total. That may be intentional; I just want to flag it for your notice.
This is intentional, in part because I think it is an important part, but
mainly because I actually do not discuss within-population variation either.

Typos and other minor points

-line 162: "data is" -> "data are"
-title of section 2.3, "F-statsitics" -> "F-statistics"
-line 219, "which" -> "that"
-line 220, "represents" -> "represent"
-line 508, "I do" -> "it does"
fixed

### 2022-01-26 fstats_pca_lewontin.tex
response no fig

@@
 My results extend $F$-statistics to non-discrete populations, moving towards more complete descriptions of human genetic variation.
+\end{abstract}
 \section{Introduction}
@@
 
+Commonly, a distinction is made between statistics estimated from data (denoted with lowercase-$f$), and theoretical quantities (defined in eq.
+1).
+I do not make this distinction, but will explicitly mention when I analyze statistics calculated from data.
+
 $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}.
@@
 PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g.
-in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis \citep{cavalli-sforza1994, patterson2006}.
+in how SNPs are standardized, how missing data are treated or whether we use individuals or populations as units of analysis \citep{cavalli-sforza1994, patterson2006}.
 The version of PCA I use here is set up such that the similarities to  $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets.
@@
 PCA allows us to approximate the points in the high-dimensional allele frequency space by a $K$-dimensional subspace of the  data.
-If all PCs are considered, $K=n-1$, in which case the data is simply rotated.
+If all PCs are considered, $K=n-1$, in which case the data are simply rotated.
 However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.
@@
 
-\subsection{Connection between PCA and $F$-statsitics}	
+\subsection{Connection between PCA and $F$-statistics}	
 \subsubsection{Principal components from $F$-statistics}
@@
 Both are based on the analyses by \cite{lazaridis2014}.
-The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs).
+The data are from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs).
 SNPs with missing data in any population are excluded.
@@
 \paragraph{``World'' data set}
-This data set is a subset of the ``World Foci'' data set of \cite{lazaridis2014}, where I removed samples which are not permitted for free reuse.
+This data set is a subset of the ``World Foci'' data set of \cite{lazaridis2014}, where I removed samples that are not permitted for free reuse.
 These populations span the globe and roughly represents global human genetic variation (638 individuals from 33 population) As adjacent sampling locations are often thousands of kilometers apart, I speculate that gene flow between these populations may not be particularly common; and their structure may therefore be well-approximated by an admixture graph.
@@
 To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.
-The right-hand side matrix of equation \ref{eq:mds} is supposed to have non-negative eigenvalues (i.e.
+The right-hand side matrix of equation \ref{eq:mds} is expected to have non-negative eigenvalues (i.e.
 $-\MC\MF\MC$ is positive-semidefinite).
-However as $F_2$-statistics are estimates, some eigenvalues might be slightly negative, which would lead to imaginary PCs.
+However when $F_2$-statistics are estimated from data, sampling noise might make some of them  slightly negative, which would lead to imaginary PCs.
 I avoid this by setting all negative eigenvalues to zero.
@@
 \paragraph{Example}
-As an example, I visualize the admixture statistic $F_3(X; \text{Sardinian}, \text{Finnish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A).
+As an example, I visualize the admixture statistic $F_3(X; \text{Sardinian}, \text{Finnish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:weu}A).
 In this case, the projected $n$-ball (light gray) and circle based on two dimensions (dark gray) have similar sizes.
@@
 This reveals that the first two PCs do not capture all the genetic variation relevant for  European population structure.
-Consequently, approximating $F_3$ by the first two or even ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.
+Consequently, approximating $F_3$ by the first two or even ten PCs (Figure \ref{fig:weu}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:weu}C we see that many higher PCs contribute to $F_3$ statistics.
 
@@
 E: Contribution of the first ten PCs to select $F_4$-statistics, with the first three PCs containing the majority of contributions.}
-	\label{fig:f3}
+	\label{fig:weu}
 \end{figure}
@@
 	}	
-	\label{fig:f4}
+	\label{fig:world}
 \end{figure}
@@
 \paragraph{Example}
-In Figure \ref{fig:f4}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, in i.e.
+In Figure \ref{fig:world}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, in i.e.
 a statistic that aims to find the population most closely related to Sardinian (a Mediterranean Island), assuming the Mbuti are an outgroup to all populations in the data set.
@@
 In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia and the Americas  project almost orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.
-We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}C), but keep the relative ordering.
+We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:weu}C), but keep the relative ordering.
 I also find that many PCs, e.g.
-PCs 3-5, 7 and 10 have almost zero contribution to all $F_3$-statistics (Figure \ref{fig:f3}D), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.
+PCs 3-5, 7 and 10 have almost zero contribution to all $F_3$-statistics (Figure \ref{fig:weu}D), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.
 
@@
 $F_4$-statistics that correspond to an internal branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot (Figure \ref{fig:geom}G), again with the caveat that we need to scale it by a constant.
-If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e.
+If the $F_4$-statistic corresponds to a branch that does not exist in the tree (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e.
 that $X_1$ and $X_2$  map to the same point on the projection axis $\overline{X_3X_4}$ (Figure \ref{fig:geom}H).
@@
 \begin{equation}
-\text{Cor}(X_1 - X_2, X_3 - X_4) =  \frac{F_4( X_1, X_2; X_3, X_4) }{\norm{X_1-X_2}\norm{X_3-X_3}} = \cos(\phi),\label{eq:angle}
+\text{Cor}(X_1 - X_2, X_3 - X_4) =  \frac{F_4( X_1, X_2; X_3, X_4) }{\norm{X_1-X_2}\norm{X_3-X_4}} = \cos(\phi),\label{eq:angle}
 \end{equation}
@@
 To illustrate the angle interpretation I return to the Western Eurasian data.
-The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:f3}A), a European gradient (from Sardinian to Finnish and Chuvash), and an Asian cline from Arab populations (top right) to the Caucasus  (bottom right).
-This is quantified in Figure \ref{fig:f3}D, where I plot the angle corresponding to $F_4(X, \text{Saudi}; \text{Sardinian}, \text{Finnish})$.
+The PCA-biplot shows two roughly parallel clines (Figure \ref{fig:weu}A), a European gradient (from Sardinian to Finnish and Chuvash), and an Asian cline from Arab populations (top right) to the Caucasus  (bottom right).
+This is quantified in Figure \ref{fig:weu}D, where I plot the angle corresponding to $F_4(X, \text{Saudi}; \text{Sardinian}, \text{Finnish})$.
 For most Asian populations, using two PCs (green points) gives an angle close to zero, corresponding to a correlation coefficient between the two clines of $r>0.9$.
 Just adding a third PC (blue), however, shows that the clines are not, in fact, parallel, and the correlation for most populations is low.
-The finding that three PCs are necessary to explain this data can also be seen from the spectrum of these $F_4$-statistics (Figure \ref{fig:f3}E), which have high contributions only from the first three PCs.
+The finding that three PCs are necessary to explain this data can also be seen from the spectrum of these $F_4$-statistics (Figure \ref{fig:weu}E), which have high contributions from the first three PCs.
+Both results indicate that adding a third PC would give a much better description of the data, and the relationship between within-European variation to Saudis in particular.
+
 \begin{figure}[!ht]
@@
 \subsubsection{Example}
-I use the World data set as an example, using Sardinian and Mbuti as references populations (Figure \ref{fig:f4}B).
-The data is the same as in  the PCA (Figure \ref{fig:f4}A), but it is now rotated such that the axis between the reference population (black line in Figure \ref{fig:f4}A) is now reflected by the x-axis.
+I use the World data set as an example, using Sardinian and Mbuti as references populations (Figure \ref{fig:world}B).
+The data are the same as in  the PCA (Figure \ref{fig:world}A), but it is now rotated such that the axis between the reference population (black line in Figure \ref{fig:world}A) is now reflected by the x-axis.
 For any pair of populations $X_1$ $X_2$, their horizontal projection distance reflects $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$ and the relative horizontal distance corresponds exactly to $F_4$-ratio admixture estimates.
@@
 My interpretation here is different in that the utility of PCA is to simplify the geometry of the data.
-One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA, but, a very skewed sampling distribution will increases the likelihood that more, or different PCs will have to be included in the analysis.
+One consequence is that the results here are not impacted by sample ascertainment, or sample sizes which are common concerns in the interpretation of PCA \citep{novembre2008a, mcvean2009, francois2010}.
+Adding more PCs will provide a gradually better approximation of $F$-statistics, and a  skewed sampling distribution would likely require more PCs to be included in an analysis.
 
-The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
-On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
-Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.
+The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
+On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
+Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs as there local differentiation, which is typically reflected on higher PCs, becomes important.
 
@@
 This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness.
-A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians.
+A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:weu}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians.
 In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.

### 2022-01-26 fstats_pca_lewontin.tex
response no fig

@@
 
-However, understanding, conceptualizing and modeling this variation is far from trivial, and misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies.
+However, understanding, conceptualizing and modeling this variation is far from trivial, particularly in a historical context in which mistaken ideas about human variation have been used to justify racist, eugenic and genocidal policies.
 Lewontin’s landmark 1972 paper on the apportionment of human genetic diversity was one of the first to quantify how little of between-population genetic variation could be attributed to ``racial'' continental-scale groupings.

### 2022-01-26 fstats_pca_lewontin.tex
proj figure

@@
 $X_y$ originates on the segment connecting $X_2$ and $X_3$, and subsequent drift may move it in a random direction.
-C: $F_3(X_x; X_2, X_3)$ and negative region (light gray circle).
-$F_4(X_1, X_x; X_3, X_4)$ will no longer be zero (compare to Figure \ref{fig:geom}D).
+C: Negative region (light gray) for $F_3(X_x; X_2, X_3)$ and for $\hat{F}_3^{(2)}(X_x; X_2, X_3)$ based on two dimensions (dark gray).
+D: $F_4(X_1, X_x; X_3, X_4)$ will no longer be zero (compare to Figure \ref{fig:geom}D).
 }
@@
 
+\begin{figure}[!ht]
+	\includegraphics[width=.4\textwidth]{figures/fig_proj.pdf}
+	\caption{\textbf{Admixture-$F_3$-statistic and Dimensionality reduction.} A: $F_3(X_x; X_1, X_2)$ in two dimensions (full data).
+Points like $X_x$ inside the circle (red) have $F_3 <0$, but points outside ($X_3, X_4$, blue) have positive $F_3$-statistics.
+B: Projection on one dimension.
+The full rejection region (light gray bar) is larger than that predicted from the distance between $X_1$ and $X_2$ (dark grey bar).
+Points with negative $F_3$ are guaranteed to lie in the light gray (but not dark gray).
+$X_3$ also projects onto the gray bar, even though its $F_3$ is positive.
+Points such as $X_4$ that project outside the rejection region have $F_3>0$.}
+	\label{fig:proj}
+\end{figure}
+
 \subsection{When are admixture-$F_3$ statistics negative?}
@@
 
-\paragraph{$F_3$ on a 2D PCA-plot.} 
+\paragraph{$F_3$ and dimensionality reduction} 
  If we project this $n$-ball on a two-dimensional plot, $\overline{X_2X_3}$ will usually not align with the PCs; thus the ball may be somewhat larger than it appears on the plot.
-This geometry is perhaps easiest visualized on a globe.
-If we look at the globe from a view point parallel to the equator, both the north and south poles are visible at the very edge of the circle.
-But if we look at it from above the north pole, the north- and south-poles will be at the very same point.
-
- Thus if $\hat{F}_3 \ll F_3$, the ``true'' circle will be bigger than what would be predicted from a 2D-plot, and populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics.
-This is because they are outside the $n$-ball in higher dimensions.
-The converse interpretation is more strict: if a population lies outside the circle on \emph{any} 2D-projection, $F_3$ is guaranteed to be bigger than 0 (see Equation \ref{eq:f3circle} in the Appendix).
+This geometry is perhaps easiest visualized on the example of projecting $F_3(X_x; X_1, X_2)$ from a two-dimensional space onto a single dimension (Figure \ref{fig:proj}).
+In that example, the distance between $X_1$ and $X_2$ has a substantial contribution from PC2, and so the negative region (light gray) is larger than what would be predicted from just one dimension (dark gray bar), but if $\hat{F}_3 \approx F_3$, the two areas would be very close.
+Thus, if considering a reduced-dimension PCA-plot, some points (such as $X_3$) may project inside the negative region, but have positive $F_3$ because they are outside the $n$-ball in higher dimensions.
+The converse interpretation is more strict: if a population lies outside the circle on \emph{any} projection, $F_3$ is guaranteed to be bigger than 0 (see Equation \ref{eq:f3circle} in the Appendix).
+An intuitive example is given by $X_4$ in Figure \ref{fig:proj}: all points projecting to the same point on Figure \ref{fig:proj}B as $X_4$ lie outside the circle.
 

### 2022-01-31 response.txt
merge

@@
 It would be great if the paper could clarify.
-TODO
+I added a short conclusion where I summarize the main points I think the paper
+contributes. I agree that it would be nice (and practical) to have a table
+describing further applications, but an issue is that I still don't fully understand
+some of the more complex of Nicks tools (such as qpgraph and qpwave), and so I do not think I can critisize
+them in a constructive way.
 
@@
 The two paragraphs about Lewontin’s paper are not really used to make the subsequent points; authors for the special issue should not feel obligated to extensively discuss Lewontin 1972 if the relationships to their papers are tenuous.
+I did rephrase this a little bit, since I agree that the criticism of qpXXX is
+not warranted. I think the point would really be that F-stats and PCA could
+allow us to study population structure by setting up orthogonality tests without
+making treeness assumptions.
 
@@
 17 “less biased” in what sense? Note that there is a “bias” in the paper in that a systematic underestimation occurs from summing only some of the PCs.
-REVISE
+TODO
 
@@
 273-282 How about a schematic figure here.
-TODO
+I did add a small figure (now Figure 3). Due to the difficulty of visualizing in
+3D, I opted to only do a 2D plot (the logic transfers)
 
@@
 394-397 Explain. Why are these concerns rendered unimportant?
-TODO
+Changed this section to explain that this is because I just aim at simplifying
+the geometry, rather than interpret it in terms of biological processes
 
@@
 430 What might this PCA-based F statistics framework look like?
+TODO
 
@@
 482 Cite Novembre 2022 here.
-TODO
+I added a tentative entry, that I will supplement as soon as I have full
+bibliographic info.
 
@@
 -lines 259-260 - this point is phrased somewhat vaguely. Please consider a more precise phrasing.
-TODO
+changed to: "In converse, populations that are distant on the first few PCs are
+guaranteed to also have a large $F_2$-distance, since the distance contributed
+from the omitted PCs cannot be negative."
 
@@
 (For example, state that in the example, the north and south poles represent population allele frequencies, the globe represents the space in which F3<=0, etc.) Another schematic of this scenario, though perhaps challenging to draw, would be useful, for example with two points in 3D space defining a ball, two orthogonal slices representing projections, and the circles (in 2d) defined by the positions of the points in the projections.
-TODO
 
@@
 It would be helpful to contrast this with an explanation of the dark grey circle as centered in the same spot, but with a diameter defined by the distance between the points in the current 2D projection, which is guaranteed to be no greater than the full distance.
-TODO
 
@@
 in 3D, the volume contained in the intersection of the cylinders defined by the circles from the 2D projections will be a bicylinder that completely contains the ball and has a bit of extra volume too.) Please address this.
-TODO
+(Response to last three comments): I agree that this is an important issue, and
+now explicitly addressed in Figure 3. Unfortunately, I could not make this work
+in 3D, but I think the 2D->1D Figure also works well, as the intuition
+translates.
 

### 2022-01-31 fstats_pca_lewontin.tex
merge

@@
 Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations, and likely more PCs need to be investigated to understand how these populations are related to each other.
-In converse, populations that are distant on a PCA-plot are guaranteed to also have a large $F_2$-distance.
+In converse, populations that are distant on the first few PCs are guaranteed to also have a large $F_2$-distance, since the distance contributed from the omitted PCs cannot be negative.
 
@@
 	\includegraphics[width=.4\textwidth]{figures/fig_proj.pdf}
-	\caption{\textbf{Admixture-$F_3$-statistic and Dimensionality reduction.} A: $F_3(X_x; X_1, X_2)$ in two dimensions (full data).
-Points like $X_x$ inside the circle (red) have $F_3 <0$, but points outside ($X_3, X_4$, blue) have positive $F_3$-statistics.
+	\caption{\textbf{Admixture-$F_3$-statistic and Dimensionality reduction.} A: The gray circle is defined by $X_1$ and $X_2$ and reflects the area for which $F_3$ is negative, i.e.
+$F_3(X_x; X_1, X_2) < 0 < F_3(X_3; X_1, X_2) \leq F_3(X_4; X_1, X_2)$.
 B: Projection on one dimension.
-The full rejection region (light gray bar) is larger than that predicted from the distance between $X_1$ and $X_2$ (dark grey bar).
-Points with negative $F_3$ are guaranteed to lie in the light gray (but not dark gray).
-$X_3$ also projects onto the gray bar, even though its $F_3$ is positive.
-Points such as $X_4$ that project outside the rejection region have $F_3>0$.}
+The full rejection region (light gray bar) is given by the boundaries of the circle in A, and is larger than that predicted from the distance between $X_1$ and $X_2$ (dark grey bar) on PC1 alone.
+Points like $X_x$ with negative $F_3$ are guaranteed to lie in the light gray (but not dark gray) region.
+$X_3$ also projects onto the gray bar, even though it is outside the circle and $F_3$ is positive, and it would also lie inside the positive area if projected onto PC2.
+However, points such as $X_4$ that project outside the rejection region have $F_3>0$, since the associated projection line does not intercept the circle.}
 	\label{fig:proj}
@@
 Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2010, francois2021}.
-My interpretation here is different in that the utility of PCA is to simplify the geometry of the data.
-One consequence is that the results here are not impacted by sample ascertainment, or sample sizes which are common concerns in the interpretation of PCA \citep{novembre2008a, mcvean2009, francois2010}.
-Adding more PCs will provide a gradually better approximation of $F$-statistics, and a  skewed sampling distribution would likely require more PCs to be included in an analysis.
+My interpretation here is different in that the utility of PCA is to simplify the geometry of the data, rather than attributing meaning to the produced PCs.
+One consequence is that the results here are less directly impacted by sample ascertainment, sample sizes or number-of-PCs, which are common concerns in the interpretation of PCA \citep{novembre2008a, mcvean2009, francois2010, francois2021}; adding more PCs will provide a successively better approximation of the $F$-statistics.
 
@@
 However, understanding, conceptualizing and modeling this variation is far from trivial, particularly in a historical context in which mistaken ideas about human variation have been used to justify racist, eugenic and genocidal policies.
-Lewontin’s landmark 1972 paper on the apportionment of human genetic diversity was one of the first to quantify how little of between-population genetic variation could be attributed to ``racial'' continental-scale groupings.
+Lewontin’s landmark 1972 paper on the apportionment of human genetic diversity was one of the first to quantify how little of between-population genetic variation could be attributed to ``racial'' continental-scale groupings \citep{novembre2022}.
 Over the last five decades, this view has been corroborated, refined and extended many times \citep{cann1987, cavalli-sforza1994, barbujani1997, rosenberg2002a}.
@@
 
+\begin{comment}
 This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}.
@@
 The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not need to assume that gene flow is rare.
+\end{comment}
+
+An alternative could be to focus on orthogonality of drift as a \emph{the} central concept for analyzing genetic variation data.
+This is often informally done in the context of PCA, where it is often observed that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}.
+These associations are not exact, and impacted e.g.
+by the size and composition of the analyzed data set.
+On the other hand, we can use $F$-statistics to formally test orthogonality.
+Distinct clades in an admixture graph would be one possibility how orthogonality could arise, but assuming a graph is not necessary.
+Motivated by PCA, we could set up $F_4$ as tests of orthogonality due to either space (distinct populations evolve independently) or scale (i.e.
+between-population diversity is independent of within-population diversity).
+This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming discrete lineages.
 
 \section{Conclusion}
-\begin{itemize}
-	\item $F$-stats and PCA are partially redundant
-	\item Population structure is sparse - trees	
-	\item Formulating hypotheses
-	\item Orthogonality as a central concept
-	\item Orthogonality of scales (within-population variation) is independent of between-population variation
-	\item Orthogonality of isolation (two locations have little exchange)
-\end{itemize}
+$F$-statistics and PCA have both proven to be tremendously useful to study, visualize and test aspects of population structure.
+Here, I show that these approaches are closely connected, and highlight a few implications.
+First, PCA and $F$-statistics should never be treated as independent analyses.
+If they agree, this can be used as a sanity check that no major assumptions about e.g.
+population groupings are violated, but the underlying biological relationships investigated are the same in both approaches.
+
+If sufficient PCs are considered, both $F_3$ and $F_4$-statistics do have simple interpretations in PCA-space.
+If used as a test for admixture, $F_3$ corresponds to a $n$-sphere in PCA-space, which manifests as a circle on a PCA-plot.
+The informal notion that an admixed population should lie between its sources on a PCA-plot is necessary, but not sufficient; unadmixed populations may also project there.
+In contrast, it is sufficient (but not necessary) to determine that a population has a positive $F_3$ statistic.
+
+Interpretations of the outgroup-$F_3$-statistics and $F_4$-statistics in a PCA-framework rely heavily on the idea of a projection: In a tree or admixture-graph framework, we can loosely interpret $F_4(A, B; C, D)$ as a projection of one the path $A-B$ onto $C-D$; only edges shared between the paths will contribute non-zero amounts to this statistics.
+This projection analogy holds when we replace the populations by points in PCA-space, which enables us to study population models beyond discrete graphs.
+In this context, $F_4$ is simply an orthogonality test, an $F_4$ of zero means that the pairs of populations are orthogonal (i.e.
+independent) of another.
+
+One difficulty in interpreting the results here is that they only hold when all PCs are considered.
+Yet, in my applications I found that two (for the World data set) and three PCs (for Western Eurasia) already provided a very good approximation for the majority of $F$-statistics, so that the results here should be applicable to at least continental-scale data sets.
+
 \section{Supplementary File Description}

### 2022-01-31 fstats_pca_lewontin.tex
stuff

@@
 \begin{abstract}
-Human  genetic diversity is shaped by our complex history.
-Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics \textit{sensu} Patterson.
-I show that these two perspectives are closely related, and I derive explicit connections between the two approaches.
+Principal Component Analysis (PCA) and $F$-statistics \textit{sensu} Patterson are two of the most widely used population genetic tools to analyze human genetic variation.
+Here, I show that these two methods are closely related, and I derive explicit connections between the two approaches.
 I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link.
-I illustrate my results on two examples, one of local, and one of global human diversity.
+I show that admixture corresponds to I illustrate my results on two examples, one of local, and one of global human diversity.
 In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and thus PCA-plots are effective to predict $F$-statistics.

### 2022-02-01 fstats_pca_lewontin.tex
rev3

@@
 
-\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
+\title{A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis}
 \author{Benjamin M Peter}
@@
 \begin{abstract}
-Principal Component Analysis (PCA) and $F$-statistics \textit{sensu} Patterson are two of the most widely used population genetic tools to analyze human genetic variation.
-Here, I show that these two methods are closely related, and I derive explicit connections between the two approaches.
-I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link.
-I show that admixture corresponds to I illustrate my results on two examples, one of local, and one of global human diversity.
-In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and thus PCA-plots are effective to predict $F$-statistics.
-Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models.
-My results extend $F$-statistics to non-discrete populations, moving towards more complete descriptions of human genetic variation.
+Principal Component Analysis (PCA) and $F$-statistics \textit{sensu} Patterson are two of the most widely used population genetic tools to study human genetic variation.
+Here, I derive explicit connections between the two approaches and  show that these two methods are closely related.
+$F$-statistics have a simple geometrical interpretation in the context of PCA, and orthogonal projections are a key concept to establish this link.
+I show that negative $F_3$ corresponds to a circle on a PCA plot, and that $F_4$ is related to an angle measurement.
+I illustrate my results on two examples, one of Western Eurasian, and one of global human diversity.
+In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and that PCA-plots are effective at predicting $F$-statistics.
+My results extend $F$-statistics to more continuous population models, moving towards a more complete descriptions of human genetic variation.
 \end{abstract}
@@
 However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}.
-In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}.
+In addition,  major historical population movements such as the out-of-Africa \citep{ramachandran2005}, Austronesian \citep{stoneking2016} or Bantu expansions \citep{li2014} lead to more gradual patterns of genetic diversity over space \citep{francois2010}.
 Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{boca2020}.
@@
 A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}.
-Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}.
+Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, jombart2009, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}.
 These methods are limited in their ability to estimate biologically meaningful parameters, but provide useful summaries and visualizations.
@@
 \subsection{Formal Definition of $F$-statistics}
-Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ biallelic loci.
+Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ biallelic loci, and for simplicity I will assume that there is no missing data.
 Let $x_{il}$ denote the frequency of an arbitrary allele at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$.
@@
 
-\subsubsection{Geometric interpretation of $F$-statistics}
+\subsection{Geometric interpretation of $F$-statistics}
 An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare.
@@
 
-The geometric approach of \cite{oteo-garcia2021} assumes a very high-dimensional space, as each SNP in a data set adds a dimension, and data sets have commonly a million or more SNPs.
-Such high-dimensional spaces are hard to visualize and analyze, however, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data \citep{patterson2006}.
+One issue with the geometric approach of \cite{oteo-garcia2021} is that  each SNP (commonly in the millions) adds a dimension, but high-dimensional spaces are hard to visualize, interpret and analyze.
+Fortunately, it has been commonly observed that population structure is often quite low-dimensional, and only a  few PCs frequently provide a good approximation of the covariance structure in human genetic variation data \citep{patterson2006}.
 Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.
@@
 These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}.
-The theory developed here suggests that developing a PCA-based $F$-statistics framework might be more powerful, as related populations could be used to fill in missing data.
+One alternative suggested by the theory developed here could be to first perform a PCA to impute missing data, and then calculate $F$-statistics using this imputed data set.
+Using this approach, the accuracy of $F$-statistics including low-coverage samples would be greatly increased, as they can ``borrow'' information from closely related samples.
 
@@
 between-population diversity is independent of within-population diversity).
-This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming discrete lineages.
+This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming discrete lineages or admixture events.
 

### 2022-02-07 fstats_pca_lewontin.tex
revision3 part 1

@@
 $F$-statistics have a simple geometrical interpretation in the context of PCA, and orthogonal projections are a key concept to establish this link.
-I show that negative $F_3$ corresponds to a circle on a PCA plot, and that $F_4$ is related to an angle measurement.
+I show that for any pair of populations, any population that is admixed as determined by an $F_3$-statistic will lie inside a circle on a PCA plot.
+Furthermore, the $F_4$-statistic is closely related to an angle measurement, and will be zero if the differences between pairs of populations intersect at a right angle in PCA space.
 I illustrate my results on two examples, one of Western Eurasian, and one of global human diversity.
 In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and that PCA-plots are effective at predicting $F$-statistics.
-My results extend $F$-statistics to more continuous population models, moving towards a more complete descriptions of human genetic variation.
+Thus, while F-statistics are commonly understood in terms of discrete populations, the geometric perspective illustrates that they can be viewed in a framework of populations that vary in a more continuous manner.
 \end{abstract}
@@
 On a PCA-plot, we can visualize this $F_3$-statistic as the projection of the vector $X_i - X_O$ onto $X_U - X_O$: 
-$$\vectorproj[X_U - X_O]{X_i - X_O} =F_3(X_O; X_U, X_i) \frac{X_U - X_O}{F_2(X_O; X_U)}.$$
+$$\vectorproj[X_U - X_O]{(X_i - X_O)} =F_3(X_O; X_U, X_i) \frac{X_U - X_O}{F_2(X_O; X_U)}.$$
 
@@
 The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
-On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
-Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs as there local differentiation, which is typically reflected on higher PCs, becomes important.
+On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs \citep{cavalli-sforza1975}.
+Thus, I would expect that approximating $F$-statistics with few PCs works well whenever local variation is unimportant, i.e.
+when calculating $F$-statistics far apart in PCA space.
+On the flip side, populations close-by on a PCA-plot are often seperated on higher PCs, and thus whenever close-by populations are included in the same $F$-statistics, these higher PCs will be required for good approximations.
 
@@
 \subsection{Assumptions}
-The other cause for disagreements between $F$-statistics and PCA are differences in assumptions.
+In addition to the selection of PCs, the other cause for disagreements between $F$-statistics and PCA are differences in assumptions.
 The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized.
@@
 \subsubsection{Missing data}
-The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing missing data have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
-Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}.
-In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation.
-These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}.
-One alternative suggested by the theory developed here could be to first perform a PCA to impute missing data, and then calculate $F$-statistics using this imputed data set.
-Using this approach, the accuracy of $F$-statistics including low-coverage samples would be greatly increased, as they can ``borrow'' information from closely related samples.
+The matrix decompositions underlying PCA assume complete data, and thus cannot be used when some data is missing \citep{jollife2013}.
+As missing data is a very common practical problem, there is a large number of algorithms for imputing missing data.
+The simplest approach is to replace missing data with zeros \citep[as implemented e.g.
+in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
+In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard error.
+
+These strategies are distinct, and reflect the original purposes of the approaches.
+For statistical tests based on $F$-statistics, we wish to isolate a set of three or four populations and get our best guess based on just that subset of data.
+In contrast methods for PCA can leverage the additional individuals, and thus will likely result in more accurate estimates.
+
+However, the way we handle missing data is not tied into the method.
+For example, we could evaluate the robustness of a PCA by resampling data.
+Similarly, the theory developed here suggests that we could obtain accurate $F$-statistics with missing data by first performing a PCA using a method that handles missing data, and then calculate $F$-statistics from these PCs.
 
@@
 \subsubsection{Individual vs. population-based analyses}
-The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.
-This is not necessary; population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
-Thus, more commonly the choice between individual-based and population-based approaches is in the goal of an analysis.
-Grouping individuals into populations introduces a major assumption, that can e.g.
-be justified using an individual-based PCA.
-In particular, since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on an individual-based PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.
+The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.However, population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
+Often, an individual-based PCA is used to justify grouping individuals into populations; i.e.
+individuals that form a tight cluster on a PCA-plot have similar relationships to everyone else in the data set, and can thus be treated as a unit of analysis.
+Thus, if the assumptions are satisfied, $F$-statistics for individual-based and population-based analyses are expected to be very similar.
+PCA, on the other hand, is strongly impacted by the number of individuals from each population \citep[e.g.][]{mcvean2009}; as each individual is weighted equally, variation related to populations with many samples will be overrepresented on the first PCs.
 

### 2022-02-08 fstats_pca_lewontin.tex
conclusion v4

--- 2022-02-01 fstats_pca_lewontin.tex
@@
 
-The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
+One difficulty in interpreting my results is that they only hold when all PCs are considered, and thus are approximations when considering only the first few PCs.
+However, for the two  data sets I analyzed here, two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E).
+Thus, we may expect that whenever the first few PCs provide a good approximation of population structure, this results in well-approximated $F_4$-statistics.
 On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
@@
 
+Thus, a more abstract way of interpreting $F$-statistics is that they reflect the relative position of population to each other in PCA-space.
+The classical model of admixture as a result of contact between long-separated and isolated populations is just one of potentially many demographic models that might result in this data pattern.
+
 \begin{comment}
@@
 
-An alternative could be to focus on orthogonality of drift as a \emph{the} central concept for analyzing genetic variation data.
+In this vein, admixture tests based on $F$-statistics evaluate if drift in different parts of the tree are orthogonal to each other.
+This orthogonality is expected to arise whenever drift in different populations is independent, which happens, for example, if we consider two isolated populations, or if we consider the changes in allele frequency in distinct time intervals.
+
+In the context of tree-like population models, this orthogonality arises from drift on non-overlapping branches.
+
+an alternative could be to focus on orthogonality of drift as a \emph{the} central concept for analyzing genetic variation data.
+
 This is often informally done in the context of PCA, where it is often observed that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}.
@@
 between-population diversity is independent of within-population diversity).
-This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming discrete lineages or admixture events.
+This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming that lineages or admixture events are discrete.
 
@@
 
-If sufficient PCs are considered, both $F_3$ and $F_4$-statistics do have simple interpretations in PCA-space.
-If used as a test for admixture, $F_3$ corresponds to a $n$-sphere in PCA-space, which manifests as a circle on a PCA-plot.
-The informal notion that an admixed population should lie between its sources on a PCA-plot is necessary, but not sufficient; unadmixed populations may also project there.
-In contrast, it is sufficient (but not necessary) to determine that a population has a positive $F_3$ statistic.
+If sufficiently many PCs are considered, both $F_3$ and $F_4$-statistics do have simple interpretations in PCA-space.
+If used as a test for admixture, $F_3$ corresponds to testing whether the admixed population lies in a $n$-sphere between the potential source population in PCA-space.
+This result makes the informal notion that an admixed population should lie between its sources on a PCA-plot more precise.
+Furthermore, I show that while it is necessary for an admixed population to lie between its sources on a PCA plot, this is not a sufficient condition and unadmixed populations may also project inside the $n$-sphere.
+In contrast, if a population falls outside the $n$-sphere on any PCA-plot, this is sufficient to determine that this population has a positive $F_3$ statistic.
 
-Interpretations of the outgroup-$F_3$-statistics and $F_4$-statistics in a PCA-framework rely heavily on the idea of a projection: In a tree or admixture-graph framework, we can loosely interpret $F_4(A, B; C, D)$ as a projection of one the path $A-B$ onto $C-D$; only edges shared between the paths will contribute non-zero amounts to this statistics.
-This projection analogy holds when we replace the populations by points in PCA-space, which enables us to study population models beyond discrete graphs.
-In this context, $F_4$ is simply an orthogonality test, an $F_4$ of zero means that the pairs of populations are orthogonal (i.e.
-independent) of another.
-
-One difficulty in interpreting the results here is that they only hold when all PCs are considered.
-Yet, in my applications I found that two (for the World data set) and three PCs (for Western Eurasia) already provided a very good approximation for the majority of $F$-statistics, so that the results here should be applicable to at least continental-scale data sets.
+Interpretations of the outgroup-$F_3$-statistics and $F_4$-statistics in a PCA-framework rely heavily on the geometric concepts of projections and orthogonality: In a tree or admixture-graph framework, we can loosely interpret $F_4(A, B; C, D)$ as a projection of one the path $A-B$ onto $C-D$; only edges shared between the paths will contribute non-zero amounts to this statistic.
+This interpretation holds when we replace the populations by points in PCA-space, which enables us to study population models beyond discrete graphs.
+Thus, the geometric framework enables us to expand the application of $F$-statistics beyond current uses, allows us to better understand the meaning of these statistics and helps us to avoid overinterpretations particularly in cases when populations structure is continuous.
 

### 2022-02-08 fstats_pca_lewontin.tex
local revs

--- 2022-02-07 fstats_pca_lewontin.tex
@@
 \subsubsection{Missing data}
-The matrix decompositions underlying PCA assume complete data, and thus cannot be used when some data is missing \citep{jollife2013}.
+The matrix decompositions underlying PCA assume complete data, and thus cannot be used when some data is missing \citep{jolliffe2013}.
 As missing data is a very common practical problem, there is a large number of algorithms for imputing missing data.
@@
 \subsubsection{Individual vs. population-based analyses}
-The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.However, population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
+The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.
+However, population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}.
 Often, an individual-based PCA is used to justify grouping individuals into populations; i.e.
@@
 \subsubsection{Summary}
-The version of PCA used here differs from that proposed by \cite{patterson2006}, and thus some care will be required to directly extend the interpretations developed here to individual-based PCAs.
+Motivated by $F$-statistics, the PCAs I consider here are based on estimated population allele frequencies, whereas most genome-scale studies of human genetic variation use observed individual-based allele frequencies that are normalized by overall allele frequencies.
+Thus, some care will be required to directly extend the interpretations developed here to individual-based PCAs.
 However, the differences are largely due to conventions, and particularly for studies where the description of population structure is a major focus, results might be easier to interpret if conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.
@@
 \subsection{The Apportionment of Human Diversity}
-Most genetic variation in humans is shared between all of us, but the around 15\% that can be explained by population structure can be leveraged to study our history and diversity in great detail \citep{dobzhansky1972, cavalli-sforza1964, reich2018a}.
+Most genetic variation in humans is shared between all of us, but the around 15\% that can be explained by population structure can be leveraged to study our history and diversity in great detail \citep{dobzhansky1972, cavalli-sforza1994, novembre2022}.
 For some data sets it is possible to predict an individuals' origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.
@@
 prediction of phenotypes \citep{berg2019, yair2021}, inference of demographic parameters, and schematic models of human genetic history \citep{patterson2012}.
-This is also the case for $F$-statistics, which are motivated by trees.
-They assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
+In a similar vein, when interpreting $F$-statistics in the context of admixture graphs, we make the  implicit assumtion that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
 However, these simplifications do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of deemphasizing people that do not rigidly fall into predefined genetic groups.

### 2022-02-09 fstats_pca_lewontin.tex

@@
 The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
-On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs \citep{cavalli-sforza1975}.
-Thus, I would expect that approximating $F$-statistics with few PCs works well whenever local variation is unimportant, i.e.
-when calculating $F$-statistics far apart in PCA space.
-On the flip side, populations close-by on a PCA-plot are often seperated on higher PCs, and thus whenever close-by populations are included in the same $F$-statistics, these higher PCs will be required for good approximations.
+Thus, we may expect that whenever the first few PCs provide a good approximation of population structure, this results in well-approximated $F_4$-statistics.
+On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
+Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs as there local differentiation, which is typically reflected on higher PCs, becomes important.
 
@@
 
-If sufficient PCs are considered, both $F_3$ and $F_4$-statistics do have simple interpretations in PCA-space.
-If used as a test for admixture, $F_3$ corresponds to a $n$-sphere in PCA-space, which manifests as a circle on a PCA-plot.
-The informal notion that an admixed population should lie between its sources on a PCA-plot is necessary, but not sufficient; unadmixed populations may also project there.
-In contrast, it is sufficient (but not necessary) to determine that a population has a positive $F_3$ statistic.
+If sufficiently many PCs are considered, both $F_3$ and $F_4$-statistics do have simple interpretations in PCA-space.
+If used as a test for admixture, $F_3$ corresponds to testing whether the admixed population lies in a $n$-sphere between the potential source population in PCA-space.
+This result makes the informal notion that an admixed population should lie between its sources on a PCA-plot more precise.
+Furthermore, I show that while it is necessary for an admixed population to lie between its sources on a PCA plot, this is not a sufficient condition and unadmixed populations may also project inside the $n$-sphere.
+In contrast, if a population falls outside the $n$-sphere on any PCA-plot, this is sufficient to determine that this population has a positive $F_3$ statistic.
 
-Interpretations of the outgroup-$F_3$-statistics and $F_4$-statistics in a PCA-framework rely heavily on the idea of a projection: In a tree or admixture-graph framework, we can loosely interpret $F_4(A, B; C, D)$ as a projection of one the path $A-B$ onto $C-D$; only edges shared between the paths will contribute non-zero amounts to this statistics.
-This projection analogy holds when we replace the populations by points in PCA-space, which enables us to study population models beyond discrete graphs.
-In this context, $F_4$ is simply an orthogonality test, an $F_4$ of zero means that the pairs of populations are orthogonal (i.e.
-independent) of another.
-
-One difficulty in interpreting the results here is that they only hold when all PCs are considered.
-Yet, in my applications I found that two (for the World data set) and three PCs (for Western Eurasia) already provided a very good approximation for the majority of $F$-statistics, so that the results here should be applicable to at least continental-scale data sets.
+Interpretations of the outgroup-$F_3$-statistics and $F_4$-statistics in a PCA-framework rely heavily on the geometric concepts of projections and orthogonality: In a tree or admixture graph framework, we can loosely interpret $F_4(A, B; C, D)$ as the overlap of the path from $A$ to $B$ onto the path from $C$ to $D$, or equivalently as the  projection of $A-B$ onto $C-D$; only edges shared between the two paths will contribute non-zero amounts to this statistic.
+This interpretation holds when we replace the populations by points in PCA-space, and enables us to study population models beyond discrete graphs.
+Thus, the geometric framework enables us to expand the application of $F$-statistics beyond current uses, allows us to better understand the meaning of these statistics and helps us to avoid overinterpretations particularly in cases when populations structure is continuous.
 

### 2022-02-09 fstats_pca_lewontin.tex
proofread

@@
 
-The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a, gopalan2022}.
-In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}.
-In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985}.
-However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}.
+The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{serre2004, rosenberg2005, bradburd2018, peter2020a, gopalan2022}.
+In particular, populations are expected to differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}.
+In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{wright1943, slatkin1985}.
+However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990}.
 In addition,  major historical population movements such as the out-of-Africa \citep{ramachandran2005}, Austronesian \citep{stoneking2016} or Bantu expansions \citep{li2014} lead to more gradual patterns of genetic diversity over space \citep{francois2010}.
@@
 A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}.
-Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, jombart2009, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}.
+Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, jombart2009, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS, ][]{lessa1990}.
 These methods are limited in their ability to estimate biologically meaningful parameters, but provide useful summaries and visualizations.
-These analyses are then complemented with methods based on explicit demographic models, which are used to estimate parameters or test hypotheses \citep{gutenkunst2009, excoffier2013, kamm2015}.
+Typically, these analyses are then complemented with methods based on explicit demographic models, which are used to estimate parameters or test hypotheses \citep{gutenkunst2009, excoffier2013, kamm2015}.
 
@@
 	\caption{\textbf{Representation of $F$-statistics on trees and 2D-PCA-plots.} The schematics show four populations and their representation using an (arbitrarily rooted) tree (top row) or a 2D-PCA plot (bottom row).
-A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space.
-B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthogonal projection of $X_1 - X_3$ onto $X_1$-$X_4$.
-C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or the orthogonal projection of $X_2 - X_3$ on $X_1 - X4$.
+A: $F_2$ represents the squared Euclidean distance between two tree leafs, and in PC-space.
+B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthogonal projection of $X_1 - X_3$ onto $X_1-X_4$.
+C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or to the orthogonal projection of $X_2 - X_3$ onto $X_1 - X4$.
 D: $F_4(X_1, X_2; X_3, X_4)$ The two paths from $X_1$ to $X_2$ and $X_3$ and $X_4$ are non-overlapping in the tree, which corresponds to orthogonal vectors in PCA-space.}
@@
 A comprehensive  treatise on PCA is given by e.g.
-\cite{jolliffe2013}, a useful primer on the mathematics is \cite{pachter2014}, and a helpful guide to interpretation is \cite{cavalli-sforza1994}.
+\cite{jolliffe2013}, a useful technical primer is \cite{pachter2014}, and a helpful guide to interpretation is \cite{cavalli-sforza1994}.
 Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.
@@
 \subsection{Geometric interpretation of $F$-statistics}
-An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare.
+An implicit assumption in the development of $F$-statistics in the context of admixture graphs has been that population lineages are mostly discrete, and that gene flow is rare.
+Recently, \cite{oteo-garcia2021} showed that this is not, in fact, necessary.
 Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$.
 In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space.
-In particular the $F$-statistics can be written as
+In this framework, the $F$-statistics can be written as
 \begin{subequations}
@@
 	\caption{\textbf{Admixture representation on 2D-PCA-plot.} The schematics show four populations and their representation using an admixture graph (A) or a 2D-PCA plot.
-A: Admixture graph, with population $X_y$ originating as an admixture of $X_2$ (with proportion $1-\alpha$)and $X_3$ (proportion ($\alpha$), with $X_2$ contributing proportion $\alpha$.
+A: Admixture graph, with population $X_y$ originating as an admixture of $X_2$ (with proportion $1-\alpha$)and $X_3$ (proportion ($\alpha$).
 Subsequent drift (red branch) will change allele frequency to sampled admixture population $X_x$.
@@
 Since the allele frequencies of $X_y$ are a linear combination of $X_2$ and $X_3$, it will lie on the line segment connecting these two populations (Figure \ref{fig:admix}B), at a location predicted by the admixture proportions.
-Subsequent drift will change the allele frequencies of $X_x$ (to say, $X_y$), and so in general it might fall on a different point on a PCA-plot.
+Subsequent drift will change the allele frequencies of $X_y$ (to say, $X_x$), and so in general it might fall on a different point on a PCA-plot.
 An exception occurs when $X_x$ (and no other populations related to $X_x$) are not part of the construction of the PCA, so that $X_x - X_y$ is orthogonal to all PCs, i.e.
@@
 Of the right-hand-side terms, only the $F_3$ term depends on the $X_i$.
-The fraction can be thought of as a normalizing constant,  so the $F_3$-statistic is proportional to the length of  the projected vector, and they can thus be interpreted similarly.
+The fraction can be thought of as a normalizing constant,  so the $F_3$-statistic is proportional to the length of the projected vector.
 This means, that the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in Figure \ref{fig:outgroupf3} this is $X_3$.
@@
 One interpretation of $F_4$ on PCA plots is similar to that of $F_3$; as a projection of one vector onto another, with the difference that now all four points may be distinct.
-$F_4$-statistics that correspond to an internal branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot (Figure \ref{fig:geom}G), again with the caveat that we need to scale it by a constant.
+$F_4$-statistics that correspond to an internal branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot, again with the caveat that we need to scale it by a constant.
 If the $F_4$-statistic corresponds to a branch that does not exist in the tree (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e.
-that $X_1$ and $X_2$  map to the same point on the projection axis $\overline{X_3X_4}$ (Figure \ref{fig:geom}H).
+that $X_1$ and $X_2$  map to the same point on the projection axis $\overline{X_3X_4}$..
 In the case of an admixture graph, this is no longer the case: Both population $X_y$ and $X_x$ in Figure \ref{fig:admix}D do \emph{not} map to the same point as $X_1$ or $X_2$ do, implying that statistics of the form $F_4(X_1, X_x; X_3, X_4) \neq 0$.
@@
 where $\phi$ is the angle between $X_1 - X_2$ and $X_3 - X_4$.
-Thus, independent drift events lead to $\cos(\phi) = 0$, so that the angle is 90 degrees, whereas an angle close to zero means $\cos(\phi)\approx 1$, which means most of the genetic drift on this branch is shared.
+Thus, independent drift events lead to $\cos(\phi) = 0$, so that the angle is 90 degrees, whereas an angle close to zero ($\cos(\phi)\approx 1$) means most of the genetic drift on this branch is shared.
 
@@
 I use the World data set as an example, using Sardinian and Mbuti as references populations (Figure \ref{fig:world}B).
-The data are the same as in  the PCA (Figure \ref{fig:world}A), but it is now rotated such that the axis between the reference population (black line in Figure \ref{fig:world}A) is now reflected by the x-axis.
+The data are the same as in  the PCA (Figure \ref{fig:world}A), but it is now rotated such that the axis between the reference population (black line in Figure \ref{fig:world}A) is aligned with the x-axis.
 For any pair of populations $X_1$ $X_2$, their horizontal projection distance reflects $F_4(\text{Sardinian}, \text{Mbuti}; X_1, X_2)$ and the relative horizontal distance corresponds exactly to $F_4$-ratio admixture estimates.
@@
 Thus, we may expect that whenever the first few PCs provide a good approximation of population structure, this results in well-approximated $F_4$-statistics.
-On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
-Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs as there local differentiation, which is typically reflected on higher PCs, becomes important.
+On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}B), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
+Thus, I would also expect other $F$-statistics including large contribution from local variation such as  $F_4$-statistics between closely related populations to require substantially more PCs.
 
@@
 In addition to the selection of PCs, the other cause for disagreements between $F$-statistics and PCA are differences in assumptions.
-The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized.
+The version of PCA I use for my analyses is chosen such that the similarities to $F$-statistics are maximized.
 In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted, and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies.
@@
 in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
-In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard error.
+In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard errors.
 
@@
 For statistical tests based on $F$-statistics, we wish to isolate a set of three or four populations and get our best guess based on just that subset of data.
-In contrast methods for PCA can leverage the additional individuals, and thus will likely result in more accurate estimates.
+In contrast, methods for PCA can leverage the additional individuals, and thus will likely result in more accurate estimates.
 
@@
 \subsubsection{Normalization}
-In PCA, SNPs are typically normalized to have expected variance of one; a step that is omitted in calculating $F$-statistics \citep{patterson2006}.
-The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable.
-which holds regardless of weighting.
+In PCA, SNPs are typically normalized to have expected variance of one; a step that is omitted in calculating $F$-statistics \citep{patterson2012}.
+The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable, which holds regardless of weighting.
 Thus, normalization of SNPs is largely a matter of convention, for  $F$-statistics the dependency on  additional samples (through  mean allele frequencies) is often unwanted, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics such as \texttt{qpAdm} \citep{patterson2012, harney2021}.
-As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but should be explored in future work \citep{felsenstein1973}.
+As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but could be explored in future work \citep{felsenstein1973}.
 
@@
 Probabilistic PCA is one class of approaches that aim to separate the population structure from sampling noise \citep[e.g.][]{agrawal2020}.
-It seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.
+It seems likely that probabilistic PCA would yield a representation of the data that is more closely aligned with $F$-statistics than regular PCA.
 

### 2022-03-07 fstats_pca_lewontin.tex
final changes

@@
 
-The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure.
-Thus, we may expect that whenever the first few PCs provide a good approximation of population structure, this results in well-approximated $F_4$-statistics.
-On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}B), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs.
-Thus, I would also expect other $F$-statistics including large contribution from local variation such as  $F_4$-statistics between closely related populations to require substantially more PCs.
+The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of the overall population structure.
+On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}B).
+This is likely due to PCA approximating the global structure in the data set; statistics that only involve distantly related populations will only require a few PCs for good approximations, whereas statistics that contain a term measuring local variation, such as $F_3$-statistics or $F_4$ between closely related populations will require more PCs for good approximations, because local variation is often found on higher PCs.
 
@@
 The simplest approach is to replace missing data with zeros \citep[as implemented e.g.
-in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015, meisner2018, meisner2021}.
+in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015,  meisner2021}.
 In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard errors.
@@
 
-Thus, a more abstract way of interpreting $F$-statistics is that they reflect the relative position of population to each other in PCA-space.
-The classical model of admixture as a result of contact between long-separated and isolated populations is just one of potentially many demographic models that might result in this data pattern.
+Thus, a more general way we could think about modeling population structure using $F$-statistics is as identifying orthogonal drift components.
+In a tree model, orthogonality arises  because changes in allele frequencies on distinct branches of the tree are independent from each other (and high-dimensional random vectors are almost surely orthogonal).
+The classical model of admixture as a result of contact between long-separated and isolated populations is one of potentially many demographic models that results in non-orthogonality; gene flow of any kind will result in correlated genetic drift, and hence in non-zero $F_3$ or $F_4$-statistics.
 
-\begin{comment}
-This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}.
-One issue with these approaches is that they are usually restricted to at most a few dozen populations.
-As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include.
-A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data.
-However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits.
-The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not need to assume that gene flow is rare.
-\end{comment}
-
-In this vein, admixture tests based on $F$-statistics evaluate if drift in different parts of the tree are orthogonal to each other.
-This orthogonality is expected to arise whenever drift in different populations is independent, which happens, for example, if we consider two isolated populations, or if we consider the changes in allele frequency in distinct time intervals.
-
-In the context of tree-like population models, this orthogonality arises from drift on non-overlapping branches.
-
-an alternative could be to focus on orthogonality of drift as a \emph{the} central concept for analyzing genetic variation data.
-
-This is often informally done in the context of PCA, where it is often observed that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}.
+Thinking of population structure in terms of orthogonal components may be abstract, but it is quite similar to how PCA is sometimes interpreted.
+In a PCA, we frequently make the informal observation that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}.
 These associations are not exact, and impacted e.g.
 by the size and composition of the analyzed data set.
-On the other hand, we can use $F$-statistics to formally test orthogonality.
-Distinct clades in an admixture graph would be one possibility how orthogonality could arise, but assuming a graph is not necessary.
+On the other hand, we could use $F$-statistics to formally test orthogonality, or to quantify correlations.
 Motivated by PCA, we could set up $F_4$ as tests of orthogonality due to either space (distinct populations evolve independently) or scale (i.e.
 between-population diversity is independent of within-population diversity).
-This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming that lineages or admixture events are discrete.
+This slight generalization of $F$-statistics could allow us to reframe many questions about gene flow or divergence that are currently asked as tests of orthogonality, without assuming that lineages or admixture events are discrete.
 

### 2022-08-26 talk_leipzig.tex
stuff

@@
 
-\begin{comment}
 {
@@
 }
-\end{comment}
-
-\begin{comment}
 
@@
 \end{frame}
-\end{comment}
 
@@
 }
+\end{frame}
+
+\begin{frame}
+	\frametitle{Now In Press!}
+			\hspace*{-1\beamerleftmargin}
+	\includegraphics[width=1.2\textwidth]{talk_figures/paper_head.png}
 \end{frame}
@@
 
+\begin{comment}
 \begin{frame}
@@
 \end{frame}
+
+\end{comment}
 
@@
 \end{columns}
+\end{frame}
+
+\begin{frame}
+\frametitle{Dimensionality Intuition}
+\begin{center}
+\includegraphics<1>[width=.5\textwidth]{figures/fig_proj_talk_0.pdf}
+\includegraphics<2>[width=.5\textwidth]{figures/fig_proj_talk_1.pdf}
+\includegraphics<3>[width=.5\textwidth]{figures/fig_proj_talk_2.pdf}
+\includegraphics<4>[width=.5\textwidth]{figures/fig_proj_talk_3.pdf}
+\includegraphics<5>[width=.5\textwidth]{figures/fig_proj_talk_4.pdf}
+\end{center}	
 \end{frame}
@@
 
-\begin{comment}
+\begin{frame}
+	\frametitle{Positions Available}
+	\begin{columns}
+		\begin{column}{0.3\textwidth}
+			\includegraphics[width=\textwidth]{talk_figures/erc.png}
+		\end{column}
+		\begin{column}{0.7\textwidth}
+			\textbf{NEADMIX}-project
+			\begin{itemize}
+				\item Goal: Study Neandertal ancestry through time
+				\item 1 PhD and 1 Postdoc position available!
+				\item Theory, method development and data analysis
+			\end{itemize}
+		\end{column}
+	\end{columns}
+\end{frame}
+
 \begin{frame}

### 2022-08-26 data/talk_figs.R

@@ -1,10 +1,11 @@
 require(tidyverse)
+require(scales)
 
-a = read_tsv("v44.3_1240K_public.anno") %>%
+a = read_tsv("v52.2_HO_public.anno") %>%
     mutate(lat=as.numeric(Lat.), long=as.numeric(Long.)) %>%
-    rename(year=6, age=10) %>%
+    rename(year=6, age=9) %>%
     mutate(year=as.numeric(year), age=as.numeric(age)) %>%
-    filter(age>0, !is.na(year), !is.na(lat), !is.na(long)) %>%
+    filter(age>0, year<=2022, !is.na(year), !is.na(lat), !is.na(long)) %>%
     select(lat, long, year, age)
 
 df_year = a %>% 
@@ -21,14 +22,28 @@ library("rnaturalearthdata")
 
 world <- ne_countries(scale = "medium", returnclass = "sf")
 
+make_fig <- function(a, fname){
 P2 = ggplot() + 
     geom_sf(data=world, color=NA) + 
     theme_minimal() + 
     geom_point(data=a, aes(color=age, x=long, y=lat)) +
     theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
-    theme(legend.position = 'none') + 
+    #theme(legend.position = 'none') + 
     xlab(NULL) + ylab(NULL) +
     coord_sf(xlim=c(-170, 170), ylim=c(-55, 70)) +
-    scale_color_viridis_c(trans='log')
+    scale_color_viridis_c(breaks=c(0000, 10000, 20000),
+    limits=c(0000,20000), oob=squish)
 
-ggsave("figures/ancient_dna_map.png", P2, width=7, height=3)
+ggsave(fname, P2, width=7, height=3)
+}
+
+a %>% filter(year <= 2010) %>% make_fig(fname='figures/ancient_dna_till2010.png')
+a %>% filter(year <= 2014) %>% make_fig(fname='figures/ancient_dna_till2014.png')
+a %>% filter(year <= 2016) %>% make_fig(fname='figures/ancient_dna_till2016.png')
+a %>% filter(year <= 2022) %>% make_fig(fname='figures/ancient_dna_till2022.png')
+
+a %>% filter(age > 40000) %>% make_fig(fname='figures/ancient_dna_older_than40k.png')
+
+
+a %>% ggplot(aes(x=age)) + geom_histogram(binwidth=1000) -> P3
+ggsave('figures/ancient_dna_by_year.png', P3, width=5, height=3)

### 2022-08-26 data/world_map_talks.R

@@ -7,7 +7,7 @@ library(ggpubr)
 BTHEME = theme_classic() + theme(legend.position="none")
 
 
-anno = read_tsv("v44.3_HO_public.anno") %>%
+anno = read_tsv("v52.2_HO_public.anno") %>%
     select(iid=2,
            age=6,
            pop=8,

### 2022-08-26 ppca.R

@@ -88,23 +88,25 @@ ppca <- function(data, n_pcs=4, n_iter=1000, tol=1e-4, verbose=T){
 
 test_ppca <- function(){
     require(glue)
-    N_SNPS = 20000
-    N_SAMPLES = 400
-    TRUE_RANK = 3
-    S1 = 1; S2 = 19
-    S3 = 3; S4 = 75
+    N_SNPS = 100000
+    N_SAMPLES = 600
+    TRUE_RANK = 5
+    S1 = 1; S2 = 456
+    S3 = 132; S4 = 561
     NOISE = 0.05
 
-    N_PCS1 = 3
+    N_PCS1 = 5
     N_PCS2 = 20
 
 
     require(MASS)
     set.seed(10)
-    points = matrix(runif(N_SAMPLES * TRUE_RANK), nrow=N_SAMPLES)
+    points = matrix(runif(N_SAMPLES * TRUE_RANK), nrow=N_SAMPLES) / 10
     test_cov = points %*% t(points)
-    data = mvrnorm(N_SNPS, mu=rep(0, N_SAMPLES), Sigma=test_cov); 
-    d2=data + mvrnorm(N_SNPS, mu=rep(0, N_SAMPLES), Sigma=diag(NOISE, N_SAMPLES))
+    data = mvrnorm(N_SNPS, mu=rep(0.5, N_SAMPLES), Sigma=test_cov); 
+    dceil = pmax(pmin(data, 1), 0)
+    #d2=data + mvrnorm(N_SNPS, mu=rep(0, N_SAMPLES), Sigma=diag(NOISE, N_SAMPLES))
+    d2 = t(apply(dceil, 1, function(p)rbinom(ncol(dceil), 2, p))) / 2
 
     x = ppca(d2, verbose=F, n_pcs=N_PCS1)
     y = ppca(d2, verbose=F, n_pcs=N_PCS2)
@@ -210,3 +212,8 @@ ppca2 = function(Y){
     X = solve(t(C) %*% C) %*% t(C) %*% t(Y)
     C = t(Y) %*% t(X) %*% solve(X %*% t(X))
 }
+
+test_ppca2 <- function(){
+    require(admixtools)
+}
+

### 2022-08-26 data/pcas.R
fstats_pca

@@ -1,0 +1,136 @@
+source("fscripts.R")
+library(ggpubr)
+library(ggrepel)
+library(glue)
+
+R = 1.5
+BTHEME = theme_classic() + theme(legend.position="none")
+
+if(T){
+f2s = admixtools::read_f2("westeurasian1/")
+pcmat = pca_from_f2s(f2s)
+pcs = pca_from_pcmat(pcmat)
+idx = "Sardinian"
+idy = "Finnish"
+
+sample_list = pcs$pop 
+
+f3 = f3(f2s, pop3=idx, pop2=idy, pop1=sample_list) #admixture
+f3pc = f3_from_pc_matrix(pcmat, idx, idy)
+#f3 = f3(f2s, pop1=idx, pop2=idy, pop3=sample_list) #outgroup
+f4 = f4(f2s, pop3=idx, pop4 = idy, pop1=sample_list, pop2=sample_list) %>% arrange(pop1, pop2)
+
+circle = pcs %>% filter(pop %in% c(idx, idy)) %>% select(-pop)
+circle_df = colMeans(circle) %>% t %>% as.data.frame
+circle_df$radius = (dist(circle) %>% as.matrix)[1,2] / 2
+circle_df$r0 = (dist(circle[,1:2]) %>% as.matrix)[1,2] / 2
+circle_df$r34 = (dist(circle[,3:4]) %>% as.matrix)[1,2] / 2
+circle_df$r23 = (dist(circle[,2:3]) %>% as.matrix)[1,2] / 2
+circle_df$r56 = (dist(circle[,5:6]) %>% as.matrix)[1,2] / 2
+
+pcs2 = pcs %>% left_join(f3, by=c(pop="pop1")) %>% 
+    mutate(col=as.factor(cut(est, c(-1,-1e-10,1e-10,2), labels=F) ))
+P1 = ggplot() + 
+    geom_circle(aes(x0=PC1, y0 =PC2, r=radius), data=circle_df, color='lightgray', fill='lightgray') +
+    geom_circle(aes(x0=PC1, y0 =PC2, r=r0), data=circle_df, color='darkgray', fill='#808080') +
+    geom_text_repel(aes(x=PC1, y=PC2, label=pop, color=col), data=pcs2) +
+    geom_point(aes(x=PC1, y=PC2, label=pop, color=col), data=pcs2) +
[…]
